### 2019-01-09 paper.tex
structure of paper

\documentclass{lmcs} %%% last changed 2014-08-20

%% mandatory lists of keywords 
\keywords{MANDATORY list of keywords}

%% read in additional TeX-packages or personal macros here:
%% e.g. \usepackage{tikz}
\usepackage{hyperref}
%%\input{myMacros.tex}
%% define non-standard environments BEYOND the ones already supplied 
%% here, for example
\theoremstyle{plain}\newtheorem{satz}[thm]{Satz} %\crefname{satz}{Satz}{S\"atze}
%% Do NOT replace the proclamation environments lready provided by
%% your own.

\def\eg{{\em e.g.}}
\def\cf{{\em cf.}}

%% due to the dependence on amsart.cls, \begin{document} has to occur
%% BEFORE the title and author information:

\begin{document}

\title[Short title TODO]{Homomorphisms from type theoretic signatures using parametricity}
\titlecomment{{\lsuper*} This is a journal version of the paper \cite{hiit}.}

\author[A.~Kaposi]{Ambrus Kaposi}	%required
\address{Department of Programming Languages and Compilers, E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}	%required
\email{akaposi@inf.elte.hu}  %optional
%\thanks{thanks 1, optional.}	%optional

\author[A.~Kov{\'a}cs]{Andr{\'a}s Kov{\'a}cs}	%optional
\address{Department of Programming Languages and Compilers, E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}	%required
\email{kovacsandras@inf.elte.hu}  %optional
\thanks{This work was supported by the European Union, co-financed by the
European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002) and COST Action
EUTypes CA15123.}	%optional

%% etc.

%% required for running head on odd and even pages, use suitable
%% abbreviations in case of long titles and many authors:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% the abstract has to PRECEDE the command \maketitle:
%% be sure not to issue the \maketitle command twice!

\begin{abstract}
  \noindent If a logical relation between two algebras is a graph of a
  function, it is an algebra homomorphism. 
\end{abstract}

\maketitle

\section*{Introduction}

In HoTT, algebraic structures can have higher equalities. What is a
homomorphism between two such algebraic structures? Example: circle,
sphere. This also answers the question: what is an elimination
principle for any HIT?

Another Motivation: usefulness of GATs for formalising programming
languages. Example: tt-in-tt with equations. Algebraic view of
progr. langs.

Method: Uday observation (inverse Reynolds), signatures for
generalised algebraic theories using domain-specific tt, parametricity
interpretation of dependent types. We put these
together. Achievements: getting induction principles for non-strictly
positive types, describing induction principles for HIITs.

This is a journal version, main new thing: relation to parametricity
(graph model), category model (extension of reflexive graph model),
homomorphisms for non strictly positive operators, relation of
syntactic translations and models for solving coherence issue, for
HIITs the syntactic translation cannot be extended to a category.

Formalisation?

\subsection*{Overview of the paper}

\subsection*{Related work}

\section*{Theory of signatures}

Brick red color. Subst.calc, U,El,Pi. We formalise the calculus as a
QIIT, but present it with variables names and judgements. Eliminating
from the QIIT needs to preserve equalities. Metatheory extensional.

\section*{From logical relations to homomorphisms}

Graph model, (A,M''), check what it gives for natural numbers

we restrict U from relations to functions and El becomes graph of the
function

now we get homomorphisms

example: homomorphism for non-strictly positive operators

problem: suc is not the usual one, we cannot compose homomorphisms:
composition does not work for Pi

\section*{Categories of algebras}

Restriction to strictly positive operators in signatures.

category model

\section*{Extensions}

Pi with metatheoretic domain and large codomain

Pi with metatheoretic domain and small codomain (infinitary)

Extensional identity

Identities between sorts

Examples

\section*{Higher equalities}

The metatheory has K, so we do not elimiante into it, but to a target
theory which does not have K. This also solves a coherence problem
when eliminating from the syntax as a QIIT in HoTT as
metatheory. Also, we add intensional Id type.

\subsection*{Source theory}

(Brick red)

\subsection*{Target theory}

(Blue)

\subsection*{Translations}

cannot be extended to a category because U[] is not strict unless we
add new computation rules.

\subsection*{Examples}

\section*{Conclusions and further work}

\section*{Acknowledgment}
  \noindent The authors wish to acknowledge fruitful discussions with
  A and B.

\bibliographystyle{plain}
\bibliography{b}

\appendix
\section{}

appendix contents

\end{document}

### 2019-01-27 fscd.tex
Notes, stuff renamings, draft structure

\documentclass[a4paper,UKenglish]{lipics-v2018}

\usepackage{microtype}
\usepackage{proof}
\usepackage[dvipsnames]{xcolor}
\usepackage{amsmath}
\usepackage{stmaryrd}

\bibliographystyle{plainurl}

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{A Syntax for Higher Inductive-Inductive Types}
%\titlerunning{} %optional, in case that the title is too long; the running title should fit into the top page column

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author{Ambrus Kaposi}{Faculty of Informatics, E{\"o}tv{\"o}s Lor{\'a}nd University, P{\'a}zm{\'a}ny P{\'e}ter s{\'e}t{\'a}ny 1/C, 1117 Budapest, Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{}
\author{András Kovács}{Faculty of Informatics, E{\"o}tv{\"o}s Lor{\'a}nd University, P{\'a}zm{\'a}ny P{\'e}ter s{\'e}t{\'a}ny 1/C, 1117 Budapest, Hungary}{kovacsandras@inf.elte.hu}{https://orcid.org/0000-0002-6375-9781}{}
\authorrunning{A. Kaposi and A. Kovács} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\Copyright{Ambrus Kaposi and András Kovács}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\subjclass{Theory of computation $\rightarrow$ Type theory}
\keywords{homotopy type theory, inductive-inductive types, higher inductive types, quotient inductive types, logical relations}% mandatory: Please provide 1-5 keywords

\funding{This work was supported by the European Union, co-financed by
  the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002), USAF
  grant FA9550-16-1-0029, and by COST Action EUTypes CA15123.}

\acknowledgements{The authors thank Thorsten Altenkirch, Simon
  Boulier, Paolo Capriotti, Péter Diviánszky, Gábor Lehel and Nicolas
  Tabareau for discussions related to this paper and the anonymous
  reviewers for their helpful comments and suggestions. We thank
  Balázs Kőműves for the idea of using a Tarski universe instead of a
  Russell universe in the theory of codes.}

% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{H\'{e}l\`{e}ne Kirchner}
\EventNoEds{1}
\EventLongTitle{3rd International Conference on Formal Structures for Computation and Deduction (FSCD 2018)}
\EventShortTitle{FSCD 2018}
\EventAcronym{FSCD}
\EventYear{2018}
\EventDate{July 9--12, 2018}
\EventLocation{Oxford, UK}
\EventLogo{}
\SeriesVolume{108}
\ArticleNo{20}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%=====================================================================

%% setting \sqcdot as HoTT-book-style transitivity
\makeatletter
\DeclareRobustCommand{\sqcdot}{\mathbin{\mathpalette\morphic@sqcdot\relax}}
\newcommand{\morphic@sqcdot}[2]{%
  \sbox\z@{$\m@th#1\centerdot$}%
  \ht\z@=.33333\ht\z@
  \vcenter{\box\z@}%
}
\makeatother

\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\REN}{\mathsf{REN}}
\newcommand{\op}{\mathsf{op}}
\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\PSh}{\mathsf{PSh}}
\newcommand{\FamPSh}{\mathsf{FamPSh}}
\renewcommand{\ll}{\llbracket}
\providecommand{\rr}{\rrbracket}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Tms}{\mathsf{Tms}}
\newcommand{\R}{\mathsf{R}}
\newcommand{\TM}{\mathsf{TM}}
\newcommand{\NE}{\mathsf{NE}}
\newcommand{\NF}{\mathsf{NF}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\renewcommand{\u}{\mathsf{u}}
\renewcommand{\ne}{\mathsf{ne}}
\newcommand{\nf}{\mathsf{nf}}
\newcommand{\lQ}{\mathsf{lQ}}
\newcommand{\lU}{\mathsf{lU}}
\renewcommand{\lq}{\mathsf{lq}}
\newcommand{\lu}{\mathsf{lu}}
\newcommand{\cul}{\ulcorner}
\newcommand{\cur}{\urcorner}
\newcommand{\norm}{\mathsf{norm}}
\newcommand{\Nf}{\mathsf{Nf}}
\newcommand{\Ne}{\mathsf{Ne}}
\newcommand{\Nfs}{\mathsf{Nfs}}
\newcommand{\Nes}{\mathsf{Nes}}
\newcommand{\ID}{\mathsf{ID}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\nat}{\,\dot{\rightarrow}\,}
%\newcommand{\nat}{\overset{\mathsf{n}}{\ra}} % this is how we denote it in the formalisation
\newcommand{\Nat}{\mathsf{Nat}}
\renewcommand{\S}{\overset{\mathsf{s}}{\ra}} % we have it with uppercase S in the formalisation
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
%\newcommand{\blank}{\!{-}\!}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\tr}[2]{\ensuremath{{}_{#1 *}\mathopen{}{#2}\mathclose{}}}
\newcommand{\C}{\mathsf{C}}
\newcommand{\Code}{\mathsf{Code}}
\newcommand{\M}{\mathsf{M}}
% from the book
%\newcommand{\M}{{\scalebox{0.6}{$\mathsf{M}$}}}
%% \newcommand{\C}{\mathcal{C}}
\newcommand{\data}{\mathsf{data}}
\newcommand{\ind}{\hspace{1em}}
\newcommand{\idP}{\mathsf{idP}}
\newcommand{\compP}{\mathsf{compP}}
\newcommand{\idF}{\mathsf{idF}}
\newcommand{\compF}{\mathsf{compF}}
\newcommand{\proj}{\mathsf{proj}}
\newcommand{\ExpPSh}{\mathsf{ExpPSh}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\Vars}{\mathsf{Vars}}
\newcommand{\wk}{\mathsf{wk}}
\newcommand{\neuU}{\mathsf{neuU}}
\newcommand{\neuEl}{\mathsf{neuEl}}
\newcommand{\var}{\mathsf{var}}
\newcommand{\natn}{\mathsf{natn}}
\newcommand{\natS}{\mathsf{natS}}
\newcommand{\LET}{\mathsf{let}}
\newcommand{\IN}{\mathsf{in}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\trans}{\mathbin{\raisebox{0.2ex}{$\displaystyle\centerdot$}}}
\newcommand{\zero}{\mathsf{zero}}
\newcommand{\suc}{\mathsf{suc}}
\newcommand{\N}{\mathbb{N}}

\newcommand\arcfrombottom{
  \begin{tikzpicture}[scale=0.03em]
    \draw (0,0) arc (0:180:0.5);
    \draw (0,0) edge[->] (0,-0.3);
    \draw (-1,0) edge (-1,-0.3);
  \end{tikzpicture}
}
\newcommand\arcfromtop{
  \begin{tikzpicture}[scale=0.03em]
    \draw (0,0) arc (180:360:0.5);
    \draw (0,0) edge[->] (0,0.3);
    \draw (1,0) edge (1,0.3);
  \end{tikzpicture}
}

\newcommand{\Elim}{\mathsf{Elim}}
\newcommand{\elim}{\mathsf{elim}}
\newcommand{\Rec}{\mathsf{Rec}}
\newcommand{\record}{\mathsf{record}}
\newcommand{\funext}{\mathsf{funext}} \newcommand{\Q}{\mathsf{Q}}
\newcommand{\T}{\mathsf{T}} \newcommand{\leaf}{\mathsf{leaf}}
\newcommand{\node}{\mathsf{node}} \newcommand{\perm}{\mathsf{perm}}
\newcommand{\coe}{\mathsf{coe}} \newcommand{\vz}{\mathsf{vz}}
\newcommand{\vs}{\mathsf{vs}} \newcommand{\untr}{\mathsf{untr}}
\newcommand{\from}{\mathsf{from}}
\newcommand{\fromeq}{{\mathsf{from}\hspace{-0.3em}\equiv}}
\newcommand{\fromsimeq}{{\mathsf{from}\hspace{-0.3em}\simeq}}
\newcommand{\Model}{\mathsf{Model}}
\newcommand{\DModel}{\mathsf{DModel}}
\newcommand{\module}{\mathsf{module}}
\newcommand{\open}{\mathsf{open}}
\renewcommand{\P}{\mathsf{P}} \newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}} \newcommand{\false}{\mathsf{false}}
\renewcommand{\not}{\mathsf{not}} \newcommand{\0}{\mathsf{0}}
\newcommand{\1}{\mathsf{1}} \renewcommand{\Pr}{\mathsf{Pr}}
\newcommand{\PrNat}{\mathsf{PrNat}} \newcommand{\J}{\mathsf{J}}
\newcommand{\wkV}{\mathsf{wkV}} \renewcommand{\r}[1]{{\P_{#1}}}
\newcommand{\stab}{\mathsf{stab}} \newcommand{\NTy}{\mathsf{NTy}}
\newcommand{\isDec}{\mathsf{isDec}} \newcommand{\dec}{\mathsf{dec}}
\newcommand{\yes}{\mathsf{yes}} \newcommand{\no}{\mathsf{no}}
\newcommand{\case}{\mathsf{case}} \newcommand{\inj}{\mathsf{inj}}
\newcommand{\K}{\mathsf{K}} \newcommand{\lb}{\langle}
\newcommand{\rb}{\rangle} \newcommand{\Decl}{\mathsf{Decl}}
\newcommand{\Core}{\mathsf{Core}} \newcommand{\IND}{\mathsf{ind}}
\newcommand{\Id}{\mathsf{Id}} \newcommand{\Base}{\mathsf{Base}}
\newcommand{\Setoid}{\mathsf{Setoid}}
\newcommand{\FamSetoid}{\mathsf{FamSetoid}}
\newcommand{\prop}{\mathsf{prop}} \newcommand{\resp}{\mathsf{resp}}
\newcommand{\transport}{\mathsf{transport}}
\newcommand{\I}{\mathsf{I}} \newcommand{\E}{\mathsf{E}}
\newcommand{\transp}{\mathsf{transp}}
\newcommand{\Transp}{\mathsf{Transp}}
\newcommand{\W}{\mathsf{W}}
\newcommand{\Fin}{\mathsf{Fin}} \newcommand{\fzero}{\mathsf{fzero}}
\newcommand{\fsuc}{\mathsf{fsuc}}
\newcommand{\inv}{\mathsf{inv}}
\newcommand{\con}{\mathsf{con}}
\newcommand{\LEFT}{\mathsf{left}}
\newcommand{\RIGHT}{\mathsf{right}}
\newcommand{\seg}{\mathsf{seg}}
\newcommand{\Int}{\mathsf{Int}}
\renewcommand{\in}{\mathbin{\hat:}}
\renewcommand{\hat}[1]{{\color{BrickRed}{#1}}}
\newcommand{\vdashh}{\mathbin{\hat\vdash}}
\newcommand{\rah}{\mathbin{\hat\ra}}
\newcommand{\commah}{\hat,\,}
\newcommand{\timesh}{\mathbin{\hat\times}}
\newcommand{\eqh}{\mathbin{\hat=}}
\newcommand{\TR}{\hat{\mathsf{tr}}}
\newcommand{\ap}{\hat{\mathsf{ap}}}
\newcommand{\apd}{\hat{\mathsf{apd}}}
\renewcommand{\tt}{\hat{\mathsf{tt}}}
\newcommand{\Tel}{\hat{\mathsf{Tel}}}
\newcommand{\Type}{\hat{\mathsf{Type}}}
\newcommand{\emptytel}{\hat{\epsilon}}
\newcommand{\telext}{\mathbin{\hat{\lhd}}}
\newcommand{\ltel}{\hat{\ll}}
\newcommand{\rtel}{\hat{\rr}}
\newcommand\pp{\ensuremath{\hat{\mathbin{+\mkern-10mu+}}}}
\newcommand{\semicol}{\hat;\,}
\newcommand{\targetass}{\hat{\Gamma}\semicol}
%\newcommand{\targetass}{;}

%=====================================================================

\begin{document}

\maketitle

%=====================================================================

\begin{abstract}
Higher inductive-inductive types (HIITs) generalise inductive types of
dependent type theories in two directions. On the one hand they allow
the simultaneous definition of multiple sorts that can be indexed over
each other. On the other hand they support equality constructors, thus
generalising higher inductive types of homotopy type theory. Examples
that make use of both features are the Cauchy reals and the well-typed
syntax of type theory where conversion rules are given as equality
constructors. In this paper we propose a general definition of HIITs
using a domain-specific type theory. A context in this small type
theory encodes a HIIT by listing the type formation rules and
constructors. The type of the elimination principle and its
$\beta$-rules are computed from the context using a variant of the
syntactic logical relation translation. We show that for indexed
W-types and various examples of HIITs the computed elimination
principles are the expected ones. Showing that the thus specified
HIITs exist is left as future work. The type theory specifying HIITs
was formalised in Agda together with the syntactic translations. A
Haskell implementation converts the types of sorts and constructors
into valid Agda code which postulates the elimination principles and
computation rules.
\end{abstract}

%=====================================================================

\section{Introduction}
\label{sec:intro}

Many dependent type theories support some form of inductive types. An
inductive type is given by its constructors, along with an elimination
principle which expresses that all inhabitants are constructed using
finitely many applications of the constructors.

For example, the inductive type of natural numbers $\Nat$ is given by
the constructors $\zero:\Nat$ and $\suc:\Nat \ra \Nat$. The
eliminator corresponds to the usual notion of mathematical induction:
\[
  \Elim\Nat:(P:\Nat \ra \mathsf{Type})(pz: P\,\zero)\big(ps:(n:\Nat)\ra P\,n\ra P\,(\suc\,n)\big)(n:\Nat)\ra P\,n
\]
$P$ is a family of types over natural numbers, which is called the
\emph{motive} of the eliminator. It can be viewed as a proof-relevant
predicate on $\Nat$. The arguments $pz$ and $ps$ are called the
\emph{methods} of the eliminator. The \emph{target} of the eliminator
is $n$ and given methods for each constructor, the eliminator provides
a witness of $P\,n$. Thus, if $P$ holds for $\zero$ and $\suc$
preserves $P$, then $P$ holds for all natural numbers. The behaviour
of the eliminator is described by a \emph{computation-rule}
($\beta$-rule) for each constructor:
\begin{alignat*}{5}
  & \Elim\Nat\,P\,pz\,ps\,\zero && \equiv pz \\
  & \Elim\Nat\,P\,pz\,ps\,(\suc\,n) && \equiv ps\,n\,(\Elim\Nat\,P\,pz\,ps\,n)
\end{alignat*}
These express that the eliminator applied to a constructor expression
reduces to an application of the corresponding induction method. From
an operational point of view, $\Elim\Nat$ replaces all the $\zero$ and
$\suc$ constructors with the given induction methods.

% A special case of the eliminator is the recursor (non-dependent
% eliminator) which allows the definition of a non-dependent function
% from the inductive type, for $\Nat$ it has the following type.
% \[
% \Rec\Nat : \forall(A:\mathsf{Type}).A \ra (A \ra A) \ra \Nat \ra A
% \]
% The eliminator can be derived from the recursor and a uniqueness
% principle.

Dependent families of types can be defined in a similar way, for
example vectors of $A$-elements $\mathsf{Vec}_A: \Nat \ra \mathsf{Type}$ which
are indexed by their length. Another generalisation of inductive types
are mutual inductive types. However, these can be reduced to indexed
families where indices classify constructors for each mutual type.
Inductive-inductive types \cite{forsberg-phd} are mutual definitions
where this reduction does not work: here a type is defined together
with a family indexed over it. An example is the following fragment of
the well-typed syntax of type theory where the second sort $\Ty$ is
indexed over the first sort $\Con$, but constructors of $\Con$ also
refer to $\Ty$:
\begin{alignat*}{3}
  & \Con && : \mathsf{Type} && \text{sort of contexts} \\
  & \Ty  && : \Con \ra \mathsf{Type} && \text{sort of types given a context} \\
  & \bullet && : \Con && \text{constructor for the empty context} \\
  & \blank\rhd\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con && \text{constructor for context extension} \\
  & \U && : (\Gamma:\Con)\ra\Ty\,\Gamma && \text{constructor for a base type} \\
  & \Pi && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\Gamma\rhd A)\ra\Ty\,\Gamma \hspace{1em} && \text{constructor for dependent functions}
\end{alignat*}
There are two eliminators for this type: one for $\Con$ and one for
$\Ty$. Both take the same arguments: two motives ($P:\Con\ra\mathsf{Type}$ and
$Q:(\Gamma:\Con)\ra P\,\Gamma\ra\Ty\,\Gamma\ra\mathsf{Type}$) and four methods
(one for each constructor, we don't list these).
\begin{alignat*}{6}
  & \Elim\Con && : (P:\dots)(Q:\dots)\ra\dots\ra(\Gamma:\Con) && \ra P\,\Gamma \\
  & \Elim\Ty  && : (P:\dots)(Q:\dots)\ra\dots\ra(A:\Ty\,\Gamma) && \ra Q\,\Gamma\,(\Elim\Con\,\Gamma)\,A
\end{alignat*}
Note that the type of $\Elim\Ty$ refers to $\Elim\Con$, which is why
this elimination principle is called recursive-recursive (analogously
to the phrase ``inductive-inductive'').

Higher inductive types (HITs, \cite[Chapter 6]{HoTTbook}) generalise
inductive types in a different way: they allow constructors expressing
equalities of elements of the type being defined. This enables, among
others, the definition of types quotiented by a relation. For example,
the type of integers $\Int$ can be given by a constructor
$\mathsf{pair}:\Nat\ra\Nat\ra\Int$ and an equality constructor
$\mathsf{eq}:(a\,b\,c\,d:\Nat)\ra a+d=_\Nat b+c\ra
\mathsf{pair}\,a\,b=_{\Int}\mathsf{pair}\,c\,d$ targetting an equality
of $\Int$. The eliminator for $\Int$ expects a motive
$P:\Int\ra\mathsf{Type}$, a method for the $\mathsf{pair}$ constructor
$p:(a\,b:\Nat)\ra P\,(\mathsf{pair}\,a\,b)$ and a method for the
equality constructor $\mathsf{path}$. This method is a proof that
given $e:a+d=_\Nat b+c$, $p\,a\,b$ is equal to $p\,c\,d$ (the types of
which are equal by $e$). Thus the method for the equality constructor
ensures that all functions defined from the quotiented type respect
the relation. Since the integers are supposed to be a set (which
means that any two equalities between the same two integers are
equal), we would need an additional higher equality constructor
$\mathsf{trunc}:(x\,y:\Int)\ra(p\,q:x=_\Int y)\ra p=_{x=_\Int y} q$.
HITs allow equality constructors at any level. With the view of types
as spaces in mind, point constructors add points to the space,
equality constructors add paths and higher constructors add homotopies
between paths.

Not all constructor expressions make sense. For example \cite[Example
  6.13.1]{HoTTbook}, given an $f:(X:\mathsf{Type})\ra X\ra X$, suppose that an
inductive type $\mathsf{Ival}$ is generated by the point constructors
$\mathsf{a}:\mathsf{Ival}$, $\mathsf{b}:\mathsf{Ival}$ and a path
constructor $\sigma:f\,\mathsf{Ival}\,\mathsf{a}
=_{\mathsf{Ival}}f\,\mathsf{Ival}\,\mathsf{b}$. The eliminator for this type
should take a motive $P:\mathsf{Ival}\ra\mathsf{Type}$, two methods $p_a :
P\,\mathsf{a}$ and $p_b : P\,\mathsf{b}$, and a path connecting
elements of $P\,(f\,\mathsf{Ival}\,\mathsf{a})$ and
$P\,(f\,\mathsf{Ival}\,\mathsf{b})$. However it is not clear what these
elements should be: we only have elements $p_a:P\,\mathsf{a}$ and
$p_b:P\,\mathsf{b}$, and there is no way in general to transform these
to have types $P\,(f\,\mathsf{Ival}\,\mathsf{a})$ and
$P\,(f\,\mathsf{Ival}\,\mathsf{b})$.

Another invalid example is an inductive type $\mathsf{Neg}$ with a
constructor $\con:(\mathsf{Neg} \ra \bot) \ra \mathsf{Neg}$ where $\bot$
is the empty type. An eliminator for this type should (at least) yield
a projection function $\proj: \mathsf{Neg} \ra (\mathsf{Neg} \ra
\bot)$. Given this, we can define $u :\equiv \con\,(\lambda x
. \proj\, x\,x):\mathsf{Neg}$ and then derive $\bot$ by
$\proj\,u\,u$. The existence of $\mathsf{Neg}$ would make the type
theory inconsistent. A common restriction to avoid such situations is
\emph{strict positivity}. It means that the type being defined cannot
occur on the left hand side of a function arrow in a parameter of a
constructor. This excludes the above constructor $\con$.

In this paper we propose a general syntax for higher
inductive-inductive types (HIITs) which includes the above positive
examples and excludes the negative ones. Our syntax for HIITs allows
any number of inductive-inductive sorts, possibly infinitary higher
constructors of any dimension and restricts constructors to strictly
positive ones. It also allows free usage of $\J$ and $\refl$ in HIIT
specifications.  We also show how to derive the types of the
eliminators and computation rules from the type formation rules and
constructors.

The core idea is to represent HIIT specifications as contexts in a
domain-specific type theory which we call the \emph{theory of
  codes}. A context in this theory can be seen as a \emph{code} for a
HIIT, similarly to how a container \cite{abbot05containers} can be
seen as a code for a simple inductive type. Type formers in the theory
of codes are restricted in order to enforce strict positivity. For
example, natural numbers are defined as the three-element context
\[
  Nat:\U,\,\,\, zero:\underline{Nat},\,\,\, suc : Nat \ra \underline{Nat}
\]
where $Nat$, $zero$ and $suc$ are simply variable names, and underlining
denotes $\El$ (decoding) for the Tarski-style universe $\U$.

We use a variant of Bernardy et al.'s logical predicate translation
\cite{bernardy2010parametricity} to derive the types of motives and
methods, and a logical relation translation to derive the types of the
eliminators and computation rules. The target of these translations is
a type theory with a predicative hierarchy of Russell-style universes
closed under $\Pi$, $\Sigma$, the equality (identity) type
$\blank=\blank$ and the unit type $\top$. The source type theory is
the target type theory extended with rules for the theory of codes.

To our knowledge, this is the first proposal for a definition of
HIITs. Proving the existence of the HIITs thus specified is left as
future work.

\subsection{Overview of the paper}

We start by describing the target type theory in Section
\ref{sec:target}. In Section \ref{sec:source}, we define the source
type theory. The source type theory is the target theory extended with
the theory of codes, i.e.\ the rules to describe HIITs. We also
provide several examples of HIIT definitions. In Section
\ref{sec:translation} we define three syntactic translations from the
source to the target theory, each depending on the previous one:
first, we compute the types of type formation rules and constructors
(Section \ref{sec:c}); then, assuming the constructors exist, we
compute the types of motives and methods (Section \ref{sec:m});
finally we compute the types of the eliminators together with their
computation rules (Section \ref{sec:e}). To illustrate these
operations, we show how they compute on a few example codes: natural
numbers, the circle, indexed W-types and the two-dimensional sphere
(Appendix \ref{sec:app}). In Section \ref{sec:adding} we add the
pieces together by specifying what it means for the target type theory
to support HIITs. Section \ref{sec:formalisation} describes the
formalisation and a Haskell implementation. We conclude in Section
\ref{sec:summary}.

\subsection{Related work}

Inductive types can be specified using external syntactic schemes or
internal codes. In the former case the type theory is extended with
derivation rules specifying inductive types. In the latter case there
is an internal type of codes such that each code represents a valid
inductive type, and actual types are produced from codes by decoding
functions. Our development uses the former approach.

External schemes for inductive families are given in
\cite{Dybjer97inductivefamilies,paulinmohring}, for
inductive-recursive types in \cite{dybjer00ir}. A symmetric scheme for
both inductive and coinductive types is given in
\cite{henning}. Basold et al. \cite{niels} define an external
syntactic scheme for higher inductive types with only 0-constructors
and compute the types of elimination principles. In \cite{nielsmsc} a
semantics is given for the same class of HITs but with no recursive
equality constructors. Dybjer and Moeneclaey define a syntactic scheme
for finitary HITs and show their existence in a groupoid model
\cite{moeneclaey}.

Internal codes for simple inductive types such as natural numbers,
lists or binary trees can be given by containers which are decoded to
W-types \cite{abbot05containers}. Morris and Altenkirch
\cite{morris09indexed} extend the notion of container to that of
indexed container which specifies indexed inductive types. Codes for
inductive-recursive types are given in
\cite{Dybjer99afinite}. Inductive-inductive types were introduced by
Forsberg together with an internal coding scheme
\cite{forsberg-phd}. Sojakova \cite{sojakova} defines a subset of HITs
called W-suspensions by an internal coding scheme similar to
W-types. She proves that the induction principle is equivalent to
homotopy initiality.

Quotient types \cite{hofmann95extensional} are precursors of higher
inductive types (HITs). The notion of HIT first appeared in
\cite{HoTTbook}, however only through examples and without a general
definition.  Lumsdaine and Shulman give a general specification of
models of type theory supporting higher inductive types
\cite{lumsdaineShulman}. They introduce the notion of cell monad with
parameters and characterise the class of models which have intial
algebras for a cell monad with parameters. Kraus \cite{krausprop} and
Van Doorn \cite{doorn} construct propositional truncation as a
sequential colimit. The schemes mentioned so far do not support higher
inductive-inductive types.

The closest to our work is the article of Altenkirch et
al. \cite{gabe} which gives a categorical specification of quotient
inductive-inductive types (QIITs), i.e.\ set-truncated higher
inductive-inductive types. Sorts are specified as a list of functors
into $\Set$ where the domain of the functor is a category constructed
from results of the previous functors, thus encoding dependencies of
later sorts on previous ones. The constructors are specified mutually
with their category of algebras and underlying carrier functor. The
specification supports set-level equality constructors. From a
specification of a QIIT they derive the type of the eliminator and
show that this corresponds to initiality.

The logical predicate syntactic translation was introduced by Bernardy
et al. \cite{bernardy2010parametricity}. The idea that a context can
be seen as a definition of an inductive type and the logical predicate
translation can be used to derive the types of motives and methods was
described in \cite[Section 5.3]{ttintt}. Logical relations are used to
derive the computation rules in \cite[Section 4.3]{kaposi-phd},
however only for closed QIITs. Syntactic translations in the context
of the calculus of inductive constructions are discussed in
\cite{next700}. Logical relations and parametricity can also be used
to justify the \emph{existence} of inductive types in a type theory
with an impredicative universe \cite{atkey}. In contrast, we only use
logical relations to \emph{describe} HIITs.

%=====================================================================

\section{Target type theory}
\label{sec:target}

In this section we describe the target type theory. It is the target
of our translations, and it also serves as a source of constants which
are external to the HIIT being defined. It has Russell-style
universes, $\Pi$, $\Sigma$, equality (identity) and unit type. Our
notation is close to Agda's: we use named variables, terms are
identified up to $\alpha$-conversion, substitution and weakening are
implicit. To distinguish notation from the theory of codes described
in the next section, we write term formers and metavariables in
{\color{BrickRed}brick red} colour. We have the following judgement
kinds.
\begin{alignat*}{4}
  & \vdashh \hat{\Gamma} && \text{$\hat{\Gamma}$ is a valid target context} \\
  & \hat{\Gamma}\vdashh \hat{t} \in \hat{A} \hspace{3em} && \text{the target term $\hat{t}$ has target type $\hat{A}$ in target context $\hat{\Gamma}$}
\end{alignat*}
We only describe the target type theory in informal English instead of
writing down all the rules, since they are standard. See
\cite[Appendix A.2]{HoTTbook} for a formal treatment.

Context extension is written $\hat{\Gamma}\commah\hat{x}\in
\hat{A}$. We have a cumulative hierarchy of universes
$\Type_{i}$.

Dependent function space is denoted $(\hat{x}\in \hat{A})\rah
\hat{B}$. We write $\hat{A}\rah \hat{B}$ if $\hat{B}$ does not depend
on $\hat{x}$, and $\rah $ associates to the right, $(\hat{x}\in
\hat{A})(\hat{y}\in \hat{B})\rah \hat{C}$ abbreviates $(\hat{x}\in
\hat{A})\rah (\hat{y}\in \hat{B})\rah \hat{C}$ and
$(\hat{x}\,\hat{y}\in \hat{A})\rah \hat{B}$ abbreviates $(\hat{x}\in
\hat{A})(\hat{y}\in \hat{A})\rah \hat{B}$. We write $\hat{\lambda}
\hat{x}. \hat{t}$ for abstraction and $\hat{t}\,\hat{u}$ for
left-associative application.

$(\hat{x}\in \hat{A})\timesh \hat{B}$ stands for $\Sigma$ types,
$\hat{A}\timesh \hat{B}$ for the non-dependent version and $\timesh$
associates to the left. The constructor for $\Sigma$ types is denoted
$(\hat{t}\commah \hat{u})$ with eliminators $\hat{\proj_1}$ and
$\hat{\proj_2}$. Both $\Pi$ and $\Sigma$ have definitional $\beta$ and
$\eta$ rules.

The equality (identity) type for a type $\hat{A}$ and elements
$\hat{t}\in \hat{A}$, $\hat{u}\in \hat{A}$ is denoted
$\hat{t}\eqh_{\hat{A}}\hat{u}$ and comes with the constructor
$\hat{\refl}_{\hat{t}}$ and eliminator $\hat{\J}$ with definitional
$\beta$-rule. The notation is
$\hat{\J}_{\hat{A}\,\hat{t}\,\hat{P}}\,\hat{pr}\,_{\hat{u}}\,\hat{eq}$
for $\hat{t}\in \hat{A}$, $\hat{P}\in (\hat{x}\in \hat{A})\rah
\hat{t}\eqh_\hat{A}\hat{x}\rah \Type_{i}$, $\hat{pr}\in
\hat{P}\,\hat{t}\,\hat{\refl}$ and $\hat{eq} \in
\hat{t}\eqh_\hat{A}\hat{u}$. Sometimes we omit parameters in subscripts.

We will use the following functions defined using $\hat{\J}$ in the
standard way. We write $\TR_{\hat{P}}\,\hat{e}\,\hat{t}\in
\hat{P}\,\hat{v}$ for transport of $\hat{t} \in \hat{P}\,\hat{u}$
along $\hat{e} \in \hat{u}\eqh \hat{v}$. We write
$\ap\,\hat{f}\,\hat{e}\in\hat{f}\,\hat{u}\eqh\hat{f}\,\hat{v}$ for
$\hat{f}:\hat{A}\rah\hat{B}$ and $\hat{e}\in\hat{u}\eqh\hat{v}$,
$\apd\,\hat{f}\,\hat{e} \in \TR_{\hat{P}}\,\hat{e}\,(\hat{f}\,\hat{u})
\eqh \hat{f}\,\hat{v}$ for $\hat{f}\in(\hat{x}\in \hat{A})\rah
\hat{B}$ and $\hat{e} \in \hat{u}\eqh \hat{v}$.

The unit type is denoted $\hat{\top}$ with constructor $\tt$.

%=====================================================================

\section{Source type theory}
\label{sec:source}

\begin{figure}
(1) Contexts and variables
\[
\begin{gathered}
  \infer{\hat{\Gamma}\vdash\cdot}{\vdashh\hat{\Gamma}}
\end{gathered}
\hspace{2em}
\begin{gathered}
  \infer{\hat{\Gamma}\vdash\Delta,x:A}{\targetass\Delta\vdash A}
\end{gathered}
\hspace{2em}
\begin{gathered}
  \infer{\targetass\Delta,x:A\vdash x : A}{\targetass\Delta\vdash A}
\end{gathered}
\hspace{2em}
\begin{gathered}
  \infer{\targetass\Delta,y:B\vdash x : A}{\targetass\Delta\vdash x : A && \targetass\Delta\vdash B}
\end{gathered}
\]
(2) Universe
\[
\begin{gathered}
  \infer{\targetass\Delta \vdash \U}{\targetass\vdash\Delta}
\end{gathered}
\hspace{2em}
\begin{gathered}
  \infer{\targetass\Delta \vdash \underline{a}}{\targetass\Delta \vdash a : \U}
\end{gathered}
\]
(3) Inductive parameters
\[
\begin{gathered}
  \infer{\targetass\Delta \vdash (x:a)\ra B}{\targetass\Delta \vdash a : \U && \targetass\Delta,x:\underline{a} \vdash B}
\end{gathered}
\hspace{2em}
\begin{gathered}
  \infer{\targetass\Delta \vdash t\, u : B[x \mapsto u]}{\targetass\Delta \vdash t : (x:a)\ra B && \targetass\Delta \vdash u : \underline{a}}
\end{gathered}
\]
(4) Equality
\[
\begin{gathered}
  \infer{\targetass\Delta \vdash t=_a u : \U}{\targetass\Delta \vdash a : \U && \targetass\Delta \vdash t : \underline{a} && \targetass\Delta \vdash u : \underline{a}}
\end{gathered}
\hspace{2em}
\begin{gathered}
  \infer{\targetass\Delta \vdash \refl : \underline{t=_a t}}{\targetass\Delta \vdash t : \underline{a}}
\end{gathered}
\]

\[
\infer{\targetass\Delta \vdash \J_{a\,t\,\,(x.z.p)}\,pr\,_u\,eq : \underline{p[x\mapsto u, z\mapsto eq]}}
      {\begin{array}{l}
          \targetass\Delta \vdash t : \underline{a} \\
          \targetass\Delta,x:\underline{a},z:\underline{t=_a x}\vdash p : \U \\
          \targetass\Delta \vdash pr : \underline{p[x\mapsto t, z\mapsto \refl]} \\
          \targetass\Delta \vdash u : \underline{a} \\
          \targetass\Delta \vdash eq : \underline{t=_a u}
      \end{array}}
      \]

\[
\infer{\targetass\Delta \vdash \J\beta_{a\,t\,\,(x.z.p)}\,pr:\underline{(\J_{a\,t\,\,(x.z.p)}\,pr\,_t\,\refl) =_{p[x\mapsto t, z\mapsto \refl]} pr}}
      {\targetass\Delta \vdash t : \underline{a}
        && \targetass\Delta,x:\underline{a},z:\underline{t =_a x}\vdash p : \U
        && \targetass\Delta \vdash pr : \underline{p[x\mapsto t, z\mapsto \refl]}
      }
\]
(5) Non-inductive parameters
\[
\begin{gathered}
  \infer{\hat{\Gamma}\semicol\Delta \vdash (\hat{x}\in \hat{A})\ra B}{\hat{\Gamma}\vdashh\hat{A}\in\Type_0 && \hat{\Gamma}\semicol\vdash\Delta && (\hat{\Gamma}\commah\hat{x}\in \hat{A})\semicol \Delta \vdash B}
\end{gathered}
\]

\[
\begin{gathered}
  \infer{\hat{\Gamma}\semicol\Delta \vdash t\, \hat{u} : B[\hat{x}\mapsto \hat{u}]}{\hat{\Gamma}\semicol\Delta \vdash t : (\hat{x}\in \hat{A})\ra B && \hat{\Gamma}\vdashh \hat{u} \in \hat{A}}
\end{gathered}
\]
(6) Infinitary parameters
\[
\begin{gathered}
  \infer{\hat{\Gamma}\semicol\Delta \vdash (\hat{x}\in \hat{A})\ra b : \U}{\hat{\Gamma}\vdashh\hat{A}\in\Type_0 && \hat{\Gamma}\semicol\vdash\Delta && (\hat{\Gamma}\commah\hat{x}\in \hat{A})\semicol \Delta \vdash b : \U}
\end{gathered}
\]

\[
\begin{gathered}
  \infer{\hat{\Gamma}\semicol\Delta \vdash t\, \hat{u} : \underline{b[\hat{x}\mapsto \hat{u}]}}
        {\hat{\Gamma}\semicol\Delta \vdash t : \underline{(\hat{x}\in \hat{A})\ra b} && \hat{\Gamma}\vdash \hat{u} \in \hat{A}}
\end{gathered}
\]
\caption{The theory of HIIT codes (part of the source type
  theory). Substitution and weakening are implicit, we assume fresh
  names everywhere and consider $\alpha$-convertible terms equal. The
  $\hat{\Gamma};$ assumptions are not used or changed in parts
  (1)--(4).}
\label{fig:codes}
\end{figure}

The source type theory is the target type theory extended with the
following judgement kinds.
\begin{alignat*}{4}
  & \hat{\Gamma}\vdash\Delta && \text{$\Delta$ is a context in the target context $\hat{\Gamma}$}\\
  & \hat{\Gamma}\semicol\Delta\vdash A && \text{$A$ is a type in context $\Delta$ and target context $\hat{\Gamma}$} \\
  & \hat{\Gamma}\semicol\Delta\vdash t : A \hspace{3em} && \text{$t$ is a term of type $A$ in context $\Delta$ and target context $\hat{\Gamma}$}
\end{alignat*}
We name the subset of rules of the source theory which derives these
judgements \emph{the theory of codes}. The derivation rules are listed
in figure \ref{fig:codes}. A context $\Delta$ for which
$\hat{\Gamma}\vdash\Delta$ can be derived represents a specification of a
HIIT.

Although every judgement is valid up to a context in the target
type theory, note that none of the rules in (1)--(4) depend on or
change these assumptions, so they can be safely ignored until part
(5). We explain the rules in order.

(1) The rules for context formation and variables are standard. We
assume fresh names everywhere to avoid name capture. Note that
weakening is implicit.

(2) There is a universe $\U$, with decoding written as an underline
(usually $\El$ in the literature). Type formation rules will target
$\U$. With this part of the syntax we can already define contexts
specifying the empty type, unit type and booleans:
\[
\cdot,\,\,\,Empty:\U \hspace{3em} \cdot,\,\,\,Unit:\U,\,\,\,tt:\underline{Unit} \hspace{3em} \cdot,\,\,\,Bool:\U,\,\,\,true:\underline{Bool},\,\,\,false:\underline{Bool}
\]

(3) We have a function space with small domain and large
codomain. This can be used to add inductive arguments to type
formation rules and constructors. As $\U$ is not closed under this
function space, these function types cannot (recursively) appear in
inductive arguments, which ensures strict positivity. When the
codomain does not depend on the domain, $a\ra B$ can be written
instead of $(x:a)\ra B$.

Now we can specify the natural numbers as a context:
\[
\cdot,\,\,\,Nat : \U,\,\,\,zero:\underline{Nat},\,\,\,suc:Nat\ra\underline{Nat}
\]
We can also encode inductive-inductive definitions such as the
fragment of the well-typed syntax of a type theory mentioned in the
introduction:
\begin{alignat*}{5}
  & \cdot,\,\,\,Con:\U,\,\,\,Ty:Con\ra\U,\,\,\,\bullet:\underline{Con},\,\,\,\blank\rhd\blank:(\Delta:Con)\ra Ty\,\Delta\ra\underline{Con}, \\
  & U : (\Delta:Con)\ra \underline{\Ty\,\Delta},\,\,\,\Pi:(\Delta:Con)(A:Ty\,\Delta)(B:Ty\,(\Delta\rhd A))\ra\underline{Ty\,\Delta}
\end{alignat*}

(4) $\U$ is closed under the equality type, with eliminator $\J$ and a
weak (non-definitional) $\beta$-rule. Weakness is required because the
syntactic translation $\blank^\E$ defined in Section \ref{sec:e} does
not preserve this $\beta$-rule strictly. Adding equality to the theory
of codes allows higher constructors and inductive equality parameters
as well. We can now define the circle HIT as the following context:
\[
\cdot,\,\,\,S^1:\U,\,\,\,base:\underline{S^1},\,\,\,loop:\underline{base =_{S^1} base}
\]
The $\J$ rule allows constructors to mention operations on paths as
well. For instance, the definition of the torus depends on path
composition, which can be defined using $\J$: given $p:\underline{t=_a
  u}$ and $q:\underline{u=_a v}$, $p \sqcdot q$ abbreviates
$\J_{a\,u\,x.z.(t=x)}\,p\,_v\,q : \underline{t=_a v}$. The torus is
given as follows.
\begin{alignat*}{5}
  & \cdot,\,\,\,T^2:\U,\,\,\,b : \underline{T^2},\,\,\, p:\underline{b =_{T^2} b},\,\,\,q:\underline{b=_{T^2} b},\,\,\, t:\underline{p\sqcdot q=_{(b=_{T^2} b)} q\sqcdot p}
\end{alignat*}
With the equality type at hand, we can define a full well-typed syntax
of type theory as given e.g. in \cite{ttintt} as an inductive type
(see the examples in the formalisation described in Section
\ref{sec:formalisation}).

So far we were only able to define closed HIITs, which excludes lists
of a given type or the integers as given in the introduction. This is
where we need the target theory to be included in the source theory. A
context $\Delta$ for which $\hat{\Gamma}\vdash\Delta$ holds can be
seen as a specification of an inductive type which depends on
$\hat{\Gamma}$. In the case of lists, $\hat{\Gamma}$ will be
$\hat{A}\in\Type_0$. In the case of integers, we need
$\hat{Nat}:\Type_0$ and $\blank\hat{\mathbin{+}}\blank:\hat{Nat}\rah
\hat{Nat}\rah \hat{Nat}$ from $\hat{\Gamma}$.

(5) We have a function space where the domain is a type in the target
theory. We distinguish it from (3) by using red brick $\in$ instead of
$:$ in the domain specification. We specify lists and the integers as follows.
\begin{alignat*}{4}
  & \hat{A}\in\Type_0 && \vdash\,\cdot,\,\, && List:\U,\,\,\,nil:\underline{List},\,\,\,cons:(\hat{x}\in \hat{A})\ra List\ra\underline{List} \\
  & \hat{\Gamma} && \vdash\,\cdot,\,\,&& Int:\U,\,\,\,pair:(\hat{x}\,\hat{y}\in\hat{Nat})\ra\underline{Int},\,\,\, \\
  & && && eq:(\hat{a}\,\hat{b}\,\hat{c}\,\hat{d}\in \hat{Nat})(\hat{p}\in \hat{a}\hat{\mathbin{+}}\hat{d}\eqh_{\hat{Nat}} \hat{b}\hat{\mathbin{+}}\hat{c})\ra \underline{pair\,\hat{a}\,\hat{b}=_{Int}\mathsf{pair}\,\hat{c}\,\hat{d}}, \\
  & && && trunc:(x y : Int)(p\,q : a=_{Int} b)\ra \underline{p=_{x=_{Int} y} q}
\end{alignat*}
In the case of integers, $\hat{\Gamma}$ is
$\hat{Nat}\in\Type_0\hat{,}\,\blank\hat{\mathbin{+}}\blank\in\hat{Nat}\rah
\hat{Nat}\rah\hat{Nat}$, or alternatively, we could require natural
numbers in the target theory. As another example, propositional
truncation for a type $\hat{A}$ is specified as follows.
\[
\hat{A}\in\Type_0\vdash\, \cdot,\,\,\,tr:\U,\,\,\,emb : (\hat{x}\in \hat{A})\ra \underline{tr},\,\,\,eq:(x\,y:tr)\ra \underline{x=_{tr} y}
\]
The smallness of $\hat{A}$ is required in (5). It is possible to
generalize the syntax of HIITs to arbitrary universe levels, but it is
not essential to the current development. Note that the (5) function
space preserves strict positivity, since in the target theory there is
no way to recursively refer to the inductive type \emph{being
  defined}. The situation is analogous to the case of $W$-types
\cite{abbot05containers}, where shapes and positions can contain
arbitrary types but they cannot recursively refer to the $W$-type
being defined.

(6) $\U$ is also closed under a function space where the domain is a
target theory type and the codomain is a small source theory type. We
overload the application notation for non-inductive parameters, as it
is usually clear from context which application is meant. The rules
allow types with infinitary constructors, for example trees containing
$\hat{A}$-elements at the leaves and branching by $\hat{B}$ (which
could be an infinite type):
\[
\hat{A}\in\Type_0\commah\hat{B}\in\Type_0\vdash\,\cdot,\,\,\,T:\U,\,\,\,leaf:(\hat{x}\in \hat{A})\ra\underline{T},\,\,\,node:((\hat{x}\in \hat{B})\ra T)\ra\underline{T}
\]
Here, $leaf$ has a function type (5) and $node$ has a function type
(3) with a function type (6) in the domain. More generally, we can
define $W$-types \cite{abbot05containers} as follows. $\hat{S}$
describes the ``shapes'' of the constructors and $\hat{P}$ the
``positions'' where recursive arguments can appear.
\[
\hat{S}\in\Type_0\commah\hat{P}:\hat{S}\rah\Type_0\vdash\,\cdot,\,\,\,W:\U,\,\,\,sup: (\hat{s} \in \hat{S})\ra((\hat{p} \in \hat{P}\,\hat{s})\ra W)\ra \underline{W}
\]
For a more complex infinitary example, see the definition of Cauchy
reals in \cite[Definition 11.3.2]{HoTTbook}. It can be also found as
an example file in our Haskell implementation.

The invalid examples $\mathsf{Ival}$ and $\mathsf{Neg}$ cannot be
encoded by the theory of codes. For $\mathsf{Ival}$, we can go as far
as
\[
\cdot,\,\,\,{Ival}:\U,\,\,\,a:\underline{{Ival}},\,\,\,b:\underline{{Ival}},\,\,\,\sigma:\underline{?  =_{{Ival}} ?}.
\]
The first argument of the function
$\hat{f}\in(\hat{X}:\Type)\rah\hat{X}\rah\hat{X}$ is a target theory
type, but we only have ${Ival}:\U$ in the theory of
codes. $\mathsf{Neg}$ cannot be typed because the first parameter of
the constructor $\mathsf{con}$ is a function from a small type to a
target theory type, and no such functions can be formed.

%=====================================================================

\section{A syntactic translation from the source to the target theory}
\label{sec:translation}

An inductive type is specified by a context in the theory of codes
defined in the previous section. In this section we define the
$\blank^\C$, $\blank^\M$ and $\blank^\E$ operations, which work as follows
on the code for natural numbers.
\begin{alignat*}{5}
  & && (\cdot,\,Nat : \U,\,zero:\underline{Nat},\,suc:Nat\ra\underline{Nat})^\C \\
  & \equiv \,\, && \hat{\top}\timesh(\hat{n}\in\Type_0)\timesh(\hat{z}\in \hat{n})\timesh(\hat{n}\rah \hat{n}) \\
  & && (\cdot,\,Nat : \U,\,zero:\underline{Nat},\,suc:Nat\ra\underline{Nat})^\M\,(\tt\commah\hat{n}\commah \hat{z}\commah \hat{s}) \\
  & \equiv \,\, && \hat{\top}\timesh(\hat{n^M}\in \hat{n} \rah \Type_{i})\timesh(\hat{z^M}\in \hat{n^M}\,\hat{z})\timesh \big((\hat{x}\in \hat{n})\rah  \hat{n^M}\,\hat{x}\rah  \hat{n^M}\,(\hat{s}\,\hat{x})\big) \\
  & && (\cdot,Nat : \U,zero:\underline{Nat},suc:Nat\ra\underline{Nat})^\E\,(\tt\commah \hat{n}\commah \hat{z}\commah \hat{s})\,(\tt\commah\hat{n^M}\commah \hat{z^M}\commah \hat{s^M}) \\
  & \equiv \,\, && \hat{\top}\timesh\big(\hat{n^E} \in (\hat{x}\in \hat{n})\rah  \hat{n^M}\,\hat{x}\big)\timesh(\hat{z^E} \in \hat{n^E}\,\hat{z}\eqh \hat{z^M}) \timesh \big((\hat{x}\in \hat{n})\rah  \hat{n^E}\,(\hat{s}\,\hat{x})\eqh \hat{n^M}\,\hat{x}\,(\hat{n^E}\,\hat{x})\big)
\end{alignat*}
The {\color{BrickRed}brick red} coloured result of $\blank^\C$ gives
the types of the type formation rule and constructors as an iterated
$\Sigma$-type. Assuming the existence of constructors, $\blank^\M$
returns the types of motives and methods. Assuming the existence of
constructors and the corresponding motives and methods, $\blank^\E$
returns the types of the eliminator and computation rules as target
theory equalities $\eqh$. The notation above denotes
\emph{left}-nested iterated $\Sigma$-types. A more precise
presentation would replace each variable with a projection from the
preceding $\Sigma$-type. We use this notation in order to reduce
visual clutter.

Each context entry in the theory of codes specifies a type formation
rule or a constructor. In general, the last component of a type in a
context entry is of three possible forms: it is either $\U$,
$\underline{a}$ for some neutral $a$, or $\underline{t=_a u}$. The
following table summarizes the results of the various translations in
the mentioned three cases:
\begin{alignat*}{5}
  & \text{return type}\hspace{1em} && \blank^\C && \blank^\M && \blank^\E \\\hline
  & \U && \text{type formation rule}\hspace{1em} && \text{motive} && \text{eliminator} \\
  & \underline{a} && \text{point constructor}\hspace{1em} && \text{method} && \text{computation rule} \\
  & \underline{t=_a u} && \text{path constructor} && \text{method expressing} && \text{higher computation rule} \\
  & && && \text{preservation of equality}\hspace{1em}
\end{alignat*}
Note that there is no syntactic distinction between the three kinds of
constructors above. Any number of them can be introduced in any order,
and each constructor can refer to any previous one. A distinguishing
feature of our approach is the utilisation of universes instead of
structural rules to introduce new sorts and to ensure strict
positivity.

The $\blank^\C$, $\blank^\M$ and $\blank^\E$ operations are defined by
induction on the derivations of the source syntax. The operations are
identity on derivations of target contexts and target terms (of the
forms $\vdashh\hat{\Gamma}$ and $\hat{\Gamma}\vdashh \hat{t}\in
\hat{A}$) and derive target terms from theory of codes contexts, types
and terms (of the forms $\hat{\Gamma}\vdash\Delta$,
$\hat{\Gamma}\semicol\Delta\vdash A$ and
$\hat{\Gamma}\semicol\Delta\vdash t:A$, respectively). We only present
the non-identity parts with pattern matching notation, describing how
a context, type or a term in the theory of codes is converted to a
term in the target theory.

All three operations respect definitional equality. This amounts to
preserving equalities of the substitution calculus, as there are no
$\beta$-rules introduced in the theory of codes.

%---------------------------------------------------------------------

\subsection{Type formation rules and constructors}
\label{sec:c}

Given a context in the theory of codes, $\blank^\C$ returns the types
of type formation rules and constructors as an iterated $\Sigma$-type
in the target theory. It is specified as follows.
\[
\infer{\hat{\Gamma}\vdashh\Delta^\C\in\Type_1}{\hat{\Gamma}\vdash\Delta}
\hspace{2em}
\infer{\hat{\Gamma}\vdashh A^\C \in \Delta^\C\rah \Type_1}{\hat{\Gamma}\semicol\Delta\vdash A}
\hspace{2em}
\infer{\hat{\Gamma}\vdashh t^\C \in (\hat{\gamma}\in\Delta^\C)\rah  A^\C\,\hat{\gamma}}{\hat{\Gamma}\semicol\Delta\vdash t : A}
\]
Given a context depending on the target context $\hat{\Gamma}$,
$\blank^\C$ returns a type in $\hat{\Gamma}$. Given a type in context
$\Delta$ it returns a family over $\Delta^\C$. A term is interpreted
by a dependent function in the target theory. The implementation of
$\blank^\C$ is essentially the standard model, where everything is
interpreted by its {\color{BrickRed}brick red} counterpart, except for
contexts which are interpreted as iterated $\Sigma$-types.\begingroup
\allowdisplaybreaks
\begin{alignat*}{5}
  & \cdot^\C && :\equiv \hat{\top} \\
  & (\Delta,x:A)^\C && :\equiv (\hat{\gamma}\in\Delta^\C)\timesh A^\C\,\hat{\gamma} \\
  & x^\C\,\hat{\gamma} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma} \\ % TODO: write De Bruijn versions
  & \U^\C\,\hat{\gamma} && :\equiv \Type_0 \\
  & (\underline{a})^\C\,\hat{\gamma} && :\equiv a^\C\,\hat{\gamma} \\
  & ((x:a)\ra B)^\C\,\hat{\gamma} && :\equiv (\hat{x}\in a^\C\,\hat{\gamma})\rah  B^\C\,(\hat{\gamma}\commah\hat{x}) \\
  & (t\,u)^\C\,\hat{\gamma} && :\equiv (t^\C\,\hat{\gamma})\,(u^\C\,\hat{\gamma}) \\
  & ((\hat{x}\in \hat{A})\ra B)^\C\,\hat{\gamma} && :\equiv (\hat{x}\in \hat{A})\rah  B^\C\,\hat{\gamma} \\
  & (t\,\hat{u})^\C\,\hat{\gamma} && :\equiv (t^\C\,\hat{\gamma})\,\hat{u} \\
  & (t=_a u)^\C\,\hat{\gamma} && :\equiv t^\C\,\hat{\gamma} \eqh u^\C\,\hat{\gamma} \\
  & \refl^\C\,\hat{\gamma} && :\equiv \hat{\refl} \\
  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\C\,\hat{\gamma} && :\equiv \hat{\J}_{(a^\C\,\hat{\gamma})\,(t^\C\,\hat{\gamma})\,(\hat{\lambda} \hat{x}\,\hat{z}.p^\C\,(\hat{\gamma}\commah\hat{x}\commah\hat{z}))}\,(pr^\C\,\hat{\gamma})\,_{(u^\C\,\hat{\gamma})}\,(eq^\C\,\hat{\gamma}) \\
  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\C\,\hat{\gamma} && :\equiv \hat{\refl} \\
  & ((\hat{x}\in \hat{A})\ra b)^\C\,\hat{\gamma} && :\equiv (\hat{x}\in \hat{A})\rah  b^\C\,\hat{\gamma} \\
  & (t\,\hat{u})^\C\,\hat{\gamma} && :\equiv (t^\C\,\hat{\gamma})\,\hat{u}
\end{alignat*}\endgroup

For example, $\blank^\C$ acts as follows on the topological circle:
\begin{alignat*}{5}
  & && (\cdot,\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\C \equiv \hat{\top}\timesh(\hat{S^1}\in\Type_0)\timesh(\hat{b}\in \hat{S^1})\timesh(\hat{loop\in b=b})
\end{alignat*}
The resulting left-nested $\Sigma$ type could be written explicitly as below. We shall keep to the more readable notation from now on.
\begin{alignat*}{5}
  & \Big(\hat{x''}\in\big(\hat{x'}\in(\hat{x}\in\hat{\top})\timesh\Type_0\big)\timesh\hat{\proj_2}\,\hat{x'}\Big)\timesh(\hat{\proj_2}\,\hat{x''}\eqh \hat{\proj_2}\,\hat{x''})
\end{alignat*}

%---------------------------------------------------------------------

\subsection{Motives and methods}
\label{sec:m}

Given a code for an inductive type and the constructors specified by
the code, the operation $\blank^\M$ returns the induction motives and methods.

$\blank^\M$ is a variant of the unary logical predicate translation of
Bernardy et al.\ \cite{bernardy12parametricity}. We fix a level $i$
for the universe we would like to eliminate into. For each context
$\Delta$, $\Delta^\M$ is a predicate over the standard interpretation
$\Delta^\C$. For a type $\Delta\vdash A$, $A^\M$ is a predicate over
$A^\C$, which also depends on $\hat{\gamma}\in\Delta^\C$ and a witness
of $\Delta^\M\,\hat{\gamma}$. All of these may refer to a target
theory context $\hat{\Gamma}$.
\[
\infer{\hat{\Gamma}\vdashh\Delta^\M \in \Delta^\C\rah \Type_{i+1}}{\hat{\Gamma}\vdash\Delta}
\hspace{2em}
\infer{\hat{\Gamma}\vdashh A^\M \in (\hat{\gamma}\in\Delta^\C)\rah \Delta^\M\,\hat{\gamma}\rah  A^\C\,\hat{\gamma}\rah \Type_{i+1}}{\hat{\Gamma}\semicol\Delta\vdash A}
\]
For a term $t$, $t^\M$ witnesses that the predicate corresponding to
its type holds for $t^\C$. This is often called a \emph{fundamental
  theorem} in the literature on logical predicates.
\[
\infer{\hat{\Gamma}\vdashh t^\M \in (\hat{\gamma}\in\Delta^\C)(\hat{\gamma^M}\in\Delta^\M\,\hat{\gamma})\rah  A^\M\,\hat{\gamma}\,\hat{\gamma^M}\,(t^\C\,\hat{\gamma})}{\hat{\Gamma}\semicol\Delta\vdash t : A}
\]
We introduce the following shorthand: $t\,\hat{\gamma}\,\hat{\gamma^M}$ is
abbreviated as $t\,\hat{\gamma^2}$ for some $t$ expression. The
implementation of $\blank^\M$ is given below.  \begingroup
\allowdisplaybreaks
\begin{alignat*}{5}
  & \cdot^\M\,\hat{\gamma} && :\equiv \hat{\top} \\
  & (\Delta,\,x:A)^\M\,(\hat{\gamma}\commah\,\hat{t}) && :\equiv (\hat{\gamma^M}\in\Delta^\M\,\hat{\gamma})\timesh A^\M\,\hat{\gamma^2}\,\hat{t} \\
  & x^\M\,\hat{\gamma^2} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^M} \\ % TODO: De Bruijn
  & \U^\M\,\hat{\gamma^2}\,\hat{A} && :\equiv \hat{A} \rah  \Type_{i} \\
  & (\underline{a})^\M\,\hat{\gamma^2}\,\hat{t} && :\equiv a^\M\,\hat{\gamma^2}\,\hat{t} \\
  & ((x:a)\ra B)^\M\,\hat{\gamma^2}\,\hat{f} && :\equiv (\hat{x}\in a^\C\,\hat{\gamma})(\hat{x^M}\in a^\M\,\hat{\gamma^2}\,\hat{x}) \rah  B^\M\,(\hat{\gamma}\commah\hat{x})\,(\hat{\gamma^M}\commah\hat{x^M})\,(\hat{f}\,\hat{x}) \\
  & (t\,u)^\M\,\hat{\gamma^2} && :\equiv (t^\M\,\hat{\gamma^2})\,( u^\C\,\hat{\gamma})\,(u^\M\,\hat{\gamma^2}) \\
  & ((\hat{x}\in \hat{A})\ra B)^\M\,\hat{\gamma^2}\,\hat{f} && :\equiv (\hat{x}\in \hat{A})\rah  B^\M\,\hat{\gamma^2}\,(\hat{f}\,\hat{x}) \\
  & (t\,\hat{u})^\M\,\hat{\gamma^2} && :\equiv t^\M\,\hat{\gamma^2}\,\hat{u} \\
  & (t=_a u)^\M\,\hat{\gamma^2}\,\hat{e} && :\equiv \TR_{(a^\M\,\hat{\gamma^2})}\,\hat{e}\,(t^\M\,\hat{\gamma^2}) \eqh u^\M\hat{\gamma^2} \\
  & (\refl_t)^\M\,\hat{\gamma^2} && :\equiv \hat{\refl}_{(t^\M\,\hat{\gamma^2})} \\
  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\M\,\hat{\gamma^2} && :\equiv \hat{\J}\,\big(\hat{\J}\,(pr^\M\,\hat{\gamma^2})\,(eq^\C\,\hat{\gamma})\big)\,(eq^\M\,\hat{\gamma^2}) \\
  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\M\,\hat{\gamma^2} && :\equiv \hat{\refl} \\
  & ((\hat{x}\in \hat{A})\ra b)^\M\,\hat{\gamma^2}\,\hat{f} && :\equiv (\hat{x}\in \hat{A})\rah  b^\M\,\hat{\gamma^2}\,(\hat{f}\,\hat{x}) \\
  & (t\,\hat{u})^\M\,\hat{\gamma^2} && :\equiv t^\M\,\hat{\gamma^2}\,\hat{u}
\end{alignat*}\endgroup
The predicate for a context is given by iterating $\blank^\M$ for its
constituent types. For a variable, the corresponding witness is looked
up from $\hat{\gamma^M}$.

The predicate for the universe, given an element of
$\hat{A}\in\U^\C\,\hat{\gamma}$ (with $\U^\C\,\hat{\gamma}\equiv\Type_0$) returns the
predicate space over $\hat{A}$. The predicate for a type $\underline{a}$ is
given by the predicate for $a$.

The predicate for a function type with small domain expresses
preservation of predicates (at the domain and codomain
types). Witnesses of application are given by recursive application of
$\blank^\M$. The definitions for the other (non-inductive) function
spaces are similar, except there is no predicate for the domain types,
and thus no witnesses are required.

The predicate for the equality type $t=_a u$, for each
$\hat{e}\in(t=_a u)^\C\,\hat{\gamma}$, i.e.\ $\hat{e}\in
t^\C\,\hat{\gamma}\eqh u^\C\,\hat{\gamma}$, says that $t^\M$ and
$u^\M$ are equal. As these have different types, we have to transport
over the original equality $\hat{e}$. The witness for $\refl$ is
reflexivity in the target theory. The interpretation of $\J$ is given
by a double $\hat{\J}$ application, which definition is sourced from
\cite{lasson}. Here, we use a shortened $\hat{\J}$ notation; we refer
to the formalisation (Section \ref{sec:formalisation}) for the
details.

Again, let us consider the circle example:
\begin{alignat*}{5}
  & && (\cdot,\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\M\,\,(\hat{\tt,\,S^1,\,b,\,loop})\\
  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{1M}}\in \hat{S^1}\rah\Type_i)\timesh(\hat{b^M}\in \hat{S^{1M}\,b})\timesh(\hat{loop^M\in \TR_{\,S^{1M}}\,loop\,b^M = b^M})
\end{alignat*}
The inputs of $\blank^\M$ here are the code for the circle (the
context in black) and a tuple of the type formation rule
$\hat{S^{1}}$, the constructor $\hat{b}$ and the equality constructor
$\hat{loop}$. It returns a family over the type $\hat{S^1}$, an
element of this family $\hat{b^M}$ at index $\hat{b}$, and and an
equality between $\hat{b^M}$ and $\hat{b^M}$ which lies over $\hat{loop}$.

%---------------------------------------------------------------------

\subsection{Eliminators and computation rules}
\label{sec:e}

The operation $\blank^\E$ yields eliminators and computation rules. It
is a generalised binary logical relation translation where the type of
the second parameter of the relation may depend on the first
parameter.

Contexts are interpreted as dependent binary relations between
constructors and methods. The universe level $i$ was previously chosen
for the $\blank^\M$ operation.
\[
\infer{\hat{\Gamma}\semicol\Delta^\E \in (\hat{\gamma}\in\Delta^\C)\rah \Delta^\M\,\hat{\gamma}\rah \Type_{i}}{\hat{\Gamma}\vdash\Delta}
\]
Types are interpreted as dependent binary relations which additionally
depend on ($\hat{\gamma}\commah\,\hat{\gamma^M}\commah\,\hat{\gamma^E}$) interpretations of the
context.
\[
\infer{\hat{\Gamma}\vdashh A^\E \in (\hat{\gamma}\in\Delta^\C)(\hat{\gamma^M}\in\Delta^\M\,\hat{\gamma})(\hat{\gamma^E}\in\Delta^\E\,\hat{\gamma}\,\hat{\gamma^M})(\hat{x}\in A^\C\,\hat{\gamma})\rah  A^\M\,\hat{\gamma}\,\hat{\gamma^M}\,\hat{x}\rah \Type_{i}}{\hat{\Gamma}\semicol\Delta\vdash A}
\]
For a term $t$, $t^\E$ witnesses that the relation corresponding to
its type holds for $t^\C$ and $t^\M$.
\[
\infer{\hat{\Gamma}\vdashh t^\E \in (\hat{\gamma}\in\Delta^\C)(\hat{\gamma^M}\in\Delta^\M\,\hat{\gamma})(\hat{\gamma^E}\in\Delta^\E\,\hat{\gamma}\,\hat{\gamma^M})\rah  A^\E\,\hat{\gamma}\,\hat{\gamma^M}\,\hat{\gamma^E}\,(t^\C\,\hat{\gamma})\,(t^\M\,\hat{\gamma}\,\hat{\gamma^M})}{\hat{\Gamma}\semicol\Delta \vdash t : A}
\]
In addition to $\hat{\gamma^2}$, we use $t\,\hat{\gamma^3}$ to abbreviate $t\,\hat{\gamma}\,\hat{\gamma^M}\,\hat{\gamma^E}$. The implementation is the following.
%% \begingroup \allowdisplaybreaks
\begin{alignat*}{5}
  & \cdot^\E\,\hat{\gamma}\,\hat{\gamma^M} && :\equiv \hat{\top} \\
  & (\Delta,\,x:A)^\E\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^M},\hat{t^M}) && :\equiv (\hat{\gamma^E}\in\Delta^\E\hat{\gamma^2})\timesh A^\E\,\hat{\gamma^3}\,\hat{t}\,\hat{t^M} \\
  & x^\E\,\hat{\gamma^3} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^E} \\ % TODO: De Bruijn
  & \U^\E\,\hat{\gamma^3}\,\hat{A}\,\hat{A^M} && :\equiv (\hat{x}\in \hat{A})\rah  \hat{A^M}\,\hat{x} \\
  & (\underline{a})^\E\,\hat{\gamma^3}\,\hat{t}\,\hat{t^M} && :\equiv a^\E\,\hat{\gamma^3}\,\hat{t} \eqh \hat{t^M} \\
  & ((x:a)\ra B)^\E\,\hat{\gamma^3}\,\hat{f}\,\hat{f^M} && :\equiv (\hat{x}\in a^\C\,\hat{\gamma})\rah  B^\E\,(\hat{\gamma,x})\,(\hat{\gamma^M,}\,a^\E\,\hat{\gamma^3}\,\hat{x})\,(\hat{\gamma^E,}\,\hat{\refl}) \\
  & && \hspace{8.4em} (\hat{f}\,\hat{x})\,\big(\hat{f^M}\,\hat{x}\,(a^\E\,\hat{\gamma^3}\,\hat{x})\big) \\
  & (t\,u)^\E\,\hat{\gamma^3} && :\equiv \hat{\J}\,(t^\E\,\hat{\gamma^3}\,( u^\C\,\hat{\gamma}))\,(u^\E\,\hat{\gamma^3}) \\
  & ((\hat{x}\in \hat{A})\ra B)^\E\,\hat{\gamma^3}\,\hat{f}\,\hat{f^M} && :\equiv (\hat{x}\in \hat{A})\rah B^\E\,\hat{\gamma^3}\,(\hat{f}\,\hat{x})\,(\hat{f^M}\,\hat{x}) \\
  & (t\,\hat{u})^\E\,\hat{\gamma^3} && :\equiv t^\E\,\hat{\gamma^3}\,\hat{u} \\
  & (t=_a u)^\E\,\hat{\gamma^3}\,\hat{e} && :\equiv \TR\,(t^\E\,\hat{\gamma^3})\big(\TR\,(u^\E\,\hat{\gamma^3})\,(\apd\,(a^\E\,\hat{\gamma^3})\,\hat{e})\big) \\
  & (\refl_t)^\E\,\hat{\gamma^3} && :\equiv \hat{\J}\,\hat{\refl}\,(t^\E\,\hat{\gamma^3}) \\
  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\E\,\hat{\gamma^3} && :\equiv \\
  & && \hspace{-10em}\hat{\J}\,\bigg(\hat{\J}\,\Big(\hat{\J}\,\big(\hat{\J}\,(\hat{\lambda}\,\hat{p^M}\,\hat{p^E}\,\hat{pr^M}\,\hat{pr^E}.\hat{pr^E})\,(t^\E\,\hat{\gamma^3})\,\\
  & && \hspace{-9em}(\hat{\mathsf{uncurry}}\,p^\M\,\hat{\gamma^2})\,(\hat{\mathsf{uncurry}}\,p^\E\,\hat{\gamma^3})\,(pr^\M\,\hat{\gamma^2})\,(pr^\E\,\hat{\gamma^3})\big)\,eq^\C\,\hat{\gamma}\Big)\,u^\E\,\hat{\gamma^3}\bigg)\,(eq^\E\,\hat{\gamma^3}) \\
  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\E\,\hat{\gamma^3} && :\equiv \\
  & && \hspace{-10em}\hat{\J}\,\big(\hat{\J}\,(\hat{\lambda}\,\hat{p^M}\,\hat{p^E}\hat{.}\,\hat{\refl})\,(t^\E\,\hat{\gamma^3})\,(\hat{\mathsf{uncurry}}\,p^\M\,\hat{\gamma^2})\,(\hat{\mathsf{uncurry}}\,p^\E\,\hat{\gamma^3})\big)\,(pr^\E\,\hat{\gamma^3})\\
  & ((\hat{x}\in \hat{A})\ra b)^\E\,\hat{\gamma^3\,f\,t} && :\equiv b^\E\,\hat{\gamma^3}\,(\hat{f}\,\hat{t}) \\
  & (t\,\hat{u})^\E\,\hat{\gamma^3} && :\equiv \ap\,(\hat{\lambda} \hat{f.}\hat{f}\,\hat{u})\,(t^\E\,\hat{\gamma^3})
\end{alignat*}
%% \endgroup

The $\U^\E$ and $(\underline{a})^\E$ definitions are the key points of
the $\blank^\E$ operation. The definitions for the other $\blank^\E$
cases are largely determined by these.

The $\U^\E$ rule yields the type of the eliminator for a type
formation rule. In the natural numbers example above, the non-indexed
$Nat : \U$ sort is interpreted as $\hat{n^E} \in (\hat{x}\in
\hat{n})\rah \hat{n^M}\,\hat{x}$. For indexed sorts, the indices are
first processed by the $\blank^\E$ cases for inductive and non-inductive
parameters, until the ultimate $\U$ return type is reached. Hence, the
eliminator for a sort is always a function.

Analogously, the $\blank^\E$ result type for a point or path constructor is
always a $\beta$-rule, i.e.\ a function type ending in an equality. To
see why, consider the $(\underline{a})^\E$ definition. It expresses
that applications of $a^\E$ eliminators must be equal to the
corresponding $\hat{t^M}$ induction methods. Hence, for path and point
constructor types, $\blank^\E$ works by first processing all inductive and
non-inductive indices, then finally returning an equality type.

Let us also consider the $((x:a)\ra B)^\E$ case for inductive
parameters. Here, we make crucial use of the fact that the domain type
$a$ is small. This provides us $a^\E\,\hat{\gamma^3\,x}$, which we use
to witness the $a^\M\,\hat{\gamma^2}\,\hat{x}$ hypothesis for
$B^\E$. Without the size restriction on inductive parameters (which
enforces strict positivity), the $\blank^\E$ operation would not be
possible at all, because $a^\E$ for a large $a$ type would be merely
an opaque relation instead of an eliminator function.

Here, we only provide abbreviated definitions for the $t\,u$, $t=_a
u$, $\refl$, $\J$ and $\J\beta$ cases. In the $\J$ case, we write
$\hat{\mathsf{uncurry}}\,p^\M$ for
$\hat{\lambda\,\gamma\,\gamma^M\,x\,x^M\,z\,z^M.}\,p^\M\,(\hat{\gamma,x,z})\,(\hat{\gamma^M,x^M,z^M})$
and analogously elsewhere. The full definitions can be found in the
Agda formalisation. The definitions are highly constrained by the
required types, and not particularly difficult to implement with the
help of a proof assistant; they all involve doing successive path
induction on all equalities available from induction hypotheses, with
appropriately generalized induction motives.

The full $(\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\E$ definition is quite
large, and, for instance, yields a very large $\beta$-rule for the
higher inductive torus definition (the reader can confirm this using
the Haskell implementation). Nevertheless, an implementation focused
on practicality may provide smaller specialized $\blank^\E$ definitions for
commonly used equality operations such as composition or inverses.

The circle example is a bit more interesting here:
\begin{alignat*}{5}
  & && (\cdot,\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\E\,\,(\hat{\tt,\,S^1,\,b,\,loop})\,(\hat{\tt,\,S^{1M},\,b^M,\,loop^M}) \\
  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{1E}}\in (\hat{x}\in\hat{S^1})\hat{\,\ra\,}\hat{S^{1M}\,x})\timesh(\hat{b^E}\in \hat{S^{1E}\,\,b = b^M})\\
  & && \hspace{0.75em}\timesh(\hat{loop^E\,\,\in\,\,} \TR\,_{(\hat{\lambda x. \TR\,_{S^{1M}}\,loop\,x\,=\,b^M})}\,\hat{b^E}\,(\TR\,_{(\hat{\lambda} \hat{x}. \TR\,_{\hat{S^{1M}}}\,\hat{loop}\,(\hat{S^{1E}\,b})\,\eqh\,\hat{x})}\,\hat{b^E}\,(\apd\,\hat{S^{1E}\,loop}))\,\\
  & && \hspace{5.5em} \hat{=\,loop^M})
\end{alignat*}
In homotopy type theory, the $\beta$-rule for $loop$ is usually just
$\apd\,\hat{S^{1E}\,loop}\,\hat{\,=\,loop^M}$, but here all
$\beta$-rules are propositional, so we need to transport with
$\hat{b^E}$ to make the equation well-typed. When computing the type
of $\hat{loop^E}$, we start with $(\underline{b
  =b})^E\,\,\hat{\gamma^3}\,\,\hat{loop}\,\,\hat{loop^M}$. Next, this
evaluates to $(b=b)^\E\,\hat{\gamma^3}\,\hat{loop}\eqh\hat{loop^M}$,
and then we unfold the left hand side to get the doubly-transported
expression in the result.

In Appendix \ref{sec:app}, we show how the elimination principle is
computed for the two-dimensional sphere.

For another example for the translations, we consider indexed W-types
which can describe a large class of inductive definitions
\cite{morris09indexed}. Suppose we are in the target context $\hat{I}
\in \Type_0\commah\hat{S} \in \Type_0\commah\hat{P} \in \hat{S} \rah
\Type_0\commah\hat{out} \in \hat{S} \rah \hat{I}\commah\hat{in} \in
(\hat{s} \in \hat{S})\rah \hat{P}\,\hat{s}\rah \hat{I}$. Then, the
code for the corresponding indexed W-type is the following:
\[
W :\equiv (\cdot,\,\,\,w:(\hat{i} \in \hat{I})\ra\U,\,\,\,sup: (\hat{s} \in \hat{S})\ra((\hat{p} \in \hat{P\,s})\ra w\,(\hat{in\,s\,p}))\ra \underline{w\,(\hat{out\,s})})
\]
We pick a universe level $j$ for elimination. The interpretations of $W$ are the following, omitting leading $\hat{\top}$ components:
\begin{alignat*}{5}
  & \rlap{$W^\C \equiv (\hat{w}\in \hat{I}\rah\Type_0)\timesh((\hat{s}\in \hat{S})\rah((\hat{p}\in \hat{P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\rah \hat{w}\,(\hat{out\,s}))$}\\
  & W^\M\,(\hat{w, sup}) && \equiv\, && (\hat{w^M} \in (\hat{i}\in \hat{I})\rah \hat{w\,i}\rah \Type_j)\\
  & && \hspace{0.23em}\timesh && \big((\hat{s\in S})(\hat{f}\in (\hat{p\in P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\\
  & && && \rah ((\hat{p\in P\,s})\rah \hat{w^M}\,(\hat{in\,s\,p})\,(\hat{f\,p}))\rah \hat{w^M}\,(\hat{out\,s})\,(\hat{sup\,s\,f})\big) \\
  & \rlap{$ W^\E\,(\hat{w, sup})\,(\hat{w^M,sup^M}) \equiv $} \\
  & && && (\hat{w^E} \in (\hat{i \in I})(\hat{x \in w\,i})\rah \hat{w^M\,i\,x})\\
  & && \hspace{0.23em}\timesh && \big((\hat{s\in S})(\hat{f}\in (\hat{p\in P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\\
  & && && \rah \hat{w^E}\,(\hat{out\,s})\,(\hat{sup\,\,s\,f}) \eqh \hat{sup^M\,s\,f}\,(\hat{\lambda} \hat{p.}\, \hat{w^E}\,(\hat{in\,s\,p})\,(\hat{f\,p}))\big)
\end{alignat*}

%=====================================================================

\section{Existence of HIITs}
\label{sec:adding}

The target type theory supports HIITs if whenever we can derive
$\hat{\Gamma}\vdash\Delta$ in the source theory, the following rules are
admissible.
\[
\infer{\hat{\Gamma}\vdashh\hat{\con}_\Delta\in \Delta^\C}{\vspace{1em}}
\hspace{5em}
\infer{\hat{\Gamma}\vdashh\hat{\elim}_\Delta\,\hat{m}\in\Delta^\E\,\hat{\con}_\Delta\,\hat{m}}{\hat{\Gamma}\vdashh \hat{m}\in \Delta^\M\,\hat{\con}_\Delta}
\]
We can add HIITs to the target theory by extending it with the theory
of codes (making the target and the source theory the same) and adding
the above rules with the additional assumption of
$\hat{\Gamma}\vdash\Delta$. However, this only adds HIITs with weak
computation rules. To make the computation rules definitional, we
would probably need a two-level target type theory with an equality
having an equality reflection rule as in Voevodsky's homotopy type
system \cite{hts} or Andromeda \cite{andromeda}.

%=====================================================================

\section{Formalisation and implementation}
\label{sec:formalisation}

There are three additional development artifacts to the current work:
a Haskell implementation, a shallow Agda formalisation and a deep Agda
formalisation. All three are available from the homepage of the first
author.

The Haskell implementation takes as input a file which contains a
representation of a $\hat{\Gamma}\vdash\Delta$ specification of a
HIIT. Then, it checks the input with respect to the rules in figure
\ref{fig:codes}, and outputs an Agda-checkable file which contains the
results of the $\blank^\C$, $\blank^\M$ and $\blank^\E$
translations. It comes with examples, including the ones in this
paper, indexed W-types \cite{morris09indexed}, the dense completion
\cite[Appendix A.1.3]{forsberg-phd} and several HITs from
\cite{HoTTbook} including the definition of Cauchy reals. It can be
checked that our implementation computes the expected elimination
principles in these cases.

The shallow Agda formalisation embeds both the source and target theories
shallowly into Agda: it represents types as Agda types, functions as
Agda functions, and so on. We also leave the $\blank^\C$ operation
implicit. We state each case of the $\blank^\M$ and $\blank^\E$
translations as Agda functions from all induction hypotheses to the
result type of the translation, which lets us ``typecheck'' the
translation. We have found that this style of formalisation is
conveniently light, but remains detailed enough to be useful. We also
generated some of the code of the Haskell implementation from this
formalisation.

The deep Agda formalisation still uses a shallow embedding of the
target type theory, but it uses a deep embedding of the source theory,
in the style of \cite{ttintt}. In the formalisation we merge the three
translations into a single model construction. This allows us to prove
strict preservation of definitional equalities in the substitution
calculus of the source theory, in contrast to the shallow
formalisation, where we cannot reason about definitional equalities of
Agda terms. Due to technical challenges, this formalisation uses
transport instead of $\J$ in the source theory, but this still covers
a rather large class of HIIT definitions.

%=====================================================================

\section{Conclusions and further work}
\label{sec:summary}

Higher inductive-inductive types are useful in defining the well-typed
syntax of type theory in an abstract way \cite{ttintt}. From a
universal algebra point of view, they provide initial algebras for
multi-sorted algebraic theories where the sorts can depend on each
other. From the perspective of homotopy type theory, they provide
synthetic versions of homotopy-theoretic constructions such as higher
dimensional spheres or cell complexes. So far, no general scheme of
HIITs have been proposed. To quote Lumsdaine and Shulman
\cite{lumsdaineShulman}:
\begin{quotation}
``The constructors of an ordinary inductive type are unrelated to each
other.  But in a higher inductive type each constructor must be able
to refer to the previous ones; specifically, the source and target of
a path-constructor generally involve the previous
point-constructors. No syntactic scheme has yet been proposed for this
that covers all cases of interest while remaining meaningful and
consistent.''
\end{quotation}
In this paper we proposed such a syntactic scheme which also includes
inductive-inductive types. We tackled the problem of complex
dependencies on previous type formation rules and constructors by a
well-known method of describing intricate dependencies: the syntax of
type theory itself. We had to limit the type formers to only allow
strictly positive definitions, but these restrictions are the only
things that a type theorist has to learn to understand our codes. Our
encoding is also \emph{direct} in the sense that the types of
constructors and eliminators are exactly as required and not merely up
to isomorphisms.

In this paper we only \emph{specified} HIITs and characterised their
induction principles. Proving their \emph{existence} is left as
further work. This would likely involve reducing HIITs to basic
building blocks such as W-types and quotient types.

The theory of codes was defined as part of the syntax of another type
theory, the target theory. An alternative way would be to define
the theory of codes internally to a type theoretic metatheory in the
style of \cite{ttintt}. However, as described in \cite[Section
  6]{ttintt}, there would be a coherence problem: for the internal
syntax to be useful, we need to truncate it to be a set (as the
$\mathsf{trunc}$ constructor did for $\Int$ in the introduction). As
the eliminator needs to respect the equality given by
$\mathsf{trunc}$, we would only be able to eliminate from the internal
type theory into a set. This would preclude the definition of even the
$\blank^\C$ operation, which corresponds to the standard model. A
possible solution to this problem would be to add all the higher
coherence laws to the syntax (e.g. the pentagon law for the
composition of substitutions) and then prove that the syntax is a
set. For this however, we would probably need a two-level metatheory
as in \cite{semisegal}.

When working in a metatheory with uniqueness of identity proofs (which
implies that all HIITs are in essence QIITs), the theory of codes
admits a category model where the interpretation of a context is the
category of algebras corresponding to the context. In the future, we
would like to prove that these categories have initial algebras given
by terms in the theory of codes.

%=====================================================================

\appendix
\section{Elimination principle computed for the two-dimensional sphere}
\label{sec:app}

The two-dimensional sphere is given by the following context in the
theory of codes.
\[
\Gamma :\equiv \Big(\cdot,\,\,\,S^2:\U,\,\,\,b:\underline{S^2},\,\,\,surf:\underline{\refl_{b} =_{(b=_{S^2} b)} \refl_{b}}\Big)
\]
The sphere-algebras are computed as follows.
\begin{alignat*}{5}
  & && \Gamma^\C \equiv \hat{\top}\timesh(\hat{S^2}\in\Type_0)\timesh(\hat{b}\in \hat{S^2})\timesh(\hat{surf\in \hat{\refl_{\hat{b}}}=_{(\hat{b}=_{\hat{S^2}}\hat{b})}\hat{\refl_{\hat{b}}}})
\end{alignat*}
Given a sphere-algebra and fixing a universe level $i$, the motives
and methods are computed as follows.
\begin{alignat*}{5}
  & && \Gamma^\M\,\,(\hat{\tt,\,S^2,\,b,\,surf})\\
  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{2M}}\in \hat{S^2}\rah\Type_i) \\
  & && \hspace{0.75em} \timesh(\hat{b^M}\in \hat{S^{2M}\,b}) \\
  & && \hspace{0.75em} \timesh(\hat{surf^M\in \TR_{(\TR_{S^{2M}}\,\blank\,b^M=b^M)}\,surf\,\refl_{b^M} = \refl_{b^M}})
\end{alignat*}
Given a sphere-algebra and the motives and methods for this algebra,
the types of the elimination principles are computed as follows.
\begin{alignat*}{5}
  & && \Gamma^\E\,\,(\hat{\tt,\,S^2,\,b,\,surf})\,(\hat{\tt,\,S^{2M},\,b^M,\,surf^M}) \\
  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{2E}}\in (\hat{x}\in\hat{S^2})\hat{\,\ra\,}\hat{S^{2M}\,x}) \\
  & && \hspace{0.75em} \timesh(\hat{b^E}\in \hat{S^{2E}\,\,b = b^M})\\
  & && \hspace{0.75em} \timesh\bigg(\hat{surf^E\,\,\in\,\,} \TR\,\Big(\hat{\J\,\refl\,b^E}\Big)\,\Big(\TR\,\big(\hat{\J\,\refl\,b^E}\big)\,\big(\hat{\apd}\,(\hat{\lambda} \hat{x}.\TR\,\hat{b^E}\,(\TR\,\hat{b^E}\,(\hat{\apd}\,\hat{S^{2E}}\,\hat{x})))\,\hat{surf}\big)\Big) \\
  & && \hspace{5.5em} \hat{=} \,\hat{surf^M}\bigg)
\end{alignat*}
Note that if $\hat{b^E}$ is a definitional equality (that is, we have
$\hat{S^{2E}\,\,b} \equiv \hat{b^M}$), the occurrences of $\hat{b^E}$
in the type of $\hat{surf^E}$ can be replaced by $\hat{\refl}$. In
this case the type of $\hat{surf^E}$ becomes the expected
$\hat{\apd}\,(\hat{\apd}\,\hat{S^{2E}})\,\hat{surf}\,\hat{=}\,\hat{surf^M}$.

%=====================================================================

\bibliography{b}

\end{document}

### 2019-01-28 paper.tex
title
test
Progress to theory of signatures section

@@
-\documentclass{lmcs} %%% last changed 2014-08-20
+\documentclass[dvipsnames]{lmcs} %%% last changed 2014-08-20
+
+%% build: latexmk -pdf -pvc -interaction=nonstopmode paper.tex
 
 %% mandatory lists of keywords
-\keywords{MANDATORY list of keywords}
+\keywords{homotopy type theory, higher inductive types, inductive types}
 
@@
 %% e.g. \usepackage{tikz}
+%% \usepackage[dvipsnames]{xcolor}
+\usepackage{microtype}
+\usepackage{proof}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{stmaryrd}
+\bibliographystyle{plainurl}
 \usepackage{hyperref}
+
+%% ABBREVS
+%% --------------------------------------------------------------------------------
+
+%% setting \sqcdot as HoTT-book-style transitivity
+\makeatletter
+\DeclareRobustCommand{\sqcdot}{\mathbin{\mathpalette\morphic@sqcdot\relax}}
+\newcommand{\morphic@sqcdot}[2]{%
+  \sbox\z@{$\m@th#1\centerdot$}%
+  \ht\z@=.33333\ht\z@
+  \vcenter{\box\z@}%
+}
+\makeatother
+
+\newcommand{\U}{\mathsf{U}}
+\newcommand{\El}{\mathsf{El}}
+\newcommand{\REN}{\mathsf{REN}}
+\newcommand{\op}{\mathsf{op}}
+\newcommand{\ra}{\rightarrow}
+\newcommand{\Ra}{\Rightarrow}
+\newcommand{\Set}{\mathsf{Set}}
+\newcommand{\PSh}{\mathsf{PSh}}
+\newcommand{\FamPSh}{\mathsf{FamPSh}}
+\renewcommand{\ll}{\llbracket}
+\providecommand{\rr}{\rrbracket}
+\newcommand{\Con}{\mathsf{Con}}
+\newcommand{\Ty}{\mathsf{Ty}}
+\newcommand{\Tm}{\mathsf{Tm}}
+\newcommand{\Tms}{\mathsf{Tms}}
+\newcommand{\R}{\mathsf{R}}
+\newcommand{\TM}{\mathsf{TM}}
+\newcommand{\NE}{\mathsf{NE}}
+\newcommand{\NF}{\mathsf{NF}}
+\newcommand{\p}{\mathsf{p}}
+\newcommand{\q}{\mathsf{q}}
+\renewcommand{\u}{\mathsf{u}}
+\renewcommand{\ne}{\mathsf{ne}}
+\newcommand{\nf}{\mathsf{nf}}
+\newcommand{\lQ}{\mathsf{lQ}}
+\newcommand{\lU}{\mathsf{lU}}
+\renewcommand{\lq}{\mathsf{lq}}
+\newcommand{\lu}{\mathsf{lu}}
+\newcommand{\cul}{\ulcorner}
+\newcommand{\cur}{\urcorner}
+\newcommand{\norm}{\mathsf{norm}}
+\newcommand{\Nf}{\mathsf{Nf}}
+\newcommand{\Ne}{\mathsf{Ne}}
+\newcommand{\Nfs}{\mathsf{Nfs}}
+\newcommand{\Nes}{\mathsf{Nes}}
+\newcommand{\ID}{\mathsf{ID}}
+\newcommand{\id}{\mathsf{id}}
+\newcommand{\nat}{\,\dot{\rightarrow}\,}
+\newcommand{\Nat}{\mathsf{Nat}}
+%% \renewcommand{\S}{\overset{\mathsf{s}}{\ra}}
+\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
+\newcommand{\lam}{\mathsf{lam}}
+\newcommand{\app}{\mathsf{app}}
+\newcommand{\tr}[2]{\ensuremath{{}_{#1 *}\mathopen{}{#2}\mathclose{}}}
+\newcommand{\C}{\mathsf{C}}
+\newcommand{\Code}{\mathsf{Code}}
+
+\newcommand{\A}{\mathsf{A}}
+\newcommand{\M}{\mathsf{M}}
+\newcommand{\D}{\mathsf{D}}
+\renewcommand{\S}{\mathsf{S}}
+
+\newcommand{\data}{\mathsf{data}}
+\newcommand{\ind}{\hspace{1em}}
+\newcommand{\idP}{\mathsf{idP}}
+\newcommand{\compP}{\mathsf{compP}}
+\newcommand{\idF}{\mathsf{idF}}
+\newcommand{\compF}{\mathsf{compF}}
+\newcommand{\proj}{\mathsf{proj}}
+\newcommand{\ExpPSh}{\mathsf{ExpPSh}}
+\newcommand{\map}{\mathsf{map}}
+\newcommand{\Var}{\mathsf{Var}}
+\newcommand{\Vars}{\mathsf{Vars}}
+\newcommand{\wk}{\mathsf{wk}}
+\newcommand{\neuU}{\mathsf{neuU}}
+\newcommand{\neuEl}{\mathsf{neuEl}}
+\newcommand{\var}{\mathsf{var}}
+\newcommand{\natn}{\mathsf{natn}}
+\newcommand{\natS}{\mathsf{natS}}
+\newcommand{\LET}{\mathsf{let}}
+\newcommand{\IN}{\mathsf{in}}
+\newcommand{\refl}{\mathsf{refl}}
+\newcommand{\trans}{\mathbin{\raisebox{0.2ex}{$\displaystyle\centerdot$}}}
+\newcommand{\zero}{\mathsf{zero}}
+\newcommand{\suc}{\mathsf{suc}}
+\newcommand{\N}{\mathbb{N}}
+
+\newcommand\arcfrombottom{
+  \begin{tikzpicture}[scale=0.03em]
+    \draw (0,0) arc (0:180:0.5);
+    \draw (0,0) edge[->] (0,-0.3);
+    \draw (-1,0) edge (-1,-0.3);
+  \end{tikzpicture}
+}
+\newcommand\arcfromtop{
+  \begin{tikzpicture}[scale=0.03em]
+    \draw (0,0) arc (180:360:0.5);
+    \draw (0,0) edge[->] (0,0.3);
+    \draw (1,0) edge (1,0.3);
+  \end{tikzpicture}
+}
+
+\newcommand{\Elim}{\mathsf{Elim}}
+\newcommand{\elim}{\mathsf{elim}}
+\newcommand{\Rec}{\mathsf{Rec}}
+\newcommand{\record}{\mathsf{record}}
+\newcommand{\funext}{\mathsf{funext}} \newcommand{\Q}{\mathsf{Q}}
+\newcommand{\T}{\mathsf{T}} \newcommand{\leaf}{\mathsf{leaf}}
+\newcommand{\node}{\mathsf{node}} \newcommand{\perm}{\mathsf{perm}}
+\newcommand{\coe}{\mathsf{coe}} \newcommand{\vz}{\mathsf{vz}}
+\newcommand{\vs}{\mathsf{vs}} \newcommand{\untr}{\mathsf{untr}}
+\newcommand{\from}{\mathsf{from}}
+\newcommand{\fromeq}{{\mathsf{from}\hspace{-0.3em}\equiv}}
+\newcommand{\fromsimeq}{{\mathsf{from}\hspace{-0.3em}\simeq}}
+\newcommand{\Model}{\mathsf{Model}}
+\newcommand{\DModel}{\mathsf{DModel}}
+\newcommand{\module}{\mathsf{module}}
+\newcommand{\open}{\mathsf{open}}
+\renewcommand{\P}{\mathsf{P}} \newcommand{\Bool}{\mathsf{Bool}}
+\newcommand{\true}{\mathsf{true}} \newcommand{\false}{\mathsf{false}}
+\renewcommand{\not}{\mathsf{not}} \newcommand{\0}{\mathsf{0}}
+\newcommand{\1}{\mathsf{1}} \renewcommand{\Pr}{\mathsf{Pr}}
+\newcommand{\PrNat}{\mathsf{PrNat}} \newcommand{\J}{\mathsf{J}}
+\newcommand{\wkV}{\mathsf{wkV}} \renewcommand{\r}[1]{{\P_{#1}}}
+\newcommand{\stab}{\mathsf{stab}} \newcommand{\NTy}{\mathsf{NTy}}
+\newcommand{\isDec}{\mathsf{isDec}} \newcommand{\dec}{\mathsf{dec}}
+\newcommand{\yes}{\mathsf{yes}} \newcommand{\no}{\mathsf{no}}
+\newcommand{\case}{\mathsf{case}} \newcommand{\inj}{\mathsf{inj}}
+\newcommand{\K}{\mathsf{K}} \newcommand{\lb}{\langle}
+\newcommand{\rb}{\rangle} \newcommand{\Decl}{\mathsf{Decl}}
+\newcommand{\Core}{\mathsf{Core}} \newcommand{\IND}{\mathsf{ind}}
+\newcommand{\Id}{\mathsf{Id}} \newcommand{\Base}{\mathsf{Base}}
+\newcommand{\transport}{\mathsf{transport}}
+\newcommand{\I}{\mathsf{I}} \newcommand{\E}{\mathsf{S}}
+\newcommand{\transp}{\mathsf{transp}}
+\newcommand{\Transp}{\mathsf{Transp}}
+\newcommand{\W}{\mathsf{W}}
+\newcommand{\Fin}{\mathsf{Fin}} \newcommand{\fzero}{\mathsf{fzero}}
+\newcommand{\fsuc}{\mathsf{fsuc}}
+\newcommand{\inv}{\mathsf{inv}}
+\newcommand{\con}{\mathsf{con}}
+\newcommand{\LEFT}{\mathsf{left}}
+\newcommand{\RIGHT}{\mathsf{right}}
+\newcommand{\seg}{\mathsf{seg}}
+\newcommand{\Int}{\mathsf{Int}}
+\renewcommand{\in}{\mathbin{\hat:}}
+\renewcommand{\hat}[1]{{\color{BrickRed}{#1}}}
+\newcommand{\vdashh}{\mathbin{\hat\vdash}}
+\newcommand{\rah}{\mathbin{\hat\ra}}
+\newcommand{\commah}{\hat,\,}
+\newcommand{\timesh}{\mathbin{\hat\times}}
+\newcommand{\eqh}{\mathbin{\hat=}}
+\newcommand{\TR}{\hat{\mathsf{tr}}}
+\newcommand{\ap}{\hat{\mathsf{ap}}}
+\newcommand{\apd}{\hat{\mathsf{apd}}}
+\renewcommand{\tt}{\hat{\mathsf{tt}}}
+\newcommand{\Tel}{\hat{\mathsf{Tel}}}
+\newcommand{\Type}{\hat{\mathsf{Type}}}
+\newcommand{\emptytel}{\hat{\epsilon}}
+\newcommand{\telext}{\mathbin{\hat{\lhd}}}
+\newcommand{\ltel}{\hat{\ll}}
+\newcommand{\rtel}{\hat{\rr}}
+\newcommand\pp{\ensuremath{\hat{\mathbin{+\mkern-10mu+}}}}
+\newcommand{\semicol}{\hat;\,}
+\newcommand{\targetass}{\hat{\Gamma}\semicol}
+\newcommand{\Alg}{\mathsf{Alg}}
+\newcommand{\DisplayedAlg}{\mathsf{DisplayedAlg}}
+\newcommand{\Morphism}{\mathsf{Morphism}}
+\newcommand{\Section}{\mathsf{Section}}
+
+%% END OF ABBREVS
+%% --------------------------------------------------------------------------------
+
 %%\input{myMacros.tex}
@@
 
-\title[Short title TODO]{Homomorphisms from type theoretic signatures using parametricity}
+\title{Signatures and Induction Principles for Higher Inductive-Inductive Types}
 \titlecomment{{\lsuper*} This is a journal version of the paper \cite{hiit}.}
@@
 
-%% etc.
-
 %% required for running head on odd and even pages, use suitable
@@
 \begin{abstract}
-  \noindent If a logical relation between two algebras is a graph of a
-  function, it is an algebra homomorphism. 
+\noindent
+Higher inductive-inductive types (HIITs) generalize inductive types of dependent
+type theories in two ways. On the one hand they allow the simultaneous
+definition of multiple sorts that can be indexed over each other. On the other
+hand they support equality constructors, thus generalizing higher inductive
+types of homotopy type theory. Examples that make use of both features are the
+Cauchy reals and the well-typed syntax of type theory where conversion rules are
+given as equality constructors. In this paper we propose a general definition of
+HIITs using a small type theory, named the theory of signatures. A context in
+this theory encodes a HIIT by listing the constructors. We also compute notions
+of induction and recursion for HIITs, by using variants of syntactic logical
+relation translations. Building full categorical semantics and constructing
+initial algebras is left for future work. The theory of HIIT signatures was
+formalised in Agda together with the syntactic translations. We also provide a
+Haskell implementation, which takes signatures as input and outputs translation
+results as valid Agda code.
 \end{abstract}
@@
 
-\section*{Introduction}
+\section{Introduction}
+\label{sec:intro}
 
-In HoTT, algebraic structures can have higher equalities. What is a
-homomorphism between two such algebraic structures? Example: circle,
-sphere. This also answers the question: what is an elimination
-principle for any HIT?
+Many dependent type theories support some form of inductive types. An
+inductive type is given by its constructors, along with an elimination
+principle which expresses that all inhabitants are constructed using
+finitely many applications of the constructors.
 
-Another Motivation: usefulness of GATs for formalising programming
-languages. Example: tt-in-tt with equations. Algebraic view of
-progr. langs.
+For example, the inductive type of natural numbers $\Nat$ is given by
+the constructors $\zero:\Nat$ and $\suc:\Nat \ra \Nat$, and has the
+well-known induction principle:
+\[
+  \Elim\Nat:(P:\Nat \ra \mathsf{Type})(pz: P\,\zero)\big(ps:(n:\Nat)\ra P\,n\ra P\,(\suc\,n)\big)(n:\Nat)\ra P\,n
+\]
+\noindent $P$ is a family of types (or: a proof-relevant predicate) over natural
+numbers, which is called the \emph{induction motive}. The arguments $pz$ and
+$ps$ are called the \emph{induction methods}. The behavior of induction is
+described by a \emph{computation-rule} ($\beta$-rule) for each constructor
+and induction method:
+\begin{alignat*}{5}
+  & \Elim\Nat\,P\,pz\,ps\,\zero && \equiv pz \\
+  & \Elim\Nat\,P\,pz\,ps\,(\suc\,n) && \equiv ps\,n\,(\Elim\Nat\,P\,pz\,ps\,n)
+\end{alignat*}
 
-Method: Uday observation (inverse Reynolds), signatures for
-generalised algebraic theories using domain-specific tt, parametricity
-interpretation of dependent types. We put these
-together. Achievements: getting induction principles for non-strictly
-positive types, describing induction principles for HIITs.
+%% These express that the eliminator applied to a constructor expression
+%% reduces to an application of the corresponding induction method. From
+%% an operational point of view, $\Elim\Nat$ replaces all the $\zero$ and
+%% $\suc$ constructors with the given induction methods.
 
-This is a journal version, main new thing: relation to parametricity
-(graph model), category model (extension of reflexive graph model),
-homomorphisms for non strictly positive operators, relation of
-syntactic translations and models for solving coherence issue, for
-HIITs the syntactic translation cannot be extended to a category.
+% A special case of the eliminator is the recursor (non-dependent
+% eliminator) which allows the definition of a non-dependent function
+% from the inductive type, for $\Nat$ it has the following type.
+% \[
+% \Rec\Nat : \forall(A:\mathsf{Type}).A \ra (A \ra A) \ra \Nat \ra A
+% \]
+% The eliminator can be derived from the recursor and a uniqueness
+% principle.
 
-Formalisation?
+Indexed families of types can be also considered, for example length-indexed
+vectors of $A$-elements $\mathsf{Vec}_A: \Nat \ra \mathsf{Type}$. Mutual
+inductive types are yet another generalization, however, they can be reduced to
+indexed families where indices classify constructors for each mutual type.
+Inductive-inductive types \cite{forsberg-phd} are mutual definitions where this
+reduction does not work: here a type can be defined together with a family
+indexed over it. An example is the following fragment of a well-typed syntax of
+a type theory, where the second $\Ty$ type constructor is indexed over $\Con$,
+but constructors of $\Con$ also refer to $\Ty$:
+\begin{alignat*}{3}
+  & \Con && : \mathsf{Type} && \text{contexts} \\
+  & \Ty  && : \Con \ra \mathsf{Type} && \text{types in contexts} \\
+  & \bullet && : \Con && \text{constructor for the empty context} \\
+  & \blank\rhd\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con && \text{constructor for context extension} \\
+  & \iota && : (\Gamma:\Con)\ra\Ty\,\Gamma && \text{constructor for a base type} \\
+  & \Pi && : (\Gamma:\Con)(A:\Ty\,\Gamma)\ra\Ty\,(\Gamma\rhd A)\ra\Ty\,\Gamma \hspace{1em} && \text{constructor for dependent functions}
+\end{alignat*}
+There are two eliminators for this type: one for $\Con$ and one for
+$\Ty$. Both take the same arguments: two motives ($P:\Con\ra\mathsf{Type}$ and
+$Q:(\Gamma:\Con)\ra P\,\Gamma\ra\Ty\,\Gamma\ra\mathsf{Type}$) and four methods
+(one for each constructor, which we omit).
+\begin{alignat*}{6}
+  & \Elim\Con && : (P:\dots)(Q:\dots)\ra\dots\ra(\Gamma:\Con) && \ra P\,\Gamma \\
+  & \Elim\Ty  && : (P:\dots)(Q:\dots)\ra\dots\ra(A:\Ty\,\Gamma) && \ra Q\,\Gamma\,(\Elim\Con\,\Gamma)\,A
+\end{alignat*}
+Note that the type of $\Elim\Ty$ refers to $\Elim\Con$; for this reason
+this elimination principle is sometimes called ``recursive-recursive''
+(analogously to ``inductive-inductive'').
 
-\subsection*{Overview of the paper}
+Higher inductive types (HITs, \cite[Chapter 6]{HoTTbook}) generalize inductive
+types in a different way: they allow constructors expressing equalities of
+elements of the type being defined. This enables, among others, the definition
+of types quotiented by a relation. For example, the type of integers $\Int$ can
+be given by a constructor $\mathsf{pair}:\Nat\ra\Nat\ra\Int$ and an equality
+constructor $\mathsf{eq}:(a\,b\,c\,d:\Nat)\ra a+d=_\Nat b+c\ra
+\mathsf{pair}\,a\,b=_{\Int}\mathsf{pair}\,c\,d$ targetting an equality of
+$\Int$. The eliminator for $\Int$ expects a motive $P:\Int\ra\mathsf{Type}$, a
+method for the $\mathsf{pair}$ constructor $p:(a\,b:\Nat)\ra
+P\,(\mathsf{pair}\,a\,b)$ and a method for the equality constructor
+$\mathsf{path}$. This method is a proof that given $e:a+d=_\Nat b+c$, $p\,a\,b$
+is equal to $p\,c\,d$ (the types of which are equal by $e$). Thus the method for
+the equality constructor ensures that all functions defined from the quotiented
+type respect the relation. Since the integers are supposed to be a set (which
+means that any two equalities between the same two integers are equal), we would
+need an additional higher equality constructor
+$\mathsf{trunc}:(x\,y:\Int)\ra(p\,q:x=_\Int y)\ra p=_{x=_\Int y} q$.  HITs may
+have constructors of iterated equality types as well. With the view of types as
+spaces in mind, point constructors add points to spaces, equality constructors
+add paths and higher constructors add homotopies between higher-dimensional paths.
 
-\subsection*{Related work}
+Not all constructor expressions make sense. For example \cite[Example
+  6.13.1]{HoTTbook}, given an $f:(X:\mathsf{Type})\ra X\ra X$, suppose that an
+inductive type $\mathsf{Ival}$ is generated by the point constructors
+$\mathsf{a}:\mathsf{Ival}$, $\mathsf{b}:\mathsf{Ival}$ and a path
+constructor $\sigma:f\,\mathsf{Ival}\,\mathsf{a}
+=_{\mathsf{Ival}}f\,\mathsf{Ival}\,\mathsf{b}$. The eliminator for this type
+should take a motive $P:\mathsf{Ival}\ra\mathsf{Type}$, two methods $p_a :
+P\,\mathsf{a}$ and $p_b : P\,\mathsf{b}$, and a path connecting
+elements of $P\,(f\,\mathsf{Ival}\,\mathsf{a})$ and
+$P\,(f\,\mathsf{Ival}\,\mathsf{b})$. However it is not clear what these
+elements should be: we only have elements $p_a:P\,\mathsf{a}$ and
+$p_b:P\,\mathsf{b}$, and there is no way in general to transform these
+to have types $P\,(f\,\mathsf{Ival}\,\mathsf{a})$ and
+$P\,(f\,\mathsf{Ival}\,\mathsf{b})$.
 
-\section*{Theory of signatures}
+Another invalid example is an inductive type $\mathsf{Neg}$ with a constructor
+$\con:(\mathsf{Neg} \ra \bot) \ra \mathsf{Neg}$ where $\bot$ is the empty
+type. An eliminator for this type should (at least) yield a projection function
+$\proj: \mathsf{Neg} \ra (\mathsf{Neg} \ra \bot)$. Given this, we can define $u
+:\equiv \con\,(\lambda x . \proj\, x\,x):\mathsf{Neg}$ and then derive $\bot$ by
+$\proj\,u\,u$. The existence of $\mathsf{Neg}$ would make the type theory
+inconsistent. A common restriction to avoid such situations is \emph{strict
+  positivity}. It means that the type being defined cannot occur on the left
+hand side of a function arrow in a parameter of a constructor. This excludes the
+above constructor $\con$.
 
-Brick red color. Subst.calc, U,El,Pi. We formalise the calculus as a
-QIIT, but present it with variables names and judgements. Eliminating
-from the QIIT needs to preserve equalities. Metatheory extensional.
+In this paper we propose a notion of signatures for higher inductive-inductive
+types (HIITs) which includes the above valid examples and excludes the invalid
+ones. Our signatures allow any number of inductive-inductive type constructors,
+possibly infinitary higher constructors of any dimension and restricts
+constructors to strictly positive ones. It also allows free usage of $\J$ (path
+induction) and $\refl$ in HIIT signatures, and allows mixing type, point and path
+constructors in any order.
 
-\section*{From logical relations to homomorphisms}
+The core idea is to represent HIIT specifications as contexts in a
+domain-specific type theory which we call the \emph{theory of signatures}. Type
+formers in the theory of codes are restricted in order to enforce strict
+positivity. For example, natural numbers are defined as the three-element
+context
+\[
+  Nat:\U,\,\,\, zero:\underline{Nat},\,\,\, suc : Nat \ra \underline{Nat}
+\]
+where $Nat$, $zero$ and $suc$ are simply variable names, and underlining
+denotes $\El$ (decoding) for the Tarski-style universe $\U$.
 
-Graph model, (A,M''), check what it gives for natural numbers
+We also show how to derive induction and recursion principles for each
+signature. We use variants of \emph{syntactic logical relation translations} to
+compute notions of \emph{algebras}, \emph{homomorphisms}, \emph{displayed
+  algebras} and \emph{displayed algebra sections}, and then define induction and
+recursion in terms of these notions.
 
-we restrict U from relations to functions and El becomes graph of the
-function
+To our knowledge, this is the first proposal for a definition of
+HIITs. However, we do not provide complete (higher) categorical semantics
+for HIITs, nor do we show that initial algebras exist for specified HIITs.
 
-now we get homomorphisms
+\subsection{Overview of the Paper}
 
-example: homomorphism for non-strictly positive operators
+We start by describing the theory of HIIT signatures in Section
+\ref{sec:signatures}. Here, we also describe the syntax for an external type
+theory, which serves as the source of constants which are external to a
+signature, like natural numbers in the case of length-indexed vectors. In
+Section \ref{sec:general}, we give a general definition of induction and
+recursion. In Section \ref{sec:coherence} we explain the choice of using
+syntactic translations in the rest of the paper. In Sections TODO to TODO, we
+describe four syntactic translations from the theory of signatures to the syntax
+of the external type theory, respectively computing algebras, homomorphisms,
+displayed algebras and sections of displayed algebras. In Section TODO, we
+consider extending the previous translations with additional components of the
+categorical semantics of HIITs (e.g.\ identity and composition for
+homomorphisms), and explain why the approach in this paper does not make this
+feasible. Section \ref{sec:formalization} describes the Agda formalization and
+the Haskell implementation. We conclude in Section \ref{sec:summary}.
 
-problem: suc is not the usual one, we cannot compose homomorphisms:
-composition does not work for Pi
+\subsection{Related Work}
 
-\section*{Categories of algebras}
+Schemes for inductive families are given in
+\cite{Dybjer97inductivefamilies,paulinmohring}, and for inductive-recursive
+types in \cite{dybjer00ir}. A symmetric scheme for both inductive and
+coinductive types is given in \cite{henning}. Basold et al. \cite{niels} define
+a syntactic scheme for higher inductive types with only 0-constructors and
+compute the types of induction principles. In \cite{nielsmsc} a semantics is
+given for the same class of HITs but with no recursive equality
+constructors. Dybjer and Moeneclaey define a syntactic scheme for finitary HITs
+and show their existence in a groupoid model \cite{moeneclaey}.
 
-Restriction to strictly positive operators in signatures.
+Internal codes for simple inductive types such as natural numbers, lists or
+binary trees can be given by containers which are decoded to W-types
+\cite{abbot05containers}. Morris and Altenkirch \cite{morris09indexed} extend
+the notion of container to that of indexed container which specifies indexed
+inductive types. Codes for inductive-recursive types are given in
+\cite{Dybjer99afinite}. Inductive-inductive types were introduced by Forsberg
+\cite{forsberg-phd}. Sojakova \cite{sojakova} defines a subset of HITs called
+W-suspensions by a coding scheme similar to W-types. She proves that the
+induction principle is equivalent to homotopy initiality.
 
-category model
+Quotient types \cite{hofmann95extensional} are precursors of higher inductive
+types (HITs). The notion of HIT first appeared in \cite{HoTTbook}, however only
+through examples and without a general definition.  Lumsdaine and Shulman give a
+general specification of models of type theory supporting higher inductive types
+\cite{lumsdaineShulman}. They introduce the notion of cell monad with parameters
+and characterize the class of models which have initial algebras for a cell
+monad with parameters. \cite{cubicalhits} develop semantics for several HITs
+(sphere, torus, suspensions, truncations, pushouts) in certain presheaf toposes,
+and extend the syntax of cubical type theory \cite{ctt} with these HITs. Kraus
+\cite{krausprop} and Van Doorn \cite{doorn} construct propositional truncation
+as a sequential colimit. The schemes mentioned so far do not support
+inductive-inductive types.
 
-\section*{Extensions}
+Cartmell's generalized algebraic theories (GATs) \cite{gat} pioneered a
+type-theoretic notion of algebraic signature. GATs can be viewed as finitary
+quotient inductive-inductive signatures (QIITs), although GATs support equalities
+between inductive types (sorts) as well, which have not been so far explored in
+other works about QIITs and HIITs.
 
-Pi with metatheoretic domain and large codomain
+The article of Altenkirch et al. \cite{gabe} gives specification and semantics
+of QIITs in a set-truncated setting. Signatures are given as lists of functors
+which can be interpreted as complete categories of algebras, and completeness is
+used to talk about notions of induction and recursion.  However, no strict
+positivity restriction is given, nor a construction of initial algebras.
 
-Pi with metatheoretic domain and small codomain (infinitary)
+Closely related to the current work is the paper by the current authors and
+Altenkirch \cite{kaposi2019constructing}, which also concerns QIITs. There,
+signatures for QIITs are essentially a restriction of the signatures given here,
+but in contrast to the current work, the restricted quotient setting enables
+building initial algebras and detailed categorical semantics.
 
-Extensional identity
+The logical predicate syntactic translation was introduced by Bernardy et
+al. \cite{bernardy2010parametricity}. The idea that a context can be seen as a
+signatures and the logical predicate translation can be used to derive the types
+of induction motives and methods was described in \cite[Section
+  5.3]{ttintt}. Logical relations are used to derive the computation rules in
+\cite[Section 4.3]{kaposi-phd}, but only for closed QIITs. Syntactic
+translations in the context of the calculus of inductive constructions are
+discussed in \cite{next700}. Logical relations and parametricity can also be
+used to justify the existence of inductive types in a type theory with an
+impredicative universe \cite{atkey}.
 
-Identities between sorts
+\section{Signatures for HIITs}
+\label{sec:signatures}
 
-Examples
+In this section we define signatures for HIITs. First, we list
+the main motivations behind our definition.
 
-\section*{Higher equalities}
+\begin{itemize}
+\item\emph{Ubiquitous type dependencies.}
+Recall the inductive-inductive $\Con$-$\Ty$ example from Section
+\ref{sec:intro}: there, types of constructors may refer back to previous
+constructors. Additionally, $\Ty$ is indexed over the previously declared $\Con$
+type constructor. This suggests that we should not try to stratify signatures,
+in contrast to previous works, and use a fully dependent type theory.
 
-The metatheory has K, so we do not elimiante into it, but to a target
-theory which does not have K. This also solves a coherence problem
-when eliminating from the syntax as a QIIT in HoTT as
-metatheory. Also, we add intensional Id type.
+\item\emph{Referring to external types.}
+We would like to mention types which are external to the signature. For example,
+length-indexed vectors refer to natural numbers which are supposed to already
+exist. Hence, we also assume a syntax for an \emph{external type theory}, which
+is the source of such types, and constructions in the theory of signatures may
+depend on a context in the external type theory.
 
-\subsection*{Source theory}
+\item\emph{Strict positivity.}
+In prior literature, schemes for inductive types usually include structural
+restrictions for this. In our case, a \emph{universe} can be used to make
+size restrictions which also entail strict positivity.
 
-(Brick red)
+\item\emph{Iterated equalities and path induction in signatures.}
+We can support this relatively easily by closing the universe under equality
+type formation, and also using a standard (although size-restricted) definition
+of path induction.
+\end{itemize}
 
-\subsection*{Target theory}
+%=====================================================================
 
-(Blue)
+\subsection{Theory of Signatures}
+\label{sec:tos}
 
-\subsection*{Translations}
+\begin{figure}
 
-cannot be extended to a category because U[] is not strict unless we
-add new computation rules.
+(1) Contexts and variables
 
-\subsection*{Examples}
+\[
+\begin{gathered}
+  \infer{\hat{\Gamma}\vdash\cdot}{\vdashh\hat{\Gamma}}
+\end{gathered}
+\hspace{2em}
+\begin{gathered}
+  \infer{\hat{\Gamma}\vdash\Delta,x:A}{\targetass\Delta\vdash A}
+\end{gathered}
+\hspace{2em}
+\begin{gathered}
+  \infer{\targetass\Delta,x:A\vdash x : A}{\targetass\Delta\vdash A}
+\end{gathered}
+\hspace{2em}
+\begin{gathered}
+  \infer{\targetass\Delta,y:B\vdash x : A}{\targetass\Delta\vdash x : A && \targetass\Delta\vdash B}
+\end{gathered}
+\]
 
-\section*{Conclusions and further work}
+\vspace{0.5em}
+(2) Universe
 
-\section*{Acknowledgment}
-  \noindent The authors wish to acknowledge fruitful discussions with
-  A and B.
+\[
+\begin{gathered}
+  \infer{\targetass\Delta \vdash \U}{\targetass\vdash\Delta}
+\end{gathered}
+\hspace{2em}
+\begin{gathered}
+  \infer{\targetass\Delta \vdash \underline{a}}{\targetass\Delta \vdash a : \U}
+\end{gathered}
+\]
+
+\vspace{0.5em}
+(3) Inductive parameters
+
+\[
+\begin{gathered}
+  \infer{\targetass\Delta \vdash (x:a)\ra B}{\targetass\Delta \vdash a : \U && \targetass\Delta,x:\underline{a} \vdash B}
+\end{gathered}
+\hspace{2em}
+\begin{gathered}
+  \infer{\targetass\Delta \vdash t\, u : B[x \mapsto u]}{\targetass\Delta \vdash t : (x:a)\ra B && \targetass\Delta \vdash u : \underline{a}}
+\end{gathered}
+\]
+
+\vspace{0.5em}
+(4) Equality
+
+\[
+\begin{gathered}
+  \infer{\targetass\Delta \vdash t=_a u : \U}{\targetass\Delta \vdash a : \U && \targetass\Delta \vdash t : \underline{a} && \targetass\Delta \vdash u : \underline{a}}
+\end{gathered}
+\hspace{2em}
+\begin{gathered}
+  \infer{\targetass\Delta \vdash \refl : \underline{t=_a t}}{\targetass\Delta \vdash t : \underline{a}}
+\end{gathered}
+\]
+
+\[
+\infer{\targetass\Delta \vdash \J_{a\,t\,\,(x.z.p)}\,pr\,_u\,eq : \underline{p[x\mapsto u, z\mapsto eq]}}
+      {\begin{array}{l}
+          \targetass\Delta \vdash t : \underline{a} \\
+          \targetass\Delta,x:\underline{a},z:\underline{t=_a x}\vdash p : \U \\
+          \targetass\Delta \vdash pr : \underline{p[x\mapsto t, z\mapsto \refl]} \\
+          \targetass\Delta \vdash u : \underline{a} \\
+          \targetass\Delta \vdash eq : \underline{t=_a u}
+      \end{array}}
+      \]
+
+\[
+\infer{\targetass\Delta \vdash \J\beta_{a\,t\,\,(x.z.p)}\,pr:\underline{(\J_{a\,t\,\,(x.z.p)}\,pr\,_t\,\refl) =_{p[x\mapsto t, z\mapsto \refl]} pr}}
+      {\targetass\Delta \vdash t : \underline{a}
+        && \targetass\Delta,x:\underline{a},z:\underline{t =_a x}\vdash p : \U
+        && \targetass\Delta \vdash pr : \underline{p[x\mapsto t, z\mapsto \refl]}
+      }
+\]
+
+\vspace{0.5em}
+(5) External parameters
+
+\[
+\begin{gathered}
+  \infer{\hat{\Gamma}\semicol\Delta \vdash (\hat{x}\in \hat{A})\ra B}{\hat{\Gamma}\vdashh\hat{A}\in\Type_{\hat{0}} && \hat{\Gamma}\semicol\vdash\Delta && \hat{(\hat{\Gamma}\commah\hat{x}\in \hat{A})}\semicol \Delta \vdash B}
+\end{gathered}
+\]
+
+\[
+\begin{gathered}
+  \infer{\hat{\Gamma}\semicol\Delta \vdash t\, \hat{u} : B[\hat{x}\mapsto \hat{u}]}{\hat{\Gamma}\semicol\Delta \vdash t : (\hat{x}\in \hat{A})\ra B && \hat{\Gamma}\vdashh \hat{u} \in \hat{A}}
+\end{gathered}
+\]
+
+\vspace{0.5em}
+(6) Infinitary parameters
+
+\[
+\begin{gathered}
+  \infer{\hat{\Gamma}\semicol\Delta \vdash (\hat{x}\in \hat{A})\ra b : \U}{\hat{\Gamma}\vdashh\hat{A}\in\Type_{\hat{0}} && \hat{\Gamma}\semicol\vdash\Delta && (\hat{\Gamma}\commah\hat{x}\in \hat{A})\semicol \Delta \vdash b : \U}
+\end{gathered}
+\]
+
+\[
+\begin{gathered}
+  \infer{\hat{\Gamma}\semicol\Delta \vdash t\, \hat{u} : \underline{b[\hat{x}\mapsto \hat{u}]}}
+        {\hat{\Gamma}\semicol\Delta \vdash t : \underline{(\hat{x}\in \hat{A})\ra b} && \hat{\Gamma}\vdash \hat{u} \in \hat{A}}
+\end{gathered}
+\]
+\caption{The theory of HIIT signatures. Weakenings are implicit, we assume fresh
+  names everywhere and consider $\alpha$-convertible terms equal. The
+  $\hat{\Gamma};$ assumptions are not used or changed in parts
+  (1)--(4).}
+\label{sigrules}
+\end{figure}
+
+We list typing rules for the theory of signatures in Figure \ref{sigrules}. We consider the
+following judgments:
+\begin{alignat*}{4}
+  & \hat{\Gamma}\vdash\Delta && \text{$\Delta$ is a context in the external context $\hat{\Gamma}$}\\
+  & \hat{\Gamma}\semicol\Delta\vdash A && \text{$A$ is a type in context $\Delta$ and external context $\hat{\Gamma}$} \\
+  & \hat{\Gamma}\semicol\Delta\vdash t : A \hspace{3em} && \text{$t$ is a term of type $A$ in context $\Delta$ and external context $\hat{\Gamma}$}
+\end{alignat*}
+
+We have the convention that constructions in the external type theory are
+notated in {\color{BrickRed}brick red} color. Although every judgement is valid
+up to a context in the external type theory, note that none of the rules in
+(1)--(4) depend on or change these assumptions, and even after that, we do not
+refer to any particular type former from the external theory. We describe the
+external theory in more detail in Section TODO. Also, the rules presented here
+are informal and optimized for readability; we describe the Agda formalizations
+in Section \ref{sec:formalization}. We explain the rules for the theory of
+signatures in order now.
+
+(1) The rules for context formation and variables are standard. We assume fresh
+names everywhere to avoid name capture, and leave weakenings implicit.
+
+(2) There is a universe $\U$, with decoding written as an underline instead of
+usual $\El$, to bolster readability. With this part of the syntax, we can already
+define contexts specifying the empty type, unit type and booleans, or in general,
+a collections of finite sets:
+\[
+\boldsymbol{\cdot},\,\,\,Empty:\U \hspace{3em} \boldsymbol{\cdot},\,\,\,Unit:\U,\,\,\,tt:\underline{Unit} \hspace{3em} \boldsymbol{\cdot},\,\,\,Bool:\U,\,\,\,true:\underline{Bool},\,\,\,false:\underline{Bool}
+\]
+
+(3) We have a function space with small domain and large codomain. This can be
+used to add inductive arguments to all kinds of constructors. As $\U$ is not
+closed under this function space, these function types cannot (recursively)
+appear in inductive arguments, which ensures strict positivity. When the
+codomain does not depend on the domain, $a\ra B$ can be written instead of
+$(x:a)\ra B$.
+
+Now we can specify the natural numbers as a context:
+\[
+\boldsymbol{\cdot},\,\,\,Nat : \U,\,\,\,zero:\underline{Nat},\,\,\,suc:Nat\ra\underline{Nat}
+\]
+We can also encode inductive-inductive definitions such as the
+fragment of the well-typed syntax of a type theory mentioned in the
+introduction:
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot},\,\,\,Con:\U,\,\,\,Ty:Con\ra\U,\,\,\,\bullet:\underline{Con},\,\,\,\blank\rhd\blank:(\Delta:Con)\ra Ty\,\Delta\ra\underline{Con}, \\
+  & U : (\Delta:Con)\ra \underline{\Ty\,\Delta},\,\,\,\Pi:(\Delta:Con)(A:Ty\,\Delta)(B:Ty\,(\Delta\rhd A))\ra\underline{Ty\,\Delta}
+\end{alignat*}
+Note that this notion of inductive-inductive types is more general than the one
+considered in previous works \cite{forsberg-phd}, as we allow any number of type
+constructors, and arbitrary mixing of type and point constructors.
+
+(4) $\U$ is closed under the equality type, with eliminator $\J$ and a weak
+(propositional) $\beta$-rule. Weakness is required because the translations in
+Sections \ref{sec:morphisms} and \ref{sec:sections} do not preserve this
+$\beta$-rule strictly. We explain this in more detail in Sections TODO and
+TODO. Adding equality to the theory of codes allows higher constructors and
+inductive equality parameters as well. We can now define the higher-inductive
+circle as the following context:
+\[
+\boldsymbol{\cdot},\,\,\,S^1:\U,\,\,\,base:\underline{S^1},\,\,\,loop:\underline{base =_{S^1} base}
+\]
+
+The $\J$ rule allows constructors to mention operations on paths as well. For
+instance, the definition of the torus depends on path composition, which can be
+defined using $\J$: given $p:\underline{t=_a u}$ and $q:\underline{u=_a v}$, $p
+\sqcdot q$ abbreviates $\J_{a\,u\,x.z.(t=x)}\,p\,_v\,q : \underline{t=_a
+  v}$. The torus is given as follows.
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot},\,\,\,T^2:\U,\,\,\,b : \underline{T^2},\,\,\, p:\underline{b =_{T^2} b},\,\,\,q:\underline{b=_{T^2} b},\,\,\, t:\underline{p\sqcdot q=_{(b=_{T^2} b)} q\sqcdot p}
+\end{alignat*}
+With the equality type at hand, we can define a full well-typed syntax of type
+theory as given e.g.\ in \cite{ttintt} as an inductive type.  Also, see the
+examples in the Agda formalization described in Section
+\ref{sec:formalization}.
+
+So far we were only able to define closed HIITs, which do not refer to external
+types. Now we add rules which include external types into signatures. A context
+$\Delta$ for which $\hat{\Gamma}\vdash\Delta$ holds can be seen as a
+specification of an inductive type which depends on $\hat{\Gamma}$. For example,
+in the case of lists for arbitrary external element types, $\hat{\Gamma}$ will
+be $\hat{A}\in\Type_{\hat{0}}$.
+
+(5) We have a function space where the domain is a type in the external theory. We
+distinguish it from (3) by using red brick $\in$ instead of $:$ in the domain
+specification. We specify lists and the integers as follows.
+\begin{alignat*}{4}
+  & \hat{A}\in\Type_{\hat{0}} && \vdash\,\boldsymbol{\cdot},\,\, && List:\U,\,\,\,nil:\underline{List},\,\,\,cons:(\hat{x}\in \hat{A})\ra List\ra\underline{List} \\
+  & \hat{\Gamma} && \vdash\,\boldsymbol{\cdot},\,\,&& Int:\U,\,\,\,pair:(\hat{x}\,\hat{y}\in\hat{Nat})\ra\underline{Int},\,\,\, \\
+  & && && eq:(\hat{a}\,\hat{b}\,\hat{c}\,\hat{d}\in \hat{Nat})(\hat{p}\in \hat{a}\hat{\mathbin{+}}\hat{d}\eqh_{\hat{Nat}} \hat{b}\hat{\mathbin{+}}\hat{c})\ra \underline{pair\,\hat{a}\,\hat{b}=_{Int}\mathsf{pair}\,\hat{c}\,\hat{d}}, \\
+  & && && trunc:(x y : Int)(p\,q : a=_{Int} b)\ra \underline{p=_{x=_{Int} y} q}
+\end{alignat*}
+In the case of integers, $\hat{\Gamma}$ is
+$\hat{Nat}\in\Type_{\hat{0}}\hat{,}\,\blank\hat{\mathbin{+}}\blank\in\hat{Nat}\rah
+\hat{Nat}\rah\hat{Nat}$, or alternatively, we could require natural
+numbers in the external theory. As another example, propositional
+truncation for a type $\hat{A}$ is specified as follows.
+\[
+\hat{A}\in\Type_{\hat{0}}\vdash\, \boldsymbol{\cdot},\,\,\,tr:\U,\,\,\,emb : (\hat{x}\in \hat{A})\ra \underline{tr},\,\,\,eq:(x\,y:tr)\ra \underline{x=_{tr} y}
+\]
+The smallness of $\hat{A}$ is required in (5). It is possible to generalize
+signatures to arbitrary universe levels, but it is not essential to the current
+development. Note that the (5) function space preserves strict positivity, since
+in the external theory there is no way to recursively refer to the inductive type
+\emph{being defined}. The situation is analogous to the case of $W$-types
+\cite{abbot05containers}, where shapes and positions can contain arbitrary types
+but they cannot recursively refer to the $W$-type being defined.
+
+(6) $\U$ is also closed under a function space where the domain is an external
+type and the codomain is a small source theory type. We overload the application
+notation for external parameters, as it is usually clear from context which
+application is meant. The rules allow types with infinitary constructors, for
+example trees containing elements of $\hat{A}$ at the leaves and branching by
+$\hat{B}$ (which could be an infinite type):
+\[
+\hat{A}\in\Type_{\hat{0}}\commah\hat{B}\in\Type_{\hat{0}}\vdash\,\boldsymbol{\cdot},\,\,\,T:\U,\,\,\,leaf:(\hat{x}\in \hat{A})\ra\underline{T},\,\,\,node:((\hat{x}\in \hat{B})\ra T)\ra\underline{T}
+\]
+Here, $leaf$ has a function type (5) and $node$ has a function type (3) with a
+function type (6) in the domain. More generally, we can define $W$-types
+\cite{abbot05containers} as follows. $\hat{S}$ describes the ``shapes'' of the
+constructors and $\hat{P}$ the ``positions'' where recursive arguments can
+appear.
+\[
+\hat{S}\in\Type_{\hat{0}}\commah\hat{P}:\hat{S}\rah\Type_{\hat{0}}\vdash\,\boldsymbol{\cdot},\,\,\,W:\U,\,\,\,sup: (\hat{s} \in \hat{S})\ra((\hat{p} \in \hat{P}\,\hat{s})\ra W)\ra \underline{W}
+\]
+For a more complex infinitary example, see the definition of Cauchy
+reals in \cite[Definition 11.3.2]{HoTTbook}. It can be also found as
+an example file in our Haskell implementation.
+
+The invalid examples $\mathsf{Ival}$ and $\mathsf{Neg}$ from Section
+\ref{sec:intro} cannot be encoded by the theory of codes. For $\mathsf{Ival}$,
+we can go as far as
+\[
+\boldsymbol{\cdot},\,\,\,{Ival}:\U,\,\,\,a:\underline{{Ival}},\,\,\,b:\underline{{Ival}},\,\,\,\sigma:\underline{?  =_{{Ival}} ?}.
+\]
+The first argument of the function
+$\hat{\hat{f}\in(\hat{X}:\Type)\rah\hat{X}\rah\hat{X}}$ is an external type, but we
+only have ${Ival}:\U$ in the theory of codes. $\mathsf{Neg}$ cannot be typed
+because the first parameter of the constructor $\mathsf{con}$ is a function from
+a small type to an external type, and no such functions can be formed.
+
+\subsection{External type theory}
+\label{sec:external}
+
+We describe the external type theory here. It serves two purposes: it is a
+source of types external to a HIIT signature, and it also serves as the target
+for the syntactic translations described in Sections TODO to TODO. It is not
+essential that we use the same theory for both purposes; we do so only because
+it simplifies the presentation. Also, we need not specify the external theory in
+formal detail, as it is a standard type theory. We only need to make some
+assumptions about supported type formers. We generally keep the notation close
+to Agda, and use {\color{BrickRed}brick red} color to distinguish from
+constructions in the theory of signatures.
+
+There is a cumulative Russell-style hierarchy of universes $\Type_{\hat{i}}$, with
+universes closed under $\Pi$, $\Sigma$, equality and unit types.
+
+Dependent function space is denoted $\hat{(\hat{x}\in \hat{A})\rah \hat{B}}$. We
+write $\hat{A}\rah \hat{B}$ if $\hat{B}$ does not depend on $\hat{x}$, and $\rah
+$ associates to the right, $\hat{(\hat{x}\in \hat{A})(\hat{y}\in \hat{B})\rah
+  \hat{C}}$ abbreviates $\hat{(\hat{x}\in \hat{A})\rah (\hat{y}\in \hat{B})\rah
+  \hat{C}}$ and $\hat{(\hat{x}\,\hat{y}\in \hat{A})\rah \hat{B}}$ abbreviates
+$\hat{(\hat{x}\in \hat{A})(\hat{y}\in \hat{A})\rah \hat{B}}$. We write
+$\hat{\lambda x. t}$ for abstraction and $\hat{t}\,\hat{u}$ for left-associative
+application.
+
+$\hat{(\hat{x}\in \hat{A})\timesh \hat{B}}$ stands for $\Sigma$ types, $\hat{A}\timesh
+\hat{B}$ for the non-dependent version. We sometimes use a short re-associated
+notation for left-nested iterated $\Sigma$ types, for example $\hat{(A :
+  \Type_{\hat{0}})\times A \times A}$ may stand for the left-nested $\hat{(x :
+  (A : \Type_{\hat{0}})\times A) \times \proj_1\,A}$. The constructor for
+$\Sigma$ is denoted $\hat{(t,\,u)}$ with eliminators $\hat{\proj_1}$
+and $\hat{\proj_2}$. Both $\Pi$ and $\Sigma$ have definitional $\beta$ and
+$\eta$ rules.
+
+The equality type for a type $\hat{A}$ and elements $\hat{t}\in \hat{A}$,
+$\hat{u}\in \hat{A}$ is denoted $\hat{t}\eqh_{\hat{A}}\hat{u}$, and we have the
+constructor $\hat{\refl}_{\hat{t}}$ and the eliminator $\hat{\J}$ with
+definitional $\beta$-rule. The notation is
+$\hat{\J}_{\hat{A}\,\hat{t}\,\hat{P}}\,\hat{pr}\,_{\hat{u}}\,\hat{eq}$ for
+$\hat{t}\in \hat{A}$, $\hat{P}\in (\hat{x}\in \hat{A})\rah
+\hat{t}\eqh_\hat{A}\hat{x}\rah \Type_{\hat{i}}$, $\hat{pr}\in
+\hat{P}\,\hat{t}\,\hat{\refl}$ and $\hat{eq} \in
+\hat{t}\eqh_\hat{A}\hat{u}$. Sometimes we omit parameters in subscripts.
+
+We will use the following functions defined using $\hat{\J}$ in the
+standard way. We write $\TR_{\hat{P}}\,\hat{e}\,\hat{t}\in
+\hat{P}\,\hat{v}$ for transport of $\hat{t} \in \hat{P}\,\hat{u}$
+along $\hat{e} \in \hat{u}\eqh \hat{v}$. We write
+$\ap\,\hat{f}\,\hat{e}\in\hat{f}\,\hat{u}\eqh\hat{f}\,\hat{v}$ for
+$\hat{f}:\hat{A}\rah\hat{B}$ and $\hat{e}\in\hat{u}\eqh\hat{v}$,
+$\hat{\apd\,\hat{f}\,\hat{e} \in \TR_{\hat{P}}\,\hat{e}\,(\hat{f}\,\hat{u})
+\eqh \hat{f}\,\hat{v}}$ for $\hat{\hat{f}\in(\hat{x}\in \hat{A})\rah
+\hat{B}}$ and $\hat{e} \in \hat{u}\eqh \hat{v}$.
+
+The unit type is denoted $\hat{\top}$ with constructor $\tt$.
+
+%=====================================================================
+
+\section{General Definitions for Induction and Recursion}
+\label{sec:general}
+
+Armed with a definition for HIIT signatures, we would like to have of notions of
+\emph{induction} and \emph{recursion} for each signature. However, instead of
+trying to directly extract them from signatures, it is more helpful to have more
+fundamental (categorical) semantic concepts: \emph{algebras},
+\emph{homomorphisms}, \emph{displayed algebras} and \emph{sections of displayed
+  algebras}. Then, we can express induction and recursion using these.
+
+Let us first consider natural numbers, and see how the usual definition of
+induction arises.
+
+For $\Nat$, algebras are simply a triple consisting of a type, a value and an
+endofunction. Below, we leave universe indices explicit, and we use the notation
+of the external type theory described in Section \ref{sec:signatures}.
+\begin{alignat*}{5}
+& \Alg : \mathsf{Type} \\
+& \Alg \equiv (N : \mathsf{Type}) \times N \times (N \ra N)
+\end{alignat*}
+
+\noindent Displayed $\Nat$-algebras (sometimes called fibered algebras, as in \cite{sojakova})
+are likewise triples, but depending on a $\Nat$-algebra.
+\begin{alignat*}{5}
+& \DisplayedAlg : \Alg \ra \mathsf{Type} && \\
+& \DisplayedAlg\,(N,\,z,\,s) \equiv && \\
+&  \hspace{3em} (N^D : N \ra \mathsf{Type})\times (z^D : N^D\,z)\times ((n : N)\ra N^D\,n\ra N^D\,(s\,n))
+\end{alignat*}
+
+\noindent  Homomorphisms, as usual in mathematics, are structure-preserving functions:
+\begin{alignat*}{5}
+& \Morphism : \Alg \ra \Alg \ra \mathsf{Type} && \\
+& \Morphism\,(N_0,\,z_0,\,s_0)\,(N_1,\,z_1,\,s_1) \equiv && \\
+&  \hspace{3em} (N^M : N_0 \ra N_1)\times (z^M : N^M\,z_0 = z_1)\times ((n : N_0)\ra N^M\,(s_0\,n) = s_1\,(N^M\,n))
+\end{alignat*}
+
+\noindent Sections of displayed algebras can be viewed as a dependently typed analogue of homomorphisms:
+\begin{alignat*}{5}
+& \Section : (\alpha : \Alg) \ra \DisplayedAlg\,\,\alpha \ra \mathsf{Type} && \\
+& \Section\,(N,\,z,\,s)\,(N^D,\,z^D,\,s^D) \equiv && \\
+&  \hspace{3em} (N^S : (n : N) \ra N^D\,n)\times (z^S : N^S\,z = z^D)\times ((n : N)\ra N^S\,(s\,n) = s^D\,(N^S\,n))
+\end{alignat*}
+
+\noindent Now, we can reformulate induction for $\Nat$. First, we assume that
+there exists a distinguished $\Nat$-algebra, named $\mathsf{InitialAlg}$. The
+induction principle has the following type:
+\[
+ \mathsf{Induction} : (M : \mathsf{\DisplayedAlg}\,\,\mathsf{InitialAlg}) \ra \Section\, M
+\]
+\noindent Unfolding the definitions, it is apparent that this is the same notion
+of $\Nat$-induction as we gave before. The initial algebra consists of type and value
+constructors, the induction motives and methods are bundled into a displayed
+algebra, and as result we get a section, containing an eliminator function together
+with its $\beta$-rules. Additionally, we can define recursion using homomorphisms:
+\[
+\mathsf{Recursion} : (\alpha : \Alg) \ra \Morphism\,\mathsf{InitialAlg}\,\alpha
+\]
+This corresponds to \emph{weak initiality} in the sense of category theory: for
+each algebra, there is a morphism from the weakly initial algebra to it. Strong initiality
+in the setting of higher inductive types is called \emph{homotopy initiality} \cite{sojakova},
+and it is defined as follows for $\Nat$:
+\begin{alignat*}{5}
+\mathsf{Initiality} : (\alpha : \Alg) \ra \mathsf{isContr}\,(\Morphism\,\mathsf{InitialAlg}\,\alpha)
+\end{alignat*}
+
+\noindent where $\mathsf{isContr}\,A \equiv (a : A)\times((a' : A)\ra a = a')$.
+Hence, there is a unique morphism from the initial algebra, but in the setting of homotopy type theory, unique inhabitation can be viewed instead as contractibility.
+
+Observe that the definitions for $\mathsf{Induction}$, $\mathsf{Recursion}$ and
+$\mathsf{Initiality}$ need not refer to natural numbers, and can be used
+similarly in cases of other structures. Thus, the task in the following is to
+derive algebras, homomorphisms, displayed algebras and sections from HIIT
+signatures, in a way which generalizes beyond the current $\Nat$ example to
+indexed types, induction-induction and higher constructors. But even in the
+general case, displayed algebras yield induction motives and methods, and
+homomorphisms and sections yield a function for each type constructor and a
+$\beta$-rule for each point or path constructor.
+
+We will compute algebras and the other notions by induction on the syntax of the
+theory of signatures. However, first we need to clarify the formal justifications
+behind these computations.
+
+\section{Coherence Problems and Syntactic Translations}
+\label{sec:coherence}
+
+TODO
+
+%% --------------------------------------------------------------------------------
+
+%% \section{A syntactic translation from the source to the target theory}
+%% \label{sec:translation}
+
+%% An inductive type is specified by a context in the theory of codes
+%% defined in the previous section. In this section we define the
+%% $\blank^\C$, $\blank^\M$ and $\blank^\E$ operations, which work as follows
+%% on the code for natural numbers.
+%% \begin{alignat*}{5}
+%%   & && (\boldsymbol{\cdot},\,Nat : \U,\,zero:\underline{Nat},\,suc:Nat\ra\underline{Nat})^\C \\
+%%   & \equiv \,\, && \hat{\top}\timesh(\hat{n}\in\Type_{\hat{0}})\timesh(\hat{z}\in \hat{n})\timesh(\hat{n}\rah \hat{n}) \\
+%%   & && (\boldsymbol{\cdot},\,Nat : \U,\,zero:\underline{Nat},\,suc:Nat\ra\underline{Nat})^\M\,(\tt\commah\hat{n}\commah \hat{z}\commah \hat{s}) \\
+%%   & \equiv \,\, && \hat{\top}\timesh(\hat{n^M}\in \hat{n} \rah \Type_{i})\timesh(\hat{z^M}\in \hat{n^M}\,\hat{z})\timesh \big((\hat{x}\in \hat{n})\rah  \hat{n^M}\,\hat{x}\rah  \hat{n^M}\,(\hat{s}\,\hat{x})\big) \\
+%%   & && (\boldsymbol{\cdot},Nat : \U,zero:\underline{Nat},suc:Nat\ra\underline{Nat})^\E\,(\tt\commah \hat{n}\commah \hat{z}\commah \hat{s})\,(\tt\commah\hat{n^M}\commah \hat{z^M}\commah \hat{s^M}) \\
+%%   & \equiv \,\, && \hat{\top}\timesh\big(\hat{n^E} \in (\hat{x}\in \hat{n})\rah  \hat{n^M}\,\hat{x}\big)\timesh(\hat{z^E} \in \hat{n^E}\,\hat{z}\eqh \hat{z^M}) \timesh \big((\hat{x}\in \hat{n})\rah  \hat{n^E}\,(\hat{s}\,\hat{x})\eqh \hat{n^M}\,\hat{x}\,(\hat{n^E}\,\hat{x})\big)
+%% \end{alignat*}
+%% The {\color{BrickRed}brick red} coloured result of $\blank^\C$ gives
+%% the types of the type formation rule and constructors as an iterated
+%% $\Sigma$-type. Assuming the existence of constructors, $\blank^\M$
+%% returns the types of motives and methods. Assuming the existence of
+%% constructors and the corresponding motives and methods, $\blank^\E$
+%% returns the types of the eliminator and computation rules as external
+%% equalities $\eqh$. The notation above denotes
+%% \emph{left}-nested iterated $\Sigma$-types. A more precise
+%% presentation would replace each variable with a projection from the
+%% preceding $\Sigma$-type. We use this notation in order to reduce
+%% visual clutter.
+
+%% Each context entry in the theory of codes specifies a type formation
+%% rule or a constructor. In general, the last component of a type in a
+%% context entry is of three possible forms: it is either $\U$,
+%% $\underline{a}$ for some neutral $a$, or $\underline{t=_a u}$. The
+%% following table summarizes the results of the various translations in
+%% the mentioned three cases:
+%% \begin{alignat*}{5}
+%%   & \text{return type}\hspace{1em} && \blank^\C && \blank^\M && \blank^\E \\\hline
+%%   & \U && \text{type formation rule}\hspace{1em} && \text{motive} && \text{eliminator} \\
+%%   & \underline{a} && \text{point constructor}\hspace{1em} && \text{method} && \text{computation rule} \\
+%%   & \underline{t=_a u} && \text{path constructor} && \text{method expressing} && \text{higher computation rule} \\
+%%   & && && \text{preservation of equality}\hspace{1em}
+%% \end{alignat*}
+%% Note that there is no syntactic distinction between the three kinds of
+%% constructors above. Any number of them can be introduced in any order,
+%% and each constructor can refer to any previous one. A distinguishing
+%% feature of our approach is the utilisation of universes instead of
+%% structural rules to introduce new sorts and to ensure strict
+%% positivity.
+
+%% The $\blank^\C$, $\blank^\M$ and $\blank^\E$ operations are defined by
+%% induction on the derivations of the source syntax. The operations are
+%% identity on derivations of external contexts and external terms (of the
+%% forms $\vdashh\hat{\Gamma}$ and $\hat{\Gamma}\vdashh \hat{t}\in
+%% \hat{A}$) and derive target terms from theory of codes contexts, types
+%% and terms (of the forms $\hat{\Gamma}\vdash\Delta$,
+%% $\hat{\Gamma}\semicol\Delta\vdash A$ and
+%% $\hat{\Gamma}\semicol\Delta\vdash t:A$, respectively). We only present
+%% the non-identity parts with pattern matching notation, describing how
+%% a context, type or a term in the theory of codes is converted to a
+%% term in the external theory.
+
+%% All three operations respect definitional equality. This amounts to
+%% preserving equalities of the substitution calculus, as there are no
+%% $\beta$-rules introduced in the theory of codes.
+
+%---------------------------------------------------------------------
+
+\section{Algebras}
+\label{sec:algebras}
+
+Given a context in the theory of signatures, $\blank^\A$ computes the
+corresponding notion of algebras as an iterated $\Sigma$-type in the external
+syntax. It is specified as follows.
+\[
+\infer{\hat{\Gamma}\vdashh\Delta^\A\in\Type_{\hat{1}}}{\hat{\Gamma}\vdash\Delta}
+\hspace{2em}
+\infer{\hat{\Gamma}\vdashh A^\A \in \Delta^\A\rah \Type_{\hat{1}}}{\hat{\Gamma}\semicol\Delta\vdash A}
+\hspace{2em}
+\infer{\hat{\Gamma}\vdashh t^\A \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah  A^\A\,\hat{\gamma}}{\hat{\Gamma}\semicol\Delta\vdash t : A}
+\]
+
+The $\blank^\A$ translation is essentially the standard interpretation of
+signatures, where every construction in the source syntax is interpreted with a
+corresponding {\color{BrickRed}brick red} construction in the target syntax. The
+only notable change is in the interpretation of contexts: a source context is
+interpreted as an iterated $\Sigma$-type.
+
+\begingroup
+\allowdisplaybreaks
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot}^\A && :\equiv \hat{\top} \\
+  & (\Delta,x:A)^\A && :\equiv \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\timesh A^\A\,\hat{\gamma} \\
+  & x^\A\,\hat{\gamma} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma} \\ % TODO: write De Bruijn versions
+  & \U^\A\,\hat{\gamma} && :\equiv \Type_{\hat{0}} \\
+  & (\underline{a})^\A\,\hat{\gamma} && :\equiv a^\A\,\hat{\gamma} \\
+  & ((x:a)\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)} \\
+  & (t\,u)^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\A\,\hat{\gamma}\hat{)} \\
+  & ((\hat{x}\in \hat{A})\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A})\rah  B^\A\,\hat{\gamma} \\
+  & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u} \\
+  & (t=_a u)^\A\,\hat{\gamma} && :\equiv t^\A\,\hat{\gamma} \eqh u^\A\,\hat{\gamma} \\
+  & \refl^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\A\,\hat{\gamma} && :\equiv \hat{\J}_{\hat{(}a^\A\,\hat{\gamma}\hat{)}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}\hat{\lambda} \hat{x}\,\hat{z}.p^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\commah\hat{z}\hat{)}\hat{)}}\,\hat{(}pr^\A\,\hat{\gamma}\hat{)}\,_{\hat{(}u^\A\,\hat{\gamma}\hat{)}}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)} \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & ((\hat{x}\in \hat{A})\ra b)^\A\,\hat{\gamma} && :\equiv (\hat{x}\in \hat{A}\hat{)}\rah  b^\A\,\hat{\gamma} \\
+  & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u}
+\end{alignat*}\endgroup
+
+For example, $\blank^\A$ acts as follows on signature for circles:
+\begin{alignat*}{5}
+  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\A \equiv \hat{\hat{\top}\timesh(\hat{S^1}\in\Type_{\hat{0}})\timesh(\hat{b}\in \hat{S^1})\timesh(\hat{loop\in b=b})}
+\end{alignat*}
+
+Note that the resulting left-nested $\Sigma$ type could be written explicitly as
+below, without syntactic sugar. We shall keep to the more readable notation
+from now on.
+\begin{alignat*}{5}
+  & \hat{\Big(\hat{x''}\in\big(\hat{x'}\in(\hat{x}\in\hat{\top})\timesh\Type_{\hat{0}}\big)\timesh\hat{\proj_2}\,\hat{x'}\Big)\timesh(\hat{\proj_2}\,\hat{x''}\eqh \hat{\proj_2}\,\hat{x''})}
+\end{alignat*}
+
+%---------------------------------------------------------------------
+
+\section{From Logical Relations to Homomorphisms}
+\label{sec:morphisms}
+
+TODO
+
+\section{Displayed Algebras}
+\label{sec:displayed}
+
+The $\blank^\D$ translation computes displayed algebras for a signature. Recall
+from Section \ref{sec:general} that displayed algebras serve as bundles of
+induction motives and methods.
+
+$\blank^\D$ is essentially an unary logical predicate translation over the
+$\blank^\A$ model. It is related to the logical predicate translation of
+Bernardy et al.\ \cite{bernardy12parametricity}, but our implementation differs
+by interpreting contexts as $\Sigma$-types instead of extended contexts.
+
+We fix a level $i$ for the universe we would like to eliminate into. For each
+context $\Delta$, $\Delta^\D$ is a predicate over the standard interpretation
+$\Delta^\A$. For a type $\Delta\vdash A$, $A^\D$ is a predicate over $A^\A$,
+which also depends on $\hat{\gamma}\in\Delta^\A$ and a witness of
+$\Delta^\D\,\hat{\gamma}$. All of these may refer to a target theory context
+$\hat{\Gamma}$.
+\[
+\infer{\hat{\Gamma}\vdashh\Delta^\D \in \Delta^\A\rah \Type_{\hat{i+1}}}{\hat{\Gamma}\vdash\Delta}
+\hspace{2em}
+\infer{
+  \hat{\Gamma}\vdashh A^\D \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah  A^\A\,\hat{\gamma}\rah \Type_{\hat{i+1}}}
+      {\hat{\Gamma}\semicol\Delta\vdash A}
+\]
+For a term $t$, $t^\D$ witnesses that the predicate corresponding to
+its type holds for $t^\A$. This is often called a \emph{fundamental
+  theorem} in the literature on logical predicates.
+\[
+\infer{\hat{\Gamma}\vdashh t^\D \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}}{\hat{\Gamma}\semicol\Delta\vdash t : A}
+\]
+We introduce the following shorthand: $t\,\hat{\gamma}\,\hat{\gamma^D}$ is
+abbreviated as $t\,\hat{\gamma^2}$ for some $t$ expression. The
+implementation of $\blank^\D$ is given below.  \begingroup
+%% \allowdisplaybreaks
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot}^\D\,\hat{\gamma} && :\equiv \hat{\top} \\
+  & (\Delta,\,x:A)^\D\,(\hat{\gamma}\commah\,\hat{t}) && :\equiv \hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\timesh A^\D\,\hat{\gamma^2}\,\hat{t} \\
+  & x^\D\,\hat{\gamma^2} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^D} \\ % TODO: De Bruijn
+  & \U^\D\,\hat{\gamma^2}\,\hat{A} && :\equiv \hat{A} \rah  \Type_{\hat{i}} \\
+  & (\underline{a})^\D\,\hat{\gamma^2}\,\hat{t} && :\equiv a^\D\,\hat{\gamma^2}\,\hat{t} \\
+  & ((x:a)\ra B)^\D\,\hat{\gamma^2}\,\hat{f} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\hat{(}\hat{x^D}\in a^\D\,\hat{\gamma^2}\,\hat{x}\hat{)} \rah  B^\D\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)}\,\hat{(}\hat{\gamma^D}\commah\hat{x^D}\hat{)}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,u)^\D\,\hat{\gamma^2} && :\equiv \hat{(}t^\D\,\hat{\gamma^2}\hat{)}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\D\,\hat{\gamma^2}\hat{)} \\
+  & ((\hat{x}\in \hat{A})\ra B)^\D\,\hat{\gamma^2}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^2}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^\D\,\hat{\gamma^2} && :\equiv t^\D\,\hat{\gamma^2}\,\hat{u} \\
+  & (t=_a u)^\D\,\hat{\gamma^2}\,\hat{e} && :\equiv \TR_{\hat{(}a^\D\,\hat{\gamma^2}\hat{)}}\,\hat{e}\,\hat{(}t^\D\,\hat{\gamma^2}\hat{)} \eqh u^\D\hat{\gamma^2} \\
+  & (\refl_t)^\D\,\hat{\gamma^2} && :\equiv \hat{\refl}_{\hat{(}t^\D\,\hat{\gamma^2}\hat{)}} \\
+  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\D\,\hat{\gamma^2} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^2}\hat{)} \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\D\,\hat{\gamma^2} && :\equiv \hat{\refl} \\
+  & ((\hat{x}\in \hat{A})\ra b)^\D\,\hat{\gamma^2}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^\D\,\hat{\gamma^2}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^\D\,\hat{\gamma^2} && :\equiv t^\D\,\hat{\gamma^2}\,\hat{u}
+\end{alignat*}\endgroup
+The predicate for a context is given by iterating $\blank^\D$ for its
+constituent types. For a variable, the corresponding witness is looked up from
+$\hat{\gamma^D}$.
+
+The translation of the universe, given an element of
+$\hat{A}\in\U^\A\,\hat{\gamma}$ (with
+$\U^\A\,\hat{\gamma}\equiv\Type_{\hat{0}}$) returns the predicate space over
+$\hat{A}$. For $\underline{a}$ types, we just return the translation of $a$.
+
+The predicate for a function type with small domain expresses preservation of
+predicates (at the domain and codomain types). Witnesses of application are
+given by recursive application of $\blank^\D$. The definitions for the other
+(non-inductive) function spaces are similar, except there is no predicate for
+the domain types, and thus no witnesses are required.
+
+The predicate for the equality type $t=_a u$, for each $\hat{e}\in(t=_a
+u)^\A\,\hat{\gamma}$, i.e.\ $\hat{e}\in t^\A\,\hat{\gamma}\eqh
+u^\A\,\hat{\gamma}$, says that $t^\D$ and $u^\D$ are equal. As these have
+different types, we have to transport over the original equality
+$\hat{e}$. Hence, induction methods for path constructors will be \emph{paths
+  over paths} in the sense of homotopy type theory.
+
+$\refl$ is interpreted with just reflexivity in the target syntax.  The
+interpretation of $\J$ is given by a double $\hat{\J}$ application, which
+definition is sourced from \cite{lasson}. Here, we use a shortened $\hat{\J}$
+notation; we refer to the formalization (Section \ref{sec:formalization}) for
+the details.
+
+Again, let us consider the circle example:
+\begin{alignat*}{5}
+  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\D\,\,(\hat{\tt,\,S^1,\,b,\,loop})\\
+  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{1D}}\in \hat{S^1}\rah\Type_i)\timesh(\hat{b^D}\in \hat{S^{1D}\,b})\timesh(\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D})
+\end{alignat*}
+The inputs of $\blank^\D$ here are the signature for the circle (the context in
+black) and an $\hat{S^{1}}$-algebra consisting of three non-$\top$ components.
+It returns a family over the type $\hat{S^1}$, an element of this family
+$\hat{b^D}$ at index $\hat{b}$, and and an equality between $\hat{b^D}$ and
+$\hat{b^D}$ which lies over $\hat{loop}$. This is the same as the usual
+induction motives and methods for the circle, e.g.\ as described in \cite{HoTTbook}.
+
+%---------------------------------------------------------------------
+
+\section{Displayed Algebra Sections}
+\label{sec:sections}
+
+TODO: revise when we have homomorphism section, in order to remove redundant
+explanations. Also revise language.
+
+The operation $\blank^\S$ yields displayed algebra sections. It is a generalized
+binary logical relation translation where the type of the second parameter of
+the relation may depend on the first parameter. Displayed algebra sections can
+be viewed as dependent homomorphisms: while homomorphisms are
+structure-preserving families of functions, sections are structure-preserving
+families of dependent functions.
+
+Contexts are interpreted as dependent binary relations between algebras and
+displayed algebras. The universe level $i$ was previously chosen for the
+$\blank^\D$ translation.
+\[
+\infer{\hat{\Gamma}\semicol\Delta^\S \in (\hat{\gamma}\in\Delta^\A)\rah \Delta^\D\,\hat{\gamma}\rah \Type_{i}}{\hat{\Gamma}\vdash\Delta}
+\]
+Types are interpreted as dependent binary relations which additionally
+depend on ($\hat{\gamma}\commah\,\hat{\gamma^D}\commah\,\hat{\gamma^S}$) interpretations of the
+context.
+\[
+\infer{\hat{\Gamma}\vdashh A^\S \in (\hat{\gamma}\in\Delta^\A)(\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma})(\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D})(\hat{x}\in A^\A\,\hat{\gamma})\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{x}\rah \Type_{i}}{\hat{\Gamma}\semicol\Delta\vdash A}
+\]
+For a term $t$, $t^\S$ witnesses that the relation corresponding to
+its type holds for $t^\A$ and $t^\D$.
+\[
+\infer{\hat{\Gamma}\vdashh t^\S \in (\hat{\gamma}\in\Delta^\A)(\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma})(\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D})\rah  A^\S\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{\gamma^S}\,(t^\A\,\hat{\gamma})\,(t^\D\,\hat{\gamma}\,\hat{\gamma^D})}{\hat{\Gamma}\semicol\Delta \vdash t : A}
+\]
+In addition to $\hat{\gamma^2}$, we use $t\,\hat{\gamma^3}$ to abbreviate $t\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{\gamma^S}$. The implementation is the following.
+%% \begingroup \allowdisplaybreaks
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot}^\S\,\hat{\gamma}\,\hat{\gamma^D} && :\equiv \hat{\top} \\
+  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv (\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2})\timesh A^\S\,\hat{\gamma^3}\,\hat{t}\,\hat{t^D} \\
+  & x^\S\,\hat{\gamma^3} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^S} \\ % TODO: De Bruijn
+  & \U^\S\,\hat{\gamma^3}\,\hat{A}\,\hat{A^D} && :\equiv (\hat{x}\in \hat{A})\rah  \hat{A^D}\,\hat{x} \\
+  & (\underline{a})^\S\,\hat{\gamma^3}\,\hat{t}\,\hat{t^D} && :\equiv a^\S\,\hat{\gamma^3}\,\hat{t} \eqh \hat{t^D} \\
+  & ((x:a)\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv (\hat{x}\in a^\A\,\hat{\gamma})\rah  B^\S\,(\hat{\gamma,x})\,(\hat{\gamma^D,}\,a^\S\,\hat{\gamma^3}\,\hat{x})\,(\hat{\gamma^S,}\,\hat{\refl}) \\
+  & && \hspace{8.4em} (\hat{f}\,\hat{x})\,\big(\hat{f^D}\,\hat{x}\,(a^\S\,\hat{\gamma^3}\,\hat{x})\big) \\
+  & (t\,u)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,(t^\S\,\hat{\gamma^3}\,( u^\A\,\hat{\gamma}))\,(u^\S\,\hat{\gamma^3}) \\
+  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv (\hat{x}\in \hat{A})\rah B^\S\,\hat{\gamma^3}\,(\hat{f}\,\hat{x})\,(\hat{f^D}\,\hat{x}) \\
+  & (t\,\hat{u})^\S\,\hat{\gamma^3} && :\equiv t^\S\,\hat{\gamma^3}\,\hat{u} \\
+  & (t=_a u)^\S\,\hat{\gamma^3}\,\hat{e} && :\equiv \TR\,(t^\S\,\hat{\gamma^3})\big(\TR\,(u^\S\,\hat{\gamma^3})\,(\apd\,(a^\S\,\hat{\gamma^3})\,\hat{e})\big) \\
+  & (\refl_t)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,\hat{\refl}\,(t^\S\,\hat{\gamma^3}) \\
+  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S\,\hat{\gamma^3} && :\equiv \\
+  & && \hspace{-10em}\hat{\J}\,\bigg(\hat{\J}\,\Big(\hat{\J}\,\big(\hat{\J}\,(\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}.\,\hat{pr^S})\,(t^\S\,\hat{\gamma^3})\,\\
+  & && \hspace{-9em}(\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2})\,(\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3})\,(pr^\D\,\hat{\gamma^2})\,(pr^\S\,\hat{\gamma^3})\big)\,eq^\A\,\hat{\gamma}\Big)\,u^\S\,\hat{\gamma^3}\bigg)\,(eq^\S\,\hat{\gamma^3}) \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^3} && :\equiv \\
+  & && \hspace{-10em}\hat{\J}\,\big(\hat{\J}\,(\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl})\,(t^\S\,\hat{\gamma^3})\,(\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2})\,(\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3})\big)\,(pr^\S\,\hat{\gamma^3})\\
+  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^3\,f\,t} && :\equiv b^\S\,\hat{\gamma^3}\,(\hat{f}\,\hat{t}) \\
+  & (t\,\hat{u})^\S\,\hat{\gamma^3} && :\equiv \ap\,(\hat{\lambda} \hat{f.}\hat{f}\,\hat{u})\,(t^\S\,\hat{\gamma^3})
+\end{alignat*}
+%% \endgroup
+
+The $\U^\S$ and $(\underline{a})^\S$ definitions are the key points of
+the $\blank^\S$ operation. The definitions for the other $\blank^\S$
+cases are largely determined by these.
+
+The $\U^\S$ rule yields the type of the eliminator for a type
+formation rule. In the natural numbers example above, the non-indexed
+$Nat : \U$ sort is interpreted as $\hat{n^S} \in (\hat{x}\in
+\hat{n})\rah \hat{n^D}\,\hat{x}$. For indexed sorts, the indices are
+first processed by the $\blank^\S$ cases for inductive and non-inductive
+parameters, until the ultimate $\U$ return type is reached. Hence, the
+eliminator for a sort is always a function.
+
+Analogously, the $\blank^\S$ result type for a point or path constructor is
+always a $\beta$-rule, i.e.\ a function type ending in an equality. To
+see why, consider the $(\underline{a})^\S$ definition. It expresses
+that applications of $a^\S$ eliminators must be equal to the
+corresponding $\hat{t^D}$ induction methods. Hence, for path and point
+constructor types, $\blank^\S$ works by first processing all inductive and
+non-inductive indices, then finally returning an equality type.
+
+Let us also consider the $((x:a)\ra B)^\S$ case for inductive
+parameters. Here, we make crucial use of the fact that the domain type
+$a$ is small. This provides us $a^\S\,\hat{\gamma^3\,x}$, which we use
+to witness the $a^\D\,\hat{\gamma^2}\,\hat{x}$ hypothesis for
+$B^\S$. Without the size restriction on inductive parameters (which
+enforces strict positivity), the $\blank^\S$ operation would not be
+possible at all, because $a^\S$ for a large $a$ type would be merely
+an opaque relation instead of an eliminator function.
+
+Here, we only provide abbreviated definitions for the $t\,u$, $t=_a
+u$, $\refl$, $\J$ and $\J\beta$ cases. In the $\J$ case, we write
+$\hat{\mathsf{uncurry}}\,p^\D$ for
+$\hat{\lambda\,\gamma\,\gamma^D\,x\,x^D\,z\,z^D.}\,p^\D\,(\hat{\gamma,x,z})\,(\hat{\gamma^D,x^D,z^D})$
+and analogously elsewhere. The full definitions can be found in the
+Agda formalization. The definitions are highly constrained by the
+required types, and not particularly difficult to implement with the
+help of a proof assistant; they all involve doing successive path
+induction on all equalities available from induction hypotheses, with
+appropriately generalized induction motives.
+
+The full $(\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S$ definition is quite
+large, and, for instance, yields a very large $\beta$-rule for the
+higher inductive torus definition (the reader can confirm this using
+the Haskell implementation). Nevertheless, an implementation focused
+on practicality may provide smaller specialized $\blank^\S$ definitions for
+commonly used equality operations such as composition or inverses.
+
+The circle example is a bit more interesting here:
+\begin{alignat*}{5}
+  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\S\,\,(\hat{\tt,\,S^1,\,b,\,loop})\,(\hat{\tt,\,S^{1M},\,b^D,\,loop^D}) \\
+  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{1E}}\in (\hat{x}\in\hat{S^1})\hat{\,\ra\,}\hat{S^{1M}\,x})\timesh(\hat{b^S}\in \hat{S^{1E}\,\,b = b^D})\\
+  & && \hspace{0.75em}\timesh(\hat{loop^S\,\,\in\,\,} \TR\,_{(\hat{\lambda x.
+\TR\,_{S^{1M}}\,loop\,x\,=\,b^D})}\,\hat{b^S}\,(\TR\,_{(\hat{\lambda} \hat{x}.
+\TR\,_{\hat{S^{1M}}}\,\hat{loop}\,(\hat{S^{1E}\,b})\,\eqh\,\hat{x})}\,\hat{b^S}\,(\apd\,\hat{S^{1E}\,loop}))\,\\
+  & && \hspace{5.5em} \hat{=\,loop^D})
+\end{alignat*}
+In homotopy type theory, the $\beta$-rule for $loop$ is usually just
+$\apd\,\hat{S^{1E}\,loop}\,\hat{\,=\,loop^D}$, but here all
+$\beta$-rules are propositional, so we need to transport with
+$\hat{b^S}$ to make the equation well-typed. When computing the type
+of $\hat{loop^S}$, we start with $(\underline{b
+  =b})^S\,\,\hat{\gamma^3}\,\,\hat{loop}\,\,\hat{loop^D}$. Next, this
+evaluates to $(b=b)^\S\,\hat{\gamma^3}\,\hat{loop}\eqh\hat{loop^D}$,
+and then we unfold the left hand side to get the doubly-transported
+expression in the result.
+
+In Appendix \ref{sec:app}, we show how the elimination principle is
+computed for the two-dimensional sphere.
+
+For another example for the translations, we consider indexed W-types
+which can describe a large class of inductive definitions
+\cite{morris09indexed}. Suppose we are in the target context $\hat{I}
+\in \Type_{\hat{0}}\commah\hat{S} \in \Type_{\hat{0}}\commah\hat{P} \in \hat{S} \rah
+\Type_{\hat{0}}\commah\hat{out} \in \hat{S} \rah \hat{I}\commah\hat{in} \in
+(\hat{s} \in \hat{S})\rah \hat{P}\,\hat{s}\rah \hat{I}$. Then, the
+code for the corresponding indexed W-type is the following:
+\[
+W :\equiv (\boldsymbol{\cdot},\,\,\,w:(\hat{i} \in \hat{I})\ra\U,\,\,\,sup: (\hat{s} \in \hat{S})\ra((\hat{p} \in \hat{P\,s})\ra w\,(\hat{in\,s\,p}))\ra \underline{w\,(\hat{out\,s})})
+\]
+We pick a universe level $j$ for elimination. The interpretations of $W$ are the following, omitting leading $\hat{\top}$ components:
+\begin{alignat*}{5}
+  & \rlap{$W^\A \equiv (\hat{w}\in \hat{I}\rah\Type_{\hat{0}})\timesh((\hat{s}\in \hat{S})\rah((\hat{p}\in \hat{P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\rah \hat{w}\,(\hat{out\,s}))$}\\
+  & W^\D\,(\hat{w, sup}) && \equiv\, && (\hat{w^D} \in (\hat{i}\in \hat{I})\rah \hat{w\,i}\rah \Type_j)\\
+  & && \hspace{0.23em}\timesh && \big((\hat{s\in S})(\hat{f}\in (\hat{p\in P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\\
+  & && && \rah ((\hat{p\in P\,s})\rah \hat{w^D}\,(\hat{in\,s\,p})\,(\hat{f\,p}))\rah \hat{w^D}\,(\hat{out\,s})\,(\hat{sup\,s\,f})\big) \\
+  & \rlap{$ W^\S\,(\hat{w, sup})\,(\hat{w^D,sup^D}) \equiv $} \\
+  & && && (\hat{w^S} \in (\hat{i \in I})(\hat{x \in w\,i})\rah \hat{w^D\,i\,x})\\
+  & && \hspace{0.23em}\timesh && \big((\hat{s\in S})(\hat{f}\in (\hat{p\in P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\\
+  & && && \rah \hat{w^S}\,(\hat{out\,s})\,(\hat{sup\,\,s\,f}) \eqh \hat{sup^D\,s\,f}\,(\hat{\lambda} \hat{p.}\, \hat{w^S}\,(\hat{in\,s\,p})\,(\hat{f\,p}))\big)
+\end{alignat*}
+
+%=====================================================================
+
+\section{On Extension to Categorical Semantics}
+\label{sec:categorical}
+
+TODO
+
+%% \section{Existence of HIITs}
+%% \label{sec:adding}
+
+%% The target type theory supports HIITs if whenever we can derive
+%% $\hat{\Gamma}\vdash\Delta$ in the source theory, the following rules are
+%% admissible.
+%% \[
+%% \infer{\hat{\Gamma}\vdashh\hat{\con}_\Delta\in \Delta^\C}{\vspace{1em}}
+%% \hspace{5em}
+%% \infer{\hat{\Gamma}\vdashh\hat{\elim}_\Delta\,\hat{m}\in\Delta^\E\,\hat{\con}_\Delta\,\hat{m}}{\hat{\Gamma}\vdashh \hat{m}\in \Delta^\M\,\hat{\con}_\Delta}
+%% \]
+%% We can add HIITs to the target theory by extending it with the theory
+%% of codes (making the target and the source theory the same) and adding
+%% the above rules with the additional assumption of
+%% $\hat{\Gamma}\vdash\Delta$. However, this only adds HIITs with weak
+%% computation rules. To make the computation rules definitional, we
+%% would probably need a two-level target type theory with an equality
+%% having an equality reflection rule as in Voevodsky's homotopy type
+%% system \cite{hts} or Andromeda \cite{andromeda}.
+
+%=====================================================================
+
+\section{Formalization and Implementation}
+\label{sec:formalization}
+
+There are three additional development artifacts to the current work:
+a Haskell implementation, a shallow Agda formalization and a deep Agda
+formalization. All three are available from the homepage of the first
+author.
+
+The Haskell implementation takes as input a file which contains a
+representation of a $\hat{\Gamma}\vdash\Delta$ specification of a
+HIIT. Then, it checks the input with respect to the rules in Figure
+\ref{sigrules}, and outputs an Agda-checkable file which contains the
+results of the $\blank^\C$, $\blank^\M$ and $\blank^\E$
+translations. It comes with examples, including the ones in this
+paper, indexed W-types \cite{morris09indexed}, the dense completion
+\cite[Appendix A.1.3]{forsberg-phd} and several HITs from
+\cite{HoTTbook} including the definition of Cauchy reals. It can be
+checked that our implementation computes the expected elimination
+principles in these cases.
+
+The shallow Agda formalization embeds both the source and target theories
+shallowly into Agda: it represents types as Agda types, functions as
+Agda functions, and so on. We also leave the $\blank^\C$ operation
+implicit. We state each case of the $\blank^\M$ and $\blank^\E$
+translations as Agda functions from all induction hypotheses to the
+result type of the translation, which lets us ``typecheck'' the
+translation. We have found that this style of formalization is
+conveniently light, but remains detailed enough to be useful. We also
+generated some of the code of the Haskell implementation from this
+formalization.
+
+The deep Agda formalization still uses a shallow embedding of the
+target type theory, but it uses a deep embedding of the source theory,
+in the style of \cite{ttintt}. In the formalization we merge the three
+translations into a single model construction. This allows us to prove
+strict preservation of definitional equalities in the substitution
+calculus of the source theory, in contrast to the shallow
+formalization, where we cannot reason about definitional equalities of
+Agda terms. Due to technical challenges, this formalization uses
+transport instead of $\J$ in the source theory, but this still covers
+a rather large class of HIIT definitions.
+
+%=====================================================================
+
+\section{Conclusions and further work}
+\label{sec:summary}
+
+Higher inductive-inductive types are useful in defining the well-typed
+syntax of type theory in an abstract way \cite{ttintt}. From a
+universal algebra point of view, they provide initial algebras for
+multi-sorted algebraic theories where the sorts can depend on each
+other. From the perspective of homotopy type theory, they provide
+synthetic versions of homotopy-theoretic constructions such as higher
+dimensional spheres or cell complexes. So far, no general scheme of
+HIITs have been proposed. To quote Lumsdaine and Shulman
+\cite{lumsdaineShulman}:
+\begin{quotation}
+``The constructors of an ordinary inductive type are unrelated to each
+other.  But in a higher inductive type each constructor must be able
+to refer to the previous ones; specifically, the source and target of
+a path-constructor generally involve the previous
+point-constructors. No syntactic scheme has yet been proposed for this
+that covers all cases of interest while remaining meaningful and
+consistent.''
+\end{quotation}
+In this paper we proposed such a syntactic scheme which also includes
+inductive-inductive types. We tackled the problem of complex
+dependencies on previous type formation rules and constructors by a
+well-known method of describing intricate dependencies: the syntax of
+type theory itself. We had to limit the type formers to only allow
+strictly positive definitions, but these restrictions are the only
+things that a type theorist has to learn to understand our codes. Our
+encoding is also \emph{direct} in the sense that the types of
+constructors and eliminators are exactly as required and not merely up
+to isomorphisms.
+
+In this paper we only \emph{specified} HIITs and characterized their
+induction principles. Proving their \emph{existence} is left as
+further work. This would likely involve reducing HIITs to basic
+building blocks such as W-types and quotient types.
+
+The theory of codes was defined as part of the syntax of another type
+theory, the target theory. An alternative way would be to define
+the theory of codes internally to a type theoretic metatheory in the
+style of \cite{ttintt}. However, as described in \cite[Section
+  6]{ttintt}, there would be a coherence problem: for the internal
+syntax to be useful, we need to truncate it to be a set (as the
+$\mathsf{trunc}$ constructor did for $\Int$ in the introduction). As
+the eliminator needs to respect the equality given by
+$\mathsf{trunc}$, we would only be able to eliminate from the internal
+type theory into a set. This would preclude the definition of even the
+$\blank^\C$ operation, which corresponds to the standard model. A
+possible solution to this problem would be to add all the higher
+coherence laws to the syntax (e.g.\ the pentagon law for the
+composition of substitutions) and then prove that the syntax is a
+set. For this however, we would probably need a two-level metatheory
+as in \cite{semisegal}.
+
+When working in a metatheory with uniqueness of identity proofs (which
+implies that all HIITs are in essence QIITs), the theory of codes
+admits a category model where the interpretation of a context is the
+category of algebras corresponding to the context. In the future, we
+would like to prove that these categories have initial algebras given
+by terms in the theory of codes.
 
@@
 \appendix
-\section{}
+\section{Elimination principle computed for the two-dimensional sphere}
+\label{sec:app}
 
-appendix contents
+The two-dimensional sphere is given by the following context in the
+theory of codes.
+\[
+\Gamma :\equiv \Big(\boldsymbol{\cdot},\,\,\,S^2:\U,\,\,\,b:\underline{S^2},\,\,\,surf:\underline{\refl_{b} =_{(b=_{S^2} b)} \refl_{b}}\Big)
+\]
+The sphere-algebras are computed as follows.
+\begin{alignat*}{5}
+  & && \Gamma^\C \equiv \hat{\top}\timesh(\hat{S^2}\in\Type_{\hat{0}})\timesh(\hat{b}\in \hat{S^2})\timesh(\hat{surf\in \hat{\refl_{\hat{b}}}=_{(\hat{b}=_{\hat{S^2}}\hat{b})}\hat{\refl_{\hat{b}}}})
+\end{alignat*}
+Given a sphere-algebra and fixing a universe level $i$, the motives
+and methods are computed as follows.
+\begin{alignat*}{5}
+  & && \Gamma^\M\,\,(\hat{\tt,\,S^2,\,b,\,surf})\\
+  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{2M}}\in \hat{S^2}\rah\Type_i) \\
+  & && \hspace{0.75em} \timesh(\hat{b^M}\in \hat{S^{2M}\,b}) \\
+  & && \hspace{0.75em} \timesh(\hat{surf^M\in \TR_{(\TR_{S^{2M}}\,\blank\,b^M=b^M)}\,surf\,\refl_{b^M} = \refl_{b^M}})
+\end{alignat*}
+Given a sphere-algebra and the motives and methods for this algebra,
+the types of the elimination principles are computed as follows.
+\begin{alignat*}{5}
+  & && \Gamma^\E\,\,(\hat{\tt,\,S^2,\,b,\,surf})\,(\hat{\tt,\,S^{2M},\,b^M,\,surf^M}) \\
+  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{2E}}\in (\hat{x}\in\hat{S^2})\hat{\,\ra\,}\hat{S^{2M}\,x}) \\
+  & && \hspace{0.75em} \timesh(\hat{b^E}\in \hat{S^{2E}\,\,b = b^M})\\
+  & && \hspace{0.75em} \timesh\bigg(\hat{surf^E\,\,\in\,\,} \TR\,\Big(\hat{\J\,\refl\,b^E}\Big)\,\Big(\TR\,\big(\hat{\J\,\refl\,b^E}\big)\,\big(\hat{\apd}\,(\hat{\lambda} \hat{x}.\TR\,\hat{b^E}\,(\TR\,\hat{b^E}\,(\hat{\apd}\,\hat{S^{2E}}\,\hat{x})))\,\hat{surf}\big)\Big) \\
+  & && \hspace{5.5em} \hat{=} \,\hat{surf^M}\bigg)
+\end{alignat*}
+Note that if $\hat{b^E}$ is a definitional equality (that is, we have
+$\hat{S^{2E}\,\,b} \equiv \hat{b^M}$), the occurrences of $\hat{b^E}$
+in the type of $\hat{surf^E}$ can be replaced by $\hat{\refl}$. In
+this case the type of $\hat{surf^E}$ becomes the expected
+$\hat{\apd}\,(\hat{\apd}\,\hat{S^{2E}})\,\hat{surf}\,\hat{=}\,\hat{surf^M}$.
+
+%=====================================================================
 

### 2019-01-28 paper.tex
RECOLOR PARENS TO RED

@@
 \begin{alignat*}{5}
-  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\D\,\,(\hat{\tt,\,S^1,\,b,\,loop})\\
-  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{1D}}\in \hat{S^1}\rah\Type_i)\timesh(\hat{b^D}\in \hat{S^{1D}\,b})\timesh(\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D})
+  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\D\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1D}}\in \hat{S^1}\rah\Type_i\hat{)}\timesh\hat{(}\hat{b^D}\in \hat{S^{1D}\,b}\hat{)}\timesh\hat{(}\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D}\hat{)}
 \end{alignat*}
@@
 
-TODO: revise when we have homomorphism section, in order to remove redundant
-explanations. Also revise language.
+TODO: REWRITE to remove redundacy from homomorphism section.
 
@@
 \[
-\infer{\hat{\Gamma}\semicol\Delta^\S \in (\hat{\gamma}\in\Delta^\A)\rah \Delta^\D\,\hat{\gamma}\rah \Type_{i}}{\hat{\Gamma}\vdash\Delta}
-\]
-Types are interpreted as dependent binary relations which additionally
-depend on ($\hat{\gamma}\commah\,\hat{\gamma^D}\commah\,\hat{\gamma^S}$) interpretations of the
+\infer{\hat{\Gamma}\semicol\Delta^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah \Type_{i}}{\hat{\Gamma}\vdash\Delta}
+\]
+Types are interpreted as dependent binary relations which additionally depend on
+$\hat{\gamma}$, $\hat{\gamma^D}$, $\hat{\gamma^S}$ interpretations of the
 context.
 \[
-\infer{\hat{\Gamma}\vdashh A^\S \in (\hat{\gamma}\in\Delta^\A)(\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma})(\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D})(\hat{x}\in A^\A\,\hat{\gamma})\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{x}\rah \Type_{i}}{\hat{\Gamma}\semicol\Delta\vdash A}
+\infer{\hat{\Gamma}\vdashh A^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\hat{(}\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}\hat{(}\hat{x}\in A^\A\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{x}\rah \Type_{i}}{\hat{\Gamma}\semicol\Delta\vdash A}
 \]
@@
 \[
-\infer{\hat{\Gamma}\vdashh t^\S \in (\hat{\gamma}\in\Delta^\A)(\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma})(\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D})\rah  A^\S\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{\gamma^S}\,(t^\A\,\hat{\gamma})\,(t^\D\,\hat{\gamma}\,\hat{\gamma^D})}{\hat{\Gamma}\semicol\Delta \vdash t : A}
+\infer{\hat{\Gamma}\vdashh t^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\hat{(}\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}\rah  A^\S\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{\gamma^S}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}t^\D\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}}{\hat{\Gamma}\semicol\Delta \vdash t : A}
 \]
 In addition to $\hat{\gamma^2}$, we use $t\,\hat{\gamma^3}$ to abbreviate $t\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{\gamma^S}$. The implementation is the following.
-%% \begingroup \allowdisplaybreaks
+\begingroup
+%% \allowdisplaybreaks
 \begin{alignat*}{5}
   & \boldsymbol{\cdot}^\S\,\hat{\gamma}\,\hat{\gamma^D} && :\equiv \hat{\top} \\
-  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv (\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2})\timesh A^\S\,\hat{\gamma^3}\,\hat{t}\,\hat{t^D} \\
+  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv \hat{(}\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2}\hat{)}\timesh A^\S\,\hat{\gamma^3}\,\hat{t}\,\hat{t^D} \\
   & x^\S\,\hat{\gamma^3} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^S} \\ % TODO: De Bruijn
-  & \U^\S\,\hat{\gamma^3}\,\hat{A}\,\hat{A^D} && :\equiv (\hat{x}\in \hat{A})\rah  \hat{A^D}\,\hat{x} \\
+  & \U^\S\,\hat{\gamma^3}\,\hat{A}\,\hat{A^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  \hat{A^D}\,\hat{x} \\
   & (\underline{a})^\S\,\hat{\gamma^3}\,\hat{t}\,\hat{t^D} && :\equiv a^\S\,\hat{\gamma^3}\,\hat{t} \eqh \hat{t^D} \\
-  & ((x:a)\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv (\hat{x}\in a^\A\,\hat{\gamma})\rah  B^\S\,(\hat{\gamma,x})\,(\hat{\gamma^D,}\,a^\S\,\hat{\gamma^3}\,\hat{x})\,(\hat{\gamma^S,}\,\hat{\refl}) \\
-  & && \hspace{8.4em} (\hat{f}\,\hat{x})\,\big(\hat{f^D}\,\hat{x}\,(a^\S\,\hat{\gamma^3}\,\hat{x})\big) \\
-  & (t\,u)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,(t^\S\,\hat{\gamma^3}\,( u^\A\,\hat{\gamma}))\,(u^\S\,\hat{\gamma^3}) \\
-  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv (\hat{x}\in \hat{A})\rah B^\S\,\hat{\gamma^3}\,(\hat{f}\,\hat{x})\,(\hat{f^D}\,\hat{x}) \\
+  & ((x:a)\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\S\,\hat{(}\hat{\gamma,x}\hat{)}\,\hat{(}\hat{\gamma^D,}\,a^\S\,\hat{\gamma^3}\,\hat{x}\hat{)}\,\hat{(}\hat{\gamma^S,}\,\hat{\refl}\hat{)} \\
+  & && \hspace{8.4em} \hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{\big(}\hat{f^D}\,\hat{x}\,\hat{(}a^\S\,\hat{\gamma^3}\,\hat{x}\hat{)}\hat{\big)} \\
+  & (t\,u)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^3}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^3}\hat{)} \\
+  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^3}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
   & (t\,\hat{u})^\S\,\hat{\gamma^3} && :\equiv t^\S\,\hat{\gamma^3}\,\hat{u} \\
-  & (t=_a u)^\S\,\hat{\gamma^3}\,\hat{e} && :\equiv \TR\,(t^\S\,\hat{\gamma^3})\big(\TR\,(u^\S\,\hat{\gamma^3})\,(\apd\,(a^\S\,\hat{\gamma^3})\,\hat{e})\big) \\
-  & (\refl_t)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,\hat{\refl}\,(t^\S\,\hat{\gamma^3}) \\
+  & (t=_a u)^\S\,\hat{\gamma^3}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^3}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^3}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
+  & (\refl_t)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,\hat{\refl}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)} \\
   & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S\,\hat{\gamma^3} && :\equiv \\
-  & && \hspace{-10em}\hat{\J}\,\bigg(\hat{\J}\,\Big(\hat{\J}\,\big(\hat{\J}\,(\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}.\,\hat{pr^S})\,(t^\S\,\hat{\gamma^3})\,\\
-  & && \hspace{-9em}(\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2})\,(\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3})\,(pr^\D\,\hat{\gamma^2})\,(pr^\S\,\hat{\gamma^3})\big)\,eq^\A\,\hat{\gamma}\Big)\,u^\S\,\hat{\gamma^3}\bigg)\,(eq^\S\,\hat{\gamma^3}) \\
+  & && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}.\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}\,\\
+  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^3}\hat{)}\hat{\big)}\,eq^\A\,\hat{\gamma}\hat{\Big)}\,u^\S\,\hat{\gamma^3}\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^3}\hat{)} \\
   & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^3} && :\equiv \\
-  & && \hspace{-10em}\hat{\J}\,\big(\hat{\J}\,(\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl})\,(t^\S\,\hat{\gamma^3})\,(\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2})\,(\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3})\big)\,(pr^\S\,\hat{\gamma^3})\\
-  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^3\,f\,t} && :\equiv b^\S\,\hat{\gamma^3}\,(\hat{f}\,\hat{t}) \\
-  & (t\,\hat{u})^\S\,\hat{\gamma^3} && :\equiv \ap\,(\hat{\lambda} \hat{f.}\hat{f}\,\hat{u})\,(t^\S\,\hat{\gamma^3})
+  & && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^3}\hat{)}\\
+  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^3\,f\,t} && :\equiv b^\S\,\hat{\gamma^3}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
+  & (t\,\hat{u})^\S\,\hat{\gamma^3} && :\equiv \ap\,\hat{(}\hat{\lambda} \hat{f.}\hat{f}\,\hat{u}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}
 \end{alignat*}
@@
 formation rule. In the natural numbers example above, the non-indexed
-$Nat : \U$ sort is interpreted as $\hat{n^S} \in (\hat{x}\in
-\hat{n})\rah \hat{n^D}\,\hat{x}$. For indexed sorts, the indices are
+$Nat : \U$ sort is interpreted as $\hat{\hat{n^S} \in (\hat{x}\in
+\hat{n})\rah \hat{n^D}\,\hat{x}}$. For indexed sorts, the indices are
 first processed by the $\blank^\S$ cases for inductive and non-inductive
@@
 \begin{alignat*}{5}
-  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\S\,\,(\hat{\tt,\,S^1,\,b,\,loop})\,(\hat{\tt,\,S^{1M},\,b^D,\,loop^D}) \\
-  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{1E}}\in (\hat{x}\in\hat{S^1})\hat{\,\ra\,}\hat{S^{1M}\,x})\timesh(\hat{b^S}\in \hat{S^{1E}\,\,b = b^D})\\
-  & && \hspace{0.75em}\timesh(\hat{loop^S\,\,\in\,\,} \TR\,_{(\hat{\lambda x.
-\TR\,_{S^{1M}}\,loop\,x\,=\,b^D})}\,\hat{b^S}\,(\TR\,_{(\hat{\lambda} \hat{x}.
-\TR\,_{\hat{S^{1M}}}\,\hat{loop}\,(\hat{S^{1E}\,b})\,\eqh\,\hat{x})}\,\hat{b^S}\,(\apd\,\hat{S^{1E}\,loop}))\,\\
-  & && \hspace{5.5em} \hat{=\,loop^D})
+  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\S\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\,\hat{(}\hat{\tt,\,S^{1M},\,b^D,\,loop^D}\hat{)} \\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1E}}\in \hat{(}\hat{x}\in\hat{S^1}\hat{)}\hat{\,\ra\,}\hat{S^{1M}\,x}\hat{)}\timesh\hat{(}\hat{b^S}\in \hat{S^{1E}\,\,b = b^D}\hat{)}\\
+  & && \hspace{0.75em}\timesh\hat{(}\hat{loop^S\,\,\in\,\,} \TR\,_{\hat{(}\hat{\lambda x.
+\TR\,_{S^{1M}}\,loop\,x\,=\,b^D}\hat{)}}\,\hat{b^S}\,\hat{(}\TR\,_{\hat{(}\hat{\lambda} \hat{x}.
+\TR\,_{\hat{S^{1M}}}\,\hat{loop}\,\hat{(}\hat{S^{1E}\,b}\hat{)}\,\eqh\,\hat{x}\hat{)}}\,\hat{b^S}\,\hat{(}\apd\,\hat{S^{1E}\,loop}\hat{)}\hat{)}\,\\
+  & && \hspace{5.5em} \hat{=\,loop^D}\hat{)}
 \end{alignat*}
@@
 which can describe a large class of inductive definitions
-\cite{morris09indexed}. Suppose we are in the target context $\hat{I}
+\cite{morris09indexed}. Suppose we are in the external context $\hat{I}
 \in \Type_{\hat{0}}\commah\hat{S} \in \Type_{\hat{0}}\commah\hat{P} \in \hat{S} \rah
 \Type_{\hat{0}}\commah\hat{out} \in \hat{S} \rah \hat{I}\commah\hat{in} \in
-(\hat{s} \in \hat{S})\rah \hat{P}\,\hat{s}\rah \hat{I}$. Then, the
+\hat{(}\hat{s} \in \hat{S}\hat{)}\rah \hat{P}\,\hat{s}\rah \hat{I}$. Then, the
 code for the corresponding indexed W-type is the following:
@@
 \begin{alignat*}{5}
-  & \rlap{$W^\A \equiv (\hat{w}\in \hat{I}\rah\Type_{\hat{0}})\timesh((\hat{s}\in \hat{S})\rah((\hat{p}\in \hat{P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\rah \hat{w}\,(\hat{out\,s}))$}\\
-  & W^\D\,(\hat{w, sup}) && \equiv\, && (\hat{w^D} \in (\hat{i}\in \hat{I})\rah \hat{w\,i}\rah \Type_j)\\
-  & && \hspace{0.23em}\timesh && \big((\hat{s\in S})(\hat{f}\in (\hat{p\in P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\\
-  & && && \rah ((\hat{p\in P\,s})\rah \hat{w^D}\,(\hat{in\,s\,p})\,(\hat{f\,p}))\rah \hat{w^D}\,(\hat{out\,s})\,(\hat{sup\,s\,f})\big) \\
-  & \rlap{$ W^\S\,(\hat{w, sup})\,(\hat{w^D,sup^D}) \equiv $} \\
-  & && && (\hat{w^S} \in (\hat{i \in I})(\hat{x \in w\,i})\rah \hat{w^D\,i\,x})\\
-  & && \hspace{0.23em}\timesh && \big((\hat{s\in S})(\hat{f}\in (\hat{p\in P\,s})\rah \hat{w}\,(\hat{in\,s\,p}))\\
-  & && && \rah \hat{w^S}\,(\hat{out\,s})\,(\hat{sup\,\,s\,f}) \eqh \hat{sup^D\,s\,f}\,(\hat{\lambda} \hat{p.}\, \hat{w^S}\,(\hat{in\,s\,p})\,(\hat{f\,p}))\big)
+  & \rlap{$W^\A \equiv \hat{(}\hat{w}\in \hat{I}\rah\Type_{\hat{0}}\hat{)}\timesh\hat{(}\hat{(}\hat{s}\in \hat{S}\hat{)}\rah\hat{(}\hat{(}\hat{p}\in \hat{P\,s}\hat{)}\rah \hat{w}\,\hat{(}\hat{in\,s\,p}\hat{)}\hat{)}\rah \hat{w}\,\hat{(}\hat{out\,s}\hat{)}\hat{)}$}\\
+  & W^\D\,\hat{(}\hat{w, sup}\hat{)} && \equiv\, && \hat{(}\hat{w^D} \in \hat{(}\hat{i}\in \hat{I}\hat{)}\rah \hat{w\,i}\rah \Type_{\hat{j}}\hat{)}\\
+  & && \hspace{0.23em}\timesh && \hat{\big(}\hat{(}\hat{s\in S}\hat{)}\hat{(}\hat{f}\in \hat{(}\hat{p\in P\,s}\hat{)}\rah \hat{w}\,\hat{(}\hat{in\,s\,p}\hat{)}\hat{)}\\
+  & && && \rah \hat{(}\hat{(}\hat{p\in P\,s}\hat{)}\rah \hat{w^D}\,\hat{(}\hat{in\,s\,p}\hat{)}\,\hat{(}\hat{f\,p}\hat{)}\hat{)}\rah \hat{w^D}\,\hat{(}\hat{out\,s}\hat{)}\,\hat{(}\hat{sup\,s\,f}\hat{)}\hat{\big)} \\
+  & \rlap{$ W^\S\,\hat{(}\hat{w, sup}\hat{)}\,\hat{(}\hat{w^D,sup^D}\hat{)} \equiv $} \\
+  & && && \hat{(}\hat{w^S} \in \hat{(}\hat{i \in I}\hat{)}\hat{(}\hat{x \in w\,i}\hat{)}\rah \hat{w^D\,i\,x}\hat{)}\\
+  & && \hspace{0.23em}\timesh && \hat{\big(}\hat{(}\hat{s\in S}\hat{)}\hat{(}\hat{f}\in \hat{(}\hat{p\in P\,s}\hat{)}\rah \hat{w}\,\hat{(}\hat{in\,s\,p}\hat{)}\hat{)}\\
+  & && && \rah \hat{w^S}\,\hat{(}\hat{out\,s}\hat{)}\,\hat{(}\hat{sup\,\,s\,f}\hat{)} \eqh \hat{sup^D\,s\,f}\,\hat{(}\hat{\lambda} \hat{p.}\, \hat{w^S}\,\hat{(}\hat{in\,s\,p}\hat{)}\,\hat{(}\hat{f\,p}\hat{)}\hat{)}\hat{\big)}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & && \Gamma^\C \equiv \hat{\top}\timesh(\hat{S^2}\in\Type_{\hat{0}})\timesh(\hat{b}\in \hat{S^2})\timesh(\hat{surf\in \hat{\refl_{\hat{b}}}=_{(\hat{b}=_{\hat{S^2}}\hat{b})}\hat{\refl_{\hat{b}}}})
+  & && \Gamma^\C \equiv \hat{\top}\timesh\hat{(}\hat{S^2}\in\Type_{\hat{0}}\hat{)}\timesh\hat{(}\hat{b}\in \hat{S^2}\hat{)}\timesh\hat{(}\hat{surf\in \hat{\refl_{\hat{b}}}=_{\hat{(}\hat{b}=_{\hat{S^2}}\hat{b}\hat{)}}\hat{\refl_{\hat{b}}}}\hat{)}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & && \Gamma^\M\,\,(\hat{\tt,\,S^2,\,b,\,surf})\\
-  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{2M}}\in \hat{S^2}\rah\Type_i) \\
-  & && \hspace{0.75em} \timesh(\hat{b^M}\in \hat{S^{2M}\,b}) \\
-  & && \hspace{0.75em} \timesh(\hat{surf^M\in \TR_{(\TR_{S^{2M}}\,\blank\,b^M=b^M)}\,surf\,\refl_{b^M} = \refl_{b^M}})
-\end{alignat*}
-Given a sphere-algebra and the motives and methods for this algebra,
-the types of the elimination principles are computed as follows.
-\begin{alignat*}{5}
-  & && \Gamma^\E\,\,(\hat{\tt,\,S^2,\,b,\,surf})\,(\hat{\tt,\,S^{2M},\,b^M,\,surf^M}) \\
-  & \equiv \,\, && \hat{\top}\timesh(\hat{S^{2E}}\in (\hat{x}\in\hat{S^2})\hat{\,\ra\,}\hat{S^{2M}\,x}) \\
-  & && \hspace{0.75em} \timesh(\hat{b^E}\in \hat{S^{2E}\,\,b = b^M})\\
-  & && \hspace{0.75em} \timesh\bigg(\hat{surf^E\,\,\in\,\,} \TR\,\Big(\hat{\J\,\refl\,b^E}\Big)\,\Big(\TR\,\big(\hat{\J\,\refl\,b^E}\big)\,\big(\hat{\apd}\,(\hat{\lambda} \hat{x}.\TR\,\hat{b^E}\,(\TR\,\hat{b^E}\,(\hat{\apd}\,\hat{S^{2E}}\,\hat{x})))\,\hat{surf}\big)\Big) \\
-  & && \hspace{5.5em} \hat{=} \,\hat{surf^M}\bigg)
+  & && \Gamma^\M\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2M}}\in \hat{S^2}\rah\Type_i\hat{)} \\
+  & && \hspace{0.75em} \timesh\hat{(}\hat{b^M}\in \hat{S^{2M}\,b}\hat{)} \\
+  & && \hspace{0.75em} \timesh\hat{(}\hat{surf^M\in \TR_{\hat{(}\TR_{S^{2M}}\,\blank\,b^M=b^M\hat{)}}\,surf\,\refl_{b^M} = \refl_{b^M}}\hat{)}
+\end{alignat*}
+Given a sphere-algebra and a displayed algebra over it, we get the type of sections:
+\begin{alignat*}{5}
+  & && \Gamma^\E\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\,\hat{(}\hat{\tt,\,S^{2M},\,b^M,\,surf^M}\hat{)} \\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2E}}\in \hat{(}\hat{x}\in\hat{S^2}\hat{)}\hat{\,\ra\,}\hat{S^{2M}\,x}\hat{)} \\
+  & && \hspace{0.75em} \timesh\hat{(}\hat{b^E}\in \hat{S^{2E}\,\,b = b^M}\hat{)}\\
+  & && \hspace{0.75em} \timesh\hat{\bigg(}\hat{surf^E\,\,\in\,\,} \TR\,\hat{\Big(}\hat{\J\,\refl\,b^E}\hat{\Big)}\,\hat{\Big(}\TR\,\hat{\big(}\hat{\J\,\refl\,b^E}\hat{\big)}\,\hat{\big(}\hat{\apd}\,\hat{(}\hat{\lambda} \hat{x}.\TR\,\hat{b^E}\,\hat{(}\TR\,\hat{b^E}\,\hat{(}\hat{\apd}\,\hat{S^{2E}}\,\hat{x}\hat{)}\hat{)}\hat{)}\,\hat{surf}\hat{\big)}\hat{\Big)} \\
+  & && \hspace{5.5em} \hat{=} \,\hat{surf^M}\hat{\bigg)}
 \end{alignat*}
@@
 this case the type of $\hat{surf^E}$ becomes the expected
-$\hat{\apd}\,(\hat{\apd}\,\hat{S^{2E}})\,\hat{surf}\,\hat{=}\,\hat{surf^M}$.
+$\hat{\apd}\,\hat{(}\hat{\apd}\,\hat{S^{2E}}\hat{)}\,\hat{surf}\,\hat{=}\,\hat{surf^M}$.
 

### 2019-01-29 paper.tex
Add section about coherence and syntactic translation

@@
 \newcommand{\Tm}{\mathsf{Tm}}
-\newcommand{\Tms}{\mathsf{Tms}}
+\newcommand{\Sub}{\mathsf{Sub}}
 \newcommand{\R}{\mathsf{R}}
@@
 \newcommand{\Section}{\mathsf{Section}}
+\newcommand{\ass}{\mathsf{ass}}
+\newcommand{\idl}{\mathsf{idl}}
+\newcommand{\idr}{\mathsf{idr}}
 
@@
 $ps$ are called the \emph{induction methods}. The behavior of induction is
-described by a \emph{computation-rule} ($\beta$-rule) for each constructor
+described by a \emph{computation rule} ($\beta$-rule) for each constructor
 and induction method:
@@
 recursion. In Section \ref{sec:coherence} we explain the choice of using
-syntactic translations in the rest of the paper. In Sections TODO to TODO, we
-describe four syntactic translations from the theory of signatures to the syntax
-of the external type theory, respectively computing algebras, homomorphisms,
-displayed algebras and sections of displayed algebras. In Section TODO, we
-consider extending the previous translations with additional components of the
-categorical semantics of HIITs (e.g.\ identity and composition for
-homomorphisms), and explain why the approach in this paper does not make this
-feasible. Section \ref{sec:formalization} describes the Agda formalization and
-the Haskell implementation. We conclude in Section \ref{sec:summary}.
+syntactic translations in the rest of the paper. In Sections \ref{sec:algebras}
+to \ref{sec:sections}, we describe four syntactic translations from the theory
+of signatures to the syntax of the external type theory, respectively computing
+algebras, homomorphisms, displayed algebras and sections of displayed
+algebras. In Section \ref{sec:categorical}, we consider extending the previous
+translations with additional components of the categorical semantics of HIITs
+(e.g.\ identity and composition for homomorphisms), and explain why the approach
+in this paper does not make this feasible. Section \ref{sec:formalization}
+describes the Agda formalization and the Haskell implementation. We conclude in
+Section \ref{sec:summary}.
 
@@
 \begin{gathered}
-  \infer{\hat{\Gamma}\vdash\cdot}{\vdashh\hat{\Gamma}}
+  \infer{\hat{\Gamma}\vdash\cdot}{\vdashh\,\hat{\Gamma}}
 \end{gathered}
@@
 refer to any particular type former from the external theory. We describe the
-external theory in more detail in Section TODO. Also, the rules presented here
-are informal and optimized for readability; we describe the Agda formalizations
-in Section \ref{sec:formalization}. We explain the rules for the theory of
-signatures in order now.
-
-(1) The rules for context formation and variables are standard. We assume fresh
-names everywhere to avoid name capture, and leave weakenings implicit.
+external theory in more detail in Section \ref{sec:external}. Also, the rules
+presented here are informal and optimized for readability; we describe the Agda
+formalizations in Section \ref{sec:formalization}. We explain the rules for the
+theory of signatures in order now.
+
+(1) The rules for context formation and variables are standard. We build
+signatures in a well-formed external context.  We assume fresh names everywhere
+to avoid name capture, and leave weakenings implicit.
 
@@
 
-We describe the external type theory here. It serves two purposes: it is a
-source of types external to a HIIT signature, and it also serves as the target
-for the syntactic translations described in Sections TODO to TODO. It is not
-essential that we use the same theory for both purposes; we do so only because
-it simplifies the presentation. Also, we need not specify the external theory in
-formal detail, as it is a standard type theory. We only need to make some
-assumptions about supported type formers. We generally keep the notation close
-to Agda, and use {\color{BrickRed}brick red} color to distinguish from
-constructions in the theory of signatures.
-
-There is a cumulative Russell-style hierarchy of universes $\Type_{\hat{i}}$, with
-universes closed under $\Pi$, $\Sigma$, equality and unit types.
+The external syntax serves two purposes: it is a source of types external to a
+HIIT signature, and it also serves as the target for the syntactic translations
+described in Sections \ref{sec:algebras} to \ref{sec:sections}. It is not
+essential that we use the same theory for both purposes; we do so only to
+simplify the presentation by skipping an additional translation or embedding
+step. Also, we do not specify the external theory in formal detail, since it is
+a standard type theory. We only make some assumptions about supported type
+formers. We generally keep the notation close to Agda, and use
+{\color{BrickRed}brick red} color to distinguish from constructions in the
+theory of signatures.
+
+There is a cumulative Russell-style hierarchy of universes $\Type_{\hat{i}}$,
+with universes closed under $\Pi$, $\Sigma$, equality and unit
+types. Importantly, we do not assume uniqueness of identity proofs.
 
@@
 endofunction. Below, we leave universe indices explicit, and we use the notation
-of the external type theory described in Section \ref{sec:signatures}.
+of the external type theory described in Section \ref{sec:external}.
 \begin{alignat*}{5}
@@
 
-\noindent Displayed $\Nat$-algebras (sometimes called fibered algebras, as in \cite{sojakova})
-are likewise triples, but depending on a $\Nat$-algebra.
+\noindent Displayed $\Nat$-algebras (sometimes called fibered algebras, as in
+\cite{sojakova}) are likewise triples, but each component depends on the
+corresponding components of a $\Nat$-algebra. We borrow the term ``displayed''
+from Ahrens and Lumsdaine \cite{displayedCategories}, as our displayed algebras
+generalize their displayed categories.
 \begin{alignat*}{5}
@@
 & \Section\,(N,\,z,\,s)\,(N^D,\,z^D,\,s^D) \equiv && \\
-&  \hspace{3em} (N^S : (n : N) \ra N^D\,n)\times (z^S : N^S\,z = z^D)\times ((n : N)\ra N^S\,(s\,n) = s^D\,(N^S\,n))
+&  \hspace{3em} (N^S : (n : N) \ra N^D\,n)\times (z^S : N^S\,z = z^D)\times ((n : N)\ra N^S\,(s\,n) = s^D\,n\,(N^S\,n))
 \end{alignat*}
@@
 We will compute algebras and the other notions by induction on the syntax of the
-theory of signatures. However, first we need to clarify the formal justifications
-behind these computations.
-
-\section{Coherence Problems and Syntactic Translations}
+theory of signatures. However, first we need to clarify the formal foundations
+of these computations.
+
+\section{The Coherence Problem and Syntactic Translations}
 \label{sec:coherence}
 
-TODO
+The next task would be to define a computation which takes as input a
+$\hat{\Gamma}\vdash\Delta$ signature, and returns the corresponding type of
+algebras in some type theory. This would behave as a ``standard'' model of
+signatures, which simply maps each construction in the syntax to its
+metatheoretical counterpart: types to types, universe to universe,
+functions to functions, and so on.
+
+However, it is important to interpret signatures into a theory without
+uniqueness of identity proofs (UIP), because we are considering \emph{higher}
+inductive types, and hence must remain compatible with higher-dimensional
+interpretations. In particular, we need to interpret type constructors in
+signatures into type universes which are not truncated to any homotopy level. In
+this setting, even the mundane $\alpha : (N : \mathsf{Type}) \times N \times (N
+\ra N)$ natural number algebras may have arbitrary higher-dimensional structure.
+
+On first look, we might think that the simplest way to formalize the algebra
+interpretation is the following:
+\begin{enumerate}
+\item Assume as metatheory a type theory without UIP.
+\item In this setting, define a formal syntax of the theory of signatures.
+\item Give a standard interpretation of signatures into the UIP-free metatheory.
+\end{enumerate}
+
+It may come as a surprise that realizing the above steps is an \emph{open
+  problem}, for any syntax of a dependent type theory. We call the problem of
+interpreting syntaxes of type theories into a UIP-free metatheory a
+\emph{coherence problem}. This issue appears to arise with all known ways of
+defining syntaxes for dependent type theories.
+
+In the following, we first consider the coherence problem in two settings: with
+intrinsically typed higher inductive-inductive syntaxes, then with conventional
+syntaxes involving preterms and inductively defined typing and conversion
+relations. Then, we present syntactic translations as a partial solution to the
+coherence problem.
+
+\subsection{Interpreting intrinsic syntax}
+Following Altenkirch and Kaposi \cite{ttintt}, one might define the syntax of a
+type theory as an initial category with families (CwF) \cite{dybjer1995internal}
+extended with additional type formers. The CwF part provides a calculus and
+equational theory for explicit substitutions, upon which one can build
+additional type structure. We present an excerpt below:
+
+\begin{alignat*}{5}
+  & \Con && : \Set && \text{contexts} \\
+  & \Ty  && : \Con\ra\Set && \text{types} \\
+  & \Sub  && : \Con\ra\Con\ra\Set && \text{substitutions} \\
+  & \Tm  && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \hspace{2em} && \text{terms} \\
+  & \cdot && : \Con && \text{empty context} \\
+  & \blank\rhd\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con && \text{context extension} \\
+  & \blank[\blank] && : \Ty\,\Delta\ra\Sub\,\Gamma\,\Delta\ra\Ty\,\Gamma && \text{type substitution} \\
+  & \id && : \Sub\,\Gamma\,\Gamma && \text{identity substitution} \\
+  & [\id] && : A[\id] = A && \text{action of}\,\,\id\,\,\text{on types}\\
+  & \blank\circ\blank && : \Sub\,\Theta\,\Delta\ra\Sub\,\Gamma\,\Theta\ra\Sub\,\Gamma\,\Delta && \text{substitution composition} \\
+  & \blank[\blank] && : \Tm\,\Delta\,A\ra(\sigma:\Sub\,\Gamma\,\Delta)\ra\Tm\,\Gamma\,(A[\sigma])\hspace{1em} && \text{term substitution} \\
+  & ... && \\
+  & \U && : \Ty\,\Gamma && \text{universe} \\
+  & {\U[]} && : \U[\sigma] = \U && \text{substituting the universe} \\
+  & \El && : \Tm\,\Gamma\,\U \ra \Ty\,\Gamma && \text{decoding} \\
+  & \Pi && : (A:\Ty\,\Gamma)\ra\Ty\,(\Gamma\rhd A)\ra\Ty\,\Gamma && \text{functions} \\
+  & ... &&
+\end{alignat*}
+
+This notion of syntax is much more compact and often much more convenient to use
+than extrinsic syntaxes. It is in essence ``merely'' a formalization of CwFs,
+which are often used in categorical semantics of type theory. However, its
+rigorous metatheory is subject to ongoing research (including the current
+paper). In a set-truncated setting, the current authors and Altenkirch have
+previously developed semantics and constructed initial algebras
+\cite{kaposi2019constructing}, but here we need to work in a non-truncated
+theory.
+
+\subsubsection{Set-truncation}
+Additionally, we need to set-truncate the syntax by adding the following constructors:
+\begin{alignat*}{5}
+  & \mathsf{setTy}  && : (A\,B : \Ty\,\Gamma)(p\,q : A = B)\ra p = q \\
+  & \mathsf{setTm}  && : (t\,u : \Tm\,\Gamma\,A)(p\,q : t = u)\ra p = q \\
+  & \mathsf{setSub} && : (\sigma\,\delta : \Sub\,\Gamma\,\Delta)(p\,q : \sigma = \delta)\ra p = q
+\end{alignat*}
+
+\noindent We can omit the rule for contexts, as it is derivable from the above ones. Set
+truncation forces definitional equality of the syntax (as defined by equality
+constructors in the HIIT signature) to be proof irrelevant. If we omit
+set-truncation, then the defined HIIT becomes very different from what we expect
+the syntax to be.
+
+We can easily show that the non-truncated syntax does not form sets. For
+example, $\U[]$ and $[\id]$ are two proofs of $\U[\id] = \U$, and they are not
+forced to be equal. Assuming univalence, we can give a model where types are
+interpreted as closed metatheoretic types, $\U$ is interpreted as metatheoretic
+$\Bool$, $[\id]$ is interpreted as $\refl : A = A$ for some metatheoretic $A$
+type, and $\U[]$ is interpreted as the $\Bool$ negation equivalence, thereby
+formally distinguishing $\U[]$ and $[\id]$. Hence, by Hedberg's theorem
+\cite{hedberg}, the syntax does not have decidable equality. This also implies
+that the non-truncated intrinsic syntax is not constructible from set quotients
+of extrinsic terms (since those always form sets). The non-truncated syntax just
+does not seem to be a sensible notion. This situation is similar to how
+categories in homotopy type theory need to have set-truncated objects
+\cite{ahrens2015univalent}.
+
+Unfortunately, set-truncation makes it impossible to interpret syntactic types
+as elements of a UIP-free metatheoretic $\mathsf{Type}_i$ universe. This is
+because we must provide interpretations for all set-truncation constructors,
+which amounts to shoving that the interpretations of $\Ty$, $\Con$, $\Sub$ are
+all sets. However, we cannot show $\mathsf{Type}_i$ to be a set without UIP.
+
+A possible solution would be to add \emph{all higher coherences} instead of
+set-truncating, which would yield something like an ($\omega$, 1)-CwF, but this
+is also an open research problem \cite{altenkirch2018towards, finster2019structure}.
+
+\subsection{Interpreting extrinsic syntax}
+An extrinsic syntax for type theory is defined the following way:
+\begin{enumerate}
+\item We inductively define a \emph{presyntax}: sets of preterms, pretypes,
+      precontexts, and possibly presubstitutions. These are not assumed to
+      be well-formed, and only serve as raw material for expressions.
+\item We give mutual inductive definitions for the following relations on presyntax:
+      well-formedness, typing and conversion.
+\end{enumerate}
+
+This is the conventional way of presenting the syntax; see
+e.g.\ \cite{winterhalter2019eliminating} for a detailed machine-checked
+formalization in this style. The main advantage compared to the intrinsic syntax
+is that this only requires conservative inductive definitions in the metatheory,
+which are also natively supported in current proof assistants, unlike HIITs. The
+main disadvantage is verbosity, lower level of abstraction, and a difficulty of
+pinning down a notion of model for the syntax.
+
+What about interpreting extrinsic syntax into a UIP-free universe? It is widely
+accepted that extrinsic syntaxes have standard interpretations in set-truncated
+metatheories, although carrying this out in formal detail is technically
+challenging. Streicher's seminal work \cite{streicher2012semantics} laid out a
+template for doing this: first we construct a family of partial functions from
+the presyntax to the semantic domain, then we prove afterwards that these
+functions are total on well-formed input.
+
+However, the coherence problem arises still: it is required that definitional
+equality is proof irrelevant. In a type-theoretic setting, this means that we
+need to propositionally truncate the conversion relation. This again prevents us
+from interpreting the syntax into a UIP-free universe. We have to interpret
+definitional equality in the syntax as propositional equality in the metatheory,
+but since the former is propositionally truncated, we can only eliminate it into
+propositions, and metatheoretic equality types are not propositions in general.
+
+Could we define a conversion relation which is propositional, but not truncated?
+For example, conversion could be defined in terms of a deterministic conversion
+checking algorithm. But then a complication is that we do not have a proof that
+conversion checking is \emph{total} and \emph{stable under substitution}, while
+still in the process of defining the syntax.
+
+Alternatively, we could first define a normalization algorithm for an extrinsic
+syntax in a UIP-free metatheory, and then try to interpret \emph{only normal
+  forms}. Abel, \"Ohman and Vezzosi demonstrated a UIP-free conversion checking
+algorithm in type theory \cite{abel2017decidability}, which suggests that
+UIP-free normalization may be possible as well. But since this normal form
+interpretation is a major technical challenge, and it has not been carried out
+yet, we cannot use it to justify constructions in the current paper.
+
+\subsection{Syntactic translations}
+We can circumvent the coherence problem in the following way:
+\begin{enumerate}
+\item Define a \emph{source} and a \emph{target} syntax in any suitable metatheory,
+      where the target theory does not have UIP. The source and target theories
+      do no necessarily need to differ.
+\item Interpret the source syntax into the target syntax.
+\end{enumerate}
+
+For extrinsic syntaxes, it is generally understood that a syntactic translation
+has to preserve definitional equalities in the source syntax. For intrinsic
+syntaxes, preservation of definitional equality is automatically enforced by the
+equality constructors. See Boulier et al.\ \cite{next700} for a showcase of
+syntactic translations.
+
+Now, we can take the source syntax to be the theory of signatures from Section
+\ref{sec:tos}, and the target syntax to be the external syntax from Section
+\ref{sec:external}. Truncation in the source syntax is not an issue here,
+because the target syntax is likewise truncated, so we can translate into it.
+
+However, using syntactic translations is also a significant restriction. As
+always, we must map equal inputs to equal outputs, but now the notion of
+equality for outputs coincides with definitional equality in the syntax of the
+target theory, which is far more restrictive than propositional equality. Recall
+the weak $\beta$-rule for $\J$ in the theory of signatures in Section
+\ref{sec:tos}: if we instead used a strict equality, then the translations in
+Sections \ref{sec:morphisms} and \ref{sec:sections} would not work, because they
+map $\J_{a\,t\,\,(x.z.p)}\,pr\,_t\,\refl$ and $pr$ to terms which are equal
+propositionally, but not definitionally. This restriction also prevents us from
+defining more translations which cover other parts of the categorical
+semantics, e.g.\ composition of homomorphisms. We return to this topic in
+Section \ref{sec:categorical}.
 

### 2019-01-30 paper.tex
Logrel + morphisms progress

@@
 \renewcommand{\hat}[1]{{\color{BrickRed}{#1}}}
+\newcommand{\blc}[1]{{\color{Black}{#1}}}
 \newcommand{\vdashh}{\mathbin{\hat\vdash}}
@@
 constructors. Additionally, $\Ty$ is indexed over the previously declared $\Con$
-type constructor. This suggests that we should not try to stratify signatures,
-in contrast to previous works, and use a fully dependent type theory.
+type constructor. This suggests that we should not attempt to stratify
+signatures, and instead use a fully dependent type theory. At this level of
+generality, stratification seems to complicate matters and remove the syntax
+further from familiar type theories.
 
@@
 signatures, which simply maps each construction in the syntax to its
-metatheoretical counterpart: types to types, universe to universe,
-functions to functions, and so on.
+counterpart: types to types, universe to universe, functions to functions, and
+so on.
 
@@
 
-Unfortunately, set-truncation makes it impossible to interpret syntactic types
-as elements of a UIP-free metatheoretic $\mathsf{Type}_i$ universe. This is
-because we must provide interpretations for all set-truncation constructors,
-which amounts to shoving that the interpretations of $\Ty$, $\Con$, $\Sub$ are
-all sets. However, we cannot show $\mathsf{Type}_i$ to be a set without UIP.
+Unfortunately, set-truncation makes it impossible to directly interpret
+syntactic types as elements of a UIP-free metatheoretic $\mathsf{Type}_i$
+universe. This is because we must provide interpretations for all set-truncation
+constructors, which amounts to shoving that the interpretations of $\Ty$,
+$\Con$, $\Sub$ are all sets. However, we cannot show $\mathsf{Type}_i$ to be a
+set without UIP.
 
@@
 but since the former is propositionally truncated, we can only eliminate it into
-propositions, and metatheoretic equality types are not propositions in general.
+propositions, and metatheoretic equality types are not generally propositions in
+the absence of UIP.
 
@@
       where the target theory does not have UIP. The source and target theories
-      do no necessarily need to differ.
+      do not necessarily need to differ.
 \item Interpret the source syntax into the target syntax.
@@
 
-%% --------------------------------------------------------------------------------
-
-%% \section{A syntactic translation from the source to the target theory}
-%% \label{sec:translation}
-
-%% An inductive type is specified by a context in the theory of codes
-%% defined in the previous section. In this section we define the
-%% $\blank^\C$, $\blank^\M$ and $\blank^\E$ operations, which work as follows
-%% on the code for natural numbers.
-%% \begin{alignat*}{5}
-%%   & && (\boldsymbol{\cdot},\,Nat : \U,\,zero:\underline{Nat},\,suc:Nat\ra\underline{Nat})^\C \\
-%%   & \equiv \,\, && \hat{\top}\timesh(\hat{n}\in\Type_{\hat{0}})\timesh(\hat{z}\in \hat{n})\timesh(\hat{n}\rah \hat{n}) \\
-%%   & && (\boldsymbol{\cdot},\,Nat : \U,\,zero:\underline{Nat},\,suc:Nat\ra\underline{Nat})^\M\,(\tt\commah\hat{n}\commah \hat{z}\commah \hat{s}) \\
-%%   & \equiv \,\, && \hat{\top}\timesh(\hat{n^M}\in \hat{n} \rah \Type_{i})\timesh(\hat{z^M}\in \hat{n^M}\,\hat{z})\timesh \big((\hat{x}\in \hat{n})\rah  \hat{n^M}\,\hat{x}\rah  \hat{n^M}\,(\hat{s}\,\hat{x})\big) \\
-%%   & && (\boldsymbol{\cdot},Nat : \U,zero:\underline{Nat},suc:Nat\ra\underline{Nat})^\E\,(\tt\commah \hat{n}\commah \hat{z}\commah \hat{s})\,(\tt\commah\hat{n^M}\commah \hat{z^M}\commah \hat{s^M}) \\
-%%   & \equiv \,\, && \hat{\top}\timesh\big(\hat{n^E} \in (\hat{x}\in \hat{n})\rah  \hat{n^M}\,\hat{x}\big)\timesh(\hat{z^E} \in \hat{n^E}\,\hat{z}\eqh \hat{z^M}) \timesh \big((\hat{x}\in \hat{n})\rah  \hat{n^E}\,(\hat{s}\,\hat{x})\eqh \hat{n^M}\,\hat{x}\,(\hat{n^E}\,\hat{x})\big)
-%% \end{alignat*}
-%% The {\color{BrickRed}brick red} coloured result of $\blank^\C$ gives
-%% the types of the type formation rule and constructors as an iterated
-%% $\Sigma$-type. Assuming the existence of constructors, $\blank^\M$
-%% returns the types of motives and methods. Assuming the existence of
-%% constructors and the corresponding motives and methods, $\blank^\E$
-%% returns the types of the eliminator and computation rules as external
-%% equalities $\eqh$. The notation above denotes
-%% \emph{left}-nested iterated $\Sigma$-types. A more precise
-%% presentation would replace each variable with a projection from the
-%% preceding $\Sigma$-type. We use this notation in order to reduce
-%% visual clutter.
-
-%% Each context entry in the theory of codes specifies a type formation
-%% rule or a constructor. In general, the last component of a type in a
-%% context entry is of three possible forms: it is either $\U$,
-%% $\underline{a}$ for some neutral $a$, or $\underline{t=_a u}$. The
-%% following table summarizes the results of the various translations in
-%% the mentioned three cases:
-%% \begin{alignat*}{5}
-%%   & \text{return type}\hspace{1em} && \blank^\C && \blank^\M && \blank^\E \\\hline
-%%   & \U && \text{type formation rule}\hspace{1em} && \text{motive} && \text{eliminator} \\
-%%   & \underline{a} && \text{point constructor}\hspace{1em} && \text{method} && \text{computation rule} \\
-%%   & \underline{t=_a u} && \text{path constructor} && \text{method expressing} && \text{higher computation rule} \\
-%%   & && && \text{preservation of equality}\hspace{1em}
-%% \end{alignat*}
-%% Note that there is no syntactic distinction between the three kinds of
-%% constructors above. Any number of them can be introduced in any order,
-%% and each constructor can refer to any previous one. A distinguishing
-%% feature of our approach is the utilisation of universes instead of
-%% structural rules to introduce new sorts and to ensure strict
-%% positivity.
-
-%% The $\blank^\C$, $\blank^\M$ and $\blank^\E$ operations are defined by
-%% induction on the derivations of the source syntax. The operations are
-%% identity on derivations of external contexts and external terms (of the
-%% forms $\vdashh\hat{\Gamma}$ and $\hat{\Gamma}\vdashh \hat{t}\in
-%% \hat{A}$) and derive target terms from theory of codes contexts, types
-%% and terms (of the forms $\hat{\Gamma}\vdash\Delta$,
-%% $\hat{\Gamma}\semicol\Delta\vdash A$ and
-%% $\hat{\Gamma}\semicol\Delta\vdash t:A$, respectively). We only present
-%% the non-identity parts with pattern matching notation, describing how
-%% a context, type or a term in the theory of codes is converted to a
-%% term in the external theory.
-
-%% All three operations respect definitional equality. This amounts to
-%% preserving equalities of the substitution calculus, as there are no
-%% $\beta$-rules introduced in the theory of codes.
-
-%---------------------------------------------------------------------
+In the following three sections we present syntactic translations yielding
+algebras, homomorphisms, displayed algebras and their sections. The presentation
+here is informal and focuses on readability, similarly to our presentation of
+signatures before.
 
@@
 
-Given a context in the theory of signatures, $\blank^\A$ computes the
-corresponding notion of algebras as an iterated $\Sigma$-type in the external
-syntax. It is specified as follows.
+We use $\blank^\A$ to the denote the translation which computes algebras. It is
+specified as follows, for contexts, types and terms in the theory of signatures.
+
 \[
@@
   & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u}
-\end{alignat*}\endgroup
+\end{alignat*}
+\endgroup
 
@@
 
-Note that the resulting left-nested $\Sigma$ type could be written explicitly as
-below, without syntactic sugar. We shall keep to the more readable notation
-from now on.
+Note that the resulting $\Sigma$ type is left-nested, and we use the reassociated notation for readability. The result could be written without syntactic sugar the following way:
 \begin{alignat*}{5}
@@
 \end{alignat*}
+We shall keep to the short notation from now on.
 
@@
 
-TODO
+In this section, we specify the $\blank^\M$ translation, which computes
+homomorphisms of algebras. However, it is illustrative to first consider a
+logical relation interpretation, and move to homomorphisms from that. For
+dependent type theories, logical relation models are well-known (see
+e.g.\ \cite{atkey}), so they are certainly applicable to our restricted syntax
+as well.
+
+Below, we list induction motives for $\blank^\M$; these remain the same as we
+move from logical relations to homomorphisms. We fix a universe level $\hat{i}$ for
+the translations.
+\[
+\infer{\hat{\Gamma}\vdashh\Delta^\M \in \Delta^\A\rah\Delta^\A\rah \Type_{\hat{i}}}
+      {\hat{\Gamma}\vdash\Delta}
+\hspace{1em}
+\infer{
+  \hat{\Gamma}\vdashh A^\D \in \hat{(}\hat{\gamma_0\,\gamma_1}\in\Delta^\A\hat{)}\rah \Delta^\M\,\hat{\gamma_0\,\gamma_1}\rah  A^\A\,\hat{\gamma_0}\rah A^\A\,\hat{\gamma_1}\rah \Type_{\hat{i}}}
+      {\hat{\Gamma}\semicol\Delta\vdash A}
+\]
+
+\[
+\infer{
+  \hat{\Gamma}\vdashh A^\D \hat{ : (\gamma_0\,\gamma_1 : \blc{\Delta^\A})(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\ra \blc{A^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,(\blc{t^\A}\,\gamma_0)\,(\blc{t^\A}\,\gamma_1)
+      }}
+      {\hat{\Gamma}\semicol\Delta\vdash t : A}
+\]
+
+Contexts are mapped to (proof-relevant) relations and types to families of relations
+depending on interpreted contexts. For terms, we get something which is often called
+a \emph{fundamental theorem} of a logical relation: every term is related to itself
+in related semantic contexts.
+
+We present below the logical relation interpretation only for contexts, variables,
+the universe and the inductive function space.
+
+\begingroup
+\allowdisplaybreaks
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot}^\M\hat{\gamma_0\,\gamma_1} && :\equiv \hat{\top} \\
+  & (\Delta,x:A)^\M\,\hat{\gamma_0\,\gamma_1} && :\equiv \hat{(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\times \blc{A^\M}\,\gamma_0\,\gamma_1\,\gamma^M} \\
+  & x^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^M} \\
+  & \U^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,A_0\,A_1} && :\equiv \hat{A_0\ra A_1 \ra \Type_0} \\
+  & (\underline{a})^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,t_0\,\,t_1} && :\equiv \hat{
+    \blc{a^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,t_0\,\,t_1}\\
+  & ((x:a)\ra B)^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,f_0\,f_1} && :\equiv
+  \hat{(x_0 : \blc{(\underline{a})^\A}\,\gamma_0)(x_1 : \blc{(\underline{a})^\A}\,\gamma_1)(x^M : \blc{(\underline{a})^\M}\,\gamma_0\,\gamma_1\,\gamma^M)}\\
+  & && \hat{\hspace{1em}\ra \blc{B^\M}\,(\gamma_0,\,x_0)\,(\gamma_1,\,x_1)\,(\gamma^M,\,x^M)\,(f_0\,x_0)\,(f_1\,x_1)}
+\end{alignat*}
+\endgroup
+
+We interpret the universe as relation space, and function types as types
+expressing pointwise relatedness of functions. This interpretation works the
+same way for unrestricted (non-strictly positive) function types as well.
+
+However, we would like to have underlying functions instead of relations in
+homomorphisms. We take hint from the fact that for classical (simply-typed)
+algebraic theories, a logical relation is equivalent to a homomorphism if and
+only if the underlying relation is the graph of a function
+\cite[pg. 5]{udayReynolds}. Thus we make the following change:
+\begin{alignat*}{5}
+  & \U^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,A_0\,A_1} && :\equiv \hat{A_0\ra A_1}
+\end{alignat*}
+This requires us to change $(\underline{a})^\M$ as well, since we need to
+produce a type as result, but $a^M$ now yields a function. We can morally keep
+the old definition, by viewing the result of $a^M$ as a functional relation, and
+relating $t_0$ and $t_1$ by its graph:
+\begin{alignat*}{5}
+  & (\underline{a})^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,t_0\,\,t_1} && :\equiv \hat{
+    \blc{a^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,t_0 = t_1}
+\end{alignat*}
+At this point we have merely restricted relations to functions, and left the
+rest of the interpretation unchanged. However, this is not strictly the desired
+notion of homomorphism. Consider the $\blank^\M$ interpretation for natural
+numbers:
+\begin{alignat*}{5}
+  & && (\boldsymbol{\cdot},Nat : \U,zero:\underline{Nat},suc:Nat\ra\underline{Nat})^\M
+    \,\hat{(\tt,\,N_0,\,z_0,\,s_0)\,(\tt,\,N_1,\,z_1,\,s_1)} \\
+    & \equiv\,\, && \hat{\top\times(N^M : N_0\ra N_1)}\\
+    & && \hat{\hspace{1em}\times\,\,(z^M : N^M\,z_0 = z_1)}\\
+    & && \hat{\hspace{1em}\times\,\,(s^M : (x_0 : N_0)(x_1 : N_1)(x^M : N^M\,x_0 = x_1)\ra N^M\,(s_0\,x_0)=s_1\,x_1)}
+\end{alignat*}
+In $\hat{s^M}$, there is a superfluous $\hat{x^M}$ equality proof. Fortunately,
+in the translation of the inductive function space, we can just singleton
+contract $\hat{x^M}$ away, yielding an equivalent, but stricter definition:
+\begin{alignat*}{5}
+  & ((x:a)\ra B)^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,f_0\,f_1} :\equiv \\
+  & \hat{\hspace{2em}(x_0 : \blc{a^\A}\,\gamma_0)\ra \blc{B^\M}\,(\gamma_0,\,x_0)\,(\gamma_1,\,\blc{a^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,x_0)\,(\gamma^M,\,\refl)\,
+    (f_0\,x_0)\,(f_1\,(\blc{a^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,x_0))}
+\end{alignat*}
+Now, the $\beta$-rule for successors is as expected:
+\begin{alignat*}{5}
+ & \hat{s^M : (x_0 : N_0)\ra N^M\,(s_0\,x_0)=s_1\,(N^M\,x_0)}
+\end{alignat*}
+
+Note that this singleton contraction is not possible for a general
+non-strictly positive function space. We rely on the domain being small:
+$(\underline{a})^\M$ yields an equation, but for general $\Gamma\vdash A$ types,
+$A^\M$ only yields an unknown relation.
+
+\subsection{The full homomorphism interpretation}
+
+TDAODKAS
+
+%% --------------------------------------------------------------------------------
 
@@
 
-The $\blank^\D$ translation computes displayed algebras for a signature. Recall
-from Section \ref{sec:general} that displayed algebras serve as bundles of
-induction motives and methods.
-
-$\blank^\D$ is essentially an unary logical predicate translation over the
-$\blank^\A$ model. It is related to the logical predicate translation of
-Bernardy et al.\ \cite{bernardy12parametricity}, but our implementation differs
-by interpreting contexts as $\Sigma$-types instead of extended contexts.
-
-We fix a level $i$ for the universe we would like to eliminate into. For each
+The $\blank^\D$ translation computes displayed algebras for a signature, which
+can be viewed as bundles of induction motives and methods. $\blank^\D$ is
+essentially an unary logical predicate translation over the $\blank^\A$
+model. It is related to the logical predicate translation of Bernardy et
+al.\ \cite{bernardy12parametricity}, but our implementation differs by
+interpreting contexts as $\Sigma$-types instead of extended contexts.
+
+For each
 context $\Delta$, $\Delta^\D$ is a predicate over the standard interpretation
@@
 \]
-For a term $t$, $t^\D$ witnesses that the predicate corresponding to
-its type holds for $t^\A$. This is often called a \emph{fundamental
-  theorem} in the literature on logical predicates.
+For a term $t$, $t^\D$ witnesses that the predicate corresponding to its type
+holds for $t^\A$; again this can be viewed as a fundamental theorem for the
+predicate interpretation.
+
 \[
@@
 
-The predicate for a function type with small domain expresses preservation of
-predicates (at the domain and codomain types). Witnesses of application are
-given by recursive application of $\blank^\D$. The definitions for the other
-(non-inductive) function spaces are similar, except there is no predicate for
-the domain types, and thus no witnesses are required.
+The predicate for a function type for inductive parameters expresses
+preservation of predicates. Witnesses of application are given by recursive
+application of $\blank^\D$. The definitions for the other (non-inductive)
+function spaces are similar, except there is no predicate for the domain types,
+and thus no witnesses are required.
 
@@
 $\refl$ is interpreted with just reflexivity in the target syntax. The
-interpretation of $\J$ is given by a double $\hat{\J}$ application, which
-definition is sourced from \cite{lasson}. Here, we use a shortened $\hat{\J}$
-notation; we refer to the formalization (Section \ref{sec:formalization}) for
-the details.
+interpretation of $\J$ is given by a double $\hat{\J}$ application, borrowing
+the definition from Lasson \cite{lasson}. Here, we use a shortened $\hat{\J}$
+notation; see the formalization (Section \ref{sec:formalization}) for details.
 
@@
   & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\D\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\\
-  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1D}}\in \hat{S^1}\rah\Type_i\hat{)}\timesh\hat{(}\hat{b^D}\in \hat{S^{1D}\,b}\hat{)}\timesh\hat{(}\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D}\hat{)}
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1D}}\in \hat{S^1}\rah\Type_{\hat{i}}\hat{)}\timesh\hat{(}\hat{b^D}\in \hat{S^{1D}\,b}\hat{)}\timesh\hat{(}\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D}\hat{)}
 \end{alignat*}
@@
 
-%---------------------------------------------------------------------
+%% --------------------------------------------------------------------------------
 
@@
 \label{sec:sections}
-
-TODO: REWRITE to remove redundacy from homomorphism section.
 
@@
 displayed algebras. The universe level $i$ was previously chosen for the
-$\blank^\D$ translation.
-\[
-\infer{\hat{\Gamma}\semicol\Delta^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah \Type_{i}}{\hat{\Gamma}\vdash\Delta}
+$\blank^\M$ and $\blank^\D$ translations.
+\[
+\infer{\hat{\Gamma}\semicol\Delta^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah \Type_{\hat{i}}}{\hat{\Gamma}\vdash\Delta}
 \]
@@
 \[
-\infer{\hat{\Gamma}\vdashh A^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\hat{(}\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}\hat{(}\hat{x}\in A^\A\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{x}\rah \Type_{i}}{\hat{\Gamma}\semicol\Delta\vdash A}
+\infer{\hat{\Gamma}\vdashh A^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\hat{(}\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}\hat{(}\hat{x}\in A^\A\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{x}\rah \Type_{\hat{i}}}{\hat{\Gamma}\semicol\Delta\vdash A}
 \]

### 2019-01-30 paper.tex
Morphisms progress

@@
 \newcommand{\idr}{\mathsf{idr}}
+\newcommand{\inv}{\mathsf{inv}}
 
@@
 
+The unit type is denoted $\hat{\top}$ with constructor $\tt$.
+
 Dependent function space is denoted $\hat{(\hat{x}\in \hat{A})\rah \hat{B}}$. We
@@
 
-We will use the following functions defined using $\hat{\J}$ in the
-standard way. We write $\TR_{\hat{P}}\,\hat{e}\,\hat{t}\in
-\hat{P}\,\hat{v}$ for transport of $\hat{t} \in \hat{P}\,\hat{u}$
-along $\hat{e} \in \hat{u}\eqh \hat{v}$. We write
-$\ap\,\hat{f}\,\hat{e}\in\hat{f}\,\hat{u}\eqh\hat{f}\,\hat{v}$ for
-$\hat{f}:\hat{A}\rah\hat{B}$ and $\hat{e}\in\hat{u}\eqh\hat{v}$,
-$\hat{\apd\,\hat{f}\,\hat{e} \in \TR_{\hat{P}}\,\hat{e}\,(\hat{f}\,\hat{u})
-\eqh \hat{f}\,\hat{v}}$ for $\hat{\hat{f}\in(\hat{x}\in \hat{A})\rah
-\hat{B}}$ and $\hat{e} \in \hat{u}\eqh \hat{v}$.
-
-The unit type is denoted $\hat{\top}$ with constructor $\tt$.
+We will use the following functions defined using $\hat{\J}$ in the standard
+way. We write $\TR_{\hat{P}}\,\hat{e}\,\hat{t}\in \hat{P}\,\hat{v}$ for
+transport of $\hat{t} \in \hat{P}\,\hat{u}$ along $\hat{e} \in \hat{u}\eqh
+\hat{v}$. We write
+$\ap\,\hat{f}\,\hat{e}\in\hat{f}\,\hat{u}\eqh\hat{f}\,\hat{v}$ where
+$\hat{f}:\hat{A}\rah\hat{B}$ and $\hat{e}\in\hat{u}\eqh\hat{v}$, also
+$\hat{\apd\,\hat{f}\,\hat{e} \in \TR_{\hat{P}}\,\hat{e}\,(\hat{f}\,\hat{u}) \eqh
+  \hat{f}\,\hat{v}}$ where $\hat{\hat{f}\in(\hat{x}\in \hat{A})\rah \hat{B}}$
+and $\hat{e} \in \hat{u}\eqh \hat{v}$. Borrowing notation from the homotopy type
+theory book \cite{HoTTbook}, we write $\hat{p \sqcdot q}$ for transitivity and
+$\hat{p^{-1}}$ for symmetry. We also make use of the groupoid law $\hat{\inv\,p
+  : p^{-1}\sqcdot p = \refl}$.
 
@@
 
-Note that this singleton contraction is not possible for a general
-non-strictly positive function space. We rely on the domain being small:
-$(\underline{a})^\M$ yields an equation, but for general $\Gamma\vdash A$ types,
-$A^\M$ only yields an unknown relation.
-
-\subsection{The full homomorphism interpretation}
-
-TDAODKAS
+Note that this singleton contraction is not possible for a general non-strictly
+positive function space. We rely on the domain being small: $(\underline{a})^\M$
+yields an equation, but for general $\hat{\Gamma;\,}\Delta\vdash A$ types, $A^\M$ only yields
+an unknown relation.
+
+\subsection{The full homomorphism interpretation} We list below the $\blank^\M$
+interpretations for the syntax. For brevity, we mostly leave $\hat{\gamma_0}$
+and $\hat{\gamma_1}$ implicit, notating only $\hat{\gamma^M}$ where necessary.
+We also omit some subscript parameters from $\hat{\J}$ applications.
+
+\begingroup
+\allowdisplaybreaks
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot}^\M\hat{\gamma_0\,\gamma_1} && :\equiv \hat{\top} \\
+  & (\Delta,x:A)^\M\,\hat{\gamma_0\,\gamma_1} && :\equiv \hat{(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\times \blc{A^\M}\,\gamma^M} \\
+  & x^\M\,\hat{\gamma^M} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^M} \\
+  & \U^\M\,\hat{\gamma^M\,A_0\,A_1} && :\equiv \hat{A_0\ra A_1} \\
+  & (\underline{a})^\M\,\hat{\gamma^M\,t_0\,\,t_1} && :\equiv \hat{
+    \blc{a^\M}\,\gamma^M\,t_0 = t_1} \\
+  & ((x:a)\ra B)^\M\,\hat{\gamma^M\,f_0\,f_1} && :\equiv \\
+  & && \hat{\hspace{-8em}(x_0 : \blc{a^\A}\,\gamma_0)\ra \blc{B^\M}\,(\gamma_0,\,x_0)\,(\gamma_1,\,\blc{a^\M}\,\gamma^M\,x_0)\,(\gamma^M,\,\refl)\,
+    (f_0\,x_0)\,(f_1\,(\blc{a^\M}\,\gamma^M\,x_0))}\\
+  & (t\,u)^\M\,\hat{\gamma^M} && :\equiv  \hat{\J\,(\blc{t^\M}\,\gamma^M\,(\blc{u^\A}\,\gamma_0))\,(\blc{u^\M}\,\gamma^\M) }\\
+  & ((\hat{x:A})\ra B)^\M\,\hat{\gamma^M\,f_0\,f_1} && :\equiv \hat{(x : A)\ra \blc{B^\M}\,\gamma^M\,(f_0\,x)\,(f_1\,x)} \\
+  & (t\,\hat{u})^\M\,\hat{\gamma^M} && :\equiv t^\M\,\hat{\gamma^M\,u} \\
+  & (t=_a u)^\M\,\hat{\gamma^M} && :\equiv
+    \hat{\lambda\,e.\,(\blc{t^\M}\,\gamma^M\,^{-1})\sqcdot\ap\,(\blc{a^\M}\,\gamma^M)\,e\sqcdot\,\blc{u^\M}\,\gamma^M}\\
+  & (\refl_t)^\M\,\hat{\gamma^M} && :\equiv \hat{\inv\,(\blc{t^\M}\,\gamma^M)}\\
+  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\M\,\hat{\gamma^M} && :\equiv\\
+  & && \hspace{-10em}\hat{\J\,\bigg(\J\,\Big(\J\,\big(\J\,(\lambda\,p_1\,p^M\,pr_1\,pr^M.\, pr^M)\,(\blc{t^\M}\,\gamma^M)}  \\
+  & && \hspace{-9em}\hat{
+      (\mathsf{uncurry}\,\blc{p^\A}\,\gamma_1)\,(\mathsf{uncurry}\,\blc{p^\M}\,\gamma^M)\,
+      (\blc{pr^\A}\,\gamma_1)\,(\blc{pr^\M}\,\gamma^M)\big)\,
+        (\blc{eq^\A}\,\gamma_0)\Big)(\blc{u^\M}\,\gamma^M)\bigg)(\blc{eq^\M}\,\gamma^M)   }\\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\M\,\hat{\gamma^M} && :\equiv\\
+  & && \hspace{-9em}\hat{
+    \J\,\big(\J\,(\lambda\,p_1\,p^M.\,\refl)\,(\blc{t^\M}\,\gamma^M)\,
+      (\mathsf{uncurry}\,\blc{p^\A}\,\gamma_1)\,
+      (\mathsf{uncurry}\,\blc{p^\M}\,\gamma^M)\big)
+      (\blc{pr^\M}\,\gamma^M)
+    }\\
+  & ((\hat{x}\in \hat{A})\ra b)^\M\,\hat{\gamma^M} && :\equiv
+       \hat{\lambda f\,x.\,\,\blc{b^\M}\,\gamma^M\,(f\,x)} \\
+  & (t\,\hat{u})^\M\,\hat{\gamma^M} && :\equiv
+       \hat{\ap\,(\lambda f.\,f\,u)\,(\blc{t^\M}\,\gamma^M)}
+\end{alignat*}
+\endgroup
 
@@
 \]
-We introduce the following shorthand: $t\,\hat{\gamma}\,\hat{\gamma^D}$ is
-abbreviated as $t\,\hat{\gamma^2}$ for some $t$ expression. The
-implementation of $\blank^\D$ is given below.  \begingroup
+The implementation of $\blank^\D$ is given below. We leave
+$\hat{\gamma}$-s mostly implicit, marking only $\hat{\gamma^D}$ witnesses.
+\begingroup
 %% \allowdisplaybreaks
 \begin{alignat*}{5}
-  & \boldsymbol{\cdot}^\D\,\hat{\gamma} && :\equiv \hat{\top} \\
-  & (\Delta,\,x:A)^\D\,(\hat{\gamma}\commah\,\hat{t}) && :\equiv \hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\timesh A^\D\,\hat{\gamma^2}\,\hat{t} \\
-  & x^\D\,\hat{\gamma^2} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^D} \\ % TODO: De Bruijn
-  & \U^\D\,\hat{\gamma^2}\,\hat{A} && :\equiv \hat{A} \rah  \Type_{\hat{i}} \\
-  & (\underline{a})^\D\,\hat{\gamma^2}\,\hat{t} && :\equiv a^\D\,\hat{\gamma^2}\,\hat{t} \\
-  & ((x:a)\ra B)^\D\,\hat{\gamma^2}\,\hat{f} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\hat{(}\hat{x^D}\in a^\D\,\hat{\gamma^2}\,\hat{x}\hat{)} \rah  B^\D\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)}\,\hat{(}\hat{\gamma^D}\commah\hat{x^D}\hat{)}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,u)^\D\,\hat{\gamma^2} && :\equiv \hat{(}t^\D\,\hat{\gamma^2}\hat{)}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\D\,\hat{\gamma^2}\hat{)} \\
-  & ((\hat{x}\in \hat{A})\ra B)^\D\,\hat{\gamma^2}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^2}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^\D\,\hat{\gamma^2} && :\equiv t^\D\,\hat{\gamma^2}\,\hat{u} \\
-  & (t=_a u)^\D\,\hat{\gamma^2}\,\hat{e} && :\equiv \TR_{\hat{(}a^\D\,\hat{\gamma^2}\hat{)}}\,\hat{e}\,\hat{(}t^\D\,\hat{\gamma^2}\hat{)} \eqh u^\D\hat{\gamma^2} \\
-  & (\refl_t)^\D\,\hat{\gamma^2} && :\equiv \hat{\refl}_{\hat{(}t^\D\,\hat{\gamma^2}\hat{)}} \\
-  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\D\,\hat{\gamma^2} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^2}\hat{)} \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\D\,\hat{\gamma^2} && :\equiv \hat{\refl} \\
-  & ((\hat{x}\in \hat{A})\ra b)^\D\,\hat{\gamma^2}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^\D\,\hat{\gamma^2}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^\D\,\hat{\gamma^2} && :\equiv t^\D\,\hat{\gamma^2}\,\hat{u}
-\end{alignat*}\endgroup
+  & \boldsymbol{\cdot}^D\,\hat{\gamma} && :\equiv \hat{\top} \\
+  & (\Delta,\,x:A)^D\,(\hat{\gamma}\commah\,\hat{t}) && :\equiv \hat{(}\hat{\gamma^D}\in\Delta^D\,\hat{\gamma}\hat{)}\timesh A^D\,\hat{\gamma^D}\,\hat{t} \\
+  & x^D\,\hat{\gamma^D} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^D} \\
+  & \U^D\,\hat{\gamma^D}\,\hat{A} && :\equiv \hat{A} \rah  \Type_{\hat{i}} \\
+  & (\underline{a})^D\,\hat{\gamma^D}\,\hat{t} && :\equiv a^\D\,\hat{\gamma^D}\,\hat{t} \\
+  & ((x:a)\ra B)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\hat{(}\hat{x^D}\in a^\D\,\hat{\gamma^D}\,\hat{x}\hat{)} \rah  B^\D\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)}\,\hat{(}\hat{\gamma^D}\commah\hat{x^D}\hat{)}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,u)^D\,\hat{\gamma^D} && :\equiv \hat{(}t^\D\,\hat{\gamma^D}\hat{)}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\D\,\hat{\gamma^D}\hat{)} \\
+  & ((\hat{x}\in \hat{A})\ra B)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u} \\
+  & (t=_a u)^D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}}\,\hat{e}\,\hat{(}t^\D\,\hat{\gamma^D}\hat{)} \eqh u^\D\hat{\gamma^D} \\
+  & (\refl_t)^D\,\hat{\gamma^D} && :\equiv \hat{\refl}_{\hat{(}t^\D\,\hat{\gamma^D}\hat{)}} \\
+  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^D\,\hat{\gamma^D} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^D}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^D}\hat{)} \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
+  & ((\hat{x}\in \hat{A})\ra b)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u}
+\end{alignat*}
+\endgroup
 The predicate for a context is given by iterating $\blank^\D$ for its
@@
 \]
-In addition to $\hat{\gamma^2}$, we use $t\,\hat{\gamma^3}$ to abbreviate $t\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{\gamma^S}$. The implementation is the following.
+Here, we mostly notate $\hat{\gamma^S}$ parameters and applications, making
+$\hat{\gamma^D}$-s implicit in addition to $\hat{\gamma}$-s. The implementation
+is the following.
 \begingroup
@@
   & \boldsymbol{\cdot}^\S\,\hat{\gamma}\,\hat{\gamma^D} && :\equiv \hat{\top} \\
-  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv \hat{(}\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2}\hat{)}\timesh A^\S\,\hat{\gamma^3}\,\hat{t}\,\hat{t^D} \\
-  & x^\S\,\hat{\gamma^3} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^S} \\ % TODO: De Bruijn
-  & \U^\S\,\hat{\gamma^3}\,\hat{A}\,\hat{A^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  \hat{A^D}\,\hat{x} \\
-  & (\underline{a})^\S\,\hat{\gamma^3}\,\hat{t}\,\hat{t^D} && :\equiv a^\S\,\hat{\gamma^3}\,\hat{t} \eqh \hat{t^D} \\
-  & ((x:a)\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\S\,\hat{(}\hat{\gamma,x}\hat{)}\,\hat{(}\hat{\gamma^D,}\,a^\S\,\hat{\gamma^3}\,\hat{x}\hat{)}\,\hat{(}\hat{\gamma^S,}\,\hat{\refl}\hat{)} \\
-  & && \hspace{8.4em} \hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{\big(}\hat{f^D}\,\hat{x}\,\hat{(}a^\S\,\hat{\gamma^3}\,\hat{x}\hat{)}\hat{\big)} \\
-  & (t\,u)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^3}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^3}\hat{)} \\
-  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^3}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^3}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^\S\,\hat{\gamma^3} && :\equiv t^\S\,\hat{\gamma^3}\,\hat{u} \\
-  & (t=_a u)^\S\,\hat{\gamma^3}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^3}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^3}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
-  & (\refl_t)^\S\,\hat{\gamma^3} && :\equiv \hat{\J}\,\hat{\refl}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)} \\
-  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S\,\hat{\gamma^3} && :\equiv \\
-  & && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}.\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}\,\\
-  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^3}\hat{)}\hat{\big)}\,eq^\A\,\hat{\gamma}\hat{\Big)}\,u^\S\,\hat{\gamma^3}\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^3}\hat{)} \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^3} && :\equiv \\
-  & && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^3}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^3}\hat{)}\\
-  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^3\,f\,t} && :\equiv b^\S\,\hat{\gamma^3}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
-  & (t\,\hat{u})^\S\,\hat{\gamma^3} && :\equiv \ap\,\hat{(}\hat{\lambda} \hat{f.}\hat{f}\,\hat{u}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^3}\hat{)}
-\end{alignat*}
-%% \endgroup
+  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv \hat{(}\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2}\hat{)}\timesh A^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} \\
+  & x^\S\,\hat{\gamma^S} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^S} \\
+  & \U^\S\,\hat{\gamma^S}\,\hat{A}\,\hat{A^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  \hat{A^D}\,\hat{x} \\
+  & (\underline{a})^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} && :\equiv a^\S\,\hat{\gamma^S}\,\hat{t} \eqh \hat{t^D} \\
+  & ((x:a)\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\S\,\hat{(}\hat{\gamma,x}\hat{)}\,\hat{(}\hat{\gamma^D,}\,a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\,\hat{(}\hat{\gamma^S,}\,\hat{\refl}\hat{)} \\
+  & && \hspace{8.4em} \hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{\big(}\hat{f^D}\,\hat{x}\,\hat{(}a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\hat{\big)} \\
+  & (t\,u)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^S}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)} \\
+  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv t^\S\,\hat{\gamma^S}\,\hat{u} \\
+  & (t=_a u)^\S\,\hat{\gamma^S}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^S}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
+  & (\refl_t)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{\refl}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)} \\
+  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S\,\hat{\gamma^S} && :\equiv \\
+  & && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}.\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\\
+  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,(eq^\A\,\hat{\gamma})\hat{\Big)}\,(u^\S\,\hat{\gamma^S})\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^S}\hat{)} \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \\
+  & && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\\
+  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^S\,f\,t} && :\equiv b^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
+  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv \ap\,\hat{(}\hat{\lambda} \hat{f.}\hat{f}\,\hat{u}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}
+\end{alignat*}
+\endgroup
 

### 2019-01-30 paper.tex
shuffling stuff into appendices

@@
 In this section, we specify the $\blank^\M$ translation, which computes
-homomorphisms of algebras. However, it is illustrative to first consider a
-logical relation interpretation, and move to homomorphisms from that. For
+homomorphisms of algebras. We do so by first considering a logical relation
+interpretation, and then refining it towards computing homomorphisms. For
 dependent type theories, logical relation models are well-known (see
@@
 positive function space. We rely on the domain being small: $(\underline{a})^\M$
-yields an equation, but for general $\hat{\Gamma;\,}\Delta\vdash A$ types, $A^\M$ only yields
-an unknown relation.
-
-\subsection{The full homomorphism interpretation} We list below the $\blank^\M$
-interpretations for the syntax. For brevity, we mostly leave $\hat{\gamma_0}$
-and $\hat{\gamma_1}$ implicit, notating only $\hat{\gamma^M}$ where necessary.
-We also omit some subscript parameters from $\hat{\J}$ applications.
+yields an equation, but for general $\hat{\Gamma;\,}\Delta\vdash A$ types,
+$A^\M$ only yields an unknown relation.
+
+$\blank^\M$ is very similar to the translation to displayed algebra sections,
+which is discussed in Section \ref{sec:sections}. In essence, $\blank^\M$ is a
+non-dependently typed version of the section translation. We will discuss in
+more detail the interpretations of equalities and the other function types in
+Section \ref{sec:sections}, and we leave the rest of the homomorphism
+translation to Appendix \ref{sec:morphismrules}.
+
+%% --------------------------------------------------------------------------------
+
+\section{Displayed Algebras}
+\label{sec:displayed}
+
+The $\blank^\D$ translation computes displayed algebras for a signature, which
+can be viewed as bundles of induction motives and methods. $\blank^\D$ is
+essentially an unary logical predicate translation over the $\blank^\A$
+model. It is related to the logical predicate translation of Bernardy et
+al.\ \cite{bernardy12parametricity}, but our implementation differs by
+interpreting contexts as $\Sigma$-types instead of extended contexts.
+
+For each
+context $\Delta$, $\Delta^\D$ is a predicate over the standard interpretation
+$\Delta^\A$. For a type $\Delta\vdash A$, $A^\D$ is a predicate over $A^\A$,
+which also depends on $\hat{\gamma}\in\Delta^\A$ and a witness of
+$\Delta^\D\,\hat{\gamma}$. All of these may refer to a target theory context
+$\hat{\Gamma}$.
+\[
+\infer{\hat{\Gamma}\vdashh\Delta^\D \in \Delta^\A\rah \Type_{\hat{i+1}}}{\hat{\Gamma}\vdash\Delta}
+\hspace{2em}
+\infer{
+  \hat{\Gamma}\vdashh A^\D \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah  A^\A\,\hat{\gamma}\rah \Type_{\hat{i+1}}}
+      {\hat{\Gamma}\semicol\Delta\vdash A}
+\]
+For a term $t$, $t^\D$ witnesses that the predicate corresponding to its type
+holds for $t^\A$; again this can be viewed as a fundamental theorem for the
+predicate interpretation.
+
+\[
+\infer{\hat{\Gamma}\vdashh t^\D \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}}{\hat{\Gamma}\semicol\Delta\vdash t : A}
+\]
+The implementation of $\blank^\D$ is given below. We leave
+$\hat{\gamma}$-s mostly implicit, marking only $\hat{\gamma^D}$ witnesses.
+\begingroup
+%% \allowdisplaybreaks
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot}^D\,\hat{\gamma} && :\equiv \hat{\top} \\
+  & (\Delta,\,x:A)^D\,(\hat{\gamma}\commah\,\hat{t}) && :\equiv \hat{(}\hat{\gamma^D}\in\Delta^D\,\hat{\gamma}\hat{)}\timesh A^D\,\hat{\gamma^D}\,\hat{t} \\
+  & x^D\,\hat{\gamma^D} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^D} \\
+  & \U^D\,\hat{\gamma^D}\,\hat{A} && :\equiv \hat{A} \rah  \Type_{\hat{i}} \\
+  & (\underline{a})^D\,\hat{\gamma^D}\,\hat{t} && :\equiv a^\D\,\hat{\gamma^D}\,\hat{t} \\
+  & ((x:a)\ra B)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\hat{(}\hat{x^D}\in a^\D\,\hat{\gamma^D}\,\hat{x}\hat{)} \rah  B^\D\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)}\,\hat{(}\hat{\gamma^D}\commah\hat{x^D}\hat{)}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,u)^D\,\hat{\gamma^D} && :\equiv \hat{(}t^\D\,\hat{\gamma^D}\hat{)}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\D\,\hat{\gamma^D}\hat{)} \\
+  & ((\hat{x}\in \hat{A})\ra B)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u} \\
+  & (t=_a u)^D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}}\,\hat{e}\,\hat{(}t^\D\,\hat{\gamma^D}\hat{)} \eqh u^\D\hat{\gamma^D} \\
+  & (\refl_t)^D\,\hat{\gamma^D} && :\equiv \hat{\refl}_{\hat{(}t^\D\,\hat{\gamma^D}\hat{)}} \\
+  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^D\,\hat{\gamma^D} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^D}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^D}\hat{)} \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
+  & ((\hat{x}\in \hat{A})\ra b)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u}
+\end{alignat*}
+\endgroup
+The predicate for a context is given by iterating $\blank^\D$ for its
+constituent types. For a variable, the corresponding witness is looked up from
+$\hat{\gamma^D}$.
+
+The translation of the universe, given an element of
+$\hat{A}\in\U^\A\,\hat{\gamma}$ (with
+$\U^\A\,\hat{\gamma}\equiv\Type_{\hat{0}}$) returns the predicate space over
+$\hat{A}$. For $\underline{a}$ types, we just return the translation of $a$.
+
+The predicate for a function type for inductive parameters expresses
+preservation of predicates. Witnesses of application are given by recursive
+application of $\blank^\D$. The definitions for the other (non-inductive)
+function spaces are similar, except there is no predicate for the domain types,
+and thus no witnesses are required.
+
+The predicate for the equality type $t=_a u$, for each $\hat{e}\in(t=_a
+u)^\A\,\hat{\gamma}$, i.e.\ $\hat{e}\in t^\A\,\hat{\gamma}\eqh
+u^\A\,\hat{\gamma}$, says that $t^\D$ and $u^\D$ are equal. As these have
+different types, we have to transport over the original equality
+$\hat{e}$. Hence, induction methods for path constructors will be \emph{paths
+  over paths} in the sense of homotopy type theory.
+
+$\refl$ is interpreted with just reflexivity in the target syntax. The
+interpretation of $\J$ is given by a double $\hat{\J}$ application, borrowing
+the definition from Lasson \cite{lasson}. Here, we use a shortened $\hat{\J}$
+notation; see the formalization (Section \ref{sec:formalization}) for details.
+
+Again, let us consider the circle example:
+\begin{alignat*}{5}
+  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\D\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1D}}\in \hat{S^1}\rah\Type_{\hat{i}}\hat{)}\timesh\hat{(}\hat{b^D}\in \hat{S^{1D}\,b}\hat{)}\timesh\hat{(}\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D}\hat{)}
+\end{alignat*}
+The inputs of $\blank^\D$ here are the signature for the circle (the context in
+black) and an $\hat{S^{1}}$-algebra consisting of three non-$\top$ components.
+It returns a family over the type $\hat{S^1}$, an element of this family
+$\hat{b^D}$ at index $\hat{b}$, and and an equality between $\hat{b^D}$ and
+$\hat{b^D}$ which lies over $\hat{loop}$. This is the same as the usual
+induction motives and methods for the circle, e.g.\ as described in \cite{HoTTbook}.
+
+%% --------------------------------------------------------------------------------
+
+\section{Displayed Algebra Sections}
+\label{sec:sections}
+
+TODO revision
+
+The operation $\blank^\S$ yields displayed algebra sections. Sections can be
+viewed as dependent homomorphisms: while homomorphisms are structure-preserving
+families of functions, sections are structure-preserving families of dependent
+functions.
+
+Contexts are interpreted as dependent relations between algebras and
+displayed algebras. The universe level $i$ was previously chosen for the
+$\blank^\M$ and $\blank^\D$ translations.
+\[
+\infer{\hat{\Gamma}\semicol\Delta^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah \Type_{\hat{i}}}{\hat{\Gamma}\vdash\Delta}
+\]
+Types are interpreted as dependent relations which additionally depend on
+$\hat{\gamma}$, $\hat{\gamma^D}$, $\hat{\gamma^S}$ interpretations of the
+context.
+\[
+\infer{\hat{\Gamma}\vdashh A^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\hat{(}\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}\hat{(}\hat{x}\in A^\A\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{x}\rah \Type_{\hat{i}}}{\hat{\Gamma}\semicol\Delta\vdash A}
+\]
+For a term $t$, $t^\S$ witnesses that the relation corresponding to
+its type holds for $t^\A$ and $t^\D$.
+\[
+\infer{\hat{\Gamma}\vdashh t^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\hat{(}\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}\rah  A^\S\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{\gamma^S}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}t^\D\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}}{\hat{\Gamma}\semicol\Delta \vdash t : A}
+\]
+We present the implementation below. Here, we make $\hat{\gamma^D}$-s implicit
+in addition to $\hat{\gamma}$-s, and mostly notate $\hat{\gamma^S}$ parameters
+and applications,
+
+\begingroup
+\allowdisplaybreaks
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot}^\S\,\hat{\gamma}\,\hat{\gamma^D} && :\equiv \hat{\top} \\
+  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv \hat{(}\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2}\hat{)}\timesh A^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} \\
+  & x^\S\,\hat{\gamma^S} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^S} \\
+  & \U^\S\,\hat{\gamma^S}\,\hat{A}\,\hat{A^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  \hat{A^D}\,\hat{x} \\
+  & (\underline{a})^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} && :\equiv a^\S\,\hat{\gamma^S}\,\hat{t} \eqh \hat{t^D} \\
+  & ((x:a)\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\S\,\hat{(}\hat{\gamma,x}\hat{)}\,\hat{(}\hat{\gamma^D,}\,a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\,\hat{(}\hat{\gamma^S,}\,\hat{\refl}\hat{)} \\
+  & && \hspace{8.4em} \hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{\big(}\hat{f^D}\,\hat{x}\,\hat{(}a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\hat{\big)} \\
+  & (t\,u)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^S}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)} \\
+  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv t^\S\,\hat{\gamma^S}\,\hat{u} \\
+  & (t=_a u)^\S\,\hat{\gamma^S}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^S}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
+  & (\refl_t)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{\refl}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)} \\
+  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S\,\hat{\gamma^S} && :\equiv \\
+  & && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}\hat{.}\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\\
+  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,(eq^\A\,\hat{\gamma})\hat{\Big)}\,(u^\S\,\hat{\gamma^S})\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^S}\hat{)} \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \\
+  & && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\\
+  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^S\,f\,t} && :\equiv b^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
+  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv \ap\,\hat{(}\hat{\lambda} \hat{f.}\hat{f}\,\hat{u}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}
+\end{alignat*}
+\endgroup
+
+The $\U^\S$ and $(\underline{a})^\S$ definitions are the key points of
+the $\blank^\S$ operation. The definitions for the other $\blank^\S$
+cases are largely determined by these.
+
+The $\U^\S$ rule yields the type of the eliminator for a type
+formation rule. In the natural numbers example above, the non-indexed
+$Nat : \U$ sort is interpreted as $\hat{\hat{n^S} \in (\hat{x}\in
+\hat{n})\rah \hat{n^D}\,\hat{x}}$. For indexed sorts, the indices are
+first processed by the $\blank^\S$ cases for inductive and non-inductive
+parameters, until the ultimate $\U$ return type is reached. Hence, the
+eliminator for a sort is always a function.
+
+Analogously, the $\blank^\S$ result type for a point or path constructor is
+always a $\beta$-rule, i.e.\ a function type ending in an equality. To
+see why, consider the $(\underline{a})^\S$ definition. It expresses
+that applications of $a^\S$ eliminators must be equal to the
+corresponding $\hat{t^D}$ induction methods. Hence, for path and point
+constructor types, $\blank^\S$ works by first processing all inductive and
+non-inductive indices, then finally returning an equality type.
+
+Let us also consider the $((x:a)\ra B)^\S$ case for inductive
+parameters. Here, we make crucial use of the fact that the domain type
+$a$ is small. This provides us $a^\S\,\hat{\gamma^3\,x}$, which we use
+to witness the $a^\D\,\hat{\gamma^2}\,\hat{x}$ hypothesis for
+$B^\S$. Without the size restriction on inductive parameters (which
+enforces strict positivity), the $\blank^\S$ operation would not be
+possible at all, because $a^\S$ for a large $a$ type would be merely
+an opaque relation instead of an eliminator function.
+
+Here, we only provide abbreviated definitions for the $t\,u$, $t=_a
+u$, $\refl$, $\J$ and $\J\beta$ cases. In the $\J$ case, we write
+$\hat{\mathsf{uncurry}}\,p^\D$ for
+$\hat{\lambda\,\gamma\,\gamma^D\,x\,x^D\,z\,z^D.}\,p^\D\,(\hat{\gamma,x,z})\,(\hat{\gamma^D,x^D,z^D})$
+and analogously elsewhere. The full definitions can be found in the
+Agda formalization. The definitions are highly constrained by the
+required types, and not particularly difficult to implement with the
+help of a proof assistant; they all involve doing successive path
+induction on all equalities available from induction hypotheses, with
+appropriately generalized induction motives.
+
+The full $(\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S$ definition is quite
+large, and, for instance, yields a very large $\beta$-rule for the
+higher inductive torus definition (the reader can confirm this using
+the Haskell implementation). Nevertheless, an implementation focused
+on practicality may provide smaller specialized $\blank^\S$ definitions for
+commonly used equality operations such as composition or inverses.
+
+The circle example is a bit more interesting here:
+\begin{alignat*}{5}
+  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\S\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\,\hat{(}\hat{\tt,\,S^{1M},\,b^D,\,loop^D}\hat{)} \\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1E}}\in \hat{(}\hat{x}\in\hat{S^1}\hat{)}\hat{\,\ra\,}\hat{S^{1M}\,x}\hat{)}\timesh\hat{(}\hat{b^S}\in \hat{S^{1E}\,\,b = b^D}\hat{)}\\
+  & && \hspace{0.75em}\timesh\hat{(}\hat{loop^S\,\,\in\,\,} \TR\,_{\hat{(}\hat{\lambda x.
+\TR\,_{S^{1M}}\,loop\,x\,=\,b^D}\hat{)}}\,\hat{b^S}\,\hat{(}\TR\,_{\hat{(}\hat{\lambda} \hat{x}.
+\TR\,_{\hat{S^{1M}}}\,\hat{loop}\,\hat{(}\hat{S^{1E}\,b}\hat{)}\,\eqh\,\hat{x}\hat{)}}\,\hat{b^S}\,\hat{(}\apd\,\hat{S^{1E}\,loop}\hat{)}\hat{)}\,\\
+  & && \hspace{5.5em} \hat{=\,loop^D}\hat{)}
+\end{alignat*}
+In homotopy type theory, the $\beta$-rule for $loop$ is usually just
+$\apd\,\hat{S^{1E}\,loop}\,\hat{\,=\,loop^D}$, but here all
+$\beta$-rules are propositional, so we need to transport with
+$\hat{b^S}$ to make the equation well-typed. When computing the type
+of $\hat{loop^S}$, we start with $(\underline{b
+  =b})^S\,\,\hat{\gamma^3}\,\,\hat{loop}\,\,\hat{loop^D}$. Next, this
+evaluates to $(b=b)^\S\,\hat{\gamma^3}\,\hat{loop}\eqh\hat{loop^D}$,
+and then we unfold the left hand side to get the doubly-transported
+expression in the result.
+
+In Appendix \ref{sec:app}, we show how the type of displayed algebra sections is
+computed for the two-dimensional sphere. In Appendix \ref{sec:wtypes}, we show
+the same for indexed W-types.
+
+%=====================================================================
+
+\section{On Extension to Categorical Semantics}
+\label{sec:categorical}
+
+TODO
+
+%% \section{Existence of HIITs}
+%% \label{sec:adding}
+
+%% The target type theory supports HIITs if whenever we can derive
+%% $\hat{\Gamma}\vdash\Delta$ in the source theory, the following rules are
+%% admissible.
+%% \[
+%% \infer{\hat{\Gamma}\vdashh\hat{\con}_\Delta\in \Delta^\C}{\vspace{1em}}
+%% \hspace{5em}
+%% \infer{\hat{\Gamma}\vdashh\hat{\elim}_\Delta\,\hat{m}\in\Delta^\E\,\hat{\con}_\Delta\,\hat{m}}{\hat{\Gamma}\vdashh \hat{m}\in \Delta^\M\,\hat{\con}_\Delta}
+%% \]
+%% We can add HIITs to the target theory by extending it with the theory
+%% of codes (making the target and the source theory the same) and adding
+%% the above rules with the additional assumption of
+%% $\hat{\Gamma}\vdash\Delta$. However, this only adds HIITs with weak
+%% computation rules. To make the computation rules definitional, we
+%% would probably need a two-level target type theory with an equality
+%% having an equality reflection rule as in Voevodsky's homotopy type
+%% system \cite{hts} or Andromeda \cite{andromeda}.
+
+%=====================================================================
+
+\section{Formalization and Implementation}
+\label{sec:formalization}
+
+There are three additional development artifacts to the current work:
+a Haskell implementation, a shallow Agda formalization and a deep Agda
+formalization. All three are available from the homepage of the first
+author.
+
+The Haskell implementation takes as input a file which contains a
+representation of a $\hat{\Gamma}\vdash\Delta$ specification of a
+HIIT. Then, it checks the input with respect to the rules in Figure
+\ref{sigrules}, and outputs an Agda-checkable file which contains the
+results of the $\blank^\C$, $\blank^\M$ and $\blank^\E$
+translations. It comes with examples, including the ones in this
+paper, indexed W-types \cite{morris09indexed}, the dense completion
+\cite[Appendix A.1.3]{forsberg-phd} and several HITs from
+\cite{HoTTbook} including the definition of Cauchy reals. It can be
+checked that our implementation computes the expected elimination
+principles in these cases.
+
+The shallow Agda formalization embeds both the source and target theories
+shallowly into Agda: it represents types as Agda types, functions as
+Agda functions, and so on. We also leave the $\blank^\C$ operation
+implicit. We state each case of the $\blank^\M$ and $\blank^\E$
+translations as Agda functions from all induction hypotheses to the
+result type of the translation, which lets us ``typecheck'' the
+translation. We have found that this style of formalization is
+conveniently light, but remains detailed enough to be useful. We also
+generated some of the code of the Haskell implementation from this
+formalization.
+
+The deep Agda formalization still uses a shallow embedding of the
+target type theory, but it uses a deep embedding of the source theory,
+in the style of \cite{ttintt}. In the formalization we merge the three
+translations into a single model construction. This allows us to prove
+strict preservation of definitional equalities in the substitution
+calculus of the source theory, in contrast to the shallow
+formalization, where we cannot reason about definitional equalities of
+Agda terms. Due to technical challenges, this formalization uses
+transport instead of $\J$ in the source theory, but this still covers
+a rather large class of HIIT definitions.
+
+%=====================================================================
+
+\section{Conclusions and further work}
+\label{sec:summary}
+
+Higher inductive-inductive types are useful in defining the well-typed
+syntax of type theory in an abstract way \cite{ttintt}. From a
+universal algebra point of view, they provide initial algebras for
+multi-sorted algebraic theories where the sorts can depend on each
+other. From the perspective of homotopy type theory, they provide
+synthetic versions of homotopy-theoretic constructions such as higher
+dimensional spheres or cell complexes. So far, no general scheme of
+HIITs have been proposed. To quote Lumsdaine and Shulman
+\cite{lumsdaineShulman}:
+\begin{quotation}
+``The constructors of an ordinary inductive type are unrelated to each
+other.  But in a higher inductive type each constructor must be able
+to refer to the previous ones; specifically, the source and target of
+a path-constructor generally involve the previous
+point-constructors. No syntactic scheme has yet been proposed for this
+that covers all cases of interest while remaining meaningful and
+consistent.''
+\end{quotation}
+In this paper we proposed such a syntactic scheme which also includes
+inductive-inductive types. We tackled the problem of complex
+dependencies on previous type formation rules and constructors by a
+well-known method of describing intricate dependencies: the syntax of
+type theory itself. We had to limit the type formers to only allow
+strictly positive definitions, but these restrictions are the only
+things that a type theorist has to learn to understand our codes. Our
+encoding is also \emph{direct} in the sense that the types of
+constructors and eliminators are exactly as required and not merely up
+to isomorphisms.
+
+In this paper we only \emph{specified} HIITs and characterized their
+induction principles. Proving their \emph{existence} is left as
+further work. This would likely involve reducing HIITs to basic
+building blocks such as W-types and quotient types.
+
+The theory of codes was defined as part of the syntax of another type
+theory, the target theory. An alternative way would be to define
+the theory of codes internally to a type theoretic metatheory in the
+style of \cite{ttintt}. However, as described in \cite[Section
+  6]{ttintt}, there would be a coherence problem: for the internal
+syntax to be useful, we need to truncate it to be a set (as the
+$\mathsf{trunc}$ constructor did for $\Int$ in the introduction). As
+the eliminator needs to respect the equality given by
+$\mathsf{trunc}$, we would only be able to eliminate from the internal
+type theory into a set. This would preclude the definition of even the
+$\blank^\C$ operation, which corresponds to the standard model. A
+possible solution to this problem would be to add all the higher
+coherence laws to the syntax (e.g.\ the pentagon law for the
+composition of substitutions) and then prove that the syntax is a
+set. For this however, we would probably need a two-level metatheory
+as in \cite{semisegal}.
+
+When working in a metatheory with uniqueness of identity proofs (which
+implies that all HIITs are in essence QIITs), the theory of codes
+admits a category model where the interpretation of a context is the
+category of algebras corresponding to the context. In the future, we
+would like to prove that these categories have initial algebras given
+by terms in the theory of codes.
+
+\bibliographystyle{plain}
+\bibliography{b}
+
+\appendix
+\section{The homomorphism interpretation}
+\label{sec:morphismrules}
+
+We list below the $\blank^\M$ interpretations for the syntax. For brevity, we
+mostly leave $\hat{\gamma_0}$ and $\hat{\gamma_1}$ implicit, notating only
+$\hat{\gamma^M}$ where necessary.  We also omit some subscript parameters from
+$\hat{\J}$ applications. The interpretation follows the same pattern as in the
+case of $\blank^\S$; the main difference is the interpretation of the universe
+as non-dependent function space. Additionally, the equality type former and
+reflexivity admit slightly nicer translations, since the lack of type dependency
+allows us to use composition and $\hat{\inv}$ (which was noted in Section
+\ref{sec:external}) instead of raw transports and path induction.
 
@@
 
-%% --------------------------------------------------------------------------------
-
-\section{Displayed Algebras}
-\label{sec:displayed}
-
-The $\blank^\D$ translation computes displayed algebras for a signature, which
-can be viewed as bundles of induction motives and methods. $\blank^\D$ is
-essentially an unary logical predicate translation over the $\blank^\A$
-model. It is related to the logical predicate translation of Bernardy et
-al.\ \cite{bernardy12parametricity}, but our implementation differs by
-interpreting contexts as $\Sigma$-types instead of extended contexts.
-
-For each
-context $\Delta$, $\Delta^\D$ is a predicate over the standard interpretation
-$\Delta^\A$. For a type $\Delta\vdash A$, $A^\D$ is a predicate over $A^\A$,
-which also depends on $\hat{\gamma}\in\Delta^\A$ and a witness of
-$\Delta^\D\,\hat{\gamma}$. All of these may refer to a target theory context
-$\hat{\Gamma}$.
-\[
-\infer{\hat{\Gamma}\vdashh\Delta^\D \in \Delta^\A\rah \Type_{\hat{i+1}}}{\hat{\Gamma}\vdash\Delta}
-\hspace{2em}
-\infer{
-  \hat{\Gamma}\vdashh A^\D \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah  A^\A\,\hat{\gamma}\rah \Type_{\hat{i+1}}}
-      {\hat{\Gamma}\semicol\Delta\vdash A}
-\]
-For a term $t$, $t^\D$ witnesses that the predicate corresponding to its type
-holds for $t^\A$; again this can be viewed as a fundamental theorem for the
-predicate interpretation.
-
-\[
-\infer{\hat{\Gamma}\vdashh t^\D \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}}{\hat{\Gamma}\semicol\Delta\vdash t : A}
-\]
-The implementation of $\blank^\D$ is given below. We leave
-$\hat{\gamma}$-s mostly implicit, marking only $\hat{\gamma^D}$ witnesses.
-\begingroup
-%% \allowdisplaybreaks
-\begin{alignat*}{5}
-  & \boldsymbol{\cdot}^D\,\hat{\gamma} && :\equiv \hat{\top} \\
-  & (\Delta,\,x:A)^D\,(\hat{\gamma}\commah\,\hat{t}) && :\equiv \hat{(}\hat{\gamma^D}\in\Delta^D\,\hat{\gamma}\hat{)}\timesh A^D\,\hat{\gamma^D}\,\hat{t} \\
-  & x^D\,\hat{\gamma^D} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^D} \\
-  & \U^D\,\hat{\gamma^D}\,\hat{A} && :\equiv \hat{A} \rah  \Type_{\hat{i}} \\
-  & (\underline{a})^D\,\hat{\gamma^D}\,\hat{t} && :\equiv a^\D\,\hat{\gamma^D}\,\hat{t} \\
-  & ((x:a)\ra B)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\hat{(}\hat{x^D}\in a^\D\,\hat{\gamma^D}\,\hat{x}\hat{)} \rah  B^\D\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)}\,\hat{(}\hat{\gamma^D}\commah\hat{x^D}\hat{)}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,u)^D\,\hat{\gamma^D} && :\equiv \hat{(}t^\D\,\hat{\gamma^D}\hat{)}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\D\,\hat{\gamma^D}\hat{)} \\
-  & ((\hat{x}\in \hat{A})\ra B)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u} \\
-  & (t=_a u)^D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}}\,\hat{e}\,\hat{(}t^\D\,\hat{\gamma^D}\hat{)} \eqh u^\D\hat{\gamma^D} \\
-  & (\refl_t)^D\,\hat{\gamma^D} && :\equiv \hat{\refl}_{\hat{(}t^\D\,\hat{\gamma^D}\hat{)}} \\
-  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^D\,\hat{\gamma^D} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^D}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^D}\hat{)} \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
-  & ((\hat{x}\in \hat{A})\ra b)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u}
-\end{alignat*}
-\endgroup
-The predicate for a context is given by iterating $\blank^\D$ for its
-constituent types. For a variable, the corresponding witness is looked up from
-$\hat{\gamma^D}$.
-
-The translation of the universe, given an element of
-$\hat{A}\in\U^\A\,\hat{\gamma}$ (with
-$\U^\A\,\hat{\gamma}\equiv\Type_{\hat{0}}$) returns the predicate space over
-$\hat{A}$. For $\underline{a}$ types, we just return the translation of $a$.
-
-The predicate for a function type for inductive parameters expresses
-preservation of predicates. Witnesses of application are given by recursive
-application of $\blank^\D$. The definitions for the other (non-inductive)
-function spaces are similar, except there is no predicate for the domain types,
-and thus no witnesses are required.
-
-The predicate for the equality type $t=_a u$, for each $\hat{e}\in(t=_a
-u)^\A\,\hat{\gamma}$, i.e.\ $\hat{e}\in t^\A\,\hat{\gamma}\eqh
-u^\A\,\hat{\gamma}$, says that $t^\D$ and $u^\D$ are equal. As these have
-different types, we have to transport over the original equality
-$\hat{e}$. Hence, induction methods for path constructors will be \emph{paths
-  over paths} in the sense of homotopy type theory.
-
-$\refl$ is interpreted with just reflexivity in the target syntax. The
-interpretation of $\J$ is given by a double $\hat{\J}$ application, borrowing
-the definition from Lasson \cite{lasson}. Here, we use a shortened $\hat{\J}$
-notation; see the formalization (Section \ref{sec:formalization}) for details.
-
-Again, let us consider the circle example:
-\begin{alignat*}{5}
-  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\D\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\\
-  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1D}}\in \hat{S^1}\rah\Type_{\hat{i}}\hat{)}\timesh\hat{(}\hat{b^D}\in \hat{S^{1D}\,b}\hat{)}\timesh\hat{(}\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D}\hat{)}
-\end{alignat*}
-The inputs of $\blank^\D$ here are the signature for the circle (the context in
-black) and an $\hat{S^{1}}$-algebra consisting of three non-$\top$ components.
-It returns a family over the type $\hat{S^1}$, an element of this family
-$\hat{b^D}$ at index $\hat{b}$, and and an equality between $\hat{b^D}$ and
-$\hat{b^D}$ which lies over $\hat{loop}$. This is the same as the usual
-induction motives and methods for the circle, e.g.\ as described in \cite{HoTTbook}.
-
-%% --------------------------------------------------------------------------------
-
-\section{Displayed Algebra Sections}
-\label{sec:sections}
-
-The operation $\blank^\S$ yields displayed algebra sections. It is a generalized
-binary logical relation translation where the type of the second parameter of
-the relation may depend on the first parameter. Displayed algebra sections can
-be viewed as dependent homomorphisms: while homomorphisms are
-structure-preserving families of functions, sections are structure-preserving
-families of dependent functions.
-
-Contexts are interpreted as dependent binary relations between algebras and
-displayed algebras. The universe level $i$ was previously chosen for the
-$\blank^\M$ and $\blank^\D$ translations.
-\[
-\infer{\hat{\Gamma}\semicol\Delta^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah \Type_{\hat{i}}}{\hat{\Gamma}\vdash\Delta}
-\]
-Types are interpreted as dependent binary relations which additionally depend on
-$\hat{\gamma}$, $\hat{\gamma^D}$, $\hat{\gamma^S}$ interpretations of the
-context.
-\[
-\infer{\hat{\Gamma}\vdashh A^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\hat{(}\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}\hat{(}\hat{x}\in A^\A\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{x}\rah \Type_{\hat{i}}}{\hat{\Gamma}\semicol\Delta\vdash A}
-\]
-For a term $t$, $t^\S$ witnesses that the relation corresponding to
-its type holds for $t^\A$ and $t^\D$.
-\[
-\infer{\hat{\Gamma}\vdashh t^\S \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\hat{(}\hat{\gamma^S}\in\Delta^\S\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}\rah  A^\S\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{\gamma^S}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}t^\D\,\hat{\gamma}\,\hat{\gamma^D}\hat{)}}{\hat{\Gamma}\semicol\Delta \vdash t : A}
-\]
-Here, we mostly notate $\hat{\gamma^S}$ parameters and applications, making
-$\hat{\gamma^D}$-s implicit in addition to $\hat{\gamma}$-s. The implementation
-is the following.
-\begingroup
-%% \allowdisplaybreaks
-\begin{alignat*}{5}
-  & \boldsymbol{\cdot}^\S\,\hat{\gamma}\,\hat{\gamma^D} && :\equiv \hat{\top} \\
-  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv \hat{(}\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2}\hat{)}\timesh A^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} \\
-  & x^\S\,\hat{\gamma^S} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^S} \\
-  & \U^\S\,\hat{\gamma^S}\,\hat{A}\,\hat{A^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  \hat{A^D}\,\hat{x} \\
-  & (\underline{a})^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} && :\equiv a^\S\,\hat{\gamma^S}\,\hat{t} \eqh \hat{t^D} \\
-  & ((x:a)\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\S\,\hat{(}\hat{\gamma,x}\hat{)}\,\hat{(}\hat{\gamma^D,}\,a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\,\hat{(}\hat{\gamma^S,}\,\hat{\refl}\hat{)} \\
-  & && \hspace{8.4em} \hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{\big(}\hat{f^D}\,\hat{x}\,\hat{(}a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\hat{\big)} \\
-  & (t\,u)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^S}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)} \\
-  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv t^\S\,\hat{\gamma^S}\,\hat{u} \\
-  & (t=_a u)^\S\,\hat{\gamma^S}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^S}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
-  & (\refl_t)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{\refl}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)} \\
-  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S\,\hat{\gamma^S} && :\equiv \\
-  & && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}.\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\\
-  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,(eq^\A\,\hat{\gamma})\hat{\Big)}\,(u^\S\,\hat{\gamma^S})\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^S}\hat{)} \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \\
-  & && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\\
-  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^S\,f\,t} && :\equiv b^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
-  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv \ap\,\hat{(}\hat{\lambda} \hat{f.}\hat{f}\,\hat{u}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}
-\end{alignat*}
-\endgroup
-
-The $\U^\S$ and $(\underline{a})^\S$ definitions are the key points of
-the $\blank^\S$ operation. The definitions for the other $\blank^\S$
-cases are largely determined by these.
-
-The $\U^\S$ rule yields the type of the eliminator for a type
-formation rule. In the natural numbers example above, the non-indexed
-$Nat : \U$ sort is interpreted as $\hat{\hat{n^S} \in (\hat{x}\in
-\hat{n})\rah \hat{n^D}\,\hat{x}}$. For indexed sorts, the indices are
-first processed by the $\blank^\S$ cases for inductive and non-inductive
-parameters, until the ultimate $\U$ return type is reached. Hence, the
-eliminator for a sort is always a function.
-
-Analogously, the $\blank^\S$ result type for a point or path constructor is
-always a $\beta$-rule, i.e.\ a function type ending in an equality. To
-see why, consider the $(\underline{a})^\S$ definition. It expresses
-that applications of $a^\S$ eliminators must be equal to the
-corresponding $\hat{t^D}$ induction methods. Hence, for path and point
-constructor types, $\blank^\S$ works by first processing all inductive and
-non-inductive indices, then finally returning an equality type.
-
-Let us also consider the $((x:a)\ra B)^\S$ case for inductive
-parameters. Here, we make crucial use of the fact that the domain type
-$a$ is small. This provides us $a^\S\,\hat{\gamma^3\,x}$, which we use
-to witness the $a^\D\,\hat{\gamma^2}\,\hat{x}$ hypothesis for
-$B^\S$. Without the size restriction on inductive parameters (which
-enforces strict positivity), the $\blank^\S$ operation would not be
-possible at all, because $a^\S$ for a large $a$ type would be merely
-an opaque relation instead of an eliminator function.
-
-Here, we only provide abbreviated definitions for the $t\,u$, $t=_a
-u$, $\refl$, $\J$ and $\J\beta$ cases. In the $\J$ case, we write
-$\hat{\mathsf{uncurry}}\,p^\D$ for
-$\hat{\lambda\,\gamma\,\gamma^D\,x\,x^D\,z\,z^D.}\,p^\D\,(\hat{\gamma,x,z})\,(\hat{\gamma^D,x^D,z^D})$
-and analogously elsewhere. The full definitions can be found in the
-Agda formalization. The definitions are highly constrained by the
-required types, and not particularly difficult to implement with the
-help of a proof assistant; they all involve doing successive path
-induction on all equalities available from induction hypotheses, with
-appropriately generalized induction motives.
-
-The full $(\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S$ definition is quite
-large, and, for instance, yields a very large $\beta$-rule for the
-higher inductive torus definition (the reader can confirm this using
-the Haskell implementation). Nevertheless, an implementation focused
-on practicality may provide smaller specialized $\blank^\S$ definitions for
-commonly used equality operations such as composition or inverses.
-
-The circle example is a bit more interesting here:
-\begin{alignat*}{5}
-  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\S\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\,\hat{(}\hat{\tt,\,S^{1M},\,b^D,\,loop^D}\hat{)} \\
-  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1E}}\in \hat{(}\hat{x}\in\hat{S^1}\hat{)}\hat{\,\ra\,}\hat{S^{1M}\,x}\hat{)}\timesh\hat{(}\hat{b^S}\in \hat{S^{1E}\,\,b = b^D}\hat{)}\\
-  & && \hspace{0.75em}\timesh\hat{(}\hat{loop^S\,\,\in\,\,} \TR\,_{\hat{(}\hat{\lambda x.
-\TR\,_{S^{1M}}\,loop\,x\,=\,b^D}\hat{)}}\,\hat{b^S}\,\hat{(}\TR\,_{\hat{(}\hat{\lambda} \hat{x}.
-\TR\,_{\hat{S^{1M}}}\,\hat{loop}\,\hat{(}\hat{S^{1E}\,b}\hat{)}\,\eqh\,\hat{x}\hat{)}}\,\hat{b^S}\,\hat{(}\apd\,\hat{S^{1E}\,loop}\hat{)}\hat{)}\,\\
-  & && \hspace{5.5em} \hat{=\,loop^D}\hat{)}
-\end{alignat*}
-In homotopy type theory, the $\beta$-rule for $loop$ is usually just
-$\apd\,\hat{S^{1E}\,loop}\,\hat{\,=\,loop^D}$, but here all
-$\beta$-rules are propositional, so we need to transport with
-$\hat{b^S}$ to make the equation well-typed. When computing the type
-of $\hat{loop^S}$, we start with $(\underline{b
-  =b})^S\,\,\hat{\gamma^3}\,\,\hat{loop}\,\,\hat{loop^D}$. Next, this
-evaluates to $(b=b)^\S\,\hat{\gamma^3}\,\hat{loop}\eqh\hat{loop^D}$,
-and then we unfold the left hand side to get the doubly-transported
-expression in the result.
-
-In Appendix \ref{sec:app}, we show how the elimination principle is
-computed for the two-dimensional sphere.
-
-For another example for the translations, we consider indexed W-types
-which can describe a large class of inductive definitions
-\cite{morris09indexed}. Suppose we are in the external context $\hat{I}
+\section{Displayed algebra sections for the two-dimensional sphere}
+\label{sec:app}
+
+The two-dimensional sphere is given by the following signature:
+
+\[
+\Gamma :\equiv \Big(\boldsymbol{\cdot},\,\,\,S^2:\U,\,\,\,b:\underline{S^2},\,\,\,surf:\underline{\refl_{b} =_{(b=_{S^2} b)} \refl_{b}}\Big)
+\]
+The sphere-algebras are computed as follows.
+\begin{alignat*}{5}
+  & && \Gamma^\C \equiv \hat{\top}\timesh\hat{(}\hat{S^2}\in\Type_{\hat{0}}\hat{)}\timesh\hat{(}\hat{b}\in \hat{S^2}\hat{)}\timesh\hat{(}\hat{surf\in \hat{\refl_{\hat{b}}}=_{\hat{(}\hat{b}=_{\hat{S^2}}\hat{b}\hat{)}}\hat{\refl_{\hat{b}}}}\hat{)}
+\end{alignat*}
+Given a sphere-algebra and fixing a universe level $i$, the motives
+and methods are computed as follows.
+\begin{alignat*}{5}
+  & && \Gamma^\M\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2M}}\in \hat{S^2}\rah\Type_{\hat{i}}\hat{)} \\
+  & && \hspace{0.75em} \timesh\hat{(}\hat{b^M}\in \hat{S^{2M}\,b}\hat{)} \\
+  & && \hspace{0.75em} \timesh\hat{(}\hat{surf^M\in \TR_{\hat{(}\TR_{S^{2M}}\,\blank\,b^M=b^M\hat{)}}\,surf\,\refl_{b^M} = \refl_{b^M}}\hat{)}
+\end{alignat*}
+Given a sphere-algebra and a displayed algebra over it, we get the type of sections:
+\begin{alignat*}{5}
+  & && \Gamma^\E\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\,\hat{(}\hat{\tt,\,S^{2M},\,b^M,\,surf^M}\hat{)} \\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2E}}\in \hat{(}\hat{x}\in\hat{S^2}\hat{)}\hat{\,\ra\,}\hat{S^{2M}\,x}\hat{)} \\
+  & && \hspace{0.75em} \timesh\hat{(}\hat{b^E}\in \hat{S^{2E}\,\,b = b^M}\hat{)}\\
+  & && \hspace{0.75em} \timesh\hat{\bigg(}\hat{surf^E\,\,\in\,\,} \TR\,\hat{\Big(}\hat{\J\,\refl\,b^E}\hat{\Big)}\,\hat{\Big(}\TR\,\hat{\big(}\hat{\J\,\refl\,b^E}\hat{\big)}\,\hat{\big(}\hat{\apd}\,\hat{(}\hat{\lambda} \hat{x}.\TR\,\hat{b^E}\,\hat{(}\TR\,\hat{b^E}\,\hat{(}\hat{\apd}\,\hat{S^{2E}}\,\hat{x}\hat{)}\hat{)}\hat{)}\,\hat{surf}\hat{\big)}\hat{\Big)} \\
+  & && \hspace{5.5em} \hat{=} \,\hat{surf^M}\hat{\bigg)}
+\end{alignat*}
+Note that if $\hat{b^E}$ is a definitional equality (that is, we have
+$\hat{S^{2E}\,\,b} \equiv \hat{b^M}$), the occurrences of $\hat{b^E}$
+in the type of $\hat{surf^E}$ can be replaced by $\hat{\refl}$. In
+this case the type of $\hat{surf^E}$ becomes the expected
+$\hat{\apd}\,\hat{(}\hat{\apd}\,\hat{S^{2E}}\hat{)}\,\hat{surf}\,\hat{=}\,\hat{surf^M}$.
+
+\section{Displayed algebra sections for indexed W-types}
+\label{sec:wtypes}
+
+Indexed W-types can describe a large class of inductive definitions
+\cite{morris09indexed}. Suppose we have the external context $\hat{I}
 \in \Type_{\hat{0}}\commah\hat{S} \in \Type_{\hat{0}}\commah\hat{P} \in \hat{S} \rah
@@
 
-%=====================================================================
-
-\section{On Extension to Categorical Semantics}
-\label{sec:categorical}
-
-TODO
-
-%% \section{Existence of HIITs}
-%% \label{sec:adding}
-
-%% The target type theory supports HIITs if whenever we can derive
-%% $\hat{\Gamma}\vdash\Delta$ in the source theory, the following rules are
-%% admissible.
-%% \[
-%% \infer{\hat{\Gamma}\vdashh\hat{\con}_\Delta\in \Delta^\C}{\vspace{1em}}
-%% \hspace{5em}
-%% \infer{\hat{\Gamma}\vdashh\hat{\elim}_\Delta\,\hat{m}\in\Delta^\E\,\hat{\con}_\Delta\,\hat{m}}{\hat{\Gamma}\vdashh \hat{m}\in \Delta^\M\,\hat{\con}_\Delta}
-%% \]
-%% We can add HIITs to the target theory by extending it with the theory
-%% of codes (making the target and the source theory the same) and adding
-%% the above rules with the additional assumption of
-%% $\hat{\Gamma}\vdash\Delta$. However, this only adds HIITs with weak
-%% computation rules. To make the computation rules definitional, we
-%% would probably need a two-level target type theory with an equality
-%% having an equality reflection rule as in Voevodsky's homotopy type
-%% system \cite{hts} or Andromeda \cite{andromeda}.
-
-%=====================================================================
-
-\section{Formalization and Implementation}
-\label{sec:formalization}
-
-There are three additional development artifacts to the current work:
-a Haskell implementation, a shallow Agda formalization and a deep Agda
-formalization. All three are available from the homepage of the first
-author.
-
-The Haskell implementation takes as input a file which contains a
-representation of a $\hat{\Gamma}\vdash\Delta$ specification of a
-HIIT. Then, it checks the input with respect to the rules in Figure
-\ref{sigrules}, and outputs an Agda-checkable file which contains the
-results of the $\blank^\C$, $\blank^\M$ and $\blank^\E$
-translations. It comes with examples, including the ones in this
-paper, indexed W-types \cite{morris09indexed}, the dense completion
-\cite[Appendix A.1.3]{forsberg-phd} and several HITs from
-\cite{HoTTbook} including the definition of Cauchy reals. It can be
-checked that our implementation computes the expected elimination
-principles in these cases.
-
-The shallow Agda formalization embeds both the source and target theories
-shallowly into Agda: it represents types as Agda types, functions as
-Agda functions, and so on. We also leave the $\blank^\C$ operation
-implicit. We state each case of the $\blank^\M$ and $\blank^\E$
-translations as Agda functions from all induction hypotheses to the
-result type of the translation, which lets us ``typecheck'' the
-translation. We have found that this style of formalization is
-conveniently light, but remains detailed enough to be useful. We also
-generated some of the code of the Haskell implementation from this
-formalization.
-
-The deep Agda formalization still uses a shallow embedding of the
-target type theory, but it uses a deep embedding of the source theory,
-in the style of \cite{ttintt}. In the formalization we merge the three
-translations into a single model construction. This allows us to prove
-strict preservation of definitional equalities in the substitution
-calculus of the source theory, in contrast to the shallow
-formalization, where we cannot reason about definitional equalities of
-Agda terms. Due to technical challenges, this formalization uses
-transport instead of $\J$ in the source theory, but this still covers
-a rather large class of HIIT definitions.
-
-%=====================================================================
-
-\section{Conclusions and further work}
-\label{sec:summary}
-
-Higher inductive-inductive types are useful in defining the well-typed
-syntax of type theory in an abstract way \cite{ttintt}. From a
-universal algebra point of view, they provide initial algebras for
-multi-sorted algebraic theories where the sorts can depend on each
-other. From the perspective of homotopy type theory, they provide
-synthetic versions of homotopy-theoretic constructions such as higher
-dimensional spheres or cell complexes. So far, no general scheme of
-HIITs have been proposed. To quote Lumsdaine and Shulman
-\cite{lumsdaineShulman}:
-\begin{quotation}
-``The constructors of an ordinary inductive type are unrelated to each
-other.  But in a higher inductive type each constructor must be able
-to refer to the previous ones; specifically, the source and target of
-a path-constructor generally involve the previous
-point-constructors. No syntactic scheme has yet been proposed for this
-that covers all cases of interest while remaining meaningful and
-consistent.''
-\end{quotation}
-In this paper we proposed such a syntactic scheme which also includes
-inductive-inductive types. We tackled the problem of complex
-dependencies on previous type formation rules and constructors by a
-well-known method of describing intricate dependencies: the syntax of
-type theory itself. We had to limit the type formers to only allow
-strictly positive definitions, but these restrictions are the only
-things that a type theorist has to learn to understand our codes. Our
-encoding is also \emph{direct} in the sense that the types of
-constructors and eliminators are exactly as required and not merely up
-to isomorphisms.
-
-In this paper we only \emph{specified} HIITs and characterized their
-induction principles. Proving their \emph{existence} is left as
-further work. This would likely involve reducing HIITs to basic
-building blocks such as W-types and quotient types.
-
-The theory of codes was defined as part of the syntax of another type
-theory, the target theory. An alternative way would be to define
-the theory of codes internally to a type theoretic metatheory in the
-style of \cite{ttintt}. However, as described in \cite[Section
-  6]{ttintt}, there would be a coherence problem: for the internal
-syntax to be useful, we need to truncate it to be a set (as the
-$\mathsf{trunc}$ constructor did for $\Int$ in the introduction). As
-the eliminator needs to respect the equality given by
-$\mathsf{trunc}$, we would only be able to eliminate from the internal
-type theory into a set. This would preclude the definition of even the
-$\blank^\C$ operation, which corresponds to the standard model. A
-possible solution to this problem would be to add all the higher
-coherence laws to the syntax (e.g.\ the pentagon law for the
-composition of substitutions) and then prove that the syntax is a
-set. For this however, we would probably need a two-level metatheory
-as in \cite{semisegal}.
-
-When working in a metatheory with uniqueness of identity proofs (which
-implies that all HIITs are in essence QIITs), the theory of codes
-admits a category model where the interpretation of a context is the
-category of algebras corresponding to the context. In the future, we
-would like to prove that these categories have initial algebras given
-by terms in the theory of codes.
-
-\bibliographystyle{plain}
-\bibliography{b}
-
-\appendix
-\section{Elimination principle computed for the two-dimensional sphere}
-\label{sec:app}
-
-The two-dimensional sphere is given by the following context in the
-theory of codes.
-\[
-\Gamma :\equiv \Big(\boldsymbol{\cdot},\,\,\,S^2:\U,\,\,\,b:\underline{S^2},\,\,\,surf:\underline{\refl_{b} =_{(b=_{S^2} b)} \refl_{b}}\Big)
-\]
-The sphere-algebras are computed as follows.
-\begin{alignat*}{5}
-  & && \Gamma^\C \equiv \hat{\top}\timesh\hat{(}\hat{S^2}\in\Type_{\hat{0}}\hat{)}\timesh\hat{(}\hat{b}\in \hat{S^2}\hat{)}\timesh\hat{(}\hat{surf\in \hat{\refl_{\hat{b}}}=_{\hat{(}\hat{b}=_{\hat{S^2}}\hat{b}\hat{)}}\hat{\refl_{\hat{b}}}}\hat{)}
-\end{alignat*}
-Given a sphere-algebra and fixing a universe level $i$, the motives
-and methods are computed as follows.
-\begin{alignat*}{5}
-  & && \Gamma^\M\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\\
-  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2M}}\in \hat{S^2}\rah\Type_i\hat{)} \\
-  & && \hspace{0.75em} \timesh\hat{(}\hat{b^M}\in \hat{S^{2M}\,b}\hat{)} \\
-  & && \hspace{0.75em} \timesh\hat{(}\hat{surf^M\in \TR_{\hat{(}\TR_{S^{2M}}\,\blank\,b^M=b^M\hat{)}}\,surf\,\refl_{b^M} = \refl_{b^M}}\hat{)}
-\end{alignat*}
-Given a sphere-algebra and a displayed algebra over it, we get the type of sections:
-\begin{alignat*}{5}
-  & && \Gamma^\E\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\,\hat{(}\hat{\tt,\,S^{2M},\,b^M,\,surf^M}\hat{)} \\
-  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2E}}\in \hat{(}\hat{x}\in\hat{S^2}\hat{)}\hat{\,\ra\,}\hat{S^{2M}\,x}\hat{)} \\
-  & && \hspace{0.75em} \timesh\hat{(}\hat{b^E}\in \hat{S^{2E}\,\,b = b^M}\hat{)}\\
-  & && \hspace{0.75em} \timesh\hat{\bigg(}\hat{surf^E\,\,\in\,\,} \TR\,\hat{\Big(}\hat{\J\,\refl\,b^E}\hat{\Big)}\,\hat{\Big(}\TR\,\hat{\big(}\hat{\J\,\refl\,b^E}\hat{\big)}\,\hat{\big(}\hat{\apd}\,\hat{(}\hat{\lambda} \hat{x}.\TR\,\hat{b^E}\,\hat{(}\TR\,\hat{b^E}\,\hat{(}\hat{\apd}\,\hat{S^{2E}}\,\hat{x}\hat{)}\hat{)}\hat{)}\,\hat{surf}\hat{\big)}\hat{\Big)} \\
-  & && \hspace{5.5em} \hat{=} \,\hat{surf^M}\hat{\bigg)}
-\end{alignat*}
-Note that if $\hat{b^E}$ is a definitional equality (that is, we have
-$\hat{S^{2E}\,\,b} \equiv \hat{b^M}$), the occurrences of $\hat{b^E}$
-in the type of $\hat{surf^E}$ can be replaced by $\hat{\refl}$. In
-this case the type of $\hat{surf^E}$ becomes the expected
-$\hat{\apd}\,\hat{(}\hat{\apd}\,\hat{S^{2E}}\hat{)}\,\hat{surf}\,\hat{=}\,\hat{surf^M}$.
-
-%=====================================================================
-
 \end{document}

### 2019-01-31 paper.tex
Stuff

@@
 \newcommand{\inv}{\mathsf{inv}}
+\newcommand{\ID}{\mathsf{\ID}}
 
@@
 does not seem to be a sensible notion. This situation is similar to how
-categories in homotopy type theory need to have set-truncated objects
+categories in homotopy type theory need to have set-truncated morphisms
 \cite{ahrens2015univalent}.
@@
 algebras, homomorphisms, displayed algebras and their sections. The presentation
-here is informal and focuses on readability, similarly to our presentation of
-signatures before.
+here, like in Section \ref{sec:tos}, is informal and focuses on readability.
+In particular, we omit interpretations for substitutions and preservation
+proofs for definitional equalities.
 
@@
   & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
-  & ((\hat{x}\in \hat{A})\ra b)^\A\,\hat{\gamma} && :\equiv (\hat{x}\in \hat{A}\hat{)}\rah  b^\A\,\hat{\gamma} \\
+  & ((\hat{x}\in \hat{A})\ra b)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^\A\,\hat{\gamma} \\
   & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u}
@@
 %---------------------------------------------------------------------
+
+\section{Displayed Algebras}
+\label{sec:displayed}
+
+The $\blank^\D$ translation computes displayed algebras for a signature, which
+can be viewed as bundles of induction motives and methods. $\blank^\D$ is an
+unary logical predicate translation over the $\blank^\A$ translation. It is
+related to the logical predicate translation of Bernardy et
+al.\ \cite{bernardy12parametricity}, but our implementation differs by
+interpreting contexts as $\Sigma$-types instead of extended contexts.
+
+We fix a universe level $\hat{i}$ for the translation. For each context
+$\Delta$, $\Delta^\D$ is a predicate over the standard interpretation
+$\Delta^\A$. For a type $\Delta\vdash A$, $A^\D$ is a predicate over $A^\A$,
+which also depends on $\hat{\gamma}\in\Delta^\A$ and a witness of
+$\Delta^\D\,\hat{\gamma}$. All of these may refer to a target theory context
+$\hat{\Gamma}$.
+\[
+\infer{\hat{\Gamma}\vdashh\Delta^\D \in \Delta^\A\rah \Type_{\hat{i+1}}}{\hat{\Gamma}\vdash\Delta}
+\hspace{2em}
+\infer{
+  \hat{\Gamma}\vdashh A^\D \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah  A^\A\,\hat{\gamma}\rah \Type_{\hat{i+1}}}
+      {\hat{\Gamma}\semicol\Delta\vdash A}
+\]
+For a term $t$, $t^\D$ witnesses that the predicate corresponding to its type
+holds for $t^\A$; this can be viewed as a \emph{fundamental theorem} for the
+predicate interpretation.
+
+\[
+\infer{\hat{\Gamma}\vdashh t^\D \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}}{\hat{\Gamma}\semicol\Delta\vdash t : A}
+\]
+The implementation of $\blank^\D$ is given below. We leave
+$\hat{\gamma}$-s mostly implicit, marking only $\hat{\gamma^D}$ witnesses.
+\begingroup
+%% \allowdisplaybreaks
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot}^\D\,\hat{\gamma} && :\equiv \hat{\top} \\
+  & (\Delta,\,x:A)^\D\,(\hat{\gamma}\commah\,\hat{t}) && :\equiv \hat{(}\hat{\gamma^D}\in\Delta^D\,\hat{\gamma}\hat{)}\timesh A^D\,\hat{\gamma^D}\,\hat{t} \\
+  & x^\D\,\hat{\gamma^D} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^D} \\
+  & \U^\D\,\hat{\gamma^D}\,\hat{A} && :\equiv \hat{A} \rah  \Type_{\hat{i}} \\
+  & (\underline{a})^\D\,\hat{\gamma^D}\,\hat{t} && :\equiv a^\D\,\hat{\gamma^D}\,\hat{t} \\
+  & ((x:a)\ra B)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\hat{(}\hat{x^D}\in a^\D\,\hat{\gamma^D}\,\hat{x}\hat{)} \rah  B^\D\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)}\,\hat{(}\hat{\gamma^D}\commah\hat{x^D}\hat{)}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,u)^\D\,\hat{\gamma^D} && :\equiv \hat{(}t^\D\,\hat{\gamma^D}\hat{)}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\D\,\hat{\gamma^D}\hat{)} \\
+  & ((\hat{x}\in \hat{A})\ra B)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^\D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u} \\
+  & (t=_a u)^\D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}}\,\hat{e}\,\hat{(}t^\D\,\hat{\gamma^D}\hat{)} \eqh u^\D\hat{\gamma^D} \\
+  & (\refl_t)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl}_{\hat{(}t^\D\,\hat{\gamma^D}\hat{)}} \\
+  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\D\,\hat{\gamma^D} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^D}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^D}\hat{)} \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
+  & ((\hat{x}\in \hat{A})\ra b)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^\D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u}
+\end{alignat*}
+\endgroup
+The predicate for a context is given by iterating $\blank^\D$ for its
+constituent types. For a variable, the corresponding witness is looked up from
+$\hat{\gamma^D}$.
+
+The translation of the universe, given an element of
+$\hat{A}\in\U^\A\,\hat{\gamma}$ (with
+$\U^\A\,\hat{\gamma}\equiv\Type_{\hat{0}}$) returns the predicate space over
+$\hat{A}$. For $\underline{a}$ types, we just return the translation of $a$.
+
+The predicate for a function type for inductive parameters expresses
+preservation of predicates. Witnesses of application are given by recursive
+application of $\blank^\D$. The definitions for the other (non-inductive)
+function spaces are similar, except there is no predicate for the domain types,
+and thus no witnesses are required.
+
+The translation for the equality type $t=_a u$, for each $\hat{e}\in(t=_a
+u)^\A\,\hat{\gamma}$, i.e.\ $\hat{e}\in t^\A\,\hat{\gamma}\eqh
+u^\A\,\hat{\gamma}$, witnesses that $t^\D$ and $u^\D$ are equal. As these have
+different types, we have to transport over the original equality
+$\hat{e}$. Hence, induction methods for path constructors will be \emph{paths
+  over paths} in the sense of homotopy type theory.
+
+$\refl$ is interpreted with just reflexivity in the target syntax. The
+interpretation of $\J$ is given by a double $\hat{\J}$ application, borrowing
+the definition from Lasson \cite{lasson}. Here, we use a shortened $\hat{\J}$
+notation; see the formalization (Section \ref{sec:formalization}) for details.
+
+Again, let us consider the circle example:
+\begin{alignat*}{5}
+  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\D\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1D}}\in \hat{S^1}\rah\Type_{\hat{i}}\hat{)}\timesh\hat{(}\hat{b^D}\in \hat{S^{1D}\,b}\hat{)}\timesh\hat{(}\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D}\hat{)}
+\end{alignat*}
+The inputs of $\blank^\D$ here are the signature for the circle (the context in
+black) and an $\hat{S^{1}}$-algebra consisting of three non-$\top$ components.
+It returns a family over the type $\hat{S^1}$, an element of this family
+$\hat{b^D}$ at index $\hat{b}$, and and an equality between $\hat{b^D}$ and
+$\hat{b^D}$ which lies over $\hat{loop}$. This is the same as the usual
+induction motives and methods for the circle, e.g.\ as described in \cite{HoTTbook}.
+
+%% --------------------------------------------------------------------------------
 
@@
 homomorphisms of algebras. We do so by first considering a logical relation
-interpretation, and then refining it towards computing homomorphisms. For
-dependent type theories, logical relation models are well-known (see
-e.g.\ \cite{atkey}), so they are certainly applicable to our restricted syntax
-as well.
+interpretation, and then refining it towards homomorphisms. For dependent type
+theories, logical relation models are well-known (see e.g.\ \cite{atkey}), so
+they are certainly applicable to our restricted syntax as well.
 
 Below, we list induction motives for $\blank^\M$; these remain the same as we
-move from logical relations to homomorphisms. We fix a universe level $\hat{i}$ for
-the translations.
+move from logical relations to homomorphisms. The universe level $\hat{i}$
+was chosen previously for the $\blank^\D$ translation.
 \[
@@
 \infer{
-  \hat{\Gamma}\vdashh A^\D \hat{ : (\gamma_0\,\gamma_1 : \blc{\Delta^\A})(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\ra \blc{A^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,(\blc{t^\A}\,\gamma_0)\,(\blc{t^\A}\,\gamma_1)
+  \hat{\Gamma}\vdashh t^\M\, \hat{ : (\gamma_0\,\gamma_1 : \blc{\Delta^\A})(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\ra \blc{A^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,(\blc{t^\A}\,\gamma_0)\,(\blc{t^\A}\,\gamma_1)
       }}
@@
 
-Contexts are mapped to (proof-relevant) relations and types to families of relations
-depending on interpreted contexts. For terms, we get something which is often called
-a \emph{fundamental theorem} of a logical relation: every term is related to itself
-in related semantic contexts.
+Contexts are mapped to (proof-relevant) relations and types to families of
+relations depending on interpreted contexts. For terms, we again get a
+fundamental theorem: every term has related standard interpretations in related
+semantic contexts.
 
@@
 
-We interpret the universe as relation space, and function types as types
-expressing pointwise relatedness of functions. This interpretation works the
-same way for unrestricted (non-strictly positive) function types as well.
+We interpret the universe as relation space, and function types as relations
+expressing pointwise relatedness of functions. This interpretation would work
+the same way for unrestricted (non-strictly positive) function types as well.
 
 However, we would like to have underlying functions instead of relations in
-homomorphisms. We take hint from the fact that for classical (simply-typed)
-algebraic theories, a logical relation is equivalent to a homomorphism if and
-only if the underlying relation is the graph of a function
+homomorphisms. We take hint from the fact that for classical (simply-typed and
+single-sorted) algebraic theories, a logical relation is equivalent to a
+homomorphism if and only if the underlying relation is the graph of a function
 \cite[pg. 5]{udayReynolds}. Thus we make the following change:
@@
 This requires us to change $(\underline{a})^\M$ as well, since we need to
-produce a type as result, but $a^M$ now yields a function. We can morally keep
-the old definition, by viewing the result of $a^M$ as a functional relation, and
-relating $t_0$ and $t_1$ by its graph:
+produce a type as result, but $a^M$ now yields a function. We can view the
+result of $a^\M$ as a functional relation, and relate $t_0$ and $t_1$ by its
+graph:
 \begin{alignat*}{5}
@@
 
-$\blank^\M$ is very similar to the translation to displayed algebra sections,
-which is discussed in Section \ref{sec:sections}. In essence, $\blank^\M$ is a
-non-dependently typed version of the section translation. We will discuss in
-more detail the interpretations of equalities and the other function types in
-Section \ref{sec:sections}, and we leave the rest of the homomorphism
-translation to Appendix \ref{sec:morphismrules}.
-
-%% --------------------------------------------------------------------------------
-
-\section{Displayed Algebras}
-\label{sec:displayed}
-
-The $\blank^\D$ translation computes displayed algebras for a signature, which
-can be viewed as bundles of induction motives and methods. $\blank^\D$ is
-essentially an unary logical predicate translation over the $\blank^\A$
-model. It is related to the logical predicate translation of Bernardy et
-al.\ \cite{bernardy12parametricity}, but our implementation differs by
-interpreting contexts as $\Sigma$-types instead of extended contexts.
-
-For each
-context $\Delta$, $\Delta^\D$ is a predicate over the standard interpretation
-$\Delta^\A$. For a type $\Delta\vdash A$, $A^\D$ is a predicate over $A^\A$,
-which also depends on $\hat{\gamma}\in\Delta^\A$ and a witness of
-$\Delta^\D\,\hat{\gamma}$. All of these may refer to a target theory context
-$\hat{\Gamma}$.
-\[
-\infer{\hat{\Gamma}\vdashh\Delta^\D \in \Delta^\A\rah \Type_{\hat{i+1}}}{\hat{\Gamma}\vdash\Delta}
-\hspace{2em}
-\infer{
-  \hat{\Gamma}\vdashh A^\D \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\rah \Delta^\D\,\hat{\gamma}\rah  A^\A\,\hat{\gamma}\rah \Type_{\hat{i+1}}}
-      {\hat{\Gamma}\semicol\Delta\vdash A}
-\]
-For a term $t$, $t^\D$ witnesses that the predicate corresponding to its type
-holds for $t^\A$; again this can be viewed as a fundamental theorem for the
-predicate interpretation.
-
-\[
-\infer{\hat{\Gamma}\vdashh t^\D \in \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\hat{(}\hat{\gamma^D}\in\Delta^\D\,\hat{\gamma}\hat{)}\rah  A^\D\,\hat{\gamma}\,\hat{\gamma^D}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}}{\hat{\Gamma}\semicol\Delta\vdash t : A}
-\]
-The implementation of $\blank^\D$ is given below. We leave
-$\hat{\gamma}$-s mostly implicit, marking only $\hat{\gamma^D}$ witnesses.
-\begingroup
-%% \allowdisplaybreaks
-\begin{alignat*}{5}
-  & \boldsymbol{\cdot}^D\,\hat{\gamma} && :\equiv \hat{\top} \\
-  & (\Delta,\,x:A)^D\,(\hat{\gamma}\commah\,\hat{t}) && :\equiv \hat{(}\hat{\gamma^D}\in\Delta^D\,\hat{\gamma}\hat{)}\timesh A^D\,\hat{\gamma^D}\,\hat{t} \\
-  & x^D\,\hat{\gamma^D} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^D} \\
-  & \U^D\,\hat{\gamma^D}\,\hat{A} && :\equiv \hat{A} \rah  \Type_{\hat{i}} \\
-  & (\underline{a})^D\,\hat{\gamma^D}\,\hat{t} && :\equiv a^\D\,\hat{\gamma^D}\,\hat{t} \\
-  & ((x:a)\ra B)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\hat{(}\hat{x^D}\in a^\D\,\hat{\gamma^D}\,\hat{x}\hat{)} \rah  B^\D\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)}\,\hat{(}\hat{\gamma^D}\commah\hat{x^D}\hat{)}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,u)^D\,\hat{\gamma^D} && :\equiv \hat{(}t^\D\,\hat{\gamma^D}\hat{)}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\D\,\hat{\gamma^D}\hat{)} \\
-  & ((\hat{x}\in \hat{A})\ra B)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u} \\
-  & (t=_a u)^D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}}\,\hat{e}\,\hat{(}t^\D\,\hat{\gamma^D}\hat{)} \eqh u^\D\hat{\gamma^D} \\
-  & (\refl_t)^D\,\hat{\gamma^D} && :\equiv \hat{\refl}_{\hat{(}t^\D\,\hat{\gamma^D}\hat{)}} \\
-  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^D\,\hat{\gamma^D} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^D}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^D}\hat{)} \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
-  & ((\hat{x}\in \hat{A})\ra b)^D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u}
-\end{alignat*}
-\endgroup
-The predicate for a context is given by iterating $\blank^\D$ for its
-constituent types. For a variable, the corresponding witness is looked up from
-$\hat{\gamma^D}$.
-
-The translation of the universe, given an element of
-$\hat{A}\in\U^\A\,\hat{\gamma}$ (with
-$\U^\A\,\hat{\gamma}\equiv\Type_{\hat{0}}$) returns the predicate space over
-$\hat{A}$. For $\underline{a}$ types, we just return the translation of $a$.
-
-The predicate for a function type for inductive parameters expresses
-preservation of predicates. Witnesses of application are given by recursive
-application of $\blank^\D$. The definitions for the other (non-inductive)
-function spaces are similar, except there is no predicate for the domain types,
-and thus no witnesses are required.
-
-The predicate for the equality type $t=_a u$, for each $\hat{e}\in(t=_a
-u)^\A\,\hat{\gamma}$, i.e.\ $\hat{e}\in t^\A\,\hat{\gamma}\eqh
-u^\A\,\hat{\gamma}$, says that $t^\D$ and $u^\D$ are equal. As these have
-different types, we have to transport over the original equality
-$\hat{e}$. Hence, induction methods for path constructors will be \emph{paths
-  over paths} in the sense of homotopy type theory.
-
-$\refl$ is interpreted with just reflexivity in the target syntax. The
-interpretation of $\J$ is given by a double $\hat{\J}$ application, borrowing
-the definition from Lasson \cite{lasson}. Here, we use a shortened $\hat{\J}$
-notation; see the formalization (Section \ref{sec:formalization}) for details.
-
-Again, let us consider the circle example:
-\begin{alignat*}{5}
-  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\D\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\\
-  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1D}}\in \hat{S^1}\rah\Type_{\hat{i}}\hat{)}\timesh\hat{(}\hat{b^D}\in \hat{S^{1D}\,b}\hat{)}\timesh\hat{(}\hat{loop^D\in \TR_{\,S^{1D}}\,loop\,b^D = b^D}\hat{)}
-\end{alignat*}
-The inputs of $\blank^\D$ here are the signature for the circle (the context in
-black) and an $\hat{S^{1}}$-algebra consisting of three non-$\top$ components.
-It returns a family over the type $\hat{S^1}$, an element of this family
-$\hat{b^D}$ at index $\hat{b}$, and and an equality between $\hat{b^D}$ and
-$\hat{b^D}$ which lies over $\hat{loop}$. This is the same as the usual
-induction motives and methods for the circle, e.g.\ as described in \cite{HoTTbook}.
+$\blank^\M$ is similar to the translation to displayed algebra sections, which
+is discussed in the next Section. We will discuss in more detail the
+interpretations of equalities and the other (external) function types there. A
+full listing for the homomorphism translation can be found in Appendix
+\ref{sec:morphismrules}.
 
@@
 \label{sec:sections}
-
-TODO revision
 
@@
 
-Contexts are interpreted as dependent relations between algebras and
-displayed algebras. The universe level $i$ was previously chosen for the
-$\blank^\M$ and $\blank^\D$ translations.
+Contexts are interpreted as dependent relations between algebras and displayed
+algebras. We again fix a universe level $\hat{i}$.
 \[
@@
 \]
-For a term $t$, $t^\S$ witnesses that the relation corresponding to
-its type holds for $t^\A$ and $t^\D$.
+For a term $t$, $t^\S$ again witnesses a fundamental theorem.
 \[
@@
 \]
-We present the implementation below. Here, we make $\hat{\gamma^D}$-s implicit
-in addition to $\hat{\gamma}$-s, and mostly notate $\hat{\gamma^S}$ parameters
-and applications,
+We present the implementation below. Here, we make $\hat{\gamma}$-s and
+$\hat{\gamma^D}$-s implicit and mostly notate $\hat{\gamma^S}$ parameters and
+applications.
 
@@
 
-The $\U^\S$ and $(\underline{a})^\S$ definitions are the key points of
-the $\blank^\S$ operation. The definitions for the other $\blank^\S$
-cases are largely determined by these.
-
-The $\U^\S$ rule yields the type of the eliminator for a type
-formation rule. In the natural numbers example above, the non-indexed
-$Nat : \U$ sort is interpreted as $\hat{\hat{n^S} \in (\hat{x}\in
-\hat{n})\rah \hat{n^D}\,\hat{x}}$. For indexed sorts, the indices are
-first processed by the $\blank^\S$ cases for inductive and non-inductive
-parameters, until the ultimate $\U$ return type is reached. Hence, the
-eliminator for a sort is always a function.
+The interpretations follow the same pattern as in the case of $\blank^\M$ up
+until $((x:a)\ra B)^\S$, with $\U^\S$ defined as a dependent function instead of
+a non-dependent one. Also, $\U^\S\,\hat{\gamma^S}\,\hat{A}\,\hat{A^D}$ is
+precisely the type of sections of the $\hat{A^D}$ type family.
+
+Let us consider now the translation in with the
+corresponding induction principles in mind, defined as $\mathsf{Induction}$ in
+Section \ref{sec:general}.
+
+The $\U^\S$ rule yields the type of the eliminator function for a type
+constructor. For natural numbers, the non-indexed $Nat : \U$ is interpreted
+as $\hat{\hat{Nat^S} \in (\hat{x}\in \hat{Nat})\rah \hat{Nat^D}\,\hat{x}}$. For
+indexed types, the indices are first processed by the $\blank^\S$ cases for
+inductive and external function parameters, until the ultimate $\U$ return type
+is reached. Hence, we always get an eliminator function for a type constructor.
 
 Analogously, the $\blank^\S$ result type for a point or path constructor is
-always a $\beta$-rule, i.e.\ a function type ending in an equality. To
-see why, consider the $(\underline{a})^\S$ definition. It expresses
-that applications of $a^\S$ eliminators must be equal to the
-corresponding $\hat{t^D}$ induction methods. Hence, for path and point
-constructor types, $\blank^\S$ works by first processing all inductive and
-non-inductive indices, then finally returning an equality type.
-
-Let us also consider the $((x:a)\ra B)^\S$ case for inductive
-parameters. Here, we make crucial use of the fact that the domain type
-$a$ is small. This provides us $a^\S\,\hat{\gamma^3\,x}$, which we use
-to witness the $a^\D\,\hat{\gamma^2}\,\hat{x}$ hypothesis for
-$B^\S$. Without the size restriction on inductive parameters (which
-enforces strict positivity), the $\blank^\S$ operation would not be
-possible at all, because $a^\S$ for a large $a$ type would be merely
-an opaque relation instead of an eliminator function.
-
-Here, we only provide abbreviated definitions for the $t\,u$, $t=_a
-u$, $\refl$, $\J$ and $\J\beta$ cases. In the $\J$ case, we write
+always a $\beta$-rule, i.e.\ a function type returning an equality. That is
+because $(\underline{a})^\S$ expresses that applications of $a^\S$ eliminators
+must be equal to the corresponding $\hat{t^D}$ induction methods. Hence, for
+path and point constructor types, $\blank^\S$ works by first processing all
+inductive and external parameters, then finally returning an equality type.
+
+Here, we only provide abbreviated definitions for the $t\,u$, $t=_a u$, $\refl$,
+$\J$ and $\J\beta$ cases. In the $\J$ case, we write
 $\hat{\mathsf{uncurry}}\,p^\D$ for
-$\hat{\lambda\,\gamma\,\gamma^D\,x\,x^D\,z\,z^D.}\,p^\D\,(\hat{\gamma,x,z})\,(\hat{\gamma^D,x^D,z^D})$
-and analogously elsewhere. The full definitions can be found in the
+$\hat{\lambda\,\gamma\,\gamma^D\,x\,x^D\,z\,z^D.}\,p^\D\,\hat{(}\hat{\gamma,x,z}\hat{)}\,\hat{(}\hat{\gamma^D,x^D,z^D}\hat{)}$
+and analogously elsewhere, to adjust for the fact that $p$ abstracts over
+additional $x$ and $z$ variables. The full definitions can be found in the
 Agda formalization. The definitions are highly constrained by the
 required types, and not particularly difficult to implement with the
-help of a proof assistant; they all involve doing successive path
+help of a proof assistant: they all involve doing successive path
 induction on all equalities available from induction hypotheses, with
@@
 
-The full $(\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S$ definition is quite
-large, and, for instance, yields a very large $\beta$-rule for the
-higher inductive torus definition (the reader can confirm this using
-the Haskell implementation). Nevertheless, an implementation focused
-on practicality may provide smaller specialized $\blank^\S$ definitions for
-commonly used equality operations such as composition or inverses.
+The full $(\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S$ definition is quite large, and,
+for instance, yields a very large $\beta$-rule for the higher inductive torus
+definition (the reader can confirm this using the Haskell implementation). One
+could have an implementation with specialized cases for commonly used operations
+such as path compositions and inverses, in order to produce smaller translation
+output.
 
@@
 
-\section{On Extension to Categorical Semantics}
+\section{Possible Extensions to Categorical Semantics}
 \label{sec:categorical}
 
-TODO
-
-%% \section{Existence of HIITs}
-%% \label{sec:adding}
-
-%% The target type theory supports HIITs if whenever we can derive
-%% $\hat{\Gamma}\vdash\Delta$ in the source theory, the following rules are
-%% admissible.
-%% \[
-%% \infer{\hat{\Gamma}\vdashh\hat{\con}_\Delta\in \Delta^\C}{\vspace{1em}}
-%% \hspace{5em}
-%% \infer{\hat{\Gamma}\vdashh\hat{\elim}_\Delta\,\hat{m}\in\Delta^\E\,\hat{\con}_\Delta\,\hat{m}}{\hat{\Gamma}\vdashh \hat{m}\in \Delta^\M\,\hat{\con}_\Delta}
-%% \]
-%% We can add HIITs to the target theory by extending it with the theory
-%% of codes (making the target and the source theory the same) and adding
-%% the above rules with the additional assumption of
-%% $\hat{\Gamma}\vdash\Delta$. However, this only adds HIITs with weak
-%% computation rules. To make the computation rules definitional, we
-%% would probably need a two-level target type theory with an equality
-%% having an equality reflection rule as in Voevodsky's homotopy type
-%% system \cite{hts} or Andromeda \cite{andromeda}.
+So far, we were able to compute algebras, displayed algebras, morphisms and
+sections, and this allows us to state recursion and induction principles.
+Reiterating Section \ref{sec:general}, for a signature
+$\hat{\Gamma\vdash}\,\Delta$ and assuming $\hat{\Gamma\vdash \mathsf{InitialAlg}
+  : \blc{\Delta^\A}}$, we have the following types for induction and recursion:
+\begin{alignat*}{5}
+& \hat{\Gamma\vdash \mathsf{Induction} :
+    (\gamma^D : \blc{\Delta^\D}\,\mathsf{InitialAlg})\ra \blc{\Delta^\S}\,\mathsf{InitialAlg}\,\gamma^D}\\
+& \hat{\Gamma\vdash \mathsf{Recursion} :
+    (\gamma : \blc{\Delta^\A})\ra \blc{\Delta^\M}\,\mathsf{InitialAlg}\,\gamma}
+\end{alignat*}
+
+However, this is not the full picture. We would also like to have a
+\emph{category} of algebras, with homomorphisms as morphisms, identity
+morphisms, composition together with category laws. This is not without
+difficulties.
+
+We are working in an UIP-free setting. Here, standard definitions of categories
+feature set-truncated morphisms \cite{HoTTbook}. But we have non-truncated
+notions of algebras and morphisms, so we cannot use the standard
+definitions. One solution is to simply use non-truncated categories; these have
+been previously called ``precategories'' or ``wild categories''
+\cite{semisegal}. Sojakova demonstrated \cite{sojakova} that working with wild
+categories internally to type theory is enough to prove equivalence of induction
+and homotopy initiality for a class of higher inductive types, which suggests
+that the same might be possible for HIITs. Still, it would be desirable to build
+semantics of HIITs in a richer ($\omega$, 1)-categorical setting.
+
+However, out current approach does not scale to the point where we have to worry
+about higher categories. We explain in the following. The logical next step
+towards a categorical semantics would be defining a $\blank^\ID$ translation,
+which computes identity homomorphisms:
+
+\[
+\infer{\hat{\Gamma}\vdashh\Delta^\ID\,\hat{:\,(\gamma : \blc{\Delta^\A})\ra \blc{\Delta^\M}\,\gamma\,\gamma}}
+      {\hat{\Gamma}\vdash\Delta}
+\hspace{1em}
+\infer{
+  \hat{\Gamma}\vdashh A^\ID\,\hat{:\,(\gamma : \blc{\Delta^\A})(t : \blc{A^\A}\,\gamma)\ra \blc{A^\M}\,\gamma\,\gamma\,(\blc{\Delta^\ID}\,\gamma)\,t\,t }}
+      {\hat{\Gamma}\semicol\Delta\vdash A}
+\]
+
+\[
+\infer{
+  \hat{\Gamma}\vdashh t^\ID\,\hat{:\, (\gamma : \blc{\Delta^\A})\ra
+       \blc{t^\M}\,\gamma\,\gamma\,(\blc{\Delta^\ID}\,\gamma) = \blc{A^\ID}\,\gamma\,(\blc{t^\A}\,\gamma) }}
+      {\hat{\Gamma}\semicol\Delta\vdash t : A}
+\]
+
+Here, the interpretation of terms witnesses functoriality: $t^\M$ maps
+identity morphisms in $\Delta$ to displayed identity morphisms in $A$.
+We translate $\U$ to identity functions:
+\[
+\U^\ID\,\hat{\gamma\,A} :\equiv \hat{\lambda\,(x : A).\, x}
+\]
+Note that $\U$ and $\U[\sigma]$ are definitionally equal in the theory of
+signatures, where $\sigma$ is a parallel substitution. The translation of the
+latter would be the following (omitting many details, including the handling of
+substitutions in the translation):
+\[
+(\U[\sigma])^\ID\,\hat{\gamma\,A} \equiv \hat{\TR_{(\lambda\,x. A\ra A)}\,(\sigma^\ID\,\gamma)\,(\lambda\,x.\, x)}
+\]
+Here, $\hat{\sigma^\ID\,\gamma}$ yields an equation for functoriality of
+$\hat{\sigma}$. Since the transport is constant, the result is propositionally
+equal to $\hat{\lambda\,x.\, x}$. However, we need to preserve $\U = \U[\sigma]$
+up to definitional equality.
 
@@
 
-There are three additional development artifacts to the current work:
-a Haskell implementation, a shallow Agda formalization and a deep Agda
-formalization. All three are available from the homepage of the first
-author.
+There are three additional development artifacts to the current work: a Haskell
+implementation, a shallow Agda formalization and a deep Agda formalization. All
+three are available from the homepage of the first author.
 
@@
 \appendix
-\section{The homomorphism interpretation}
+\section{The homomorphism translation}
 \label{sec:morphismrules}

### 2019-01-31 paper.tex
stuff

@@
 However, this is not the full picture. We would also like to have a
-\emph{category} of algebras, with homomorphisms as morphisms, identity
-morphisms, composition together with category laws. This is not without
-difficulties.
-
-We are working in an UIP-free setting. Here, standard definitions of categories
-feature set-truncated morphisms \cite{HoTTbook}. But we have non-truncated
-notions of algebras and morphisms, so we cannot use the standard
-definitions. One solution is to simply use non-truncated categories; these have
-been previously called ``precategories'' or ``wild categories''
+\emph{category} of algebras, with homomorphisms as morphisms. This is not
+without difficulties.
+
+We are working in an UIP-free setting. In such setting, standard definitions of
+categories feature set-truncated morphisms \cite{HoTTbook}. But we have
+non-truncated notions of algebras and homomorphisms, so we cannot use the
+standard definitions. One solution is to simply use non-truncated categories;
+these have been previously called ``precategories'' or ``wild categories''
 \cite{semisegal}. Sojakova demonstrated \cite{sojakova} that working with wild
@@
 However, out current approach does not scale to the point where we have to worry
-about higher categories. We explain in the following. The logical next step
+about higher categories. We explain in the following. The natural next step
 towards a categorical semantics would be defining a $\blank^\ID$ translation,
@@
 $\hat{\sigma}$. Since the transport is constant, the result is propositionally
-equal to $\hat{\lambda\,x.\, x}$. However, we need to preserve $\U = \U[\sigma]$
-up to definitional equality.
+equal to $\hat{\lambda\,x.\, x}$. However, that is not enough, since we need to
+preserve $\U = \U[\sigma]$ up to definitional equality.
+
+There are numerous unavoidable instances of such failures of strict preservation
+of definitional equalities for $\blank^\ID$, and they similarly arise when we
+try to construct composition of homomorphisms. We consider three potential
+solutions for this strictness problem:
+\begin{enumerate}
+\item Solving the coherence problem. This would allow us to interpret signatures
+      into the metatheory, allowing preservation of definitional equality up
+      to propositional metatheoretic equality.
+\item Reformulating the syntax of signatures so that definitional equalities
+      become weak propositional equalities. We already use weak $\beta$ for $\J$,
+      can we do so elsewhere? However, this would result in an unusual and very
+      inconvenient syntax of signatures, because we would need to weaken even
+      basic substitution rules.
+\item Instead of interpreting signatures into an UIP-free type theory, we
+      interpret them into classical combinatorial structures for higher groupoids
+      and categories, e.g.\ into simplicial sets. The drawback is that now we do not
+      have the convenient synthetic notion of higher structures, provided by the syntax
+      of type theory, and instead we have to manually build up these
+      structures. Needless to say, this makes machine-checked formalization much
+      more difficult. We have found mechanized formalization invaluable for the
+      current paper, and it would be painful to abandon it in further research.
+\end{enumerate}
+
+In summary, solutions for the strictness problems require significant deviation
+from the approach of the current paper, of require significant further research.
+
+%% Nevertheless, the current approach of using syntactic translations is a sweet
+%% spot for computing induction principles in a relatively simple way.
 
@@
 
-\section{Conclusions and further work}
+\section{Conclusions}
 \label{sec:summary}
 
-Higher inductive-inductive types are useful in defining the well-typed
-syntax of type theory in an abstract way \cite{ttintt}. From a
-universal algebra point of view, they provide initial algebras for
-multi-sorted algebraic theories where the sorts can depend on each
-other. From the perspective of homotopy type theory, they provide
-synthetic versions of homotopy-theoretic constructions such as higher
-dimensional spheres or cell complexes. So far, no general scheme of
+Higher inductive-inductive types are useful in defining the well-typed syntax of
+type theory in an abstract way \cite{ttintt}. From a universal algebra point of
+view, they provide initial algebras for multi-sorted algebraic theories where
+the sorts can depend on each other. From the perspective of homotopy type
+theory, they provide synthetic versions of homotopy-theoretic constructions such
+as higher dimensional spheres or cell complexes. So far, no general scheme of
 HIITs have been proposed. To quote Lumsdaine and Shulman
@@
 In this paper we proposed such a syntactic scheme which also includes
-inductive-inductive types. We tackled the problem of complex
-dependencies on previous type formation rules and constructors by a
-well-known method of describing intricate dependencies: the syntax of
-type theory itself. We had to limit the type formers to only allow
-strictly positive definitions, but these restrictions are the only
-things that a type theorist has to learn to understand our codes. Our
-encoding is also \emph{direct} in the sense that the types of
-constructors and eliminators are exactly as required and not merely up
-to isomorphisms.
-
-In this paper we only \emph{specified} HIITs and characterized their
-induction principles. Proving their \emph{existence} is left as
-further work. This would likely involve reducing HIITs to basic
-building blocks such as W-types and quotient types.
-
-The theory of codes was defined as part of the syntax of another type
-theory, the target theory. An alternative way would be to define
-the theory of codes internally to a type theoretic metatheory in the
-style of \cite{ttintt}. However, as described in \cite[Section
-  6]{ttintt}, there would be a coherence problem: for the internal
-syntax to be useful, we need to truncate it to be a set (as the
-$\mathsf{trunc}$ constructor did for $\Int$ in the introduction). As
-the eliminator needs to respect the equality given by
-$\mathsf{trunc}$, we would only be able to eliminate from the internal
-type theory into a set. This would preclude the definition of even the
-$\blank^\C$ operation, which corresponds to the standard model. A
-possible solution to this problem would be to add all the higher
-coherence laws to the syntax (e.g.\ the pentagon law for the
-composition of substitutions) and then prove that the syntax is a
-set. For this however, we would probably need a two-level metatheory
+inductive-inductive types. We tackled the problem of complex dependencies on
+previous type formation rules and constructors by a well-known method of
+describing intricate dependencies: the syntax of type theory itself. We had to
+limit the type formers to only allow strictly positive definitions, but these
+restrictions are the only things that a type theorist has to learn to understand
+our codes. Our encoding is also \emph{direct} in the sense that the types of
+constructors and eliminators are exactly as required and not merely up to
+isomorphisms.
+
+In this paper we only \emph{specified} HIITs and characterized their induction
+principles. Proving their \emph{existence} is left as further work. This would
+likely involve reducing HIITs to basic building blocks such as W-types and
+quotient types.
+
+The theory of codes was defined as part of the syntax of another type theory,
+the target theory. An alternative way would be to define the theory of codes
+internally to a type theoretic metatheory in the style of
+\cite{ttintt}. However, as described in \cite[Section 6]{ttintt}, there would be
+a coherence problem: for the internal syntax to be useful, we need to truncate
+it to be a set (as the $\mathsf{trunc}$ constructor did for $\Int$ in the
+introduction). As the eliminator needs to respect the equality given by
+$\mathsf{trunc}$, we would only be able to eliminate from the internal type
+theory into a set. This would preclude the definition of even the $\blank^\C$
+operation, which corresponds to the standard model. A possible solution to this
+problem would be to add all the higher coherence laws to the syntax (e.g.\ the
+pentagon law for the composition of substitutions) and then prove that the
+syntax is a set. For this however, we would probably need a two-level metatheory
 as in \cite{semisegal}.
 
-When working in a metatheory with uniqueness of identity proofs (which
-implies that all HIITs are in essence QIITs), the theory of codes
-admits a category model where the interpretation of a context is the
-category of algebras corresponding to the context. In the future, we
-would like to prove that these categories have initial algebras given
-by terms in the theory of codes.
+When working in a metatheory with uniqueness of identity proofs (which implies
+that all HIITs are in essence QIITs), the theory of codes admits a category
+model where the interpretation of a context is the category of algebras
+corresponding to the context. In the future, we would like to prove that these
+categories have initial algebras given by terms in the theory of codes.
 

### 2019-01-31 paper.tex
Fix latexmk errors

@@
 \newcommand{\Nes}{\mathsf{Nes}}
-\newcommand{\ID}{\mathsf{ID}}
+%% \newcommand{\ID}{\mathsf{ID}}
 \newcommand{\id}{\mathsf{id}}
@@
 \newcommand{\idr}{\mathsf{idr}}
-\newcommand{\inv}{\mathsf{inv}}
-\newcommand{\ID}{\mathsf{\ID}}
+\renewcommand{\inv}{\mathsf{inv}}
 
@@
 Sections \ref{sec:morphisms} and \ref{sec:sections} do not preserve this
-$\beta$-rule strictly. We explain this in more detail in Sections TODO and
-TODO. Adding equality to the theory of codes allows higher constructors and
-inductive equality parameters as well. We can now define the higher-inductive
-circle as the following context:
+$\beta$-rule strictly. We explain this in more detail in Sections
+\ref{sec:coherence} and \ref{sec:categorical}. Adding equality to the theory of
+codes allows higher constructors and inductive equality parameters as well. We
+can now define the higher-inductive circle as the following context:
 \[
@@
   & (\Delta,x:A)^\A && :\equiv \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\timesh A^\A\,\hat{\gamma} \\
-  & x^\A\,\hat{\gamma} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma} \\ % TODO: write De Bruijn versions
+  & x^\A\,\hat{\gamma} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma} \\
   & \U^\A\,\hat{\gamma} && :\equiv \Type_{\hat{0}} \\
@@
 about higher categories. We explain in the following. The natural next step
-towards a categorical semantics would be defining a $\blank^\ID$ translation,
+towards a categorical semantics would be defining a $\blank^{\mathsf{ID}}$ translation,
 which computes identity homomorphisms:
@@
 \[
-\infer{\hat{\Gamma}\vdashh\Delta^\ID\,\hat{:\,(\gamma : \blc{\Delta^\A})\ra \blc{\Delta^\M}\,\gamma\,\gamma}}
+\infer{\hat{\Gamma}\vdashh\Delta^{\mathsf{ID}}\,\hat{:\,(\gamma : \blc{\Delta^\A})\ra \blc{\Delta^\M}\,\gamma\,\gamma}}
       {\hat{\Gamma}\vdash\Delta}
@@
 \infer{
-  \hat{\Gamma}\vdashh A^\ID\,\hat{:\,(\gamma : \blc{\Delta^\A})(t : \blc{A^\A}\,\gamma)\ra \blc{A^\M}\,\gamma\,\gamma\,(\blc{\Delta^\ID}\,\gamma)\,t\,t }}
+  \hat{\Gamma}\vdashh A^{\mathsf{ID}}\,\hat{:\,(\gamma : \blc{\Delta^\A})(t : \blc{A^\A}\,\gamma)\ra \blc{A^\M}\,\gamma\,\gamma\,(\blc{\Delta^{\mathsf{ID}}}\,\gamma)\,t\,t }}
       {\hat{\Gamma}\semicol\Delta\vdash A}
@@
 \infer{
-  \hat{\Gamma}\vdashh t^\ID\,\hat{:\, (\gamma : \blc{\Delta^\A})\ra
-       \blc{t^\M}\,\gamma\,\gamma\,(\blc{\Delta^\ID}\,\gamma) = \blc{A^\ID}\,\gamma\,(\blc{t^\A}\,\gamma) }}
+  \hat{\Gamma}\vdashh t^{\mathsf{ID}}\,\hat{:\, (\gamma : \blc{\Delta^\A})\ra
+       \blc{t^\M}\,\gamma\,\gamma\,(\blc{\Delta^{\mathsf{ID}}}\,\gamma) = \blc{A^{\mathsf{ID}}}\,\gamma\,(\blc{t^\A}\,\gamma) }}
       {\hat{\Gamma}\semicol\Delta\vdash t : A}
@@
 \[
-\U^\ID\,\hat{\gamma\,A} :\equiv \hat{\lambda\,(x : A).\, x}
+\U^{\mathsf{ID}}\,\hat{\gamma\,A} :\equiv \hat{\lambda\,(x : A).\, x}
 \]
@@
 \[
-(\U[\sigma])^\ID\,\hat{\gamma\,A} \equiv \hat{\TR_{(\lambda\,x. A\ra A)}\,(\sigma^\ID\,\gamma)\,(\lambda\,x.\, x)}
-\]
-Here, $\hat{\sigma^\ID\,\gamma}$ yields an equation for functoriality of
+(\U[\sigma])^{\mathsf{ID}}\,\hat{\gamma\,A} \equiv \hat{\TR_{(\lambda\,x. A\ra A)}\,(\sigma^{\mathsf{ID}}\,\gamma)\,(\lambda\,x.\, x)}
+\]
+Here, $\hat{\sigma^{\mathsf{ID}}\,\gamma}$ yields an equation for functoriality of
 $\hat{\sigma}$. Since the transport is constant, the result is propositionally
@@
 There are numerous unavoidable instances of such failures of strict preservation
-of definitional equalities for $\blank^\ID$, and they similarly arise when we
+of definitional equalities for $\blank^{\mathsf{ID}}$, and they similarly arise when we
 try to construct composition of homomorphisms. We consider three potential
@@
 
-%% Nevertheless, the current approach of using syntactic translations is a sweet
-%% spot for computing induction principles in a relatively simple way.
-
 %=====================================================================
@@
 implementation, a shallow Agda formalization and a deep Agda formalization. All
-three are available from the homepage of the first author.
-
-The Haskell implementation takes as input a file which contains a
-representation of a $\hat{\Gamma}\vdash\Delta$ specification of a
-HIIT. Then, it checks the input with respect to the rules in Figure
-\ref{sigrules}, and outputs an Agda-checkable file which contains the
-results of the $\blank^\C$, $\blank^\M$ and $\blank^\E$
-translations. It comes with examples, including the ones in this
-paper, indexed W-types \cite{morris09indexed}, the dense completion
-\cite[Appendix A.1.3]{forsberg-phd} and several HITs from
-\cite{HoTTbook} including the definition of Cauchy reals. It can be
-checked that our implementation computes the expected elimination
+three are available from \url{https://github.com/akaposi/hiit-signatures}.
+
+The Haskell implementation takes as input a file which contains a of a
+$\hat{\Gamma}\vdash\Delta$ signature. Then, it checks the input with respect to
+the rules in Figure \ref{sigrules}, and outputs an Agda-checkable file which
+contains algebras, homomorphisms, displayed algebras and sections for the
+input signature.
+
+It comes with examples, including the ones in this paper, the
+inductive-inductive dense completion \cite[Appendix A.1.3]{forsberg-phd} and
+several HITs from \cite{HoTTbook} including the definition of Cauchy reals. It
+can be checked that our implementation computes the expected elimination
 principles in these cases.
@@
 The shallow Agda formalization embeds both the source and target theories
-shallowly into Agda: it represents types as Agda types, functions as
-Agda functions, and so on. We also leave the $\blank^\C$ operation
-implicit. We state each case of the $\blank^\M$ and $\blank^\E$
-translations as Agda functions from all induction hypotheses to the
-result type of the translation, which lets us ``typecheck'' the
-translation. We have found that this style of formalization is
-conveniently light, but remains detailed enough to be useful. We also
-generated some of the code of the Haskell implementation from this
-formalization.
-
-The deep Agda formalization still uses a shallow embedding of the
-target type theory, but it uses a deep embedding of the source theory,
-in the style of \cite{ttintt}. In the formalization we merge the three
-translations into a single model construction. This allows us to prove
-strict preservation of definitional equalities in the substitution
-calculus of the source theory, in contrast to the shallow
-formalization, where we cannot reason about definitional equalities of
-Agda terms. Due to technical challenges, this formalization uses
-transport instead of $\J$ in the source theory, but this still covers
-a rather large class of HIIT definitions.
-
-%=====================================================================
+shallowly into Agda: it represents types as Agda types, functions as Agda
+functions, and so on. We also leave the $\blank^\A$ operation implicit. We state
+each case of translations as Agda functions from all induction hypotheses to the
+result type of the translation, which lets us ``typecheck'' the translation. We
+have found that this style of formalization is conveniently light, but remains
+detailed enough to be useful. We also generated some of the code of the Haskell
+implementation from this formalization.
+
+The deep Agda formalization deeply embeds the theory of signatures as a CwF with
+additional structure, in the style of \cite{ttintt}. However, it embeds the
+external type theory shallowly as Agda, and we model dependency on external
+$\hat{\Gamma}$ contexts with Agda functions. We formalize $\blank^\A$,
+$\blank^\D$, $\blank^\M$ and $\blank^\S$ translations as a single model
+construction in Agda. This setup greatly simplifies formalization, since we do
+not have to reason explicitly about definitional equality in the external
+syntax, but we can still reason directly about preservation of definitional
+equalities in the theory of signatures. This semi-deep formalization can be in
+principle converted into a fully formal syntactic translation, because we prove
+all preservations of definitional equalities by $\refl$. Due to technical
+challenges, this formalization uses transport instead of $\J$ in the source
+theory, but this still covers a rather large class of HIIT definitions.
+
+%--------------------------------------------------------------------------------
 
@@
 Higher inductive-inductive types are useful in defining the well-typed syntax of
-type theory in an abstract way \cite{ttintt}. From a universal algebra point of
+type theory in an abstract way \cite{ttintt}. From a universal algebraic point of
 view, they provide initial algebras for multi-sorted algebraic theories where
@@
 theory, they provide synthetic versions of homotopy-theoretic constructions such
-as higher dimensional spheres or cell complexes. So far, no general scheme of
+as higher-dimensional spheres or cell complexes. So far, no general scheme of
 HIITs have been proposed. To quote Lumsdaine and Shulman
@@
 restrictions are the only things that a type theorist has to learn to understand
-our codes. Our encoding is also \emph{direct} in the sense that the types of
-constructors and eliminators are exactly as required and not merely up to
-isomorphisms.
-
-In this paper we only \emph{specified} HIITs and characterized their induction
-principles. Proving their \emph{existence} is left as further work. This would
-likely involve reducing HIITs to basic building blocks such as W-types and
-quotient types.
-
-The theory of codes was defined as part of the syntax of another type theory,
-the target theory. An alternative way would be to define the theory of codes
-internally to a type theoretic metatheory in the style of
-\cite{ttintt}. However, as described in \cite[Section 6]{ttintt}, there would be
-a coherence problem: for the internal syntax to be useful, we need to truncate
-it to be a set (as the $\mathsf{trunc}$ constructor did for $\Int$ in the
-introduction). As the eliminator needs to respect the equality given by
-$\mathsf{trunc}$, we would only be able to eliminate from the internal type
-theory into a set. This would preclude the definition of even the $\blank^\C$
-operation, which corresponds to the standard model. A possible solution to this
-problem would be to add all the higher coherence laws to the syntax (e.g.\ the
-pentagon law for the composition of substitutions) and then prove that the
-syntax is a set. For this however, we would probably need a two-level metatheory
-as in \cite{semisegal}.
-
-When working in a metatheory with uniqueness of identity proofs (which implies
-that all HIITs are in essence QIITs), the theory of codes admits a category
-model where the interpretation of a context is the category of algebras
-corresponding to the context. In the future, we would like to prove that these
-categories have initial algebras given by terms in the theory of codes.
+our signatures. Our encoding is also direct in the sense that notions of
+induction and recursion are computed exactly as required and not merely up to
+equivalences or isomorphisms, and we also demonstrated that this computation is
+feasible to implement in computer programs.
+
+We developed an approach where syntactic translations are used to provide
+semantics for HIIT signatures. Our approach seems to be a sweet spot for
+computing notions of induction and recursion in a formally verifiable and
+relatively simple way.
+
+However, there is a coherence problem in the formal treatment of syntaxes and
+models of type theories internally to type theory. We sidestepped this by
+considering syntactic translations, but the problem remains, and it prevents
+extending the current approach to categorical semantics. Our impression is that
+the true solution will be the development of higher syntaxes and models of type
+theory in type theory. This may also require adding new features to the
+meta type theory. In particular, convenient formalization of higher categories
+seems elusive in conventional homotopy type theory, and we may need two-level
+type theories \cite{semisegal}, or even directed type-theories with native
+notions of higher categories \cite{nuyts2015towards, riehl2017type}.
+
+foobar
+
+%% In this paper we specified HIITs and provide part of their semantics, by
+%% computing notions of recursion and induction. Pro
+
+%% In this paper we only \emph{specified} HIITs and characterized their induction
+%% principles. Proving their \emph{existence} is left as further work. This would
+%% likely involve reducing HIITs to basic building blocks such as W-types and
+%% quotient types.
+
+%% The theory of codes was defined as part of the syntax of another type theory,
+%% the target theory. An alternative way would be to define the theory of codes
+%% internally to a type theoretic metatheory in the style of
+%% \cite{ttintt}. However, as described in \cite[Section 6]{ttintt}, there would be
+%% a coherence problem: for the internal syntax to be useful, we need to truncate
+%% it to be a set (as the $\mathsf{trunc}$ constructor did for $\Int$ in the
+%% introduction). As the eliminator needs to respect the equality given by
+%% $\mathsf{trunc}$, we would only be able to eliminate from the internal type
+%% theory into a set. This would preclude the definition of even the $\blank^\C$
+%% operation, which corresponds to the standard model. A possible solution to this
+%% problem would be to add all the higher coherence laws to the syntax (e.g.\ the
+%% pentagon law for the composition of substitutions) and then prove that the
+%% syntax is a set. For this however, we would probably need a two-level metatheory
+%% as in \cite{semisegal}.
+
+%% When working in a metatheory with uniqueness of identity proofs (which implies
+%% that all HIITs are in essence QIITs), the theory of codes admits a category
+%% model where the interpretation of a context is the category of algebras
+%% corresponding to the context. In the future, we would like to prove that these
+%% categories have initial algebras given by terms in the theory of codes.
 

### 2019-01-31 paper.tex
Add conf paper expansion note

@@
 
+The present paper is an expanded version of our conference paper \cite{hiit}.
+In this version, we additionally compute notions of homomorphisms from
+signatures, and compare them to logical relations, in Section
+\ref{sec:morphisms}. We also explain a coherence problem in interpreting
+syntaxes of type theories, and how this influenced the current paper, in Section
+\ref{sec:coherence} and Section \ref{sec:categorical}.
+
 \subsection{Overview of the Paper}
@@
 In summary, solutions for the strictness problems require significant deviation
-from the approach of the current paper, of require significant further research.
+from the approach of the current paper, or require significant further research.
 
@@
 notions of higher categories \cite{nuyts2015towards, riehl2017type}.
-
-foobar
 

### 2019-01-31 elims-demo/Examples.hs
Add Haskell implementation

@@ -1,0 +1,145 @@
+{-# language UnicodeSyntax #-}
+
+{-|
+A couple of examples.
+
+Load this module in ghci, and entr "hiitToAgda example" to check the HIIT definitions and
+get the types of induction methods and eliminators as Agda output.
+
+-}
+
+module Examples where
+
+import Impl
+
+nat ∷ String
+nat = "constructors Nat: U; zero: Nat; suc: Nat;"
+
+fin ∷ String
+fin =
+     "assume        Nat: U; zero: Nat; suc: Nat → Nat;"
+  ++ "constructors  Fin: Nat → U; fzero: (n : Nat) → Fin (suc n); fsuc: (n: Nat) → Fin n → Fin (suc n);"
+
+vec ∷ String
+vec =
+     "assume Nat: U; zero: Nat; suc: Nat → Nat; A: U;"
+  ++ "constructors Vec: Nat → U; nil: Vec zero; cons: (n: Nat) → A → Vec n → Vec (suc n);"
+
+-- | A fragment of an inductive-inductive type theory syntax.
+conTyTm ∷ String
+conTyTm =
+     "constructors "
+  ++ "Con: U; Ty: (Γ : Con) → U; Tm: (Γ: Con)(A : Ty Γ) → U;"
+  ++ "nil: Con; cons: (Γ : Con) → Ty Γ → Con;"
+  ++ "u: (Γ : Con) → Ty Γ; El: (Γ : Con) → Tm Γ (u Γ) → Ty Γ;"
+
+wType ∷ String
+wType =
+     "assume S: U; P : S → U;"
+  ++ "constructors W: U; con: (s: S) → (P s → W) → W;"
+
+ixWType ∷ String
+ixWType =
+     "assume "
+  ++ "  I   : U;"
+  ++ "  S   : U;"
+  ++ "  P   : S → U;"
+  ++ "  Out : S → I;"
+  ++ "  In  : (s : S) → P s → I;"
+  ++ "constructors "
+  ++ "  W   : I → U;"
+  ++ "  sup : (s : S) → ((p : P s) → W (In s p)) → W (Out s);"
+
+-- | Inductive-inductive dense completion of a relation.
+denseCompletion ∷ String
+denseCompletion =
+     "assume "
+  ++   "A : U;"
+  ++   "R : A → A → U;"
[…]

### 2019-01-31 elims-demo/Impl.hs

@@ -1,0 +1,878 @@
+{-# language
+  Strict, LambdaCase, ViewPatterns, TupleSections,
+  UndecidableInstances, UnicodeSyntax, OverloadedStrings, CPP,
+  MultiParamTypeClasses, AllowAmbiguousTypes, PatternSynonyms,
+  TypeApplications, ScopedTypeVariables, FlexibleInstances,
+  FunctionalDependencies, InstanceSigs #-}
+{-# options_ghc -fwarn-incomplete-patterns #-}
+
+module Impl (hiitToAgda, runHIITDefs) where
+
+-- TODO:
+--  - universes in target TT
+--  - apd in output
+
+import Control.Arrow
+import Control.Monad hiding (ap)
+import Control.Monad.Except hiding (ap)
+import Data.Foldable
+import Data.Function
+import Data.IORef
+import Data.Char
+import Data.List
+import Data.List.Split
+import Data.Maybe
+import Data.String
+import Prelude hiding (ap)
+import System.IO.Unsafe
+
+import Text.Megaparsec.Pos
+import Text.Megaparsec.Error
+
+import Presyntax
+
+-- Errors
+--------------------------------------------------------------------------------
+
+{-# noinline currPos #-}
+currPos ∷ IORef SourcePos
+currPos = unsafeDupablePerformIO (newIORef (initialPos ""))
+
+reportError ∷ String → Elab a
+reportError msg =
+  let pos = unsafeDupablePerformIO (readIORef currPos)
+  in Left (sourcePosPretty pos ++ ":\n\n" ++ msg ++ "\n")
+
+updPos ∷ Posed a → Posed a
+updPos (p, a) = seq (unsafeDupablePerformIO (writeIORef currPos p)) (p, a)
+
+--------------------------------------------------------------------------------
+
+type Bind a  = (Name, a)
+type Bind2 a = (Name, Name, a)
+type Sub a   = [Bind a]
+
+data Tm
+  = Var Name
+  | App Tm Tm
+  | Lam (Bind Tm)
+  | Pi Tm (Bind Tm)
+  | Id (Maybe Tm) Tm Tm  -- Id A t u
+  | J Tm Tm Tm           -- J {a : U}{t : a}(P : (u : a)
+                         --   → t ≡ u → U)(pr : P t refl){u : a}(eq : Id t u)
[…]

### 2019-01-31 elims-demo/Main.hs

@@ -1,0 +1,13 @@
+{-# language UnicodeSyntax #-}
+
+module Main where
+
+import System.Environment
+import Impl
+
+main ∷ IO ()
+main = do
+  args ← getArgs
+  case args of
+    path:_ → runHIITDefs path
+    _ → putStrLn "hiit-defs: no input file"

### 2019-01-31 elims-demo/Presyntax.hs

@@ -1,0 +1,170 @@
+{-# language UnicodeSyntax, LambdaCase, TupleSections, ViewPatterns, Strict,
+    PartialTypeSignatures #-}
+{-# options_ghc -Wincomplete-patterns -Wno-partial-type-signatures #-}
+
+module Presyntax (
+    Name
+  , Tmᴾ
+  , Tmᴾ'(..)
+  , parseTmᴾ
+  , parseInput
+  , Posed
+  ) where
+
+import Control.Monad
+import Data.List
+import Data.Void
+
+import qualified Data.HashSet               as HS
+
+import           Text.Megaparsec
+import qualified Text.Megaparsec.Char  as C
+import qualified Text.Megaparsec.Char.Lexer as L
+
+--------------------------------------------------------------------------------
+
+type Name = String
+type Tmᴾ  = (SourcePos, Tmᴾ')
+
+data Tmᴾ'
+  = Varᴾ Name
+  | Appᴾ Tmᴾ Tmᴾ
+  | Lamᴾ Name Tmᴾ
+  | Piᴾ Name Tmᴾ Tmᴾ
+  | Reflᴾ Tmᴾ
+  | Jᴾ Tmᴾ Tmᴾ Tmᴾ
+  | Idᴾ Tmᴾ Tmᴾ
+  | Uᴾ
+
+-- Parser
+--------------------------------------------------------------------------------
+
+type Parser = Parsec Void String
+
+sc ∷ Parser ()
+sc = L.space C.space1 (L.skipLineComment "--") (L.skipBlockComment "{-" "-}")
+
+lexeme = L.lexeme sc
+symbol = L.symbol sc
+char c = lexeme (C.char c)
+
+parens p   = lexeme (char '(' *> p <* char ')')
+keywords   = HS.fromList ["assume", "constructors", "refl", "J", "Id", "U"]
+keychars   = "ᴱᴹ'" ∷ String
+
+pIdent = try $ lexeme $ do
+  w ← some C.alphaNumChar
+  when (HS.member w keywords) $
+    fail ("can't use keyword for identifier: " ++ w)
+  forM_ w $ \c → when (elem c keychars) $ fail ("Can't use reserved character: " ++ [c])
+  pure w
+
+type Posed a = (SourcePos, a)
+
[…]

### 2019-02-01 paper.tex
Add logrel Nat example
typos

@@
 types of homotopy type theory. Examples that make use of both features are the
-Cauchy reals and the well-typed syntax of type theory where conversion rules are
-given as equality constructors. In this paper we propose a general definition of
-HIITs using a small type theory, named the theory of signatures. A context in
-this theory encodes a HIIT by listing the constructors. We also compute notions
-of induction and recursion for HIITs, by using variants of syntactic logical
-relation translations. Building full categorical semantics and constructing
-initial algebras is left for future work. The theory of HIIT signatures was
-formalised in Agda together with the syntactic translations. We also provide a
-Haskell implementation, which takes signatures as input and outputs translation
-results as valid Agda code.
+Cauchy real numbers and the well-typed syntax of type theory where conversion
+rules are given as equality constructors. In this paper we propose a general
+definition of HIITs using a small type theory, named the theory of signatures. A
+context in this theory encodes a HIIT by listing the constructors. We also
+compute notions of induction and recursion for HIITs, by using variants of
+syntactic logical relation translations. Building full categorical semantics and
+constructing initial algebras is left for future work. The theory of HIIT
+signatures was formalised in Agda together with the syntactic translations. We
+also provide a Haskell implementation, which takes signatures as input and
+outputs translation results as valid Agda code.
 \end{abstract}
@@
 domain-specific type theory which we call the \emph{theory of signatures}. Type
-formers in the theory of codes are restricted in order to enforce strict
+formers in the theory of signatures are restricted in order to enforce strict
 positivity. For example, natural numbers are defined as the three-element
@@
 \ref{sec:coherence} and \ref{sec:categorical}. Adding equality to the theory of
-codes allows higher constructors and inductive equality parameters as well. We
+signatures allows higher constructors and inductive equality parameters as well. We
 can now define the higher-inductive circle as the following context:
@@
 The invalid examples $\mathsf{Ival}$ and $\mathsf{Neg}$ from Section
-\ref{sec:intro} cannot be encoded by the theory of codes. For $\mathsf{Ival}$,
+\ref{sec:intro} cannot be encoded by the theory of signatures. For $\mathsf{Ival}$,
 we can go as far as
@@
 $\hat{\hat{f}\in(\hat{X}:\Type)\rah\hat{X}\rah\hat{X}}$ is an external type, but we
-only have ${Ival}:\U$ in the theory of codes. $\mathsf{Neg}$ cannot be typed
+only have ${Ival}:\U$ in the theory of signatures. $\mathsf{Neg}$ cannot be typed
 because the first parameter of the constructor $\mathsf{con}$ is a function from
@@
 the same way for unrestricted (non-strictly positive) function types as well.
+For an example, this yields the following definition of logical relations for
+natural number algebras:
+\begin{alignat*}{5}
+  & && (\boldsymbol{\cdot},Nat : \U,zero:\underline{Nat},suc:Nat\ra\underline{Nat})^\M
+    \,\hat{(\tt,\,N_0,\,z_0,\,s_0)\,(\tt,\,N_1,\,z_1,\,s_1)} \\
+    & \equiv\,\, && \hat{\top\times(N^M : N_0\ra N_1 \ra \Type_0)}\\
+    & && \hat{\hspace{1em}\times\,\,(z^M : N^M\,z_0\,z_1)}\\
+    & && \hat{\hspace{1em}\times\,\,(s^M : (x_0 : N_0)(x_1 : N_1)(x^M : N^M\,x_0\,x_1)\ra N^M\,(s_0\,x_0)\,(s_1\,x_1))}
+\end{alignat*}
 
@@
 rest of the interpretation unchanged. However, this is not strictly the desired
-notion of homomorphism. Consider the $\blank^\M$ interpretation for natural
+notion of homomorphism. Consider now again the $\blank^\M$ interpretation for natural
 numbers:
@@
 \hat{(}\hat{s} \in \hat{S}\hat{)}\rah \hat{P}\,\hat{s}\rah \hat{I}$. Then, the
-code for the corresponding indexed W-type is the following:
+signature for the corresponding indexed W-type is the following:
 \[

### 2019-02-01 paper.tex
fixes

@@
 
-%% ABBREVS
+%% ABBREVIATIONS
 %% --------------------------------------------------------------------------------
@@
 
-%% These express that the eliminator applied to a constructor expression
-%% reduces to an application of the corresponding induction method. From
-%% an operational point of view, $\Elim\Nat$ replaces all the $\zero$ and
-%% $\suc$ constructors with the given induction methods.
-
-% A special case of the eliminator is the recursor (non-dependent
-% eliminator) which allows the definition of a non-dependent function
-% from the inductive type, for $\Nat$ it has the following type.
-% \[
-% \Rec\Nat : \forall(A:\mathsf{Type}).A \ra (A \ra A) \ra \Nat \ra A
-% \]
-% The eliminator can be derived from the recursor and a uniqueness
-% principle.
-
 Indexed families of types can be also considered, for example length-indexed
@@
 of signatures to the syntax of the external type theory, respectively computing
-algebras, homomorphisms, displayed algebras and sections of displayed
+algebras, displayed algebras, homomorphisms, and sections of displayed
 algebras. In Section \ref{sec:categorical}, we consider extending the previous
-translations with additional components of the categorical semantics of HIITs
+translations with additional components of a categorical semantics
 (e.g.\ identity and composition for homomorphisms), and explain why the approach
@@
 formalizations in Section \ref{sec:formalization}. We explain the rules for the
-theory of signatures in order now.
+theory of signatures in order below.
 
@@
 (2) There is a universe $\U$, with decoding written as an underline instead of
-usual $\El$, to bolster readability. With this part of the syntax, we can already
+usual $\El$, to improve readability. With this part of the syntax, we can already
 define contexts specifying the empty type, unit type and booleans, or in general,
-a collections of finite sets:
+finite sets of finite sets:
 \[
@@
 
-(3) We have a function space with small domain and large codomain. This can be
-used to add inductive arguments to all kinds of constructors. As $\U$ is not
-closed under this function space, these function types cannot (recursively)
-appear in inductive arguments, which ensures strict positivity. When the
-codomain does not depend on the domain, $a\ra B$ can be written instead of
-$(x:a)\ra B$.
+(3) We have a function space with small domain and large codomain, which we call
+the inductive function space. This can be used to add inductive parameters to all
+kinds of constructors. As $\U$ is not closed under this function space, these
+function types cannot (recursively) appear in inductive arguments, which ensures
+strict positivity. When the codomain does not depend on the domain, $a\ra B$ can
+be written instead of $(x:a)\ra B$.
 
@@
 signatures allows higher constructors and inductive equality parameters as well. We
-can now define the higher-inductive circle as the following context:
+can now define the higher inductive circle as the following context:
 \[
@@
 So far we were only able to define closed HIITs, which do not refer to external
-types. Now we add rules which include external types into signatures. A context
+types. We add rules which include external types into signatures. A context
 $\Delta$ for which $\hat{\Gamma}\vdash\Delta$ holds can be seen as a
-specification of an inductive type which depends on $\hat{\Gamma}$. For example,
-in the case of lists for arbitrary external element types, $\hat{\Gamma}$ will
-be $\hat{A}\in\Type_{\hat{0}}$.
-
-(5) We have a function space where the domain is a type in the external theory. We
-distinguish it from (3) by using red brick $\in$ instead of $:$ in the domain
+specification of an inductive type which depends on an external $\hat{\Gamma}$
+signature. For example, in the case of lists for arbitrary external element
+types, $\hat{\Gamma}$ will be $\hat{A}\in\Type_{\hat{0}}$.
+
+(5) We have a function space where the domain is a type in the external
+theory. We distinguish it from (3) by using red brick color in the domain
 specification. We specify lists and the integers as follows.
@@
 
-This notion of syntax is much more compact and often much more convenient to use
+This notion of syntax is much more compact and often more convenient to use
 than extrinsic syntaxes. It is in essence ``merely'' a formalization of CwFs,
@@
 constructors, which amounts to shoving that the interpretations of $\Ty$,
-$\Con$, $\Sub$ are all sets. However, we cannot show $\mathsf{Type}_i$ to be a
+$\Tm$, and $\Sub$ are all sets. However, we cannot show $\mathsf{Type}_i$ to be a
 set without UIP.
@@
 \ref{sec:external}. Truncation in the source syntax is not an issue here,
-because the target syntax is likewise truncated, so we can translate into it.
+because the target syntax is likewise truncated.
 
@@
 the same way for unrestricted (non-strictly positive) function types as well.
-For an example, this yields the following definition of logical relations for
-natural number algebras:
+For an example, this yields the following definition of logical relations
+between natural number algebras:
 \begin{alignat*}{5}
@@
 produce a type as result, but $a^M$ now yields a function. We can view the
-result of $a^\M$ as a functional relation, and relate $t_0$ and $t_1$ by its
-graph:
+result of $a^\M$ as a functional relation, and use its graph to relate $t_0$ and $t_1$:
 \begin{alignat*}{5}
@@
 \begin{alignat*}{5}
-  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\S\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\,\hat{(}\hat{\tt,\,S^{1M},\,b^D,\,loop^D}\hat{)} \\
-  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1E}}\in \hat{(}\hat{x}\in\hat{S^1}\hat{)}\hat{\,\ra\,}\hat{S^{1M}\,x}\hat{)}\timesh\hat{(}\hat{b^S}\in \hat{S^{1E}\,\,b = b^D}\hat{)}\\
+  & && (\boldsymbol{\cdot},\,S^1 : \U,\,b:\underline{S^1},\,loop: \underline{b=b})^\S\,\,\hat{(}\hat{\tt,\,S^1,\,b,\,loop}\hat{)}\,\hat{(}\hat{\tt,\,S^{1D},\,b^D,\,loop^D}\hat{)} \\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{1S}}\in \hat{(}\hat{x}\in\hat{S^1}\hat{)}\hat{\,\ra\,}\hat{S^{1D}\,x}\hat{)}\timesh\hat{(}\hat{b^S}\in \hat{S^{1S}\,\,b = b^D}\hat{)}\\
   & && \hspace{0.75em}\timesh\hat{(}\hat{loop^S\,\,\in\,\,} \TR\,_{\hat{(}\hat{\lambda x.
-\TR\,_{S^{1M}}\,loop\,x\,=\,b^D}\hat{)}}\,\hat{b^S}\,\hat{(}\TR\,_{\hat{(}\hat{\lambda} \hat{x}.
-\TR\,_{\hat{S^{1M}}}\,\hat{loop}\,\hat{(}\hat{S^{1E}\,b}\hat{)}\,\eqh\,\hat{x}\hat{)}}\,\hat{b^S}\,\hat{(}\apd\,\hat{S^{1E}\,loop}\hat{)}\hat{)}\,\\
+\TR\,_{S^{1D}}\,loop\,x\,=\,b^D}\hat{)}}\,\hat{b^S}\,\hat{(}\TR\,_{\hat{(}\hat{\lambda} \hat{x}.
+\TR\,_{\hat{S^{1D}}}\,\hat{loop}\,\hat{(}\hat{S^{1S}\,b}\hat{)}\,\eqh\,\hat{x}\hat{)}}\,\hat{b^S}\,\hat{(}\apd\,\hat{S^{1S}\,loop}\hat{)}\hat{)}\,\\
   & && \hspace{5.5em} \hat{=\,loop^D}\hat{)}
@@
 \]
-Note that $\U$ and $\U[\sigma]$ are definitionally equal in the theory of
-signatures, where $\sigma$ is a parallel substitution. The translation of the
+Assume that $\sigma$ is a parallel substitution, and note that $\U$ and $\U[\sigma]$
+are definitionally equal in the theory of signatures. The translation of the
 latter would be the following (omitting many details, including the handling of
@@
       can we do so elsewhere? However, this would result in an unusual and very
-      inconvenient syntax of signatures, because we would need to weaken even
-      basic substitution rules.
+      inconvenient syntax of signatures, because we would need to weaken even basic
+      substitution rules. In contrast, weak $\beta$ for $\J$ seems harmless, because
+      we are not aware of any HIIT signature in the literature that involves
+      any $\J$ computation.
 \item Instead of interpreting signatures into an UIP-free type theory, we
@@
 seems elusive in conventional homotopy type theory, and we may need two-level
-type theories \cite{semisegal}, or even directed type-theories with native
+type theories \cite{semisegal}, or directed type-theories with native
 notions of higher categories \cite{nuyts2015towards, riehl2017type}.
-
-%% In this paper we specified HIITs and provide part of their semantics, by
-%% computing notions of recursion and induction. Pro
-
-%% In this paper we only \emph{specified} HIITs and characterized their induction
-%% principles. Proving their \emph{existence} is left as further work. This would
-%% likely involve reducing HIITs to basic building blocks such as W-types and
-%% quotient types.
-
-%% The theory of codes was defined as part of the syntax of another type theory,
-%% the target theory. An alternative way would be to define the theory of codes
-%% internally to a type theoretic metatheory in the style of
-%% \cite{ttintt}. However, as described in \cite[Section 6]{ttintt}, there would be
-%% a coherence problem: for the internal syntax to be useful, we need to truncate
-%% it to be a set (as the $\mathsf{trunc}$ constructor did for $\Int$ in the
-%% introduction). As the eliminator needs to respect the equality given by
-%% $\mathsf{trunc}$, we would only be able to eliminate from the internal type
-%% theory into a set. This would preclude the definition of even the $\blank^\C$
-%% operation, which corresponds to the standard model. A possible solution to this
-%% problem would be to add all the higher coherence laws to the syntax (e.g.\ the
-%% pentagon law for the composition of substitutions) and then prove that the
-%% syntax is a set. For this however, we would probably need a two-level metatheory
-%% as in \cite{semisegal}.
-
-%% When working in a metatheory with uniqueness of identity proofs (which implies
-%% that all HIITs are in essence QIITs), the theory of codes admits a category
-%% model where the interpretation of a context is the category of algebras
-%% corresponding to the context. In the future, we would like to prove that these
-%% categories have initial algebras given by terms in the theory of codes.
 
@@
 \begin{alignat*}{5}
-  & && \Gamma^\M\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\\
-  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2M}}\in \hat{S^2}\rah\Type_{\hat{i}}\hat{)} \\
-  & && \hspace{0.75em} \timesh\hat{(}\hat{b^M}\in \hat{S^{2M}\,b}\hat{)} \\
-  & && \hspace{0.75em} \timesh\hat{(}\hat{surf^M\in \TR_{\hat{(}\TR_{S^{2M}}\,\blank\,b^M=b^M\hat{)}}\,surf\,\refl_{b^M} = \refl_{b^M}}\hat{)}
+  & && \Gamma^\D\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2D}}\in \hat{S^2}\rah\Type_{\hat{i}}\hat{)} \\
+  & && \hspace{0.75em} \timesh\hat{(}\hat{b^D}\in \hat{S^{2D}\,b}\hat{)} \\
+  & && \hspace{0.75em} \timesh\hat{(}\hat{surf^D\in \TR_{\hat{(}\TR_{S^{2D}}\,\blank\,b^D=b^D\hat{)}}\,surf\,\refl_{b^D} = \refl_{b^D}}\hat{)}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & && \Gamma^\E\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\,\hat{(}\hat{\tt,\,S^{2M},\,b^M,\,surf^M}\hat{)} \\
-  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2E}}\in \hat{(}\hat{x}\in\hat{S^2}\hat{)}\hat{\,\ra\,}\hat{S^{2M}\,x}\hat{)} \\
-  & && \hspace{0.75em} \timesh\hat{(}\hat{b^E}\in \hat{S^{2E}\,\,b = b^M}\hat{)}\\
-  & && \hspace{0.75em} \timesh\hat{\bigg(}\hat{surf^E\,\,\in\,\,} \TR\,\hat{\Big(}\hat{\J\,\refl\,b^E}\hat{\Big)}\,\hat{\Big(}\TR\,\hat{\big(}\hat{\J\,\refl\,b^E}\hat{\big)}\,\hat{\big(}\hat{\apd}\,\hat{(}\hat{\lambda} \hat{x}.\TR\,\hat{b^E}\,\hat{(}\TR\,\hat{b^E}\,\hat{(}\hat{\apd}\,\hat{S^{2E}}\,\hat{x}\hat{)}\hat{)}\hat{)}\,\hat{surf}\hat{\big)}\hat{\Big)} \\
-  & && \hspace{5.5em} \hat{=} \,\hat{surf^M}\hat{\bigg)}
-\end{alignat*}
-Note that if $\hat{b^E}$ is a definitional equality (that is, we have
-$\hat{S^{2E}\,\,b} \equiv \hat{b^M}$), the occurrences of $\hat{b^E}$
-in the type of $\hat{surf^E}$ can be replaced by $\hat{\refl}$. In
-this case the type of $\hat{surf^E}$ becomes the expected
-$\hat{\apd}\,\hat{(}\hat{\apd}\,\hat{S^{2E}}\hat{)}\,\hat{surf}\,\hat{=}\,\hat{surf^M}$.
+  & && \Gamma^\S\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\,\hat{(}\hat{\tt,\,S^{2D},\,b^D,\,surf^D}\hat{)} \\
+  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2S}}\in \hat{(}\hat{x}\in\hat{S^2}\hat{)}\hat{\,\ra\,}\hat{S^{2D}\,x}\hat{)} \\
+  & && \hspace{0.75em} \timesh\hat{(}\hat{b^S}\in \hat{S^{2S}\,\,b = b^D}\hat{)}\\
+  & && \hspace{0.75em} \timesh\hat{\bigg(}\hat{surf^S\,\,\in\,\,} \TR\,\hat{\Big(}\hat{\J\,\refl\,b^S}\hat{\Big)}\,\hat{\Big(}\TR\,\hat{\big(}\hat{\J\,\refl\,b^S}\hat{\big)}\,\hat{\big(}\hat{\apd}\,\hat{(}\hat{\lambda} \hat{x}\hat{.}\TR\,\hat{b^S}\,\hat{(}\TR\,\hat{b^S}\,\hat{(}\hat{\apd}\,\hat{S^{2S}}\,\hat{x}\hat{)}\hat{)}\hat{)}\,\hat{surf}\hat{\big)}\hat{\Big)} \\
+  & && \hspace{5.5em} \hat{=} \,\hat{surf^D}\hat{\bigg)}
+\end{alignat*}
+Note that if $\hat{b^S}$ is a definitional equality (that is, we have
+$\hat{S^{2S}\,\,b} \hat{\equiv} \hat{b^D}$), the occurrences of $\hat{b^S}$
+in the type of $\hat{surf^S}$ can be replaced by $\hat{\refl}$. In
+this case the type of $\hat{surf^S}$ becomes the more usual
+$\hat{\apd}\,\hat{(}\hat{\apd}\,\hat{S^{2S}}\hat{)}\,\hat{surf}\,\hat{=}\,\hat{surf^D}$.
 

### 2019-02-04 paper.tex
remove duplicate note on journal version
IdU formalization
add syntax for =U

@@
 \title{Signatures and Induction Principles for Higher Inductive-Inductive Types}
-\titlecomment{{\lsuper*} This is a journal version of the paper \cite{hiit}.}
+%\titlecomment{{\lsuper*} This is a journal version of the paper \cite{hiit}.}
 
@@
 \vspace{0.5em}
-(4) Equality
+(4) Paths between point and path constructors
 
@@
 \infer{\targetass\Delta \vdash \J_{a\,t\,\,(x.z.p)}\,pr\,_u\,eq : \underline{p[x\mapsto u, z\mapsto eq]}}
-      {\begin{array}{l}
-          \targetass\Delta \vdash t : \underline{a} \\
-          \targetass\Delta,x:\underline{a},z:\underline{t=_a x}\vdash p : \U \\
-          \targetass\Delta \vdash pr : \underline{p[x\mapsto t, z\mapsto \refl]} \\
-          \targetass\Delta \vdash u : \underline{a} \\
-          \targetass\Delta \vdash eq : \underline{t=_a u}
+      {\begin{array}{l l l}
+          & \targetass\Delta,x:\underline{a},z:\underline{t=_a x}\vdash p : \U & \targetass\Delta \vdash u : \underline{a} \\
+          \targetass\Delta \vdash t : \underline{a} \hspace{1.5em} & \targetass\Delta \vdash pr : \underline{p[x\mapsto t, z\mapsto \refl]} \hspace{1.5em} & \targetass\Delta \vdash eq : \underline{t=_a u}
       \end{array}}
@@
 \vspace{0.5em}
-(5) External parameters
+(5) Paths between type constructors
+
+\[
+\begin{gathered}
+  \infer{\targetass\Delta \vdash a=_\U b}{\targetass\Delta \vdash a : \U && \targetass\Delta \vdash b : \U}
+\end{gathered}
+\hspace{2em}
+\begin{gathered}
+  \infer{\targetass\Delta \vdash \refl : a=_\U a}{\targetass\Delta \vdash a : \U}
+\end{gathered}
+\]
+
+\[
+\infer{\targetass\Delta \vdash \J_{a\,\,(x.z.p)}\,pr\,_b\,eq : \underline{p[x\mapsto b, z\mapsto eq]}}
+      {\begin{array}{l l l}
+          & \targetass\Delta,x:\U,z:a=_\U x\vdash p : \U & \targetass\Delta \vdash b : \U \\
+          \targetass\Delta \vdash a : \U \hspace{1.5em} & \targetass\Delta \vdash pr : \underline{p[x\mapsto a, z\mapsto \refl]} \hspace{1.5em} & \targetass\Delta \vdash eq : a=_\U b
+      \end{array}}
+      \]
+
+\[
+\infer{\targetass\Delta \vdash \J\beta_{a\,\,(x.z.p)}\,pr:\underline{(\J_{a\,\,(x.z.p)}\,pr\,_a\,\refl) =_{p[x\mapsto a, z\mapsto \refl]} pr}}
+      {\targetass\Delta \vdash a : \U
+        && \targetass\Delta,x:\U,z:t =_\U x\vdash p : \U
+        && \targetass\Delta \vdash pr : \underline{p[x\mapsto a, z\mapsto \refl]}
+      }
+\]
+\caption{The theory of HIIT signatures, parts (1)--(5). Weakenings are
+  implicit, we assume fresh names everywhere and consider
+  $\alpha$-convertible terms equal. The $\hat{\Gamma};$ assumptions
+  are not used or changed in parts (1)--(5), only in parts (6)--(7),
+  see Figure \ref{sigrules2}.}
+\label{sigrules1}
+\end{figure}
+
+\begin{figure}
+
+(6) External parameters
 
@@
 \end{gathered}
-\]
-
-\[
+\hspace{2em}
 \begin{gathered}
@@
 \vspace{0.5em}
-(6) Infinitary parameters
+(7) Infinitary parameters
 
@@
 \]  
-\caption{The theory of HIIT signatures. Weakenings are implicit, we assume fresh
-  names everywhere and consider $\alpha$-convertible terms equal. The
-  $\hat{\Gamma};$ assumptions are not used or changed in parts
-  (1)--(4).}
-\label{sigrules}
+\caption{The theory of HIIT signatures, parts (6)--(7). Parts (1)--(5)
+  are given in Figure \ref{sigrules1}.}
+\label{sigrules2}
 \end{figure}
 
-We list typing rules for the theory of signatures in Figure \ref{sigrules}. We consider the
+We list typing rules for the theory of signatures in Figures \ref{sigrules1} and \ref{sigrules2}. We consider the
 following judgments:
@@
 \begingroup
-%% \allowdisplaybreaks
+\allowdisplaybreaks
 \begin{alignat*}{5}
@@
 $\hat{\Gamma}\vdash\Delta$ signature. Then, it checks the input with respect to
-the rules in Figure \ref{sigrules}, and outputs an Agda-checkable file which
+the rules in Figures \ref{sigrules1} and \ref{sigrules2}, and outputs an Agda-checkable file which
 contains algebras, homomorphisms, displayed algebras and sections for the

### 2019-02-04 paper.tex
small change
stuff

--- 2019-02-01 paper.tex
@@
 The present paper is an expanded version of our conference paper \cite{hiit}.
-In this version, we additionally compute notions of homomorphisms from
-signatures, and compare them to logical relations, in Section
-\ref{sec:morphisms}. We also explain a coherence problem in interpreting
-syntaxes of type theories, and how this influenced the current paper, in Section
+In this version, we extend signatures with paths between type constructors, and
+in Section \ref{sec:morphisms} we also compute notions of homomorphisms from
+signatures. We also explain a coherence problem in interpreting syntaxes of type
+theories, and how it influenced the current paper, in Section
 \ref{sec:coherence} and Section \ref{sec:categorical}.
@@
 Cartmell's generalized algebraic theories (GATs) \cite{gat} pioneered a
-type-theoretic notion of algebraic signature. GATs can be viewed as finitary
-quotient inductive-inductive signatures (QIITs), although GATs support equalities
-between inductive types (sorts) as well, which have not been so far explored in
-other works about QIITs and HIITs.
+type-theoretic notion of algebraic signature. GATs can be viewed as a variant of
+finitary quotient inductive-inductive signatures (QIITs), although GATs also
+support equations between sorts (type constructors), which so far have not been
+considered in QIIT and HIT literature.
 
@@
 
+(5) There is a different equality type former, which can be used to express
+paths between type constructors. To our knowledge, this has not been considered
+so far in HIT literature, although Cartmell considered equations between sorts
+in generalized algebraic theories \cite{gat}. For an example, paths between type
+constructors allow a compact definition of integers:
+\begin{alignat*}{5}
+  & \boldsymbol{\cdot},\,\,Int : U,\,\,zero : \underline{Int},\,p : Int =_\U Int
+\end{alignat*}
+Here, successor and predecessor functions can be recovered from the $p : Int
+=_\U Int$ equality, by transporting an $Int$ along $p$ or $p^{-1}$. This
+signature could be unfolded to a larger one, by replacing $p : Int =_\U Int$
+with an explicit $suc : Int \ra \underline{Int}$ constructor and additional
+constructors expressing that $suc$ is an equivalence. The definition of $Int$
+with a successor equivalence was previously suggested by Altenkirch, Pinyo
+[TODO] and Cavallo [TODO]. Our signature can be viewed as a shorthand for the
+unfolded version.
+
 So far we were only able to define closed HIITs, which do not refer to external
@@
 
-(5) We have a function space where the domain is a type in the external
+(6) We have a function space where the domain is a type in the external
 theory. We distinguish it from (3) by using red brick color in the domain
-specification. We specify lists and the integers as follows.
+specification. We specify lists and integers as follows, with integers now
+given as quotients of pairs of natural numbers:
 \begin{alignat*}{4}
@@
 
-(6) $\U$ is also closed under a function space where the domain is an external
+(7) $\U$ is also closed under a function space where the domain is an external
 type and the codomain is a small source theory type. We overload the application
@@
   & && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}\hat{.}\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\\
-  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,(eq^\A\,\hat{\gamma})\hat{\Big)}\,(u^\S\,\hat{\gamma^S})\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^S}\hat{)} \\
+  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\Big)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^S}\hat{)} \\
   & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \\

### 2019-02-04 paper.tex

@@
 \vspace{0.5em}
-(5) External parameters
+(5) Paths between type constructors
+
+\[
+\begin{gathered}
+  \infer{\targetass\Delta \vdash a=_\U b}{\targetass\Delta \vdash a : \U && \targetass\Delta \vdash b : \U}
+\end{gathered}
+\hspace{2em}
+\begin{gathered}
+  \infer{\targetass\Delta \vdash \refl : a=_\U a}{\targetass\Delta \vdash a : \U}
+\end{gathered}
+\]
+
+\[
+\infer{\targetass\Delta \vdash \J_{a\,\,(x.z.p)}\,pr\,_b\,eq : \underline{p[x\mapsto b, z\mapsto eq]}}
+      {\begin{array}{l l l}
+          & \targetass\Delta,x:\U,z:a=_\U x\vdash p : \U & \targetass\Delta \vdash b : \U \\
+          \targetass\Delta \vdash a : \U \hspace{1.5em} & \targetass\Delta \vdash pr : \underline{p[x\mapsto a, z\mapsto \refl]} \hspace{1.5em} & \targetass\Delta \vdash eq : a=_\U b
+      \end{array}}
+      \]
+
+\[
+\infer{\targetass\Delta \vdash \J\beta_{a\,\,(x.z.p)}\,pr:\underline{(\J_{a\,\,(x.z.p)}\,pr\,_a\,\refl) =_{p[x\mapsto a, z\mapsto \refl]} pr}}
+      {\targetass\Delta \vdash a : \U
+        && \targetass\Delta,x:\U,z:t =_\U x\vdash p : \U
+        && \targetass\Delta \vdash pr : \underline{p[x\mapsto a, z\mapsto \refl]}
+      }
+\]
+\caption{The theory of HIIT signatures, parts (1)--(5). Weakenings are
+  implicit, we assume fresh names everywhere and consider
+  $\alpha$-convertible terms equal. The $\hat{\Gamma};$ assumptions
+  are only used in parts (6)--(7), see Figure \ref{sigrules2}.}
+\label{sigrules1}
+\end{figure}
+
+\begin{figure}
+
+(6) External parameters
 

### 2019-02-04 paper.tex
added =U to ^A operation

--- 2019-02-04 paper.tex
@@
   $\alpha$-convertible terms equal. The $\hat{\Gamma};$ assumptions
-  are not used or changed in parts (1)--(5), only in parts (6)--(7),
-  see Figure \ref{sigrules2}.}
+  are only used in parts (6)--(7), see Figure \ref{sigrules2}.}
 \label{sigrules1}
@@
   & (t\,u)^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\A\,\hat{\gamma}\hat{)} \\
-  & ((\hat{x}\in \hat{A})\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A})\rah  B^\A\,\hat{\gamma} \\
-  & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u} \\
   & (t=_a u)^\A\,\hat{\gamma} && :\equiv t^\A\,\hat{\gamma} \eqh u^\A\,\hat{\gamma} \\
@@
   & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & (a=_\U b)^\A\,\hat{\gamma} && :\equiv a^\A\,\hat{\gamma} \eqh b^\A\,\hat{\gamma} \\
+  & \refl^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\A\,\hat{\gamma} && :\equiv \hat{\J}_{\Type_{\hat{0}}\,\hat{(}a^\A\,\hat{\gamma}\hat{)}\,\hat{(}\hat{\lambda} \hat{x}\,\hat{z}.p^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\commah\hat{z}\hat{)}\hat{)}}\,\hat{(}pr^\A\,\hat{\gamma}\hat{)}\,_{\hat{(}b^\A\,\hat{\gamma}\hat{)}}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)} \\
+  & (\J\beta_{a\,(x.z.p)}\,pr)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & ((\hat{x}\in \hat{A})\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\A\,\hat{\gamma} \\
+  & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u} \\
   & ((\hat{x}\in \hat{A})\ra b)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^\A\,\hat{\gamma} \\

### 2019-02-04 paper.tex
=U for ^D

@@
   & (t\,u)^\D\,\hat{\gamma^D} && :\equiv \hat{(}t^\D\,\hat{\gamma^D}\hat{)}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\D\,\hat{\gamma^D}\hat{)} \\
-  & ((\hat{x}\in \hat{A})\ra B)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^\D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u} \\
   & (t=_a u)^\D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}}\,\hat{e}\,\hat{(}t^\D\,\hat{\gamma^D}\hat{)} \eqh u^\D\hat{\gamma^D} \\
@@
   & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
+  & (a=_\U b)^\D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(\lambda A.A\ra\Type_0)}}\,\hat{e}\,\hat{(}a^\D\,\hat{\gamma^D}\hat{)} \eqh b^\D\hat{\gamma^D} \\
+  & (\refl_a)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl}_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}} \\
+  & (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\D\,\hat{\gamma^D} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^D}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^D}\hat{)} \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
+  & ((\hat{x}\in \hat{A})\ra B)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^\D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u} \\
   & ((\hat{x}\in \hat{A})\ra b)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\

### 2019-02-04 paper.tex
stuff

--- 2019-02-04 paper.tex
@@
 \usepackage{hyperref}
+\usepackage{url}
 
@@
 constructors expressing that $suc$ is an equivalence. The definition of $Int$
-with a successor equivalence was previously suggested by Altenkirch, Pinyo
-[TODO] and Cavallo [TODO]. Our signature can be viewed as a shorthand for the
-unfolded version.
+with a successor equivalence was previously suggested by Altenkirch and Pinyo
+\cite{pinyo2018integers} and Cavallo and M\"ortberg \cite{cavallointegers}. Our
+signature can be viewed as a shorthand for the unfolded version.
+
+For another example, we may have a HIIT definition for a type theory where
+Russell-style universes are compactly specified with an equality:
+\begin{alignat*}{5}
+  & ...,\,\, russell : (\Gamma : Con)\ra Tm\,\Gamma\,U =_\U Ty\,\Gamma,\,\, ...
+\end{alignat*}
 

### 2019-02-04 paper.tex
=U for ^M

--- 2019-02-04 paper.tex
@@
 
-We will use the following functions defined using $\hat{\J}$ in the standard
-way. We write $\TR_{\hat{P}}\,\hat{e}\,\hat{t}\in \hat{P}\,\hat{v}$ for
-transport of $\hat{t} \in \hat{P}\,\hat{u}$ along $\hat{e} \in \hat{u}\eqh
-\hat{v}$. We write
+We will use the following functions defined using $\hat{\J}$ in the
+standard way. We write $\TR_{\hat{P}}\,\hat{e}\,\hat{t}\in
+\hat{P}\,\hat{v}$ for transport of $\hat{t} \in \hat{P}\,\hat{u}$
+along $\hat{e} \in \hat{u}\eqh \hat{v}$ and $\hat{\coe\,e\,t : B}$ for
+coercion of $\hat{t:A}$ along $\hat{e : A =_\Type B}$. We write
 $\ap\,\hat{f}\,\hat{e}\in\hat{f}\,\hat{u}\eqh\hat{f}\,\hat{v}$ where
 $\hat{f}:\hat{A}\rah\hat{B}$ and $\hat{e}\in\hat{u}\eqh\hat{v}$, also
-$\hat{\apd\,\hat{f}\,\hat{e} \in \TR_{\hat{P}}\,\hat{e}\,(\hat{f}\,\hat{u}) \eqh
-  \hat{f}\,\hat{v}}$ where $\hat{\hat{f}\in(\hat{x}\in \hat{A})\rah \hat{B}}$
-and $\hat{e} \in \hat{u}\eqh \hat{v}$. Borrowing notation from the homotopy type
-theory book \cite{HoTTbook}, we write $\hat{p \sqcdot q}$ for transitivity and
-$\hat{p^{-1}}$ for symmetry. We also make use of the groupoid law $\hat{\inv\,p
-  : p^{-1}\sqcdot p = \refl}$.
+$\hat{\apd\,\hat{f}\,\hat{e} \in
+  \TR_{\hat{P}}\,\hat{e}\,(\hat{f}\,\hat{u}) \eqh \hat{f}\,\hat{v}}$
+where $\hat{\hat{f}\in(\hat{x}\in \hat{A})\rah \hat{B}}$ and $\hat{e}
+\in \hat{u}\eqh \hat{v}$. Borrowing notation from the homotopy type
+theory book \cite{HoTTbook}, we write $\hat{p \sqcdot q}$ for
+transitivity and $\hat{p^{-1}}$ for symmetry. We also make use of the
+groupoid law $\hat{\inv\,p : p^{-1}\sqcdot p = \refl}$.
 
@@
 \begin{alignat*}{5}
-  & (\underline{a})^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,t_0\,\,t_1} && :\equiv \hat{
-    \blc{a^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,t_0 = t_1}
+  & (\underline{a})^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,t_0\,\,t_1} && :\equiv \hat{(
+    \blc{a^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,t_0 = t_1)}
 \end{alignat*}
@@
 
-Let us consider now the translation in with the
+Let us consider now the translation in the
 corresponding induction principles in mind, defined as $\mathsf{Induction}$ in
@@
 
-We list below the $\blank^\M$ interpretations for the syntax. For brevity, we
+The specification of the $\blank^\M$ interpretation is as following.
+\[
+\infer{\hat{\Gamma}\vdashh\Delta^\M \in \Delta^\A\rah\Delta^\A\rah \Type_{\hat{i}}}
+      {\hat{\Gamma}\vdash\Delta}
+\hspace{1em}
+\infer{
+  \hat{\Gamma}\vdashh A^\D \in \hat{(}\hat{\gamma_0\,\gamma_1}\in\Delta^\A\hat{)}\rah \Delta^\M\,\hat{\gamma_0\,\gamma_1}\rah  A^\A\,\hat{\gamma_0}\rah A^\A\,\hat{\gamma_1}\rah \Type_{\hat{i}}}
+      {\hat{\Gamma}\semicol\Delta\vdash A}
+\]
+
+\[
+\infer{
+  \hat{\Gamma}\vdashh t^\M\, \hat{ : (\gamma_0\,\gamma_1 : \blc{\Delta^\A})(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\ra \blc{A^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,(\blc{t^\A}\,\gamma_0)\,(\blc{t^\A}\,\gamma_1)
+      }}
+      {\hat{\Gamma}\semicol\Delta\vdash t : A}
+\]
+
+We list below the implementation of $\blank^\M$ on the full syntax. For brevity, we
 mostly leave $\hat{\gamma_0}$ and $\hat{\gamma_1}$ implicit, notating only
@@
   & (t\,u)^\M\,\hat{\gamma^M} && :\equiv  \hat{\J\,(\blc{t^\M}\,\gamma^M\,(\blc{u^\A}\,\gamma_0))\,(\blc{u^\M}\,\gamma^\M) }\\
-  & ((\hat{x:A})\ra B)^\M\,\hat{\gamma^M\,f_0\,f_1} && :\equiv \hat{(x : A)\ra \blc{B^\M}\,\gamma^M\,(f_0\,x)\,(f_1\,x)} \\
-  & (t\,\hat{u})^\M\,\hat{\gamma^M} && :\equiv t^\M\,\hat{\gamma^M\,u} \\
   & (t=_a u)^\M\,\hat{\gamma^M} && :\equiv
@@
     }\\
+  & (a=_\U b)^\M\,\hat{\gamma^M} && :\equiv
+    \hat{\lambda\,e_0\,e_1.\,(\lambda x.\blc{b^\M}\,\gamma^M\,(\coe\,e_0\,x)) = (\lambda x.\coe\,e_1\,(\blc{a^\M}\,\gamma^M\,x))}\\
+  & (\refl_a)^\M\,\hat{\gamma^M} && :\equiv \hat{\refl} \\
+  & (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\M\,\hat{\gamma^M} && :\equiv\\
+  & && \hspace{-10em}\hat{\J\,\bigg(\lambda b^M\,eq^M.\J\,\Big(\lambda b^M\,eq^M. \J\,(\lambda P^M\,pr^M.pr^M)\,eq^M\,(\mathsf{uncurry}\,\blc{p^\M}\,\gamma^M)\,(\blc{pr^\M}\,\gamma^M)\Big)}  \\
+  & && \hspace{-9em}\hat{\,(\blc{eq^\A}\,\gamma_1)\,b^M\,eq^M\bigg)\,(\blc{eq^\M}\,\gamma_0)\,(\blc{b^\M}\,\gamma^M)\,(\blc{eq^\M}\,\gamma^M) } \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\M\,\hat{\gamma^M} && :\equiv \hat{\inv\,(\blc{pr^\M}\,\gamma^M)} \\
+  & ((\hat{x:A})\ra B)^\M\,\hat{\gamma^M\,f_0\,f_1} && :\equiv \hat{(x : A)\ra \blc{B^\M}\,\gamma^M\,(f_0\,x)\,(f_1\,x)} \\
+  & (t\,\hat{u})^\M\,\hat{\gamma^M} && :\equiv t^\M\,\hat{\gamma^M\,u} \\
   & ((\hat{x}\in \hat{A})\ra b)^\M\,\hat{\gamma^M} && :\equiv

### 2019-02-04 paper.tex
stuff

--- 2019-02-04 paper.tex
@@
 
-The translation for the equality type $t=_a u$, for each $\hat{e}\in(t=_a
-u)^\A\,\hat{\gamma}$, i.e.\ $\hat{e}\in t^\A\,\hat{\gamma}\eqh
+The translation for paths between $a : \U$ elements $t=_a u$, for each
+$\hat{e}\in(t=_a u)^\A\,\hat{\gamma}$, i.e.\ $\hat{e}\in t^\A\,\hat{\gamma}\eqh
 u^\A\,\hat{\gamma}$, witnesses that $t^\D$ and $u^\D$ are equal. As these have
@@
 notation; see the formalization (Section \ref{sec:formalization}) for details.
+
+For $a =_\U b$ paths, the interpretation is essentially the same as with the
+other path type.
 
@@
 
-Here, we only provide abbreviated definitions for the $t\,u$, $t=_a u$, $\refl$,
-$\J$ and $\J\beta$ cases. In the $\J$ case, we write
-$\hat{\mathsf{uncurry}}\,p^\D$ for
-$\hat{\lambda\,\gamma\,\gamma^D\,x\,x^D\,z\,z^D.}\,p^\D\,\hat{(}\hat{\gamma,x,z}\hat{)}\,\hat{(}\hat{\gamma^D,x^D,z^D}\hat{)}$
+In the case of $t =_a u$ equalities, we only provide abbreviated definitions for
+the $t\,u$, $t=_a u$, $\refl$, $\J$ and $\J\beta$ cases. In the $\J$ case, we
+write $\hat{\mathsf{uncurry}}\,p^\D$ for
+$\hat{\lambda\,\gamma\,\gamma^D\,x\,x^D\,z\,z^D.}$ $\,p^\D\,\hat{(}\hat{\gamma,x,z}\hat{)}\,\hat{(}\hat{\gamma^D,x^D,z^D}\hat{)}$
 and analogously elsewhere, to adjust for the fact that $p$ abstracts over
-additional $x$ and $z$ variables. The full definitions can be found in the
-Agda formalization. The definitions are highly constrained by the
-required types, and not particularly difficult to implement with the
-help of a proof assistant: they all involve doing successive path
-induction on all equalities available from induction hypotheses, with
-appropriately generalized induction motives.
+additional $x$ and $z$ variables. The full definitions can be found in the Agda
+formalization. The definitions are highly constrained by the required types, and
+not particularly difficult to implement with the help of a proof assistant: they
+all involve doing successive path induction on all equalities available from
+induction hypotheses, with appropriately generalized induction motives.
 
@@
 \end{alignat*}
+
 In homotopy type theory, the $\beta$-rule for $loop$ is usually just
@@
 expression in the result.
+
+For $a =_\U b$ equalities, let us consider here only the $(a =_\U b)^\S$ case
+for the type former. The interpretation here is inspried by the univalence
+axiom, although the translation does not require it. If we assume univalence,
+then $\hat{a =_{\Type_i} b}$ equalities in the external theory are equivalent to
+equivalences, which contain a function $\hat{f : a \ra b}$ and a proof
+$\hat{\mathsf{isEqv}\,f}$. Since $\hat{\mathsf{isEqv}}$ is propositional (proof
+irrelevant), we only need to say something about the underlying $\hat{f}$
+function in the translation.
+
+Let us consider $\blank^\S$ for higher inductive integers with the type
+equality:
+\begin{alignat*}{5}
+  & && (\boldsymbol{\cdot},\,\,Int : U,\,\,zero : \underline{Int},\,p : Int =_\U Int)^\S\,
+     \hat{(\tt,\,Int,\,z,\,p)\,(\tt,\,Int^D,\,z^D,\,p^D)} \\
+  & \equiv \,\, && \hat{
+     \top \times (Int^S : (x : Int)\ra Int^D\,x) \times (z^S : Int^S\,z = z^D)}\\
+  & && \hat{\hspace{1em}\times\,
+       (p^S : (\lambda\,x.\,p^s\,(\coe\,p\,x)) = (\lambda\,x.\,\TR\,p^D\,(\J\,(p^s\,x)\,p)))}
+\end{alignat*}
 

### 2019-02-04 paper.tex
stuff

@@
 
-\item\emph{Iterated equalities and path induction in signatures.}
-We can support this relatively easily by closing the universe under equality
-type formation, and also using a standard (although size-restricted) definition
-of path induction.
+\item\emph{Iterated equalities and type constructor equalities with path
+indcution.} We can support iterated equalities of point and path constructors
+by closing the universe under equality type formation of point and path
+constructors, together with a standard (although size-restricted) definition
+of path induction. We can also introduce a different type former for
+equalities between type constructors.
+
 \end{itemize}
@@
 paths between type constructors. To our knowledge, this has not been considered
-so far in HIT literature, although Cartmell considered equations between sorts
+in previous HIT literature, although Cartmell considered equations between sorts
 in generalized algebraic theories \cite{gat}. For an example, paths between type
@@
 \]
-The smallness of $\hat{A}$ is required in (5). It is possible to generalize
+The smallness of $\hat{A}$ is required in (6). It is possible to generalize
 signatures to arbitrary universe levels, but it is not essential to the current
-development. Note that the (5) function space preserves strict positivity, since
+development. Note that the (6) function space preserves strict positivity, since
 in the external theory there is no way to recursively refer to the inductive type
@@
 \]
-Here, $leaf$ has a function type (5) and $node$ has a function type (3) with a
-function type (6) in the domain. More generally, we can define $W$-types
+Here, $leaf$ has a function type (6) and $node$ has a function type (3) with a
+function type (7) in the domain. More generally, we can define $W$-types
 \cite{abbot05containers} as follows. $\hat{S}$ describes the ``shapes'' of the
@@
   & (t\,u)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^S}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)} \\
-  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv t^\S\,\hat{\gamma^S}\,\hat{u} \\
   & (t=_a u)^\S\,\hat{\gamma^S}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^S}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
@@
   & && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\\
+  & (a=_\U b)^\S\,\hat{\gamma^S} && :\equiv
+    \hat{\lambda\,e\,e^D.\,(\lambda\,x.\, \blc{b^\S}\,\gamma^S\, (\coe\,e\,x))
+     =
+    (\lambda\,x.\, \TR\,e^D\,(\J\,(\blc{a^\S}\,\gamma^S\,x)\,e))}\\
+  & (\refl_a)^\S\,\hat{\gamma^S} && :\equiv \hat{\refl} \\
+  & (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\S\,\hat{\gamma^S} && :\equiv\\
+  & && \hspace{-10em}\hat{\J\,\bigg(\lambda\,b^D\,\,b^S\,eq^D.\,\J\,\Big(\lambda\,b^S\,eq^S.\,\J\,(\lambda P^S\,pr^S.\,pr^S)\,eq^S\,(\mathsf{uncurry}\,\blc{p^\S}\,\gamma^S)\,(\blc{pr^\S}\,\gamma^S)\Big)}  \\
+  & && \hspace{-9em}\hat{\,(\blc{eq^\D}\,\gamma^D)\,b^S\bigg)\,(\blc{eq^\A}\,\gamma)\,(\blc{b^\D}\,\gamma^D)\,(\blc{b^\S}\,\gamma^S)\,(\blc{eq^\D}\,\gamma^D)\,(\blc{eq^\S}\,\gamma^S) } \\
+  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \hat{\J\,\refl\,(\blc{pr^\S}\,\gamma^S)} \\
+  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
+  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv t^\S\,\hat{\gamma^S}\,\hat{u} \\
   & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^S\,f\,t} && :\equiv b^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
@@
 
-For $a =_\U b$ equalities, let us consider here only the $(a =_\U b)^\S$ case
-for the type former. The interpretation here is inspried by the univalence
-axiom, although the translation does not require it. If we assume univalence,
-then $\hat{a =_{\Type_i} b}$ equalities in the external theory are equivalent to
-equivalences, which contain a function $\hat{f : a \ra b}$ and a proof
-$\hat{\mathsf{isEqv}\,f}$. Since $\hat{\mathsf{isEqv}}$ is propositional (proof
-irrelevant), we only need to say something about the underlying $\hat{f}$
-function in the translation.
-
-Let us consider $\blank^\S$ for higher inductive integers with the type
-equality:
+For $a =_\U b$ equalities, let us examine here only the $(a =_\U b)^\S$ case.
+The interpretation here is inspired by the univalence axiom, although the
+translation does not require it. If we assume univalence in the external theory,
+then $\hat{a =_{\Type_i} b}$ equalities in are equivalent to equivalences, which
+contain $\hat{f : a \ra b}$ functions together with $\hat{\mathsf{isEqv}\,f}$
+proofs. Hence, we can view the $\hat{e : \blc{a^\A}\,\gamma =
+  \blc{a^\A}\,\gamma}$ and $\hat{e^D : \TR_{(\lambda\,A.\,A \ra
+    \mathsf{Type_0})}\,e\,(\blc{a^\D}\,\gamma^D) = \blc{b^\D}\,\gamma^D}$ proofs
+as function bundled together with witnesses of equivalence. Since
+$\hat{\mathsf{isEqv}}$ is proof irrelevant, we only need to relate the
+underlying functions. For $\hat{e}$, the underlying function is given just by
+$\hat{\coe\,e}$. For $\hat{e^D}$, the underlying function is a bit more
+complicated, since $\hat{e^D}$ is a path over a path. We can observe this in the
+example for higher inductive integers:
 \begin{alignat*}{5}
@@
   & && \hat{\hspace{1em}\times\,
-       (p^S : (\lambda\,x.\,p^s\,(\coe\,p\,x)) = (\lambda\,x.\,\TR\,p^D\,(\J\,(p^s\,x)\,p)))}
-\end{alignat*}
-
-In Appendix \ref{sec:app}, we show how the type of displayed algebra sections is
-computed for the two-dimensional sphere. In Appendix \ref{sec:wtypes}, we show
-the same for indexed W-types.
+       (p^S : (\lambda\,x.\,p^S\,(\coe\,p\,x)) = (\lambda\,x.\,\TR\,p^D\,(\J\,(p^S\,x)\,p)))}
+\end{alignat*}
+
+In Appendix \ref{sec:app}, we additionally show how the type of displayed
+algebra sections is computed for the two-dimensional sphere. In Appendix
+\ref{sec:wtypes}, we show the same for indexed W-types.
 
@@
   & (a=_\U b)^\M\,\hat{\gamma^M} && :\equiv
-    \hat{\lambda\,e_0\,e_1.\,(\lambda x.\blc{b^\M}\,\gamma^M\,(\coe\,e_0\,x)) = (\lambda x.\coe\,e_1\,(\blc{a^\M}\,\gamma^M\,x))}\\
+    \hat{\lambda\,e_0\,e_1.\,(\lambda\,x.\,\blc{b^\M}\,\gamma^M\,(\coe\,e_0\,x)) = (\lambda\,x.\,\coe\,e_1\,(\blc{a^\M}\,\gamma^M\,x))}\\
   & (\refl_a)^\M\,\hat{\gamma^M} && :\equiv \hat{\refl} \\
   & (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\M\,\hat{\gamma^M} && :\equiv\\
-  & && \hspace{-10em}\hat{\J\,\bigg(\lambda b^M\,eq^M.\J\,\Big(\lambda b^M\,eq^M. \J\,(\lambda P^M\,pr^M.pr^M)\,eq^M\,(\mathsf{uncurry}\,\blc{p^\M}\,\gamma^M)\,(\blc{pr^\M}\,\gamma^M)\Big)}  \\
-  & && \hspace{-9em}\hat{\,(\blc{eq^\A}\,\gamma_1)\,b^M\,eq^M\bigg)\,(\blc{eq^\M}\,\gamma_0)\,(\blc{b^\M}\,\gamma^M)\,(\blc{eq^\M}\,\gamma^M) } \\
+  & && \hspace{-10em}\hat{\J\,\bigg(\J\,\Big(\lambda\,b^M\,eq^M.\,\J\,(\lambda\,P^M\,pr^M.\,pr^M)\,eq^M\,(\mathsf{uncurry}\,\blc{p^\M}\,\gamma^M)\,(\blc{pr^\M}\,\gamma^M)\Big)}  \\
+  & && \hspace{-9em}\hat{\,(\blc{eq^\A}\,\gamma_1)\bigg)\,(\blc{eq^\A}\,\gamma_0)\,(\blc{b^\M}\,\gamma^M)\,(\blc{eq^\M}\,\gamma^M) } \\
   & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\M\,\hat{\gamma^M} && :\equiv \hat{\inv\,(\blc{pr^\M}\,\gamma^M)} \\

### 2019-02-05 paper.tex
Fixes

@@
 possibly infinitary higher constructors of any dimension and restricts
-constructors to strictly positive ones. It also allows free usage of $\J$ (path
-induction) and $\refl$ in HIIT signatures, and allows mixing type, point and path
-constructors in any order.
+constructors to strictly positive ones. It also allows equalities between type
+constructors, free usage of $\J$ (path induction) and $\refl$ in HIIT
+signatures, and allows mixing type, point and path constructors in any order.
 
@@
 In this section we define signatures for HIITs. First, we list
-the main motivations behind our definition.
+the main considerations behind our definition.
 
@@
 
-\item\emph{Iterated equalities and type constructor equalities with path
-indcution.} We can support iterated equalities of point and path constructors
-by closing the universe under equality type formation of point and path
-constructors, together with a standard (although size-restricted) definition
-of path induction. We can also introduce a different type former for
-equalities between type constructors.
+\item\emph{Iterated equalities and type constructor equalities, with path
+induction.} We can support iterated equalities by closing the universe under
+equality type formation of point and path constructors, together with a
+standard (although size-restricted) definition of path induction. We can also
+introduce a different type former for equalities between type constructors.
 
@@
 
-\subsection{External type theory}
+\subsection{External Type Theory}
 \label{sec:external}
@@
 
-\subsection{Interpreting intrinsic syntax}
+\subsection{Interpreting Intrinsic Syntax}
 Following Altenkirch and Kaposi \cite{ttintt}, one might define the syntax of a
@@
 
-\subsection{Interpreting extrinsic syntax}
+\subsection{Interpreting Extrinsic Syntax}
 An extrinsic syntax for type theory is defined the following way:
@@
 
-\subsection{Syntactic translations}
+\subsection{Syntactic Translations}
 We can circumvent the coherence problem in the following way:
@@
 It returns a family over the type $\hat{S^1}$, an element of this family
-$\hat{b^D}$ at index $\hat{b}$, and and an equality between $\hat{b^D}$ and
+$\hat{b^D}$ at index $\hat{b}$, and a path between $\hat{b^D}$ and
 $\hat{b^D}$ which lies over $\hat{loop}$. This is the same as the usual
@@
 For $a =_\U b$ equalities, let us examine here only the $(a =_\U b)^\S$ case.
-The interpretation here is inspired by the univalence axiom, although the
-translation does not require it. If we assume univalence in the external theory,
-then $\hat{a =_{\Type_i} b}$ equalities in are equivalent to equivalences, which
+The interpretation is inspired by the univalence axiom, although the translation
+does not require it. If we assume univalence in the external theory, then
+$\hat{a =_{\Type_i} b}$ equalities in are equivalent to equivalences, which
 contain $\hat{f : a \ra b}$ functions together with $\hat{\mathsf{isEqv}\,f}$
 proofs. Hence, we can view the $\hat{e : \blc{a^\A}\,\gamma =
-  \blc{a^\A}\,\gamma}$ and $\hat{e^D : \TR_{(\lambda\,A.\,A \ra
+  \blc{b^\A}\,\gamma}$ and $\hat{e^D : \TR_{(\lambda\,A.\,A \ra
     \mathsf{Type_0})}\,e\,(\blc{a^\D}\,\gamma^D) = \blc{b^\D}\,\gamma^D}$ proofs
-as function bundled together with witnesses of equivalence. Since
+as functions bundled together with witnesses of equivalence. Since
 $\hat{\mathsf{isEqv}}$ is proof irrelevant, we only need to relate the
-underlying functions. For $\hat{e}$, the underlying function is given just by
-$\hat{\coe\,e}$. For $\hat{e^D}$, the underlying function is a bit more
-complicated, since $\hat{e^D}$ is a path over a path. We can observe this in the
-example for higher inductive integers:
+underlying functions in the translation. For $\hat{e}$, the underlying function
+is given just by $\hat{\coe\,e}$. For $\hat{e^D}$, the underlying function is a
+bit more complicated, since $\hat{e^D}$ is a path over a path. We can observe
+this in the example for higher inductive integers:
 \begin{alignat*}{5}
@@
 principle converted into a fully formal syntactic translation, because we prove
-all preservations of definitional equalities by $\refl$. Due to technical
-challenges, this formalization uses transport instead of $\J$ in the source
-theory, but this still covers a rather large class of HIIT definitions.
+all preservations of definitional equalities by $\refl$. Due to technical issues
+and performance issues in Agda, this deeper formalization uses transport instead
+of $\J$ for $t =_a u$ equalities, and it does not cover elimination for $a
+=_\U b$ equalities.
 
@@
   & (t=_a u)^\M\,\hat{\gamma^M} && :\equiv
-    \hat{\lambda\,e.\,(\blc{t^\M}\,\gamma^M\,^{-1})\sqcdot\ap\,(\blc{a^\M}\,\gamma^M)\,e\sqcdot\,\blc{u^\M}\,\gamma^M}\\
+    \hat{\lambda\,e.\,(\blc{t^\M}\,\gamma^M)^{-1}\sqcdot\ap\,(\blc{a^\M}\,\gamma^M)\,e\sqcdot\,\blc{u^\M}\,\gamma^M}\\
   & (\refl_t)^\M\,\hat{\gamma^M} && :\equiv \hat{\inv\,(\blc{t^\M}\,\gamma^M)}\\

### 2019-02-05 elims-demo/Impl.hs
sync with extended paper

@@ -73,40 +73,40 @@ data Val
   | VId (Maybe Val) Val Val
   | VJ Val ~Val Val
 
-data Code
-  = CVar Name
-  | CPiI Code (Bind Code)
-  | CAppI Code (Bind Code) Code Code -- app A (x.B) t u
-  | CPiNI Tm (Bind Code)
-  | CAppNI Code Tm
-  | CPiSNI Tm (Bind Code)
-  | CAppSNI Tm (Bind Code) Code Tm
-  | CId Code Code Code                        -- Id A t u
-  | CJ Code Code (Bind2 Code) Code Code Code  -- J A t P pr u eq
-  | CRefl Code Code
-  | CU
-  | CEl Code
+data Sig
+  = SVar Name
+  | SPiI Sig (Bind Sig)
+  | SAppI Sig (Bind Sig) Sig Sig -- app A (x.B) t u
+  | SPiNI Tm (Bind Sig)
+  | SAppNI Sig Tm
+  | SPiSNI Tm (Bind Sig)
+  | SAppSNI Tm (Bind Sig) Sig Tm
+  | SId Sig Sig Sig                        -- Id A t u
+  | SJ Sig Sig (Bind2 Sig) Sig Sig Sig  -- J A t P pr u eq
+  | SRefl Sig Sig
+  | SU
+  | SEl Sig
 
-data CVal
-  = CVVar Name
-  | CVPiI CVal (Bind (CVal → CVal))
-  | CVAppI CVal (Bind (CVal → CVal)) CVal ~CVal
-  | CVPiNI Val (Bind (Val → CVal))
-  | CVAppNI CVal ~Val
-  | CVPiSNI Val (Bind (Val → CVal))
-  | CVAppSNI Val (Bind (Val → CVal)) CVal ~Val
-  | CVId CVal CVal CVal                                       -- Id A t u
-  | CVJ CVal CVal (Bind2 (CVal → CVal → CVal)) CVal CVal CVal  -- J A t P pr u eq
-  | CVRefl CVal CVal
-  | CVU
-  | CVEl CVal
+data SVal
+  = SVVar Name
+  | SVPiI SVal (Bind (SVal → SVal))
+  | SVAppI SVal (Bind (SVal → SVal)) SVal ~SVal
+  | SVPiNI Val (Bind (Val → SVal))
+  | SVAppNI SVal ~Val
+  | SVPiSNI Val (Bind (Val → SVal))
+  | SVAppSNI Val (Bind (Val → SVal)) SVal ~Val
[…]

### 2019-08-13 paper.tex
vdash should be metatheoretic
add (uncommented) infinitary lambda
fix typo
Revisions addressing reviewer 2

@@
 \item\emph{Strict positivity.}
-In prior literature, schemes for inductive types usually include structural
-restrictions for this. In our case, a \emph{universe} can be used to make
-size restrictions which also entail strict positivity.
+In prior literature, schemes for inductive types usually include
+structural restrictions for this. In our case, a \emph{universe} is
+used to make size restrictions which also entail strict positivity.
 
 \item\emph{Iterated equalities and type constructor equalities, with path
-induction.} We can support iterated equalities by closing the universe under
+induction.} We support iterated equalities by closing the universe under
 equality type formation of point and path constructors, together with a
-standard (although size-restricted) definition of path induction. We can also
-introduce a different type former for equalities between type constructors.
+standard (although size-restricted) definition of path induction. We also
+introduce an additional type former for equalities between type constructors.
 
@@
   implicit, we assume fresh names everywhere and consider
-  $\alpha$-convertible terms equal. The $\hat{\Gamma};$ assumptions
+  $\alpha$-convertible terms equal. The $\hat{\Gamma;}$ assumptions
   are only used in parts (6)--(7), see Figure \ref{sigrules2}.}
@@
   \infer{\hat{\Gamma}\semicol\Delta \vdash t\, \hat{u} : \underline{b[\hat{x}\mapsto \hat{u}]}}
-        {\hat{\Gamma}\semicol\Delta \vdash t : \underline{(\hat{x}\in \hat{A})\ra b} && \hat{\Gamma}\vdash \hat{u} \in \hat{A}}
+        {\hat{\Gamma}\semicol\Delta \vdash t : \underline{(\hat{x}\in \hat{A})\ra b} && \hat{\Gamma}\,\hat{\vdash}\, \hat{u} \in \hat{A}}
 \end{gathered}
@@
 
-(2) There is a universe $\U$, with decoding written as an underline instead of
-usual $\El$, to improve readability. With this part of the syntax, we can already
-define contexts specifying the empty type, unit type and booleans, or in general,
-finite sets of finite sets:
+(2) There is a universe $\U$, with decoding written as an underline
+instead of the usual $\El$, to improve readability. With this part of
+the syntax, we can already define contexts specifying the empty type,
+unit type and booleans, or in general, finite sets of finite sets:
 \[
@@
 constructors, and arbitrary mixing of type and point constructors.
+
+We do not include a $\lambda$ in the syntax. The reason is twofold; first,
 
@@
   & (t=_a u)^\A\,\hat{\gamma} && :\equiv t^\A\,\hat{\gamma} \eqh u^\A\,\hat{\gamma} \\
-  & \refl^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & (\refl_t)^\S\,\hat{\gamma} && :\equiv \hat{\refl} \\
   & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\A\,\hat{\gamma} && :\equiv \hat{\J}_{\hat{(}a^\A\,\hat{\gamma}\hat{)}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}\hat{\lambda} \hat{x}\,\hat{z}.p^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\commah\hat{z}\hat{)}\hat{)}}\,\hat{(}pr^\A\,\hat{\gamma}\hat{)}\,_{\hat{(}u^\A\,\hat{\gamma}\hat{)}}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)} \\
@@
   & (a=_\U b)^\A\,\hat{\gamma} && :\equiv a^\A\,\hat{\gamma} \eqh b^\A\,\hat{\gamma} \\
-  & \refl^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & (\refl_a)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
   & (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\A\,\hat{\gamma} && :\equiv \hat{\J}_{\Type_{\hat{0}}\,\hat{(}a^\A\,\hat{\gamma}\hat{)}\,\hat{(}\hat{\lambda} \hat{x}\,\hat{z}.p^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\commah\hat{z}\hat{)}\hat{)}}\,\hat{(}pr^\A\,\hat{\gamma}\hat{)}\,_{\hat{(}b^\A\,\hat{\gamma}\hat{)}}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)} \\
@@
 \infer{
-  \hat{\Gamma}\vdashh A^\D \in \hat{(}\hat{\gamma_0\,\gamma_1}\in\Delta^\A\hat{)}\rah \Delta^\M\,\hat{\gamma_0\,\gamma_1}\rah  A^\A\,\hat{\gamma_0}\rah A^\A\,\hat{\gamma_1}\rah \Type_{\hat{i}}}
+  \hat{\Gamma}\vdashh A^\M \in \hat{(}\hat{\gamma_0\,\gamma_1}\in\Delta^\A\hat{)}\rah \Delta^\M\,\hat{\gamma_0\,\gamma_1}\rah  A^\A\,\hat{\gamma_0}\rah A^\A\,\hat{\gamma_1}\rah \Type_{\hat{i}}}
       {\hat{\Gamma}\semicol\Delta\vdash A}
@@
 does not require it. If we assume univalence in the external theory, then
-$\hat{a =_{\Type_i} b}$ equalities in are equivalent to equivalences, which
+$\hat{a =_{\Type_i} b}$ equalities are equivalent to equivalences, which
 contain $\hat{f : a \ra b}$ functions together with $\hat{\mathsf{isEqv}\,f}$
@@
   & && \hat{\hspace{1em}\times\,
-       (p^S : (\lambda\,x.\,p^S\,(\coe\,p\,x)) = (\lambda\,x.\,\TR\,p^D\,(\J\,(p^S\,x)\,p)))}
+       (p^S : (\lambda\,x.\,Int^S\,(\coe\,p\,x)) = (\lambda\,x.\,\TR\,p^D\,(\J\,(Int^S\,x)\,p)))}
 \end{alignat*}
@@
 \[
-(\U[\sigma])^{\mathsf{ID}}\,\hat{\gamma\,A} \equiv \hat{\TR_{(\lambda\,x. A\ra A)}\,(\sigma^{\mathsf{ID}}\,\gamma)\,(\lambda\,x.\, x)}
+(\U[\sigma])^{\mathsf{ID}}\,\hat{\gamma\,A} \equiv \hat{\TR_{(\lambda\,x. A\ra A)}\,(\blc{\sigma^{\mathsf{ID}}}\,\gamma)\,(\lambda\,x.\, x)}
 \]
@@
 
-There are numerous unavoidable instances of such failures of strict preservation
-of definitional equalities for $\blank^{\mathsf{ID}}$, and they similarly arise when we
-try to construct composition of homomorphisms. We consider three potential
-solutions for this strictness problem:
+Alternatively, we could try changing the $\blank^\mathsf{ID}$ interpretation of
+terms and substitutions to get \emph{definitional} equalities, instead of
+propositional equalities internal to the target syntax. This would not be a
+purely syntactic translation anymore, since for each
+$\hat{\Gamma}\semicol\Delta\vdash t : A$ we would get a universally quantified
+metatheoretic statement expressing that for each $\hat{\gamma}$ term,
+$\hat{\blc{t^\M}\,\gamma\,\gamma\,(\blc{\Delta^{\mathsf{ID}}}\,\gamma)}$ is
+definitionally equal to
+$\hat{\blc{A^{\mathsf{ID}}}\,\gamma\,(\blc{t^\A}\,\gamma)}$. This would solve
+the strictness problem in the case of $\U[\sigma]$, since there would be no need
+to transport over definitional equalities.
+
+Unfortunately, while this approach repairs $\blank^{\mathsf{ID}}$ in some cases,
+other cases become unfeasible, because we are not able to produce definitional
+equations. For example, the interpretation of the identity type would be as follows:
+\[
+(t =_a u)^{\mathsf{ID}}\,\hat{\gamma \,\,\blc{:}\,\, \blc{(t =_a u)^\M}\,\gamma\,\gamma\,(\blc{\Gamma^{\mathsf{ID}}}\,\gamma)\,(\blc{(t =_a u)^\A}\,\gamma)} \equiv \hat{\blc{\U^{\mathsf{ID}}}\,\gamma\,(\blc{(t =_a u)^\A}\,\gamma)}
+\]
+The goal type can be reduced along definitions, and using $t^{\mathsf{ID}}$ and $u^{\mathsf{ID}}$, to
+\[
+\hat{(\lambda\,e.\,\,\refl\,\sqcdot\,\ap\,(\lambda\,x.\,x)\,e\,\sqcdot\,\refl)\,\blc{\equiv}\,(\lambda\,e.\,e)}
+\]
+This definitional equality is clearly not provable, although $(t =_a
+u)^{\mathsf{ID}}$ does work when we interpret terms with weak internal
+equalities. Hence, there are strictness problems both with weak and strict
+equations for functoriality.
+
+We consider three potential solutions:
 \begin{enumerate}
@@
 \begin{alignat*}{5}
-  & && \Gamma^\C \equiv \hat{\top}\timesh\hat{(}\hat{S^2}\in\Type_{\hat{0}}\hat{)}\timesh\hat{(}\hat{b}\in \hat{S^2}\hat{)}\timesh\hat{(}\hat{surf\in \hat{\refl_{\hat{b}}}=_{\hat{(}\hat{b}=_{\hat{S^2}}\hat{b}\hat{)}}\hat{\refl_{\hat{b}}}}\hat{)}
-\end{alignat*}
-Given a sphere-algebra and fixing a universe level $i$, the motives
-and methods are computed as follows.
+  & && \Gamma^\A \equiv \hat{\top}\timesh\hat{(}\hat{S^2}\in\Type_{\hat{0}}\hat{)}\timesh\hat{(}\hat{b}\in \hat{S^2}\hat{)}\timesh\hat{(}\hat{surf\in \hat{\refl_{\hat{b}}}=_{\hat{(}\hat{b}=_{\hat{S^2}}\hat{b}\hat{)}}\hat{\refl_{\hat{b}}}}\hat{)}
+\end{alignat*}
+Given a sphere-algebra and fixing a universe level $\hat{i}$, the displayed algebra is the following.
 \begin{alignat*}{5}
@@
 Note that if $\hat{b^S}$ is a definitional equality (that is, we have
-$\hat{S^{2S}\,\,b} \hat{\equiv} \hat{b^D}$), the occurrences of $\hat{b^S}$
+$\hat{S^{2S}\,\,b}\,\hat{\equiv}\,\hat{b^D}$), the occurrences of $\hat{b^S}$
 in the type of $\hat{surf^S}$ can be replaced by $\hat{\refl}$. In
@@
 \]
-We pick a universe level $j$ for elimination. The interpretations of $W$ are the following, omitting leading $\hat{\top}$ components:
+We pick a universe level $j$ for elimination. The $\blank^\A$, $\blank^\D$ and $\blank^\S$ interpretations of $W$ are the following, omitting leading $\hat{\top}$ components:
 \begin{alignat*}{5}

### 2019-08-13 paper.tex
stuff

@@
 with a successor equivalence was previously suggested by Altenkirch and Pinyo
-\cite{pinyo2018integers} and Cavallo and M\"ortberg \cite{cavallointegers}. Our
-signature can be viewed as a shorthand for the unfolded version.
+\cite{pinyo2018integers} and Cavallo and M\"ortberg \cite{cavallointegers}.
 
@@
 \end{alignat*}
+
+In the presence of the univalence axiom, type constructor equations can be
+always equivalently represented using usual path constructors, by adding
+explicit equivalences to a signature. In this case, type constructor equations
+serve as a shorthand for equivalences.
 

### 2019-08-14 paper.tex
revision stuff

--- 2019-08-13 paper.tex
@@
 \noindent
-Higher inductive-inductive types (HIITs) generalize inductive types of dependent
-type theories in two ways. On the one hand they allow the simultaneous
-definition of multiple sorts that can be indexed over each other. On the other
-hand they support equality constructors, thus generalizing higher inductive
-types of homotopy type theory. Examples that make use of both features are the
-Cauchy real numbers and the well-typed syntax of type theory where conversion
-rules are given as equality constructors. In this paper we propose a general
-definition of HIITs using a small type theory, named the theory of signatures. A
-context in this theory encodes a HIIT by listing the constructors. We also
-compute notions of induction and recursion for HIITs, by using variants of
-syntactic logical relation translations. Building full categorical semantics and
-constructing initial algebras is left for future work. The theory of HIIT
-signatures was formalised in Agda together with the syntactic translations. We
-also provide a Haskell implementation, which takes signatures as input and
-outputs translation results as valid Agda code.
+Higher inductive-inductive types (HIITs) generalize inductive types of
+dependent type theories in two ways. On the one hand they allow the
+simultaneous definition of multiple sorts that can be indexed over
+each other. On the other, hand they support equality constructors,
+thus generalizing higher inductive types of homotopy type
+theory. Examples that make use of both features are the Cauchy real
+numbers and the well-typed syntax of type theory where conversion
+rules are given as equality constructors. In this paper we propose a
+general definition of HIITs using a small type theory, named the
+theory of signatures. A context in this theory encodes a HIIT by
+listing the constructors. We also compute notions of induction and
+recursion for HIITs by using variants of syntactic logical relation
+translations. Building full categorical semantics and constructing
+initial algebras is left for future work. The theory of HIIT
+signatures was formalised in Agda together with the syntactic
+translations. We also provide a Haskell implementation, which takes
+signatures as input and outputs translation results as valid Agda
+code.
 \end{abstract}
@@
 \]
-\noindent $P$ is a family of types (or: a proof-relevant predicate) over natural
+\noindent $P$ is a family of types (i.e.\ a proof-relevant predicate) over natural
 numbers, which is called the \emph{induction motive}. The arguments $pz$ and
@@
 
-Indexed families of types can be also considered, for example length-indexed
-vectors of $A$-elements $\mathsf{Vec}_A: \Nat \ra \mathsf{Type}$. Mutual
-inductive types are yet another generalization, however, they can be reduced to
-indexed families where indices classify constructors for each mutual type.
-Inductive-inductive types \cite{forsberg-phd} are mutual definitions where this
-reduction does not work: here a type can be defined together with a family
-indexed over it. An example is the following fragment of a well-typed syntax of
-a type theory, where the second $\Ty$ type constructor is indexed over $\Con$,
-but constructors of $\Con$ also refer to $\Ty$:
+Indexed families of types can be also considered, such as
+length-indexed vectors of $A$-elements $\mathsf{Vec}_A: \Nat \ra
+\mathsf{Type}$. Mutual inductive types are yet another generalization,
+but they can be reduced to indexed families where indices classify
+constructors for each mutual type.  Inductive-inductive types
+\cite{forsberg-phd} are mutual definitions where this reduction does
+not work: here a type can be defined together with a family indexed
+over it. An example is the following fragment of a well-typed syntax
+of a type theory, where the second $\Ty$ type constructor is indexed
+over $\Con$, but constructors of $\Con$ also refer to $\Ty$:
 \begin{alignat*}{3}
@@
 
-Closely related to the current work is the paper by the current authors and
-Altenkirch \cite{kaposi2019constructing}, which also concerns QIITs. There,
-signatures for QIITs are essentially a restriction of the signatures given here,
-but in contrast to the current work, the restricted quotient setting enables
-building initial algebras and detailed categorical semantics.
+Closely related to the current work is the paper by the current
+authors and Altenkirch \cite{kaposi2019constructing}, which also
+concerns QIITs. There, signatures for QIITs are essentially a
+restriction of the signatures given here, but in contrast to the
+current work, the restricted set-truncated setting enables building
+initial algebras and detailed categorical semantics.
 
@@
 \end{alignat*}
+
+In the presence of the univalence axiom, type constructor equations
+can be always equivalently represented using (4) path constructors, by
+adding explicit equivalences to a signature. In this case, type
+constructor equations serve as a shorthand for equivalences.
 
@@
 
-Let us first consider natural numbers, and see how the usual definition of
-induction arises.
-
-For $\Nat$, algebras are simply a triple consisting of a type, a value and an
-endofunction. Below, we leave universe indices explicit, and we use the notation
-of the external type theory described in Section \ref{sec:external}.
+Let us first consider natural numbers, and see how the usual
+definition of induction arises. For $\Nat$, algebras are simply a
+triple consisting of a type, a value and an endofunction. Below, we
+leave universe indices explicit, and we use the notation of the
+external type theory described in Section \ref{sec:external}.
 \begin{alignat*}{5}
@@
 
-We will compute algebras and the other notions by induction on the syntax of the
-theory of signatures. However, first we need to clarify the formal foundations
-of these computations.
+We compute algebras and the other notions by induction on the syntax
+of the theory of signatures. However, first we need to clarify the
+formal foundations of these computations.
 
@@
 The next task would be to define a computation which takes as input a
-$\hat{\Gamma}\vdash\Delta$ signature, and returns the corresponding type of
-algebras in some type theory. This would behave as a ``standard'' model of
-signatures, which simply maps each construction in the syntax to its
-counterpart: types to types, universe to universe, functions to functions, and
-so on.
+$\hat{\Gamma}\vdash\Delta$ signature, and returns the corresponding
+type of algebras in some type theory. This would behave as a
+``standard'' model of signatures, which simply maps each construction
+in the syntax to its counterpart: types to types, universe to
+universe, functions to functions, and so on.
 
@@
 
-It may come as a surprise that realizing the above steps is an \emph{open
-  problem}, for any syntax of a dependent type theory. We call the problem of
-interpreting syntaxes of type theories into a UIP-free metatheory a
-\emph{coherence problem}. This issue appears to arise with all known ways of
-defining syntaxes for dependent type theories.
+It may come as a surprise that realizing the above steps is an
+\emph{open problem}, for any syntax of a dependent type theory. We
+call the problem of interpreting syntaxes of type theories into a
+UIP-free metatheory a \emph{coherence problem}.
+
+This issue appears to arise with all known ways of defining syntaxes
+for dependent type theories. Shulman previously discussed this problem
+in \cite{hottshouldeat}. It is also related to the problem of
+constructing semisimplicial types in homotopy type theory; as
+explained in \cite{hottshouldeat} solving our coherence problem
+enables a construction of semisimplicial types.
 
@@
   & \boldsymbol{\cdot}^\M\hat{\gamma_0\,\gamma_1} && :\equiv \hat{\top} \\
-  & (\Delta,x:A)^\M\,\hat{\gamma_0\,\gamma_1} && :\equiv \hat{(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\times \blc{A^\M}\,\gamma_0\,\gamma_1\,\gamma^M} \\
+  & (\Delta,x:A)^\M\,\hat{(\gamma_0,\alpha_0)\,(\gamma_1,\alpha_1)} && :\equiv \hat{(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\times \blc{A^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,\alpha_0\,\alpha_1} \\
   & x^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^M} \\
@@
   & \boldsymbol{\cdot}^\M\hat{\gamma_0\,\gamma_1} && :\equiv \hat{\top} \\
-  & (\Delta,x:A)^\M\,\hat{\gamma_0\,\gamma_1} && :\equiv \hat{(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\times \blc{A^\M}\,\gamma^M} \\
+  & (\Delta,x:A)^\M\,\hat{(\gamma_0,\alpha_0)\,(\gamma_1,\alpha_1)} && :\equiv \hat{(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\times \blc{A^\M}\,\gamma^M\,\alpha_0\,\alpha_1} \\
   & x^\M\,\hat{\gamma^M} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^M} \\

### 2019-08-14 paper.tex
infinitary lambda comment

@@
 
-We do not include a $\lambda$ in the syntax. The reason is twofold; first,
-
 (4) $\U$ is closed under the equality type, with eliminator $\J$ and a weak
@@
 
-(7) $\U$ is also closed under a function space where the domain is an external
-type and the codomain is a small source theory type. We overload the application
-notation for external parameters, as it is usually clear from context which
-application is meant. The rules allow types with infinitary constructors, for
-example trees containing elements of $\hat{A}$ at the leaves and branching by
-$\hat{B}$ (which could be an infinite type):
-\[
-\hat{A}\in\Type_{\hat{0}}\commah\hat{B}\in\Type_{\hat{0}}\vdash\,\boldsymbol{\cdot},\,\,\,T:\U,\,\,\,leaf:(\hat{x}\in \hat{A})\ra\underline{T},\,\,\,node:((\hat{x}\in \hat{B})\ra T)\ra\underline{T}
-\]
-Here, $leaf$ has a function type (6) and $node$ has a function type (3) with a
-function type (7) in the domain. More generally, we can define $W$-types
-\cite{abbot05containers} as follows. $\hat{S}$ describes the ``shapes'' of the
-constructors and $\hat{P}$ the ``positions'' where recursive arguments can
-appear.
+(7) $\U$ is also closed under a function space where the domain is an
+external type and the codomain is a small source theory type. We
+overload the application notation for external parameters, as it is
+usually clear from context which application is meant. The rules allow
+types with infinitary constructors, for example, trees branching by a
+possibly infinite external type $\hat{A}$:
+\[
+\hat{A}\in\Type_{\hat{0}}\vdash\,\boldsymbol{\cdot},\,\,\,T:\U,\,\,\,leaf:\underline{T},\,\,\,node:((\hat{x}\in \hat{A})\ra T)\ra\underline{T}
+\]
+Here, $node$ has a function type (3) with a function type (7) in the
+domain. More generally, we can define $W$-types
+\cite{abbot05containers} as follows. $\hat{S}$ describes the
+``shapes'' of the constructors and $\hat{P}$ the ``positions'' where
+recursive arguments can appear.
 \[
@@
 an example file in our Haskell implementation.
+
+Note that we do not include a $\lambda$ for infinitary function types,
+although we could possibly use it when describing paths between such
+functions. The reason is that $\lambda$ can be always represented in
+signatures by introducing additional function parameters which are
+constrained by pointwise equalities. For example, consider adding the
+following path constructor to the previously described
+$\hat{A}$-branching trees:
+\[
+eq : \underline{node\,(\lambda\,\hat{x}.\,leaf) =_T leaf}
+\]
+This can be rewritten without $\lambda$ as follows:
+\[
+eq : (f : \hat{A}\ra T)\ra ((\hat{x : A})\ra f \hat{x} =_T leaf)\ra\underline{node\,f =_T leaf}
+\]
+The benefit of supporting $\lambda$ would be less encoding noise in signatures. However,
+we observe that such usage of $\lambda$ is rather rare in HITs in practice, and hence omit
+$\lambda$ for the sake of simplicity.
 

### 2019-08-14 paper.tex
add numbering to operations

--- 2019-08-14 paper.tex
@@
   & \El && : \Tm\,\Gamma\,\U \ra \Ty\,\Gamma && \text{decoding} \\
-  & \Pi && : (A:\Ty\,\Gamma)\ra\Ty\,(\Gamma\rhd A)\ra\Ty\,\Gamma && \text{functions} \\
+  & ... && \\
+  & \Pi && : (a:\Tm\,\Gamma\,\U)\ra\Ty\,(\Gamma\rhd \E\,a)\ra\Ty\,\Gamma && \text{functions} \\
   & ... &&
@@
 \begin{alignat*}{5}
-  & \boldsymbol{\cdot}^\A && :\equiv \hat{\top} \\
-  & (\Delta,x:A)^\A && :\equiv \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\timesh A^\A\,\hat{\gamma} \\
-  & x^\A\,\hat{\gamma} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma} \\
-  & \U^\A\,\hat{\gamma} && :\equiv \Type_{\hat{0}} \\
-  & (\underline{a})^\A\,\hat{\gamma} && :\equiv a^\A\,\hat{\gamma} \\
-  & ((x:a)\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)} \\
-  & (t\,u)^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\A\,\hat{\gamma}\hat{)} \\
-  & (t=_a u)^\A\,\hat{\gamma} && :\equiv t^\A\,\hat{\gamma} \eqh u^\A\,\hat{\gamma} \\
-  & (\refl_t)^\S\,\hat{\gamma} && :\equiv \hat{\refl} \\
-  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\A\,\hat{\gamma} && :\equiv \hat{\J}_{\hat{(}a^\A\,\hat{\gamma}\hat{)}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}\hat{\lambda} \hat{x}\,\hat{z}.p^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\commah\hat{z}\hat{)}\hat{)}}\,\hat{(}pr^\A\,\hat{\gamma}\hat{)}\,_{\hat{(}u^\A\,\hat{\gamma}\hat{)}}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)} \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
-  & (a=_\U b)^\A\,\hat{\gamma} && :\equiv a^\A\,\hat{\gamma} \eqh b^\A\,\hat{\gamma} \\
-  & (\refl_a)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
-  & (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\A\,\hat{\gamma} && :\equiv \hat{\J}_{\Type_{\hat{0}}\,\hat{(}a^\A\,\hat{\gamma}\hat{)}\,\hat{(}\hat{\lambda} \hat{x}\,\hat{z}.p^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\commah\hat{z}\hat{)}\hat{)}}\,\hat{(}pr^\A\,\hat{\gamma}\hat{)}\,_{\hat{(}b^\A\,\hat{\gamma}\hat{)}}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)} \\
-  & (\J\beta_{a\,(x.z.p)}\,pr)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
-  & ((\hat{x}\in \hat{A})\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\A\,\hat{\gamma} \\
-  & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u} \\
-  & ((\hat{x}\in \hat{A})\ra b)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^\A\,\hat{\gamma} \\
-  & (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u}
+  & (1)\hspace{1em} && \boldsymbol{\cdot}^\A && :\equiv \hat{\top} \\
+  & && (\Delta,x:A)^\A && :\equiv \hat{(}\hat{\gamma}\in\Delta^\A\hat{)}\timesh A^\A\,\hat{\gamma} \\
+  & && x^\A\,\hat{\gamma} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma} \\
+  & (2) && \U^\A\,\hat{\gamma} && :\equiv \Type_{\hat{0}} \\
+  & && (\underline{a})^\A\,\hat{\gamma} && :\equiv a^\A\,\hat{\gamma} \\
+  & (3) && ((x:a)\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)} \\
+  & && (t\,u)^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\A\,\hat{\gamma}\hat{)} \\
+  & (4) && (t=_a u)^\A\,\hat{\gamma} && :\equiv t^\A\,\hat{\gamma} \eqh u^\A\,\hat{\gamma} \\
+  & && (\refl_t)^\S\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & && (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\A\,\hat{\gamma} && :\equiv \hat{\J}_{\hat{(}a^\A\,\hat{\gamma}\hat{)}\,\hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{(}\hat{\lambda} \hat{x}\,\hat{z}.p^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\commah\hat{z}\hat{)}\hat{)}}\,\hat{(}pr^\A\,\hat{\gamma}\hat{)}\,_{\hat{(}u^\A\,\hat{\gamma}\hat{)}}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)} \\
+  & && (\J\beta_{a\,t\,(x.z.p)}\,pr)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & (5) && (a=_\U b)^\A\,\hat{\gamma} && :\equiv a^\A\,\hat{\gamma} \eqh b^\A\,\hat{\gamma} \\
+  & && (\refl_a)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & && (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\A\,\hat{\gamma} && :\equiv \hat{\J}_{\Type_{\hat{0}}\,\hat{(}a^\A\,\hat{\gamma}\hat{)}\,\hat{(}\hat{\lambda} \hat{x}\,\hat{z}.p^\A\,\hat{(}\hat{\gamma}\commah\hat{x}\commah\hat{z}\hat{)}\hat{)}}\,\hat{(}pr^\A\,\hat{\gamma}\hat{)}\,_{\hat{(}b^\A\,\hat{\gamma}\hat{)}}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)} \\
+  & && (\J\beta_{a\,(x.z.p)}\,pr)^\A\,\hat{\gamma} && :\equiv \hat{\refl} \\
+  & (6) && ((\hat{x}\in \hat{A})\ra B)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\A\,\hat{\gamma} \\
+  & && (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u} \\
+  & (7) && ((\hat{x}\in \hat{A})\ra b)^\A\,\hat{\gamma} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^\A\,\hat{\gamma} \\
+  & && (t\,\hat{u})^\A\,\hat{\gamma} && :\equiv \hat{(}t^\A\,\hat{\gamma}\hat{)}\,\hat{u}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \boldsymbol{\cdot}^\D\,\hat{\gamma} && :\equiv \hat{\top} \\
-  & (\Delta,\,x:A)^\D\,(\hat{\gamma}\commah\,\hat{t}) && :\equiv \hat{(}\hat{\gamma^D}\in\Delta^D\,\hat{\gamma}\hat{)}\timesh A^D\,\hat{\gamma^D}\,\hat{t} \\
-  & x^\D\,\hat{\gamma^D} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^D} \\
-  & \U^\D\,\hat{\gamma^D}\,\hat{A} && :\equiv \hat{A} \rah  \Type_{\hat{i}} \\
-  & (\underline{a})^\D\,\hat{\gamma^D}\,\hat{t} && :\equiv a^\D\,\hat{\gamma^D}\,\hat{t} \\
-  & ((x:a)\ra B)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\hat{(}\hat{x^D}\in a^\D\,\hat{\gamma^D}\,\hat{x}\hat{)} \rah  B^\D\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)}\,\hat{(}\hat{\gamma^D}\commah\hat{x^D}\hat{)}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,u)^\D\,\hat{\gamma^D} && :\equiv \hat{(}t^\D\,\hat{\gamma^D}\hat{)}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\D\,\hat{\gamma^D}\hat{)} \\
-  & (t=_a u)^\D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}}\,\hat{e}\,\hat{(}t^\D\,\hat{\gamma^D}\hat{)} \eqh u^\D\hat{\gamma^D} \\
-  & (\refl_t)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl}_{\hat{(}t^\D\,\hat{\gamma^D}\hat{)}} \\
-  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\D\,\hat{\gamma^D} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^D}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^D}\hat{)} \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
-  & (a=_\U b)^\D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(\lambda A.A\ra\Type_0)}}\,\hat{e}\,\hat{(}a^\D\,\hat{\gamma^D}\hat{)} \eqh b^\D\hat{\gamma^D} \\
-  & (\refl_a)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl}_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}} \\
-  & (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\D\,\hat{\gamma^D} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^D}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^D}\hat{)} \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
-  & ((\hat{x}\in \hat{A})\ra B)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^\D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u} \\
-  & ((\hat{x}\in \hat{A})\ra b)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^\D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u}
+  & (1)\hspace{1em} && \boldsymbol{\cdot}^\D\,\hat{\gamma} && :\equiv \hat{\top} \\
+  & && (\Delta,\,x:A)^\D\,(\hat{\gamma}\commah\,\hat{t}) && :\equiv \hat{(}\hat{\gamma^D}\in\Delta^D\,\hat{\gamma}\hat{)}\timesh A^D\,\hat{\gamma^D}\,\hat{t} \\
+  & && x^\D\,\hat{\gamma^D} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^D} \\
+  & (2) && \U^\D\,\hat{\gamma^D}\,\hat{A} && :\equiv \hat{A} \rah  \Type_{\hat{i}} \\
+  & && (\underline{a})^\D\,\hat{\gamma^D}\,\hat{t} && :\equiv a^\D\,\hat{\gamma^D}\,\hat{t} \\
+  & (3) && ((x:a)\ra B)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\hat{(}\hat{x^D}\in a^\D\,\hat{\gamma^D}\,\hat{x}\hat{)} \rah  B^\D\,\hat{(}\hat{\gamma}\commah\hat{x}\hat{)}\,\hat{(}\hat{\gamma^D}\commah\hat{x^D}\hat{)}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & && (t\,u)^\D\,\hat{\gamma^D} && :\equiv \hat{(}t^\D\,\hat{\gamma^D}\hat{)}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\,\hat{(}u^\D\,\hat{\gamma^D}\hat{)} \\
+  & (4) && (t=_a u)^\D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}}\,\hat{e}\,\hat{(}t^\D\,\hat{\gamma^D}\hat{)} \eqh u^\D\hat{\gamma^D} \\
+  & && (\refl_t)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl}_{\hat{(}t^\D\,\hat{\gamma^D}\hat{)}} \\
+  & && (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\D\,\hat{\gamma^D} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^D}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^D}\hat{)} \\
+  & && (\J\beta_{a\,t\,(x.z.p)}\,pr)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
+  & (5) && (a=_\U b)^\D\,\hat{\gamma^D}\,\hat{e} && :\equiv \TR_{\hat{(\lambda A.A\ra\Type_0)}}\,\hat{e}\,\hat{(}a^\D\,\hat{\gamma^D}\hat{)} \eqh b^\D\hat{\gamma^D} \\
+  & && (\refl_a)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl}_{\hat{(}a^\D\,\hat{\gamma^D}\hat{)}} \\
+  & && (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\D\,\hat{\gamma^D} && :\equiv \hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}pr^\D\,\hat{\gamma^D}\hat{)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\big)}\,\hat{(}eq^\D\,\hat{\gamma^D}\hat{)} \\
+  & && (\J\beta_{a\,t\,(x.z.p)}\,pr)^\D\,\hat{\gamma^D} && :\equiv \hat{\refl} \\
+  & (6) && ((\hat{x}\in \hat{A})\ra B)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  B^\D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & && (t\,\hat{u})^\D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u} \\
+  & (7) && ((\hat{x}\in \hat{A})\ra b)^\D\,\hat{\gamma^D}\,\hat{f} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  b^D\,\hat{\gamma^D}\,\hat{(}\hat{f}\,\hat{x}\hat{)} \\
+  & && (t\,\hat{u})^\D\,\hat{\gamma^D} && :\equiv t^\D\,\hat{\gamma^D}\,\hat{u}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \boldsymbol{\cdot}^\M\hat{\gamma_0\,\gamma_1} && :\equiv \hat{\top} \\
-  & (\Delta,x:A)^\M\,\hat{(\gamma_0,\alpha_0)\,(\gamma_1,\alpha_1)} && :\equiv \hat{(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\times \blc{A^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,\alpha_0\,\alpha_1} \\
-  & x^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^M} \\
-  & \U^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,A_0\,A_1} && :\equiv \hat{A_0\ra A_1 \ra \Type_0} \\
-  & (\underline{a})^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,t_0\,\,t_1} && :\equiv \hat{
+  & (1)\hspace{1em} && \boldsymbol{\cdot}^\M\hat{\gamma_0\,\gamma_1} && :\equiv \hat{\top} \\
+  & && (\Delta,x:A)^\M\,\hat{(\gamma_0,\alpha_0)\,(\gamma_1,\alpha_1)} && :\equiv \hat{(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\times \blc{A^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,\alpha_0\,\alpha_1} \\
+  & && x^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^M} \\
+  & (2) && \U^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,A_0\,A_1} && :\equiv \hat{A_0\ra A_1 \ra \Type_0} \\
+  & && (\underline{a})^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,t_0\,\,t_1} && :\equiv \hat{
     \blc{a^\M}\,\gamma_0\,\gamma_1\,\gamma^M\,t_0\,\,t_1}\\
-  & ((x:a)\ra B)^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,f_0\,f_1} && :\equiv
+  & (3) && ((x:a)\ra B)^\M\,\hat{\gamma_0\,\gamma_1\,\gamma^M\,f_0\,f_1} && :\equiv
   \hat{(x_0 : \blc{(\underline{a})^\A}\,\gamma_0)(x_1 : \blc{(\underline{a})^\A}\,\gamma_1)(x^M : \blc{(\underline{a})^\M}\,\gamma_0\,\gamma_1\,\gamma^M)}\\
-  & && \hat{\hspace{1em}\ra \blc{B^\M}\,(\gamma_0,\,x_0)\,(\gamma_1,\,x_1)\,(\gamma^M,\,x^M)\,(f_0\,x_0)\,(f_1\,x_1)}
+  & && && \hat{\hspace{1.7em}\ra \blc{B^\M}\,(\gamma_0,\,x_0)\,(\gamma_1,\,x_1)\,(\gamma^M,\,x^M)\,(f_0\,x_0)\,(f_1\,x_1)}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \boldsymbol{\cdot}^\S\,\hat{\gamma}\,\hat{\gamma^D} && :\equiv \hat{\top} \\
-  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv \hat{(}\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2}\hat{)}\timesh A^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} \\
-  & x^\S\,\hat{\gamma^S} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^S} \\
-  & \U^\S\,\hat{\gamma^S}\,\hat{A}\,\hat{A^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  \hat{A^D}\,\hat{x} \\
-  & (\underline{a})^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} && :\equiv a^\S\,\hat{\gamma^S}\,\hat{t} \eqh \hat{t^D} \\
-  & ((x:a)\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\S\,\hat{(}\hat{\gamma,x}\hat{)}\,\hat{(}\hat{\gamma^D,}\,a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\,\hat{(}\hat{\gamma^S,}\,\hat{\refl}\hat{)} \\
-  & && \hspace{8.4em} \hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{\big(}\hat{f^D}\,\hat{x}\,\hat{(}a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\hat{\big)} \\
-  & (t\,u)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^S}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)} \\
-  & (t=_a u)^\S\,\hat{\gamma^S}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^S}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
-  & (\refl_t)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{\refl}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)} \\
-  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S\,\hat{\gamma^S} && :\equiv \\
-  & && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}\hat{.}\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\\
-  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\Big)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^S}\hat{)} \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \\
-  & && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\\
-  & (a=_\U b)^\S\,\hat{\gamma^S} && :\equiv
+  & (1)\hspace{1em} && \boldsymbol{\cdot}^\S\,\hat{\gamma}\,\hat{\gamma^D} && :\equiv \hat{\top} \\
+  & && (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv \hat{(}\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2}\hat{)}\timesh A^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} \\
+  & && x^\S\,\hat{\gamma^S} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^S} \\
+  & (2) && \U^\S\,\hat{\gamma^S}\,\hat{A}\,\hat{A^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  \hat{A^D}\,\hat{x} \\
+  & && (\underline{a})^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} && :\equiv a^\S\,\hat{\gamma^S}\,\hat{t} \eqh \hat{t^D} \\
+  & (3) && ((x:a)\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\S\,\hat{(}\hat{\gamma,x}\hat{)}\,\hat{(}\hat{\gamma^D,}\,a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\,\hat{(}\hat{\gamma^S,}\,\hat{\refl}\hat{)} \\
+  & && && \hspace{8.4em} \hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{\big(}\hat{f^D}\,\hat{x}\,\hat{(}a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\hat{\big)} \\
+  & && (t\,u)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^S}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)} \\
+  & (4) && (t=_a u)^\S\,\hat{\gamma^S}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^S}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
+  & && (\refl_t)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{\refl}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)} \\
+  & && (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S\,\hat{\gamma^S} && :\equiv \\
+  & && && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}\hat{.}\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\\
+  & && && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\Big)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^S}\hat{)} \\
+  & && (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \\
+  & && && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\\
+  & (5) && (a=_\U b)^\S\,\hat{\gamma^S} && :\equiv
     \hat{\lambda\,e\,e^D.\,(\lambda\,x.\, \blc{b^\S}\,\gamma^S\, (\coe\,e\,x))
@@
     (\lambda\,x.\, \TR\,e^D\,(\J\,(\blc{a^\S}\,\gamma^S\,x)\,e))}\\
-  & (\refl_a)^\S\,\hat{\gamma^S} && :\equiv \hat{\refl} \\
-  & (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\S\,\hat{\gamma^S} && :\equiv\\
-  & && \hspace{-10em}\hat{\J\,\bigg(\lambda\,b^D\,\,b^S\,eq^D.\,\J\,\Big(\lambda\,b^S\,eq^S.\,\J\,(\lambda P^S\,pr^S.\,pr^S)\,eq^S\,(\mathsf{uncurry}\,\blc{p^\S}\,\gamma^S)\,(\blc{pr^\S}\,\gamma^S)\Big)}  \\
-  & && \hspace{-9em}\hat{\,(\blc{eq^\D}\,\gamma^D)\,b^S\bigg)\,(\blc{eq^\A}\,\gamma)\,(\blc{b^\D}\,\gamma^D)\,(\blc{b^\S}\,\gamma^S)\,(\blc{eq^\D}\,\gamma^D)\,(\blc{eq^\S}\,\gamma^S) } \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \hat{\J\,\refl\,(\blc{pr^\S}\,\gamma^S)} \\
-  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
-  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv t^\S\,\hat{\gamma^S}\,\hat{u} \\
-  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^S\,f\,t} && :\equiv b^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
-  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv \ap\,\hat{(}\hat{\lambda} \hat{f.}\hat{f}\,\hat{u}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}
+  & && (\refl_a)^\S\,\hat{\gamma^S} && :\equiv \hat{\refl} \\
+  & && (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\S\,\hat{\gamma^S} && :\equiv\\
+  & && && \hspace{-10em}\hat{\J\,\bigg(\lambda\,b^D\,\,b^S\,eq^D.\,\J\,\Big(\lambda\,b^S\,eq^S.\,\J\,(\lambda P^S\,pr^S.\,pr^S)\,eq^S\,(\mathsf{uncurry}\,\blc{p^\S}\,\gamma^S)\,(\blc{pr^\S}\,\gamma^S)\Big)}  \\
+  & && && \hspace{-9em}\hat{\,(\blc{eq^\D}\,\gamma^D)\,b^S\bigg)\,(\blc{eq^\A}\,\gamma)\,(\blc{b^\D}\,\gamma^D)\,(\blc{b^\S}\,\gamma^S)\,(\blc{eq^\D}\,\gamma^D)\,(\blc{eq^\S}\,\gamma^S) } \\
+  & && (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \hat{\J\,\refl\,(\blc{pr^\S}\,\gamma^S)} \\
+  & (6) && ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
+  & && (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv t^\S\,\hat{\gamma^S}\,\hat{u} \\
+  & (7) && ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^S\,f\,t} && :\equiv b^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
+  & && (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv \ap\,\hat{(}\hat{\lambda} \hat{f.}\hat{f}\,\hat{u}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}
 \end{alignat*}
@@
 \begin{alignat*}{5}
-  & \boldsymbol{\cdot}^\M\hat{\gamma_0\,\gamma_1} && :\equiv \hat{\top} \\
-  & (\Delta,x:A)^\M\,\hat{(\gamma_0,\alpha_0)\,(\gamma_1,\alpha_1)} && :\equiv \hat{(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\times \blc{A^\M}\,\gamma^M\,\alpha_0\,\alpha_1} \\
-  & x^\M\,\hat{\gamma^M} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^M} \\
-  & \U^\M\,\hat{\gamma^M\,A_0\,A_1} && :\equiv \hat{A_0\ra A_1} \\
-  & (\underline{a})^\M\,\hat{\gamma^M\,t_0\,\,t_1} && :\equiv \hat{
+  & (1)\hspace{1em} && \boldsymbol{\cdot}^\M\hat{\gamma_0\,\gamma_1} && :\equiv \hat{\top} \\
+  & && (\Delta,x:A)^\M\,\hat{(\gamma_0,\alpha_0)\,(\gamma_1,\alpha_1)} && :\equiv \hat{(\gamma^M : \blc{\Delta^\M}\,\gamma_0\,\gamma_1)\times \blc{A^\M}\,\gamma^M\,\alpha_0\,\alpha_1} \\
+  & && x^\M\,\hat{\gamma^M} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^M} \\
+  & (2) && \U^\M\,\hat{\gamma^M\,A_0\,A_1} && :\equiv \hat{A_0\ra A_1} \\
+  & && (\underline{a})^\M\,\hat{\gamma^M\,t_0\,\,t_1} && :\equiv \hat{
     \blc{a^\M}\,\gamma^M\,t_0 = t_1} \\
-  & ((x:a)\ra B)^\M\,\hat{\gamma^M\,f_0\,f_1} && :\equiv \\
-  & && \hat{\hspace{-8em}(x_0 : \blc{a^\A}\,\gamma_0)\ra \blc{B^\M}\,(\gamma_0,\,x_0)\,(\gamma_1,\,\blc{a^\M}\,\gamma^M\,x_0)\,(\gamma^M,\,\refl)\,
+  & (3) && ((x:a)\ra B)^\M\,\hat{\gamma^M\,f_0\,f_1} && :\equiv \\
+  & && && \hat{\hspace{-8em}(x_0 : \blc{a^\A}\,\gamma_0)\ra \blc{B^\M}\,(\gamma_0,\,x_0)\,(\gamma_1,\,\blc{a^\M}\,\gamma^M\,x_0)\,(\gamma^M,\,\refl)\,
     (f_0\,x_0)\,(f_1\,(\blc{a^\M}\,\gamma^M\,x_0))}\\
-  & (t\,u)^\M\,\hat{\gamma^M} && :\equiv  \hat{\J\,(\blc{t^\M}\,\gamma^M\,(\blc{u^\A}\,\gamma_0))\,(\blc{u^\M}\,\gamma^\M) }\\
-  & (t=_a u)^\M\,\hat{\gamma^M} && :\equiv
+  & && (t\,u)^\M\,\hat{\gamma^M} && :\equiv  \hat{\J\,(\blc{t^\M}\,\gamma^M\,(\blc{u^\A}\,\gamma_0))\,(\blc{u^\M}\,\gamma^\M) }\\
+  & (4) && (t=_a u)^\M\,\hat{\gamma^M} && :\equiv
     \hat{\lambda\,e.\,(\blc{t^\M}\,\gamma^M)^{-1}\sqcdot\ap\,(\blc{a^\M}\,\gamma^M)\,e\sqcdot\,\blc{u^\M}\,\gamma^M}\\
-  & (\refl_t)^\M\,\hat{\gamma^M} && :\equiv \hat{\inv\,(\blc{t^\M}\,\gamma^M)}\\
-  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\M\,\hat{\gamma^M} && :\equiv\\
-  & && \hspace{-10em}\hat{\J\,\bigg(\J\,\Big(\J\,\big(\J\,(\lambda\,p_1\,p^M\,pr_1\,pr^M.\, pr^M)\,(\blc{t^\M}\,\gamma^M)}  \\
-  & && \hspace{-9em}\hat{
+  & && (\refl_t)^\M\,\hat{\gamma^M} && :\equiv \hat{\inv\,(\blc{t^\M}\,\gamma^M)}\\
+  & && (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\M\,\hat{\gamma^M} && :\equiv\\
+  & && && \hspace{-10em}\hat{\J\,\bigg(\J\,\Big(\J\,\big(\J\,(\lambda\,p_1\,p^M\,pr_1\,pr^M.\, pr^M)\,(\blc{t^\M}\,\gamma^M)}  \\
+  & && && \hspace{-9em}\hat{
       (\mathsf{uncurry}\,\blc{p^\A}\,\gamma_1)\,(\mathsf{uncurry}\,\blc{p^\M}\,\gamma^M)\,
@@
         (\blc{eq^\A}\,\gamma_0)\Big)(\blc{u^\M}\,\gamma^M)\bigg)(\blc{eq^\M}\,\gamma^M)   }\\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\M\,\hat{\gamma^M} && :\equiv\\
-  & && \hspace{-9em}\hat{
+  & && (\J\beta_{a\,t\,(x.z.p)}\,pr)^\M\,\hat{\gamma^M} && :\equiv\\
+  & && && \hspace{-9em}\hat{
     \J\,\big(\J\,(\lambda\,p_1\,p^M.\,\refl)\,(\blc{t^\M}\,\gamma^M)\,
@@
     }\\
-  & (a=_\U b)^\M\,\hat{\gamma^M} && :\equiv
+  & (5) && (a=_\U b)^\M\,\hat{\gamma^M} && :\equiv
     \hat{\lambda\,e_0\,e_1.\,(\lambda\,x.\,\blc{b^\M}\,\gamma^M\,(\coe\,e_0\,x)) = (\lambda\,x.\,\coe\,e_1\,(\blc{a^\M}\,\gamma^M\,x))}\\
-  & (\refl_a)^\M\,\hat{\gamma^M} && :\equiv \hat{\refl} \\
-  & (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\M\,\hat{\gamma^M} && :\equiv\\
-  & && \hspace{-10em}\hat{\J\,\bigg(\J\,\Big(\lambda\,b^M\,eq^M.\,\J\,(\lambda\,P^M\,pr^M.\,pr^M)\,eq^M\,(\mathsf{uncurry}\,\blc{p^\M}\,\gamma^M)\,(\blc{pr^\M}\,\gamma^M)\Big)}  \\
-  & && \hspace{-9em}\hat{\,(\blc{eq^\A}\,\gamma_1)\bigg)\,(\blc{eq^\A}\,\gamma_0)\,(\blc{b^\M}\,\gamma^M)\,(\blc{eq^\M}\,\gamma^M) } \\
-  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\M\,\hat{\gamma^M} && :\equiv \hat{\inv\,(\blc{pr^\M}\,\gamma^M)} \\
-  & ((\hat{x:A})\ra B)^\M\,\hat{\gamma^M\,f_0\,f_1} && :\equiv \hat{(x : A)\ra \blc{B^\M}\,\gamma^M\,(f_0\,x)\,(f_1\,x)} \\
-  & (t\,\hat{u})^\M\,\hat{\gamma^M} && :\equiv t^\M\,\hat{\gamma^M\,u} \\
-  & ((\hat{x}\in \hat{A})\ra b)^\M\,\hat{\gamma^M} && :\equiv
+  & && (\refl_a)^\M\,\hat{\gamma^M} && :\equiv \hat{\refl} \\
+  & && (\J_{a\,(x.z.p)}\,pr\,_b\,eq)^\M\,\hat{\gamma^M} && :\equiv\\
+  & && && \hspace{-10em}\hat{\J\,\bigg(\J\,\Big(\lambda\,b^M\,eq^M.\,\J\,(\lambda\,P^M\,pr^M.\,pr^M)\,eq^M\,(\mathsf{uncurry}\,\blc{p^\M}\,\gamma^M)\,(\blc{pr^\M}\,\gamma^M)\Big)}  \\
+  & && && \hspace{-9em}\hat{\,(\blc{eq^\A}\,\gamma_1)\bigg)\,(\blc{eq^\A}\,\gamma_0)\,(\blc{b^\M}\,\gamma^M)\,(\blc{eq^\M}\,\gamma^M) } \\
+  & && (\J\beta_{a\,t\,(x.z.p)}\,pr)^\M\,\hat{\gamma^M} && :\equiv \hat{\inv\,(\blc{pr^\M}\,\gamma^M)} \\
+  & (6) && ((\hat{x:A})\ra B)^\M\,\hat{\gamma^M\,f_0\,f_1} && :\equiv \hat{(x : A)\ra \blc{B^\M}\,\gamma^M\,(f_0\,x)\,(f_1\,x)} \\
+  & && (t\,\hat{u})^\M\,\hat{\gamma^M} && :\equiv t^\M\,\hat{\gamma^M\,u} \\
+  & (7) && ((\hat{x}\in \hat{A})\ra b)^\M\,\hat{\gamma^M} && :\equiv
        \hat{\lambda f\,x.\,\,\blc{b^\M}\,\gamma^M\,(f\,x)} \\
-  & (t\,\hat{u})^\M\,\hat{\gamma^M} && :\equiv
+  & && (t\,\hat{u})^\M\,\hat{\gamma^M} && :\equiv
        \hat{\ap\,(\lambda f.\,f\,u)\,(\blc{t^\M}\,\gamma^M)}

### 2019-08-14 paper.tex
stuff

@@
 inductive type is given by its constructors, along with an elimination
-principle which expresses that all inhabitants are constructed using
-finitely many applications of the constructors.
+principle which expresses that it is enough to consider the
+constructors when defining a function out of the inductive type.
 
@@
 
-The article of Altenkirch et al. \cite{gabe} gives specification and semantics
+The article of Altenkirch et al.\ \cite{gabe} gives specification and semantics
 of QIITs in a set-truncated setting. Signatures are given as lists of functors
@@
 The logical predicate syntactic translation was introduced by Bernardy et
-al. \cite{bernardy2010parametricity}. The idea that a context can be seen as a
+al.\ \cite{bernardy2010parametricity}. The idea that a context can be seen as a
 signatures and the logical predicate translation can be used to derive the types
@@
 
-\noindent Now, we can reformulate induction for $\Nat$. First, we assume that
-there exists a distinguished $\Nat$-algebra, named $\mathsf{InitialAlg}$. The
-induction principle has the following type:
-\[
- \mathsf{Induction} : (M : \mathsf{\DisplayedAlg}\,\,\mathsf{InitialAlg}) \ra \Section\, M
+\noindent Now, we can reformulate induction for $\Nat$. First, we
+assume that there exists a distinguished $\Nat$-algebra, named
+$\Nat^*$. The induction principle for this algebra has the following
+type:
+\[
+ \mathsf{Induction} : (M : \mathsf{\DisplayedAlg}\,\,\Nat^*) \ra \Section\, M
 \]
@@
 \[
-\mathsf{Recursion} : (\alpha : \Alg) \ra \Morphism\,\mathsf{InitialAlg}\,\alpha
+\mathsf{Recursion} : (\alpha : \Alg) \ra \Morphism\,\Nat^* \,\alpha
 \]
@@
 \begin{alignat*}{5}
-\mathsf{Initiality} : (\alpha : \Alg) \ra \mathsf{isContr}\,(\Morphism\,\mathsf{InitialAlg}\,\alpha)
+\mathsf{Initiality} : (\alpha : \Alg) \ra \mathsf{isContr}\,(\Morphism\,\Nat^*\,\alpha)
 \end{alignat*}
@@
 \subsection{Interpreting Intrinsic Syntax}
-Following Altenkirch and Kaposi \cite{ttintt}, one might define the syntax of a
-type theory as an initial category with families (CwF) \cite{dybjer1995internal}
-extended with additional type formers. The CwF part provides a calculus and
+Following Altenkirch and Kaposi \cite{ttintt}, one might define the
+syntax of a type theory as a category with families (CwF)
+\cite{dybjer1995internal} extended with additional type formers, which
+supports an induction principle. The CwF part provides a calculus and
 equational theory for explicit substitutions, upon which one can build
@@
 
-So far, we were able to compute algebras, displayed algebras, morphisms and
-sections, and this allows us to state recursion and induction principles.
-Reiterating Section \ref{sec:general}, for a signature
-$\hat{\Gamma\vdash}\,\Delta$ and assuming $\hat{\Gamma\vdash \mathsf{InitialAlg}
-  : \blc{\Delta^\A}}$, we have the following types for induction and recursion:
+So far, we were able to compute algebras, displayed algebras,
+morphisms and sections, and this allows us to state recursion and
+induction principles.  Reiterating Section \ref{sec:general}, for a
+signature $\hat{\Gamma\vdash}\,\Delta$ and assuming $\hat{\Gamma\vdash
+  \Delta^* : \blc{\Delta^\A}}$ as a candidate algebra for the HIIT, we
+have the following types for induction and recursion:
 \begin{alignat*}{5}
 & \hat{\Gamma\vdash \mathsf{Induction} :
-    (\gamma^D : \blc{\Delta^\D}\,\mathsf{InitialAlg})\ra \blc{\Delta^\S}\,\mathsf{InitialAlg}\,\gamma^D}\\
+    (\gamma^D : \blc{\Delta^\D}\,\Delta^*)\ra \blc{\Delta^\S}\,\Delta^*\,\gamma^D}\\
 & \hat{\Gamma\vdash \mathsf{Recursion} :
-    (\gamma : \blc{\Delta^\A})\ra \blc{\Delta^\M}\,\mathsf{InitialAlg}\,\gamma}
+    (\gamma : \blc{\Delta^\A})\ra \blc{\Delta^\M}\,\Delta^*\,\gamma}
 \end{alignat*}

### 2019-08-15 response.txt
rename revision.txt

Response to the first reviewer:

- Extending to n-Type:
  As we note in section 2, generalizing external types in signatures to Type-i
  is possible and not difficult. We omit this generalization from the paper
  mainly in order to sync with the existing Agda formalization.

- Page 1: initiality vs finite generation
  We revised the wording.

- Lambdas in signatures:

  The only function space where it makes sense to have lambdas is the
  (7) infinitary function space, since we can only mention these in
  path types. But such lambdas can be always eliminated by introducing
  extra function parameters which are constrained by pointwise
  equalities. In the case of the mentioned example,

    z : H
    n : (A → H) → H
    q : n (λ x. z) = z

  We may have the following equivalent λ-free type for q:

    q : (f : A → H) → ((x : A) → f x = z) → n f = z

  In general, if we have
    Γ;Δ ⊢ λx. t : El ((x : a) → b)
  Then instead we may have
    Γ;Δ, f : El ((x : a) → b), p : (x : a) → f x = t ⊢ f : El ((x : a) → b)

  We include this explanation in the revision.

- Page 7, 8 item (5), about type equalities and UA

  We include this in the revision.

- Page 10, about initial NatAlg
  We use Nat* instead of initial Nat-algebra in the revised version.

- Page 11, about coherence issue, references

  In the revision we add a reference to the following discussion:

    https://homotopytypetheory.org/2014/03/03/hott-should-eat-itself/

  As discussed there, our coherence problem is related to the better
  known problem of constructing semisimplicial types; solving the
  former implies a solution of the latter.

- Page 12, about initial CwF

  We removed the problematic "initial" occurrence. To say a bit more
  here, if we set-truncate CwFs then there is a strict category of
  CwFs, and if we don't truncate then we conjecture that there is a
  wild category of wild CwF-s, but in any case we are referring to a
  1-category for the purpose of induction and recursion. We think that
  the subtleties involving these categories are not very relevant to
  the discussion in that section, so we avoid it in the revision.

- Page 13, "would merely decidable equality also be sufficient?"

  For the purpose of elaborating raw syntax into intrinsic syntax
  (which is in our view an important motivation for decidability), it
  would not be sufficient, because we need to eliminate from decided
  equalities to proof-relevant intrinsic terms.

--------------------------------------------------------------------------------

Response to the second reviewer:

- On the suggestion for identity homomorphisms:

  This suggestion does solve the U[] case. However, it still does not
  work; some cases (like U[]) become provable, where we can make use
  of strict equality hypotheses, but other cases fail, where we need
  to provide strict equalities. For example, assuming the suggested
  strict tᴵᴰ equations, we have

    (t =ₐ u)ᴵᴰ γ : (t = u)ᴹ γ γ (Γᴵᴰ γ) ≡ Uᴵᴰ γ ((t = u)ᴬ γ)

  The type can be reduced along definitions, and using tᴵᴰ and uᴵᴰ, to:

    (λ e. refl ◾ ap (λ x. x) e ◾ refl) ≡ (λ e. e)

  Here, e : tᴬ γ = uᴬ γ, and it is necessarily a non-strict equality. Hence,
  this case is not provable with strict tᴵᴰ equalities but provable with weak
  ones. We include this explanation in the revision.

- Addressing comments:
  - page 7, Figure 2, about external parameters:

    We do not construct HIIT definitions, only signatures. HIIT signatures can
    only depend on an HIIT specified by a previous signature if we assume the
    existence of that HIIT in the external type theory (i.e. as part of the
    context Γ). Also, the codomain types of external (6) and infinitary (7)
    functions can depend on the additional (x : A) external binder. For example,
    if we have Γ = (Nat: Type₀, zero: Nat, suc: Nat → Nat), and we write the
    signature for length-indexed A-vectors, then we have "cons : (n : Nat) → A →
    Vec n → Vec (suc n)", where the "suc n" expression refers back to the
    external "n : Nat" binder. The only way to extend the external Γ is by using
    (6) and (7) functions, otherwise, we can only refer to Γ entries which are
    assumed on the "top level", such as the A : Type₀ in the example for lists
    on page 8 in Section 2.1.

  - page 12: for the intrinsic syntax, do you also assume an induction
    principle for it, e.g. that it is initial?

    Yes. Of course, the induction principle for the intrinsic syntax would be
    something that is formalized in the current paper, so if we want to avoid
    circularity, then we need to say that the induction principle is given in an
    ad-hoc fashion. This is probably acceptable, since such ad-hoc HIT induction
    principles have been common in previous HoTT literature.

  - page 12, line -9: you say that you "need to" set-truncate; is this
    rather a strong belief that you have to?

    In the next paragraphs we justify the need for truncation. We
    think that undecidability of conversion and impossibility of
    extrinsic term models are solid justifications.

  - page 21, line -16: is there an expected equation if all computation rules
    were definitional?

    We think that if we follow "book" HoTT practice, p^s would be a
    propositional computation rule, because it involves "coe" which is defined
    externally, similarly to how path constructor computation rules involving
    "ap" are also propositional.

### 2019-12-02 paper.tex
typo discovered by Thorsten
revision 2 updates

@@
 simultaneous definition of multiple sorts that can be indexed over
-each other. On the other, hand they support equality constructors,
+each other. On the other hand they support equality constructors,
 thus generalizing higher inductive types of homotopy type
@@
 
-(6) We have a function space where the domain is a type in the external
-theory. We distinguish it from (3) by using red brick color in the domain
-specification. We specify lists and integers as follows, with integers now
-given as quotients of pairs of natural numbers:
+(6) is a function space where the domain is a type in the external theory. We
+distinguish it from (3) by using red brick color in the domain specification. We
+specify lists and integers as follows, with integers now given as quotients of
+pairs of natural numbers:
 \begin{alignat*}{4}
@@
 signatures to arbitrary universe levels, but it is not essential to the current
-development. Note that the (6) function space preserves strict positivity, since
-in the external theory there is no way to recursively refer to the inductive type
-\emph{being defined}. The situation is analogous to the case of $W$-types
+development.
+
+Note that we can assume arbitrary structures in the external $\hat{\Gamma}$
+context, which in particular allows us to specify HIITs depending on other
+(external) HIITs. We can do this by first specifying a HIIT, then using the
+methods in this paper to compute notions of algebras and induction, then assume
+the HIIT in the external context of another HIIT signature.
+
+The (6) function space preserves strict positivity, since in the external theory
+there is no way to recursively refer to the inductive type \emph{being
+  defined}. The situation is analogous to the case of $W$-types
 \cite{abbot05containers}, where shapes and positions can contain arbitrary types
-but they cannot recursively refer to the $W$-type being defined.
+but they cannot recursively refer to the $W$-type being defined. This setup rules
+out some signatures; for example, rose trees cannot be specified as follows, because
+we cannot apply the external $\hat{List}$ to the inductive $T$:
+\[
+\hat{A : \Type_0,\,\,List : \Type_0 \ra \Type_0} \vdash
+     \boldsymbol{\cdot},\,\,T : \U,\,\,node : \hat{A} \ra \hat{List}\,T \ra \underline{T}
+\]
+Analogously nested HIT examples are the ``hubs and spokes'' definitions in
+\cite[Section 6.7]{HoTTbook}. To allow such definitions, we would have to
+analyze external constructions to check whether they preserve strict
+positivity. This is out of the scope of the current paper.
 
@@
 \subsubsection{Set-truncation}
-Additionally, we need to set-truncate the syntax by adding the following constructors:
+Additionally, we prefer to set-truncate the syntax by adding the following
+constructors, for reasons shortly explained:
 \begin{alignat*}{5}
@@
 
-\noindent We can omit the rule for contexts, as it is derivable from the above ones. Set
-truncation forces definitional equality of the syntax (as defined by equality
-constructors in the HIIT signature) to be proof irrelevant. If we omit
+\noindent We can omit the rule for contexts, as it is derivable from the above
+ones. Set truncation forces definitional equality of the syntax (as defined by
+equality constructors in the HIIT signature) to be proof irrelevant. If we omit
 set-truncation, then the defined HIIT becomes very different from what we expect
@@
 formally distinguishing $\U[]$ and $[\id]$. Hence, by Hedberg's theorem
-\cite{hedberg}, the syntax does not have decidable equality. This also implies
-that the non-truncated intrinsic syntax is not constructible from set quotients
-of extrinsic terms (since those always form sets). The non-truncated syntax just
-does not seem to be a sensible notion. This situation is similar to how
-categories in homotopy type theory need to have set-truncated morphisms
-\cite{ahrens2015univalent}.
+\cite{hedberg}, the syntax does not have decidable equality, and hence it does
+not support decidable type checking. This also implies that the non-truncated
+intrinsic syntax is not constructible from set quotients of extrinsic terms
+(since those always form sets). The non-truncated syntax just does not seem to
+be a sensible notion. This situation is similar to how categories in homotopy
+type theory need to have set-truncated morphisms \cite{ahrens2015univalent}.
 
@@
 Note that if $\hat{b^S}$ is a definitional equality (that is, we have
-$\hat{S^{2S}\,\,b}\,\hat{\equiv}\,\hat{b^D}$), the occurrences of $\hat{b^S}$
+$\hat{S^{2S}\,\,b}\,\hat{\,\equiv\,}\,\hat{b^D}$), the occurrences of $\hat{b^S}$
 in the type of $\hat{surf^S}$ can be replaced by $\hat{\refl}$. In

### 2019-12-02 paper.tex

@@
 (external) HIITs. We can do this by first specifying a HIIT, then using the
-methods in this paper to compute notions of algebras and induction, then assume
-the HIIT in the external context of another HIIT signature.
+translations in Sections \ref{sec:algebras}-\ref{sec:sections} to compute
+notions of algebras and induction, then assume the HIIT in the external context
+of another HIIT signature.
 
@@
 \[
-\hat{S}\in\Type_{\hat{0}}\commah\hat{P}:\hat{S}\rah\Type_{\hat{0}}\vdash\,\boldsymbol{\cdot},\,\,\,W:\U,\,\,\,sup: (\hat{s} \in \hat{S})\ra((\hat{p} \in \hat{P}\,\hat{s})\ra W)\ra \underline{W}
+\hat{S : \Type_0, P : S \ra \Type_0} \vdash\,\boldsymbol{\cdot},\,\,\,W:\U,\,\,\,sup: (\hat{s} \in \hat{S})\ra((\hat{p} \in \hat{P}\,\hat{s})\ra W)\ra \underline{W}
 \]

### 2019-12-03 paper.tex
revision

@@
 
-%% build: latexmk -pdf -pvc -interaction=nonstopmode paper.tex
+%% build: latexmk -pdf -pvc paper.tex
 
@@
 
+The question may arise whether our $\J$ is sufficient to define all
+constructions on paths which we want to express, as we do not have
+$\Sigma$-types in signatures, and we only have a weak version of $\Pi$. First,
+as far as we know, all path constructions which occur in HITs in the literature
+are expressible using our $\J$, so it is certainly adequate in this
+sense. Second, we provide a sketch in our Agda formalization (see Section \ref{sec:formalization}) that a Frobenius
+variant of $\J$ (first considered by Garner in \cite[p.~13]{garner2009two}) is
+derivable from the $\J$ given in Figure \ref{sigrules1}, which suggests that our
+$\J$ does not lose expressiveness because of the lack of $\Sigma$ and
+$\Pi$-types.
+
 (5) There is a different equality type former, which can be used to express
@@
 
-\begingroup
-\allowdisplaybreaks
+%% \begingroup
+%% \allowdisplaybreaks
 \begin{alignat*}{5}
@@
 \end{alignat*}
-\endgroup
+%% \endgroup
 
@@
 
-There are three additional development artifacts to the current work: a Haskell
-implementation, a shallow Agda formalization and a deep Agda formalization. All
-three are available from \url{https://github.com/akaposi/hiit-signatures}.
+There are additional development artifacts to the current work: a Haskell
+implementation, two Agda formalizations of the syntactic translations (a shallow
+and a deeper version) and an Agda formalization for deriving a Frobenius $\J$
+rule from $\J$. All are available from
+\url{https://github.com/akaposi/hiit-signatures}.
 
@@
 $\hat{\Gamma}\vdash\Delta$ signature. Then, it checks the input with respect to
-the rules in Figures \ref{sigrules1} and \ref{sigrules2}, and outputs an Agda-checkable file which
-contains algebras, homomorphisms, displayed algebras and sections for the
-input signature.
+the rules in Figures \ref{sigrules1} and \ref{sigrules2}, and outputs an
+Agda-checkable file which contains algebras, homomorphisms, displayed algebras
+and sections for the input signature.
 
